### 2019-02-02 conclusion.tex
Initial ShareLaTeX Import

\section{Conclusions}

Propositional linear temporal logic (LTL) and automata are two common specification formalisms for software and hardware systems. While temporal logic has a more declarative
flavor, automata are more operational, describing how the specified system
progresses. There has been several proposed extensions to LTL that extend
its expressive power to that of related automata formalisms. We proposed here
a simple extension for propositional LTL that is based on adding propositions that summarize the prefix of an execution. Conceptually, this extension puts the specification in between propositional LTL and automata, as the additional
variables can be seen as representing the state of an automaton
that is synchronized with the temporal property. It is shown to have
the same expressive power as automata models, and is in particular
appealing for runtime verification of past (i.e., safety) temporal properties,
which already are based on summarizing the value of subformulas over
observed prefixes.

We then demonstrated that first-order linear temporal logic (FLTL), which
can be used to assert properties about systems with data, also has expressiveness deficiencies, and extended it with relations that summarize prefixes.
We showed that for the first-order case, unlike the propositional case,
this extension is not identical to the addition of dynamic (i.e., state dependent) quantification.

We presented a monitoring
algorithm for propositional past time temporal logic, extending a classical algorithm, and similarly presented an algorithm for 
first-order past temporal logic.
Finally we described the implementation of this extension in the \dejavu{} tool and provided experimental results.

### 2019-02-02 implementation.tex
Initial ShareLaTeX Import

\section{Implementation}

\dejavu{} is implemented in \scalalang{}.   
\dejavu{} takes as input a specification file containing one  or 
more  properties, and synthesizes the monitor as a self-contained \scalalang{} 
program.
This program takes as input the 
trace file and analyzes it.
The tool uses the JavaBDD library for BDD manipulations \cite{javabdd}.

\noindent {\bf Example properties.}
%
Figure \ref{fig:properties} shows three properties in the 
input format of the tool, which are not expressible in the original first-order logic of \dejavu{}. The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express Wolper's example (\cite{Wolper}) that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when the radio is toggled on.

\begin{center}
\begin{figure}
\begin{lstlisting}[language=dsl,frame=single,linewidth=0.95\textwidth,backgroundcolor=\color{white},linewidth=\columnwidth,breaklines=true,basicstyle=\small]
prop telemetry1: 
  Forall x . closed(x) -> !telem(x)
  where closed(x) := toggle(x) <-> @!closed(x)

prop telemetry2: 
  Forall x . closed(x) -> !telem(x)
  where
    closed(x) :=
        (!@true & !toggle(x))
      | (@closed(x) & !toggle(x))
      | (@open(x) & toggle(x)),
    open(x) :=
        (@open(x) & !toggle(x))
      | (@closed(x) & toggle(x))

prop spawning : 
  Forall x . Forall y .
  spawn(x,y) -> 
    (P main(x) | Exists z . (P main(z) & leadsto(z,y)))
  where
    leadsto(z,y) := 
        spawn(z,y) 
      | @ leadsto(z,y) 
      | Exists x . (@leadsto(z,x) & spawn(x,y))
\end{lstlisting}
\caption{Three properties stated in \dejavu's logic}
\label{fig:properties}
\end{figure}
\end{center}

The second property, \idsl{telemetry2}, expresses the same 
property as \idsl{telemetry1}, but in this case using two 
rules, more closely reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
\idsl{closed(x)} and \idsl{open(x)}. The rule \idsl{closed(x)} is defined as a disjunction between three alternatives. The first states that this predicate is true if we are in the initial state 
(the only state where \idsl{@true} is false), and there is no \idsl{toggle(x)} event. The next alternative states that
if \idsl{closed(x)} was true in the previous state and there is
no \idsl{toggle(x)} event, we stay in the \idsl{closed(x)} state.
The third alternative states that if we in the previous state were
in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} event, we transition to the \idsl{closed(x)} state. Similarly for the \idsl{open(x)} rule.

The third property, \idsl{spawning}, expresses a property
about threads being spawned in an operating system. We want to 
ensure that all threads are spawned by the main program or are spawned by a thread that transitively can be traced back to a thread being spawned by the main program. The events are
\idsl{main(x)} (the main program spawns thread \idsl{x}) and
\idsl{spawn(x,y)} (thread \idsl{x} spawns thread \idsl{y}).
For this we need to compute a transitive closure of spawning relationships, here expressed with the rule \idsl{leadsto(z,y)}.

\noindent {\bf Monitor synthesis.}
%
Consider the first \idsl{telemetry1} property. The property-specific 
part\footnote{An additional 600+ lines of property independent 
boilerplate code is generated.} of the synthesized monitor, is shown 
in Figure \ref{fig:monitor}. This function is called for each new event. The function operates on the two arrays holding BDDs, indexed by subformula indexes: \iscala{pre} for the previous state and \iscala{now} for the current state. For each observed event, the function \iscala{evaluate()} computes the \iscala{now} array, evaluating any sub-formula before any formula containing it,
and returns true (property is satisfied in this position of the trace)  iff \idsl{now(0)} is not $\bfalse$.

\begin{figure}
%\centering
\begin{center}
\begin{tabular}{c}
\begin{lstlisting}[language=scala,basicstyle=\small\sffamily]
override def evaluate(): Boolean = {
  // event nodes:
  now(4) = build("telem")(V("x"))
  now(6) = build("toggle")(V("x"))
  // rule nodes not below @:
  now(7) = pre(8)
  now(5) = now(6).biimp(now(7))
  // rule calls:
  now(2) = now(5)
  now(9) = now(5)
  // rest of rules below @:
  now(8) = now(9).not()
  // rest of main formula:
  now(3) = now(4).not()
  now(1) = now(2).not().or(now(3))
  now(0) = now(1).forAll(var_x.quantvar)

  val error = now(0).isZero
  if (error) monitor.recordResult()
  tmp = now; now = pre; pre = tmp
  !error
}
\end{lstlisting}
\end{tabular}
\end{center}
\caption{Monitor evaluation function for telemetry1}
\label{fig:monitor}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=7cm]{figures/ast.pdf}
\caption{Annotated abstract syntax tree for telemetry1}
\label{fig:ast}
\end{figure}

The indexing scheme is guided by an enumeration of the annotated abstract syntax tree of the formula, shown in Figure \ref{fig:ast}. The tree has been annotated by pointing out (blue node 5) the rule body and calls to the rule (purple nodes 2 and 9).
The assignments are evaluated in five steps. First the leaf nodes corresponding to events (4 and 6). Then rule nodes not below the previous-operator \idsl{@} (7 and 5). Then calls of rules (2 and 9)
which can now be calculated since the rule bodies have been calculated. Then the rest of the rules below \idsl{@}  (8). Finally the rest of the main formula (3, 1, 0). At composite subformula nodes, BDD operators are applied. For example for subformula  1, the new  value is \iscala{now(2).not().or(now(3))}, 
which is the interpretation of the formula 
\idsl{closed(x) -> !telem(x)} corresponding to the Boolean law:
$p \rightarrow q = \neg p \vee q$.

\noindent {\bf Example execution.}
%
As an example of how a trace is processed consider the following simple (correct) trace containing two events, turning on low and high frequency channels:
$\mathit{toggle}(``low").\mathit{toggle}(``high")$. The value $``low"$ is assigned the binary enumeration $110$ in node 6, shown in Figure \ref{fig:bdd1}. A dashed line leaving a node means the node has value 0, and a fully drawn line means it has value 1. The BDD represents all assignments to the bit positions 0, 1 and 2 that lead to 1 (shaded square leaf node), in this case 110, read from left to right. Likewise, Figure \ref{fig:bdd2} shows the BBD assigned to node 6 for the second value $``high"$, corresponding to the enumeration 101. Finally Figure \ref{fig:bdd3} shows the BDD representing the rule \idsl{closed(x)} in node 5 after the two events. Since channels $``low"$ and $``high"$, corresponding to enumerations 110 and 101, are now open, this BDD represents all other enumerations. That is, all enumerations except 110 and 110 lead to 1 in this BDD.

\begin{figure}
    \centering   
    \begin{subfigure}[b]{0.15\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd1.pdf}
        \caption{toggle(``low'')}
        \label{fig:bdd1}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.15\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd2.pdf}
        \caption{toggle(``high'')}
        \label{fig:bdd2}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.15\textwidth}
        \includegraphics[width=\textwidth]{figures/bdd3.pdf}
        \caption{closed(x)}
        \label{fig:bdd3}
    \end{subfigure}
    \caption{Selected BDDs from trace evaluation}
    \label{fig:bdds}
\end{figure}

\noindent {\bf Evaluation}
%
We evaluated the three properties in Figure 
\ref{fig:properties}. Table \ref{tab:evaluation} shows the  
analysis time (excluding time to compile the generated 
monitor) for different traces (format is `trace length : seconds'). The telemetry traces will repeatedly open hundreds of channels and send telemetry over them, and the spawning traces will spawn a new thread for each event in the trace. The transitive closure of the spawning property needs to process these, therefore the more extensive processing time.

\ignore{
  // repeat: repeat the following
  // repeat_toggle: open this many channels
  // repeat_telem: send this many messages on each channel
}

\ignore{
  // threads: how many threads initially spawned by main program
  // repeat: how many times new threads from new threads are spawned
  // number of spawned threads correspond to number of events
}

\begin{table}[h!]
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Property   & Trace 1 & Trace 2 & Trace 3 \\
\hline
telemetry1 & 1,200,001\,:\,3.3s & 5,200,001\,:\,7.1s & 10,200,001\,:\,12.8s \\
\hline
telemetry2 & 1,200,001\,:\,4.4s & 5,200,001\,:\,9.9s & 10,200,001\,:\,17.9s \\
\hline
spawning   & 4,950\,:\,9.4s & 10,000\,:\,88.4s & 19,900\,:\,488.8s \\
\hline
\end{tabular}
\caption{Evaluation}
\label{tab:evaluation}
\end{table}

### 2019-02-02 main.tex
Initial ShareLaTeX Import

\documentclass{llncs}
%\documentclass[converence,10pt]{IEEEtran}
%\IEEEoverridecommandlockouts
%\usepackage{cite}
%\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{algorithmic}
%\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{bm}
%\usepackage{xcolor}
%\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\newcommand{\Exists}{\bm{\exists}\kern-0.6em\bm{\exists}}
\newcommand{\Forall}{\bm{\forall}\kern-0.6em\bm{\forall}}

\input{macros}

\begin{document}

\title{An Extension of LTL with Rules 
and its Application to Runtime Verification}

\ignore{\author{
\begin{tabular}{ccc}
Klaus Havelund & \vspace{0.2cm} & 
Doron Peled \\
NASA Jet Propulsion Laboratory, Pasadena, USA & & 
Bar Ilan University, Ramat Gan, Israel \\
klaus.havelund@jpl.nasa.gov & & doron.peled@gmail.com
\end{tabular}
}
}

\ignore{
\author{\IEEEauthorblockN{Klaus Havelund} 
\IEEEauthorblockA{\textit{Laboratory for Reliable 
Software} \\
\textit{NASA Jet Propulsion Laboratory}\\
Pasadena, USA \\
klaus.havelund@jpl.nasa.gov}\\\mbox{}\\
\and
\IEEEauthorblockN{Doron Peled} 
\IEEEauthorblockA{\textit{Dep. of Computer Science} \\
\textit{Bar Ilan University}\\
Ramat Gan, Israel \\
doron.peled@gmail.com}
}
}

%\thanks{Copyright 2018. All rights reserved.}

\author{ 
Klaus Havelund~\thanks{The research performed by this author was carried out at Jet Propulsion Laboratory, California Institute of Technology, under a contract with the National Aeronautics and Space Administration.} 
\and
Doron Peled~\thanks{The research performed by this author was partially funded by Israeli Science Foundation grant 2239/15: 
``Runtime Measuring and Checking of Cyber Physical Systems''.
}
}

\institute{
Jet Propulsion Laboratory, \\
California Institute of Technology, USA 
\and
Department of Computer Science \\
Bar Ilan University, Israel
}

%\hfill
\maketitle

\begin{abstract}

Runtime Verification (RV) consists of analyzing execution 
traces using formal techniques. RV is scalable and 
therefore has a big potential for infusion in software 
engineering practices. It includes e.g. monitoring the 
execution of a system against properties formulated in 
Linear Temporal Logic (LTL). 
LTL offers a succinct notation for writing useful 
specification properties. However, it is limited in 
expressiveness in the propositional case, and several 
theoretic (non-operational) extensions have therefore been proposed. Furthermore, for many practical cases, there is a need to monitor properties 
that carry data, where one can use formalisms like 
first-order LTL. We show that 
first-order LTL has similar expressive
limitations as the propositional version. We suggest here 
two related extensions for increasing the expressive power: 
one for propositional LTL and one for 
first-order LTL. These extensions have a simple incremental operational 
semantics that is more suitable for RV than previously 
suggested extensions. We show that the 
propositional extension has the same expressiveness as the 
classical extensions, and demonstrate its
adoption for runtime verification of first-order safety 
properties. Finally, we expand the BDD-based runtime verification 
tool \dejavu{}  to support our extension and we perform some 
experiments.

\end{abstract}

\input{rule-ltl}

\input{implementation}

\input{conclusion}

\input{prop.bib}

\appendix

\noindent {\bf Sketch of proof of Theorem~\ref{theo2}.}
As we defined the semantics of
EFLTL by translating into QFLTL, QFLTL is at least as
expressive as EFLTL. We show here that 
the expressive power of EFLTL is strictly smaller than that of
QFLTL through the following example. It consists of
a temporal relation\footnote{We abuse notation and
use the same name for the relation and for its
denotation in the logic. Also, $r [ i ]$ has zero
parameters; as noted before, this is a natural extension to the syntax and semantics presented above.} $r$ with only one parameter, so $r [ i ]$
has a Boolean value. We will henceforth say that $r$ occurs
in state $i$. Another relation is $v$, with
two parameters. For each $i$, if $r$ does not appear in the
state, 
then there is a value $a$ from some (infinite) domain,
such that $v ( a, i )$, 
i.e., $v(a)$ appears in $i$.
The property requires that there are no two sets of values appearing within the relation $v$ (as $v(a)$) between successive occurrences of $r$ that are the same.

Hence, the sequence of states 
\[ \{ r \} \{ v(1) \}  \{ v(2) \} \{ v(3) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(3) \} \{  r \} \] is a model of this formula, while \[ \{ r \} \{ v(1) \}  \{ v(2) \}  \{ v(2) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(5) \}  \{  r \}\] is not.
The formula is a bit involved, hence given in several parts (macros).
\begin{itemize}
\item $add(P) ::= \forall x \, (P(x) \leftrightarrow (v(x) \vee \ominus P(x))$ [The set $P$ differs in the current state from the previous one only by values $a$ that appear in $v(a)$.]
\item $same(P) ::= \forall x \, (P(x) \leftrightarrow \ominus P(x))$ 
[The set of elements $x$ for which
$P (x)$ holds
contains the same elements as in the previous state.]
\item $empty (P) ::= \forall x \neg P(x)$ [The set $P$ is empty.]
\item $equal (P,Q) ::= \forall x (P(x) \leftrightarrow Q(x))$ 
[The sets represented by the relations
$P$ and $Q$ contain the same elements {\em in the current state}.]
\item $collect(P, \varphi, \psi) ::=
       q \wedge same(P) \wedge \varphi \wedge 
       \ominus ((add(P)\wedge \neg q \wedge \varphi) \since (q \wedge empty(P) \wedge \psi))$  [Collect elements $a$ in $v(a)$, while $\varphi$ holds, 
       since the previous state where $q$ 
       and $\psi$ held. ]
\item $init ::= r \wedge \Box (r \leftrightarrow \neg \exists x \, v(x))$ [$r$ holds in the first state, and $v(a)$ with some value $a$ holds in states where $q$ does not hold.]
\item $init \wedge \neg \Exists A \Exists B (equal (A, B) \wedge \\
collect(B, same(A),\, same(A) \since collect(A, true, true))$
[There are no two sequences of states where $v(a)$ holds
with various values $a$, between a pair of states satisfying $r$ that contain exactly the same elements (regardless of repetition).]
\end{itemize}
We prove that the above property cannot be expressed in EFLTL.
We use a combinatorial argument.
In EFLTL, after every prefix of the input, it is
sufficient to keep a finite set of (non temporal) relations over the
data elements seen in the input, which
summarizes the prefix. This follows from the set
semantics, which appears in Section~\ref{EPLTLRV}, 
where the interpretation of
each subformula, and the auxiliary relations,
at a particular state is a relation, and the interpretation 
of a property a prefix of length $i$ only depends on
these relations. Now, the number of such possible relations
for a fixed number of data elements $N$ is polynomial
in $N$. However, the above property checks whether
a set of values that appear within the relation $v$, between two consecutive appearances
of $r$'s, are the same. For that, one needs to keep a set of
subsets of the elements that appeared in the input so far. 
But the number of such subsets is exponential in $N$. \qed

\iffalse
{\bf Comments.} First, note that the reason that QFLTL {\em can} express the property in Theorem~\ref{theo2}, is that it allows to augment the input relations with auxiliary relations that are not fixed. That is, for an auxiliary relation $R$,
$R [ i ]$ is not fixed, and can change per different
prefixes (of length at least $i$), or even in different
subformulas. Second, given a simple extension to EFLTL, where 
one can use a variable that represents an updatable
counter (or a generator of distinct values), one can express the property used in~\ref{theo2}, by
keeping a relation where values between different occurrences of the event $r$ are paired with distinct values, hence, representing an unbounded set of sets.
\fi

\end{document}

### 2019-02-02 rule-ltl.tex
Initial ShareLaTeX Import

\section{Introduction}

Runtime verification (RV)  \cite{bartocci18,havelund-rv-data-2018} very generally refers to the use of rigorous (formal) 
techniques for {\em processing} execution traces emitted by a system being observed. The purpose is typically
%, again generally viewed, 
to evaluate the state of the observed system. Processing can take numerous
forms. We focus here on {\em specification-based} runtime verification, where an execution trace is checked against a property expressed in a formal logic, in our case variants of linear temporal logic.

Linear Temporal Logic (LTL) is a common specification formalism for reactive and concurrent systems. It is often used in e.g. model checking and runtime verification. Another formalism that is used for the same purpose is finite automata, often over infinite words. This includes B\"{u}chi, Rabin, Street, Muller and Parity automata~\cite{Thomas}, all having the same expressive power. In fact, model checking of an LTL specification
is usually performed by first translating the specification into a B\"{u}chi automaton. The automata formalisms are more expressive than LTL, with a classical example By Wolper~\cite{Wolper} showing that
it is not possible to express in LTL that every even state in
the sequence satisfies $p$. This has motivated extending LTL
in various ways.
We are mainly motivated by runtime verification of executions with data, for which a first-order temporal logic is
appropriate. There, too, we show that expressiveness can be extended.

We present first an extension to the propositional case that is based on using additional (auxiliary) propositions, not appearing in the execution, which summarize the past 
of the current execution. These propositions obtain their value in a state as a function of the prefix of the execution up and including that state through a past temporal formula. This can be seen as adding
summary states to the temporal specification, resulting in a specification that is between logic and finite automata representation. This extension, which we believe is conceptually simpler
and more intuitive than other suggested extensions, is shown to be
equivalent to the other common specification formalism:
B\"{u}chi automata, regular expressions and quantified LTL.

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL.
We present such an extension where additional (auxiliary) relations are
added. We show that this extension is more expressive than
first-order LTL, e.g., by repeating Wolper's classical algorithm
with data. On the other hand, we show that for the first-order version, this has less 
expressive power than adding quantification over relations, in distinction from the propositional case.

These extensions to the logics are in particular natural for
runtime verification of past (safety) temporal properties, since
the values of propositional and first-order LTL are based
on calculating and updating a summary of the prefix of the execution.
We provide runtime verification algorithms for the extended
logics and present an extension of the 
\dejavu{} tool (https://github.com/havelund/dejavu) that
realizes the extended first-order past temporal logic verification.
Finally, we provide some experimental results.

\noindent {\bf Related work.}
%
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds
state-dependent quantification over propositions to LTL.

Addition of relations to temporal databases is done
for aggregation~\cite{Libkin}, where relations are
calculated, based on the current state, and are used for further calculating
functions (sums, maxima) over the sequence of states. Aggregations were also used in the runtime verification tool 
{\sf MonPoly}~\cite{agrebasin}.
%
% --- other less directly related work: ---
%
Numerous other systems have been produced for monitoring
execution traces against formal specifications. 
These include e.g. \cite{Meredith2011} \cite{larva} \cite{Reger2015} \cite{halle-beepbeep-ieee-12} \cite{LOLA}.

% ----------------

\iffalse
The contributions of the papers are both theoretical, and
practical. We study the extensions of propositional temporal logic (both full version and past/safety version) and suggest a restricted (and, to our opinion, an intuitive) extension that is equivalent to the power of quantified LTL or B\"{u}chi/finite automata. We then study 
The corresponding extension to first-order LTL, replacing propositions with relations, and allowing variables and constant values. We propose, for the first-order version, an extension similar to the one we propose for the propositional case. But then we show that for the first-order case, this extension has
less expressive power than adding quantification in the style of propositional temporal logic.

The extensions we suggest are in particular easy to implement, as an extension to existing propositional/first-order runtime verification. We extend our tool \dejavu, experiment with new examples that take advantage of the extension.
\fi

\section{Preliminaries}
\label{sec:prelim}

In this paper, we treat several versions of the logic LTL. To simplify
the presentation, we prefix LTL with letters as follows:

\begin{itemize}
\item{P} restricts to {\em past} temporal operators.
\item{F} allows first-order quantification over
variables and relations over variables and constants.
\item{Q} adds quantification over predicates (for the propositional versions)
or over relations (for the first-order version). These objects can have different values at different states.
\item{E} allows extending the objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases).
\end{itemize}

%\klaus{The explanation of Q and E could be improved.}

\newcommand\eventty{\mathbb{E}}
\newcommand\setof[1]{\mathcal{P}(#1)}

\section{Propositional Linear Temporal Logic}

The classical definition of linear temporal logic~\cite{MP} has the following syntax:
\[ \varphi ::= \true \, | \,  p \, | \, ( \varphi \wedge \varphi ) \, |  \, \neg  \varphi \, |   \, \bigcirc \varphi \, |  
\, ( \varphi \until \varphi ) \,|  \, \ominus \varphi |
   \, ( \varphi \since \psi ) \,  \]
where $p$ is a proposition from a finite set of propositions $P$, and with $\bigcirc$, $\until$, $\ominus$, $\since$ standing for {\em next-time}, 
{\em until}, {\em previous-time} and {\em since}, respectively.

The models for LTL formulas 
are infinite sequence of states.
$\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
where $s_i \subseteq P$ for each $i\geq 0$. These are the propositions that {\em hold} in that state.
We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
LTL's semantics is defined as follows:
\begin{itemize}
\item $( \sigma, {i}) \models \mathit{true}$.
\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$.
\item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
\item $(\sigma, {i}) \models ( \varphi \land \psi)$ iff $(\sigma, {i}) \models \varphi$ and $(\sigma, {i} ) \models \psi$.
\item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
\item $(\sigma, {i}) \models ( \varphi\ \cal{U}\ \psi )$ iff for some 
$j\geq i$, $(\sigma, {j}) \models \psi$,
and for all $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
\item $(\sigma , i) \models \ominus \varphi$ iff $| \sigma | > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $1 \leq k \leq i$ such that
$(\sigma , i) \models \psi$  and for each
$j < k \leq i$, 
$(\sigma , k)  \models \psi$.
\end{itemize}
Then $\sigma \models \varphi$ when $( \sigma ,0 ) \models \varphi$.
We can use the following abbreviations:
$\false = \neg \true$, 
$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$, 
$(\varphi \rightarrow \psi) = ( \neg \varphi \vee \psi)$,
$(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
$\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
${\bf P}\  \varphi = (\true \; \since \; \varphi)$, and 
${\bf H}\ \varphi = \neg {\bf P}\ \neg \varphi$.

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} logics where the temporal propositions
correspond to unary predicates (over time). A full overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions with the following operators: constant letters, concatenation and complementation, but no star (or $\omega$-star for infinite words).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer n such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
leaves the same expressive power and is commonly used.

%\klaus{Is that also the true for the first-order case? Doron: I have no idea. But at this point, I would not ask to ask Gabbay.}

An important subset of LTL, called here PLTL, allows only past temporal
operators: $\since$, $\ominus$ and the ones derived
from them $\mathbf{H}$ and $\mathbf{P}$. The logic is often
(and also here) interpreted over finite sequences, 
where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
Alternatively,
it is also a common practice to prefix the property with a single $\Box$ (always) operator, where each of the prefixes has to satisfy $\varphi$. This later
form expresses {\em safety} properties~\cite{AS} over infinite sequences.

When PLTL is interpreted over finite sequences, its
expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.

\section{Making propositional LTL more Expressive}

Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
to some important properties. As a canonical example, he demonstrated that it is impossible to
express that all the states with even indexes of a sequence
satisfy $p$ (this is
different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states). In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
propositions.

Adding quantification over propositions allows writing a formula of the form
$\Exists q \, \varphi$, where $q$ is a proposition that can appear in $\varphi$.
Let the propositions $P$ be partitioned into two disjoint sets: 
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$. 
%A model for such a formula 
%has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.
We also allow $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
To define the semantics, let $X$ be a subset of 
the propositional variables appearing in the states of the sequence $\sigma = s_1 s_2 \ldots$ and 
denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
Since all the auxiliary variables appear in the scope of
quantifiers, it holds that $\sigma \models \varphi$
(in $\varphi$, auxiliary variables appear only within
the scope of quantifiers)
iff $\sigma |_A \models \varphi$.

This kind of quantification is considered to be ``dynamic'', since the quantified variables can have different truth values depending on the states, or, "second order", the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.

\noindent
Wolper's property can be expressed in QLTL as follows:
\begin{equation} \label{form0} \Exists q \,
( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}
Wolper's example property cannot be expressed in PLTL either.

If we allow
quantification over propositions, we obtain the logic PQLTL
(QLTL restricted to the past modalities), with the following formula:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
PQLTL has the same expressive
power as regular expressions and finite automata.
Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
PQLTL property into a deterministic
automaton.

We present now an extension of LTL, which we call ELTL. It has the same
expressive power as  QLTL.
It has a form that puts it, conceptually, in between temporal
logic and automata: on the one hand, it is possible to simply represent
a pure PLTL formula, and on the other hand, one can represent a specification that resembles a finite automaton. As will be showed later, this extension is
is very natural for runtime verification.

We partition again, as in QLTL, the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$. A model for an ELTL formula has truth assignments only over the propositions in $B$.
An ELTL property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_i$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a PLTL property where propositions from $B$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$ as the rules. In text, rules
 will be separated from each other by a comma (",").

\begin{tabbing}
\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ such that \\
 \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
\end{tabbing}

\noindent
Wolper's example can be written in ELTL
as follows:
\begin{equation} 
\label{form3}
\Box ( q \rightarrow p ) \, \mathrm{where} \, 
 ( q := \ominus \neg q)
\end{equation}
where $A = \{ q\}$ and $B = \{ p \}$.
The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
and each {\em even} state will satisfy $p$.  As $\ominus \varphi \models \false$ in the initial state, regardless of $\varphi$, then
$q$ is $\false$ in the initial state, and then it subsequently alternates.

ELTL extends the set of propositional variables with new variables whose values at a state are
based on the prefix of the model. We say that
the extension of the states are {\em well founded},
as the added propositions are dependent and effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in QLTL, where the auxiliary variables do not have to extend the execution
states in a fixed way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to guarantee that the values
of the auxiliary variables are not conflicting, as in
$a_1 = \neg a_2$ and $a_2 = a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$.

\begin{lemma} \label{fourone}
Let $\varphi$ be an ELTL formula as in~(\ref{ELTL}),
over the auxiliary propositions $A$ and basic propositions $B$ 
and a model $\sigma$ with states over the $B$, such that
$\sigma \models \psi$.
Then there is a unique
model $\sigma'$ such that $\sigma' |_A = \sigma$.

\end{lemma}

%\klaus{I think this lemma needs to tie the trace to the formula. Doron: I do not understand this comment.}

\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states. A PLTL
property depends on a state only on the prefix
of the model that precedes and includes that state.
Restricting the auxiliary variables to appear only within the scope of a previous-time operator guarantees that
their values at the current state cannot be conflicting. \qed

\begin{theorem}
\label{theo1}
The expressive power of 
ELTL is the same as QLTL.
\end{theorem}
\noindent {\bf Proof.}
Each ELTL formula $\eta$, as defined above,
is expressible in QLTL using the following formula:
\[   \Exists a_1 \ldots \Exists a_n  ( \varphi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
\]
For the other direction, one can translate any QLTL property into an $\omega$ automaton\footnote{The past operators of a QLTL property do not change this fact, since the translation can be done first to second order monadic logic, which can be translated into an $\omega$-automaton}.
We assume that the property is given as a deterministic Muller automaton $M = (S, i , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $i \in S$ is the initial state, the set $\Sigma$ consists of the inputs, each represented as a Boolean combination of the variables $B$, $\Delta : S \times \Sigma \mapsto S$, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set, 
where an accepted input needs to pass infinitely often in exactly
one of the specified sets. To formulate an ELTL formula,
he auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.

Then we translate $M$ into an ELTL formula
of the form \begin{tabbing}
$\bigvee_{F \in {\cal F}} (\bigwedge_{s \in F} \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
where \\
\ \ \ \ \ $a_s := \vee _{\Delta ( s'
, \alpha ) = s } 
\alpha \wedge \ominus a_{s'} )
\wedge (s=i \rightarrow \neg \ominus \true)
\, :\,  s \in S
$
\end{tabbing}
With $(s=i)$ is $\true$ when $s$ is $i$, and $\false$ otherwise.
Basically, the assignments to the auxiliary variables are used to simulate the change of states\footnote{It is possible to use only $\lceil \log ( | S | ) \rceil$
variables, at the expense of longer and more complicated formulas.}, and the main formula expresses the Muller acceptance condition. \qed

The use of the auxiliary variables $a_1 \ldots a_n$ in the
rules of an ELTL specification 
$\eta$ can be viewed as representing a
state machine, in particular, in
particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an ELTL specification $\eta$ as consisting with two synchronous parts: an LTL property 
(the statement) $\psi$
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary variables, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to QLTL.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary variables. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

We can extend PLTL into EPLTL by disallowing the future temporal operators (except a prefixing $\Box$, when interpreted over infinite sequences). We obtain
a formalism that is equivalent to finite automata. In this case, 
we replace the first part of the construction in the proof of Theorem~\ref{theo1}
with the simpler formula $\vee_{s \in A} a_s$, where $A$ are the accepting states of the finite automaton.

For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.
\begin{lemma}
\label{simplecase}
Any safety QPLTL property can be expressed using the following restricted
ELTL form:
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}  \] 
 with $p$ being one of the auxiliary variables $a_j$.
\end{lemma}
\noindent {\bf Proof.} The proof of the Lemma follows from the
axiomatization of QLTL in~\cite{MP}, e.g.,
$\varphi \since \psi \leftrightarrow ( \psi \vee ( \varphi \wedge \varphi \since \psi) ) )$. \qed

While this Lemma makes the use of the temporal operators
besides $\ominus$ redundant, we believe that it is still
more convenient to use a combination of
temporal operators when possible, and
auxiliary variables definitions.
The relative expressive power between the propositional temporal logics discussed here
appear in Figure~\ref{prop}.

\begin{figure}
\begin{center}
\includegraphics[height=1.1in,width=2.7in]{PROP.jpg}
\caption{\label{prop} An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness}
\end{center}
\end{figure}

\section{Runtime Verification for PLTL and its extensions}
\label{LTLruntime}

Runtime verification of temporal temporal specification often
concentrates on the past portion of the logic.  Past specifications have the important property that one can distinguish when they are violated. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.

The algorithm for PLTL, first presented in \cite{HR}, 
is based on the observation that the semantics of the 
past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
in the previous step $i - 1$ of a subformula.
To demonstrate this, we repeat here the definition of the previous-time
operator $\ominus$, and rewrite the definition
of $\since$ in an equivalent form that is
more directly amenable for runtime verification.

\begin{itemize}
\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
$( \sigma , i)  \models \varphi$ and 
$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}

The runtime verification exploits the fact that
the semantic definition is recursive in both the structure of the property, where subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, operates on two vectors (arrays) of values indexed by subformulas:  $\old$ that summarizes the truth values of the
subformulas just {\em before} the new state, and $\current$ for the {\em current} state.

\begin{enumerate}
\item Initially, for each subformula $\varphi$,
%$B ( \varphi , \current ) = \bfalse $.
$\current ( \varphi ) := \false$.
\item Observe a new event (as a set of ground predicates) $s$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.

\begin{itemize}
  \item $\current ( \true ) := \true$.
  \item $\current (  \varphi \wedge \psi  ) := 
  \current ( \varphi )\  and\ \current ( \psi )$.
  \item $\current ( \neg \varphi  ) := not\ \current ( \varphi )$.
  \item $\current (  \varphi \since \psi  ) :=  
  \current ( \psi  )\ or\ ( \current ( \varphi ) \ and\ 
      \old ( ( \varphi \since  \psi )))$.
%DAP: I removed the \; around \since, since it penetrated the 2nd column!
  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.

\end{itemize}
\item Goto step~2.
\end{enumerate}

Runtime verification of QPLTL can be done by translating a
formula into a deterministic finite automaton. Then the sequence of states to be checked forms the input for this automaton. QPLTL can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

The well foundedness of the formulas defining the new auxiliary variables in EPLTL makes the extension of the above
runtime algorithm simple.
In order to extend the algorithm to perform runtime verification of an EPLTL formula $\psi$, as defined in~(\ref{ELTL}),  we need to  add to the above
algorithm memory for $\current ( a_j )$ and $\old ( a_j )$
and perform, between steps 3 and 4, the following, per each $a_j$: \\

\ \ \ \ -\ $\current ( a_j ) = \current ( \varphi_j )$ \\

\noindent {\bf Evaluation order.} The order of calculating $\current$ is subtle here, because
auxiliary variables can appear both on the left and right of their
definition. For PLTL, one can start calculating the subformulas
bottom up or top down. But for EPLTL, this would not work. To see this, consider, for example, the formula~(\ref{form3}). It contains
the definition $q := \ominus \neg q$. Now, we cannot calculate this
bottom up, since the left-hand $q$ is not calculated yet.
If we calculate it top down, we will encounter the same problem.
However, notice that the top down calculation does not need the
value of $q$ to calculate $\ominus \neg q$; this value should already be available as $\old ( \neg q )$.

Under {\em mixed evaluation order}, one calculates $\current$ in the following order. The calculation within each of the four categories can be performed top down or bottom up.
\begin{itemize}
\item Calculate 
$\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of rule
$a_j := \varphi_j$ but not within the scope of
a $\ominus$ operator. (This is not
a problem since for $\delta = \ominus \gamma$,
$\current ( \delta )$ is set to $\old ( \gamma )$.)
\item Set 
$\current ( a_j )$ to $\current ( \varphi_j )$ for each $j$.
\item Calculate $\current ( \delta )$ for each subformulas $\delta$
that appear in $\varphi_j$ of a rule
$a_j = \varphi_j$ {\em  within} the scope of
a $\ominus$ operator.
\item Calculate $now ( \psi )$ (using $\current ( a_j )$) for
the statement $\psi$.
\end{itemize}

\section{First-Order Linear Temporal Logic FLTL}

%section*{Syntax and Semantics}
\label{sec:syntax-semantics}

Assume a finite set of infinite domains\footnote{Finite domains are handled with some minor changes, see~\cite{HPU}.}
$D_1, D_2, \ldots$,
e.g., integers or strings. 
Let $V$ be a finite set of {\em variables}, 
with typical instances $x$, $y$, $z$.
%We denote by $x: D$ the fact that the variable $x$ has the domain $D$.
An {\em assignment} over a set of variables $V$
maps each variable $x \in V$ to a value from
its associated domain $\domain ( x )$, where multiple variables (or all of them)
can be related to the same domain. For example
$[ x \rightarrow 5 , y \rightarrow \text{``abc''} ]$ assigns
the values $5$ to $x$ and the value ``abc'' to $y$.
Let $T$ a set of {\em relation} 
with typical instances $R$.
Each relation is associated with some position-dependent domains.

We defined models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is,
their last parameter
is a natural number. For a relation $R$, $R [ i ]$ is the relation obtained from $R$ by
restricting it to the value $i$ in the last parameter. For simplicity, we will describe henceforth
the logic with relations $R$ that have exactly two parameters
, hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$; this domain is fixed over all the value of the second parameter $i$. The the definition of the logic over other number of parameters is quite straightforward, and our implementation of runtime verification fully supports it.

\noindent {\bf Syntax.} 
The formulas of the core FLTL logic are 
defined by the following grammar,
where $p$ denotes to a relation $R$,
$a$ is a constant representing a value in $d ( R )$ and $x$ is a variable with $\domain ( x ) = d ( p )$.
\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
   \bigcirc \varphi \; | \; 
   ( \varphi \; \until \; \varphi ) \; | $ \\
   $ \ominus \; \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}

%We interpret a predicate $p(a)$ as a subformula that
%represents the occurrence of some
%For example, if we open
%a file named $\text{``xyz''}$, we may have $\mathit{open}
%(\text{``xyz''})$, where the
%domain of the predicate ${\mathit{open}}$ is over strings representing
%file names.

\noindent

Additional operators are defined as in the propositional logic. In addition we define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past ones
($\since$, $\ominus$ and the ones obtained from them) 
forms the logic PFLTL.
The quantification over values of variables here is `static' in the sense that
the assigned value is independent of the state in the execution. We
denote static quantification with $\exists$ and $\forall$, while
dynamic quantification uses $\Exists$ and $\Forall$.

\noindent {\bf Semantics.} 
A model is a set of
 relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$, where each of these relations has a last parameter $i$ that is a natural number.

It it standard, and more intuitive (in particular for runtime verification) to cast models of temporal logic in terms of sequences states rather than temporal relations. To do that,
 we denote by $R_k [ i ]$ the restriction of
 the relation $R_k$ to the value $i$ of the last parameter. A model ${\cal R}$ is then a sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$, where
 each {\em state} ${\cal R} [ i ] = {\cal R} [i] = \{ R_1 [i], \ldots , R_m [ i ] \}$  (a set of relations).
 The denotation $p_k$ will correspond to the relation $R_k$.

\iffalse
At a given state the formula
$p(\text{``a''})$ means that $p (\text{``a''} )$ holds
in the current state,
that is, $p (\text{``a''} )$ is among 
the ground predicates of the state.
Consider now the formula $p ( x )$, for a variable $x \in V$.
We interpret it such that $x$ is assigned any value ``$a$'' where
$p ( \text{``a''} )$ appears in the current state. 
%DP: Added parentheses on the outside level.
Thus, for interpreting $(p ( x ) \wedge q ( y ))$ in a state that
has the predicates
$p ( \text{``a''} )$ and $q ( 3 )$,
we have the assignment $[ x \mapsto \text{``a''} , y \mapsto 3 ]$.
%The past time temporal operators have the following intuitive meaning.
The formula $(\varphi_1\; \since\; \varphi_2)$ 
(reads $\varphi_1$ {\em since} $\varphi_2$)
means that $\varphi_2$ occurred in the past (including now)
and since then (beyond that state) $\varphi_1$ has been true. This is the 
past dual of the commonly used %DP: changed "common" to "commonly used".
future time  {\em until} modality~\cite{MP}. 
The property $\ominus \; \varphi$ means that $\varphi$ is true 
in the previous state.
This is the past dual of the %DP: removed "common".
future time {\em next} modality.
We can also define the following additional temporal operators:
$P \; \varphi = (\true \, \since \, \varphi)$ (``previously''),
and $H \varphi = \neg P \neg \varphi$ (``always in the past'' or ``historically'').
The operator $[\varphi_1,\varphi_2)$, borrowed from \cite{MaC}, 
has the same meaning as $(\neg \varphi_2\; \since\; \varphi_1)$, but reads more naturally as
a semi-open interval.

%Consider now the formula $P \neg p ( v )$, meaning that
%sometimes in the past $\neg p ( v )$. The semantics of $P$
%will be the union on the sets of values in the previous sets,
%thus, we need to take the union of $\neg p ( v )$ interpreted
%over all past states (including the current one).
%
%Suppose that we have the predicate $p ( i )$ at state $s_i$.
%There are several ways to interpret that.
%\begin{enumerate}
%\item We complement with respect to
%all the values in $domain ( p )$.
%Thus, the complementation will be $N \setminus \{ i \}$ at state $i$,
%where $N$ is the domain natural numbers, and the 
%$N \setminus \{ i \}$ from $1$ to $n$ is $N$.
%
%\item We complement with respect to
%the values in $domain ( p )$ that appeared so far in the sequence.
%Thus, at state $s_i$ we already saw $\{1 , \ldots , i \}$
%and as we see $p ( i )$ in the current state, the complementation 
%at state $s_i$ is $\{ 1, \ldots , i-1 \}$. The union 
%up to the $n$th state will give $\{1, \ldots , n-1 \}$.
%
%\item When new values appear later in the sequence, we 
%update the complementation related to previous states to include
%newly appearing values.
%Then we have to take the union of $n$ sets
%that do not include one distinct element each. This
%gives us at the $n$th state $\{ 1 , \ldots , n \}$.
%\end{enumerate}

%In our implementation we will take the first interpretation.
%However, we will assume that we have a reasonable
%limit on the number of different values that can appear in
%each domain. As it will cost us very little to assume quite
%a large size of values (the size of memory and amount of
%time of the implementation will be propositional to the
%{\em actual} number of values seen and not to this limit), we deem
%this assumption quite reasonable. %We will explain how to 
%remove this constraint (however, this requires quite a
%complicated implementation). It is also quite easy to see
%how to change the semantics to conform with case $1$ and
%how to implement it.

Let $\gamma$ be an assignment to the variables that appear
free in a formula $\varphi$.
Then $( \gamma ,  \sigma, i ) \models \varphi$
if $\varphi$ holds for the prefix $s_1 s_2 \ldots s_i$ of 
the trace $\sigma$
with the assignment $\gamma$. This is a standard definition,
agreeing, e.g., with~\cite{Basin}. Note that by using past %DP: appearing in, instead of appearing with.
operators, the semantics is not affected by
states $s_j$ for $j>i$.
%
\fi

Let $\vars ( \varphi )$ be the set of free (i.e., unquantified) variables of a
subformula $\varphi$. 
%The interpretation of a subformula
%$\varphi$ over some finite sequence is as a set of assignments
%to the variables $\vars ( \varphi)$.
We denote by $\gamma |_{\vars ( \varphi )}$ the restriction (projection) of
an assignment $\gamma$ to the free variables appearing in $\varphi$.
%We are careful in combining the sets of assignments
%between different
%subformulas, e.g., through union and intersection (corresponding
%to disjunction and conjunction, respectively) and the quantification.
%This is done using the projection and extension functions over
%assignments defined above.
%
Let $\epsilon$ be the empty assignment. 
%
In any of the following cases, $( \gamma,  {\cal R} , i ) \models \varphi$
is defined when $\gamma$ is an
assignment over $\vars ( \varphi )$, and $i\ge 1$.

\begin{itemize}

\item $( \epsilon , {\cal R} , i ) \models \true$.

\item $( \epsilon ,  {\cal R} , i )
 \models p_k ( a ) $ if $R_k ( a, i )$, where $a$ denotes a constant from  $d(p)$.

\item $( [ x \mapsto a ] , {\cal R} , i ) \models p_k ( x )$ if $R_k ( a , i )$.

\item $( \gamma,  {\cal R} , i ) \models ( \varphi \wedge \psi ) $~if~$(
\gamma |_{\vars  ( \varphi )} , {\cal R} , i ) \models \varphi$~and  \\
$( \gamma |_{\vars ( \psi ) } , {\cal R} , i ) \models \psi$.

\item $( \gamma , {\cal R} , i ) \models \neg \varphi$ if not $( \gamma , {\cal R} , i ) 
\models \varphi$.

\item $( \gamma , {\cal R} , i) \models \bigcirc \varphi$ if 
$(\gamma , {\cal R} , i + 1) \models
\varphi$.

\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \until \; \psi )$ if
for some $j > i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
\models \psi $ and for all $i \leq k < j $,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.

\item $( \gamma , {\cal R} , i) \models \ominus \varphi$ if $i>1$ and
$(\gamma , {\cal R} , i-1) \models
\varphi$.

\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \since \; \psi )$ if
for some $1 \le j \leq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
\models \psi $ and for all $j < k \leq i$,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.

\item $( \gamma , {\cal R} , i) \models \exists x \; \varphi$ if
there exists $a \in \domain ( x )$ such that\footnote{$\gamma \, [ x \mapsto a ]$
is the overriding of $\gamma$ with the binding $[ x \mapsto a ]$.}
$( \gamma \, [ x \mapsto a ], \sigma , i) \models \varphi$.

\end{itemize}

\iffalse
The definition of the {\em since} operator $S$ can be simplified in a standard
way such that it refers only to the positions $i$ and $i-1$ in
the sequence $\sigma$. This is based on the
fact that according to the semantics of {\em since},
$(\varphi \, \since \, \psi) = ( \psi \vee (\varphi \wedge \ominus 
(\varphi \, \since \, \psi )))$.
This will serve in the
implementation to work with only two versions of
the sets of assignments, for the current and previous state:

\begin{itemize}
\item $( \gamma , {\cal R} , i ) \models ( \varphi \, \since \, \psi )$ if
$( \gamma |_{\vars ( \psi )} , {\cal R} , i ) \models \psi$ or $i>1$,
$( \gamma |_{\vars ( \varphi )} , {\cal R} , i ) \models \varphi$, and \\
$( \gamma , {\cal R} , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}
\fi

%We can also define the following useful operators:
%$P \varphi = (\true \, \since \, \varphi)$ (for ``previously''),
%$( \varphi \, R \, \psi ) = \neg ( \neg \varphi \, \since \, \neg \psi )$ 
%(the dual of the Since operator), and
%$H \varphi = ( \false \, R \, \varphi )$ (for ``always in the past'').

%We define $\mathbf{F}$ and $\mathbf{T}$ as special constants
%in order to interpret formulas without free variables.
%They are the complement of each other.
%When combined with set operators (complementation, union, intersection),
%$\mathbf{F}$ behaves as the empty set, hence is idempotent to
%union. Accordingly, $\mathbf{T}$ is idempotent to intersection.

\iffalse
We denote by $A_{\vars ( \varphi )}$ the set of all possible assignments
of values to the variables that appear free
in $\varphi$. Thus,
$I [ \varphi, \sigma , i] \subseteq A_{\vars ( \varphi )}$.
To simplify definitions, we add
a dummy position $0$ for sequence $\sigma$ (which starts with $s_1$), 
where every formula is interpreted as an empty set.
Observe that the value $\emptyset$ and $\{ \epsilon \}$, behave
as the Boolean constants $0$ and $1$, respectively.
The set semantics is defined as follows, where $i \ge 1$.

\begin{itemize}
\item $I [ \varphi , \sigma , 0 ] = \emptyset$.
\item $I [ \true , \sigma , i ] = \{ \epsilon \}$.
\item $I [ p ( a ) , \sigma , i ] =$ if $p ( a ) \in \sigma [ i ]$ then
$\{ \epsilon \}$ else $\emptyset$.
\item $I [ p ( v ) , \sigma , i ] = \{ [ v \mapsto a ] \; | \; p ( a ) \in
\sigma [ i ] \}$.
%\item $I [ \true , \sigma , i ] = D^n$.
%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
I [ \varphi , \sigma , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
\item $I [ \neg \varphi , \sigma , i ] = 
A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , \sigma , i ]$.
\item $I [ ( \varphi \; \since \; \psi ) , \sigma , i ] = 
I [ \psi , \sigma , i ] \; \bigcup \;
( I [ \varphi , \sigma , i ] \; \bigcap \; 
I [ (\varphi \since \psi ) , \sigma , i - 1 ] )$.
\item $I [ \ominus \varphi , \sigma , i ] = I [ \varphi , \sigma , i-1 ]$.
\item $I [ \exists x \; \varphi , \sigma , i ] = 
\proj ( I [ \varphi , \sigma , i ] , \{ x \} )$.
\end{itemize}

\noindent
As before, the interpretation for the rest of the operators can
be obtained from the above using the connections between the operators,
e.g., $I [ P \, \varphi  , \sigma , i ] = 
I [ ( \true \, \since \, \varphi ) , \sigma , i ]$.
The correspondence between this set based semantics 
and the previous semantics, namely that
$\gamma \in I [ \varphi , \sigma, i ]$ iff
$( \gamma , \sigma , i ) \models \varphi$
can be proved by a simple structural induction on
the size of the formulas.
 \fi

Then for an FLTL (PFLTL) formula with no free variables, ${\cal R} \models \varphi$ iff $( \epsilon , {\cal R} , 1 ) \models \varphi$.
We will henceforth sometimes abuse notation, as is very often done in logic, and use the same
symbols both for the relations (semantics) and their representation in
the logic (syntax).

\section{Extending First-Order Temporal Logic}

%%this is specification formalism that is very expressive and yet monitorable under run time verification~\cite{HPU}.

We first demonstrate that the lack of expressiveness
carries over from LTL (PLTL) to FLTL (PFLTL).

\noindent {\bf Example 1.}
Let $p$ and $q$ be temporal reations, where their
restrictions to the state $s_i$ is $p[i]$ and $q[i]$, respectively, are unary relations $q(x)$ and $p(x)$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). This is an extension of
Wolper's example over LTL, carried over to the first-order case. To show that this is not expressible in FLTL (and PFLTL),
consider models (executions) where only one data element $a$ appears. We can prove that in this case, by a simple structural induction on (sub)formulas, that for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, respectively. However, by~\cite{Wolper},
such a formula cannot express the said property.

This example demonstrates that the so called ``non-counting''~\cite{Thomas} property of LTL carries over
to FLTL. This lack of expressiveness can be resolved by
adding propositional variables (i.e., predicates that are
only state dependent) with dynamic quantification over them.
Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015} can also express this property.

\noindent {\bf Example 2.}
%%%%

Extending FLTL (PFLTL) with quantifiers over relations, we obtain QFLTL
(and the past-restricted version QPFLTL, respectively). 
The syntax includes $\Exists p \, \varphi$, 
where $p$ denotes an auxiliary relation.
We also allow $\Forall p \varphi = \neg \Exists \neg \varphi$. The semantics is as follows.

\begin{itemize}
\item $(\gamma ,  {\cal R}, i) \models \Exists q \, \varphi$ iff there exists
${\cal R}'$ such that ${\cal R}' |_{\{ q \} } = {\cal R}$ and 
$( \gamma, {\cal R}' , i) \models \varphi$.
\end{itemize}
Consequently, quantification over relations effectively extends the model ${\cal R}$ into a model ${\cal R}'$ within the scope of the quantifier. Note that quantification here is dynmaic (as in QLTL and QPLTL).

We now extend FLTL into EFLTL in a way that was motivated by the the propositional extension from LTL (PLTL) to ELTL (PELTL). We thus allow the following formula:
\begin{equation}
\label{EFLTL}
\psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \end{equation}
\begin{enumerate}
\item $\psi$, the {\em statement}, is a FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formula with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relations with $2$ parameters; the first parameter is of the same 
type as $x$ and the second one is, as usual in this paper,
a natural number that is omitted in the temporal formulas. 
An auxiliary relations $r_j$ can appear within $\psi$. It can also appears in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, only within the
scope\footnote{This restricted has the same motivation as
the propositional version ELTL (and EPLTL): to avoid making the
formula $\false$ just by means of the definition of the auxiliary relations.}
of a previous-time operator $\ominus$. 
\end{enumerate}

Again, it
is simple to extend the syntax and semantics to include
relations $r_j$ with different number of arguments, and, correspondingly, formulas $\varphi_j$ with corresponding number of free
variables\footnote{We chose not to do that here (and in the definition of the semantics of FLTL above), because it
requires multi-indexing notation that is trivial and adds no further insight. Our implementation indeed uses
this more general capabilities.}.
The logic EPFLTL is obtain by restricting the temporal modalities of EFLTL to the past ones:
$\since$ and $\ominus$, and those obtained from them.

\begin{lemma} \label{samesame}
The auxiliary temporal relations $r_1, \ldots r_n$ are
uniquely defined from ${\cal R}$.
\end{lemma}
\noindent {\bf Proof.} by a simple induction, as in Lemma~\ref{fourone}. \qed

The following formula expresses the property that was shown to be 
not expressible using FLTL, at the beginning of this section.
\begin{equation}
\Box \forall x \, (r(x)\rightarrow p(x)) \mathrm{\ where\ }
r(x) = ( q(x) \leftrightarrow \ominus \neg  r(x)) 
\label{eq:wolper-first-order}
\end{equation}

We define\footnote{This is done for lack of space. Formal semantics 
can be given by constructing a set of temporal relation extended
with the auxiliary ones
inductively, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL specification by using the following equivalent
QFLTL formula:
\begin{equation} \label{equiv}
\Exists r_1  \, \ldots \, \Exists r_n \,  ( \psi \wedge \Box \bigwedge_{j \in \{ 1, \ldots , n\}} (r_j ( x_j )  \leftrightarrow 
\varphi_i ( x_j) )
\end{equation}

The analogous of Lemma~\ref{simplecase} shows that EPFLTL does not
need the past temporal operators besides $\ominus$.
\begin{lemma}
The full expressive power of EPFLTL can be obtained by allowing only 
specifications of the form
\[ \Box \psi \mathrm{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \]
where $\psi$ consists of a fully quantified single occurrence of some
auxiliary relation $r_j$.
\end{lemma}
\noindent {\bf Proof.} Basically, the proof is the same as Lemma~\ref{simplecase}, based on the axiomatization of the
past temporal operators. \qed

%\klaus{Something seems wrong above.}

%\klaus{Is the formula above really correct?}

We showed in Theorem~\ref{theo1} that although ELTL
can be translated into QLTL, both have the same
expressive power. We study now the relationship between
EFLTL and QFLTL.

\begin{theorem}
\label{theo2}
The expressive power of EFLTL is stricktly weaker than that of QFLTL.
\end{theorem}
The proof of this theorem, which includes the encoding of a nontrivial property, and a combinatorial argument about the amount of information
needed after a given prefix, appears in an appendix.
The relative expressive power between the first-order temporal logics discussed here
appear in Figure~\ref{fsfr}.

It follows from~(\ref{equiv}) that EFLTL (EPFLTL, respectively) has the same
expressive power as QFLTL (EPFLTL, respectively) restricted to having its
dynamic 
quantification to being existential, and prefixing the
rest of the formula. Thus, we immediately obtain from Theorem~\ref{theo2} the following:

\begin{corollary}
Constraining the dynamic quantification of QFLTL (QPFLTL) to existential quantification prefixing the
formula strictly weakens its expressive power.
\end{corollary}

\begin{figure}
\begin{center}
\includegraphics[height=1.1in,width=2.7in]{FIRST.jpg}
\caption{\label{fsfr} An arrow from A to B means that logic  B is more expressive than A}
\end{center}
\end{figure}

\section{A Runtime Verification Algorithm for EPFLTL}
\label{EPLTLRV}

Runtime verification of FLTL is performed on an
input that consists of {\em events} in the form of
tuples of relations. In our notation, the input
consists of a sequence ${\cal R}[1] \, {\cal R}[2]  \ldots$,
which we earlier identified with states, where each
${\cal R}[i]$ consists of some relations. A typical use
of runtime verification restricts these relations to have,
per each state just some finite tuples, and, quite commonly, have a single tuple from a single relation per each state.
A set of assignments over the same variables, which has some fixed order
(say alphabetical) also forms a relation.

Consider again the EPFLTL form
of specification, described in~(\ref{EFLTL}), restricted to the past operators. As syntax and semantics of various versions
of LTL are given in this paper, we repeat now the concrete definitions for EPFLTL. The syntax is as follows:

\begin{center}
$\varphi ::= \true  \; | \;
    p ( a ) \; | \;
    p ( x ) \; | \;
    ( \varphi \wedge \varphi ) \;  |   \;
   \neg \varphi \; | \;
    \ominus \varphi \; | \;
    ( \varphi  \; \since  \; \varphi ) \; | \;
    \exists x \; \varphi$
\end{center}
\[ 
\psi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
\]

\iffalse
\noindent {\bf Set Semantics.}
We now refine the semantics of the logic. Under the new definition, 
$I [ \varphi , \sigma, i ]$ is a function that returns
a set of assignments such that $\gamma \in I [ \varphi , \sigma, i ]$ 
iff $( \gamma , \sigma , i ) \models \varphi$.
This redefinition will later lead to a simple implementation 
using BDDs, where each set of assignments will be represented
as a BDD, and the Boolean operators will correspond directly 
to Boolean operators on BDDs.

In order to deal with subformulas with different sets of free
variables (hence, different domains for assignments), 
we apply a projection and an extension operator to assignments
over a subset of the variables. Let $\Gamma$ be
a set of assignments over the variables $W$, and
$U \subseteq W$.
Then $\proj ( \Gamma , U )$ (for ``projecting {\em out}'' or  ``hiding''
the variables $U$)
is the largest set of assignments over 
$W \setminus U$, 
each agreeing with some
assignment of $\Gamma$ on all the variables in $W \setminus U$.
Let $U \cap W = \emptyset$, then
$\ext ( \Gamma , U )$ is the largest set of assignments
over $W \cup U$, where each such assignment agrees 
with some assignment in $\Gamma$ on the
values assigned to the variables $W$. 
This means that we extend $\Gamma$ by
adding arbitrary values to the variables in $U$ from their domains.
Then $\proj ( \ext ( \Gamma , U ) , U ) = \Gamma$ holds.
We define the union and intersection operators on sets of
assignments, even if they are defined over non identical
sets of variables. 
In this case, the assignments are extended
over the union of the variables. Thus, if $\Gamma$ is a
set of assignments over $W$ and $\Gamma'$ is
a set of assignments over $W'$, then
$\Gamma \; \bigcup \;  \Gamma'$ is defined as $\ext (\Gamma , W' \setminus W ) \cup
\ext ( \Gamma' , W \setminus W' )$ and
$\Gamma \; \bigcap \; \Gamma'$ is $\ext (\Gamma , W' \setminus W ) \cap
\ext ( \Gamma' , W \setminus W' )$.  Hence, both are defined
over the set of variables $W \cup W'$.

%We define $\mathbf{F}$ and $\mathbf{T}$ as special constants
%in order to interpret formulas without free variables.
%They are the complement of each other.
%When combined with set operators (complementation, union, intersection),
%$\mathbf{F}$ behaves as the empty set, hence is idempotent to
%union. Accordingly, $\mathbf{T}$ is idempotent to intersection.
\fi

We denote by $A_{\vars ( \varphi )}$ the set of all possible assignments
of values to the variables that appear free
in $\varphi$. Thus,
$I [ \varphi, {\cal R} , i] \subseteq A_{\vars ( \varphi )}$.
To simplify definitions, we add
a dummy position $0$ for sequence $\sigma$ (which starts with $s_1$), 
where every formula is interpreted as an empty set.
Observe that the value $\emptyset$ and $\{ \epsilon \}$, behave
as the Boolean constants $0$ and $1$, respectively.
The set semantics is defined as follows, where $i \ge 1$.

\begin{itemize}
\item $I [ \varphi , {\cal R} , 0 ] = \emptyset$.
\item $I [ \true , {\cal R} , i ] = \{ \epsilon \}$.
\item $I [ p ( a ) , {\cal R} , i ] =$ if $p ( a ) \in \sigma [ i ]$ then
$\{ \epsilon \}$ else $\emptyset$.
\item $I [ p ( x ) , {\cal R} , i ] = \{ [ x \mapsto a ] | p ( a ) \in
\sigma [ i ] \}$.
%\item $I [ \true , \sigma , i ] = D^n$.
%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
\item $I [ \neg \varphi , {\cal R} , i ] = 
A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , {\cal R} , i ]$.
\item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
I [ \psi , {\cal R}, i ] \; \bigcup \;
( I [ \varphi , {\cal R} , i ] \bigcap 
I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
\item $I [ \ominus \varphi , {\cal R} , i ] = I [ \varphi , {\cal R} , i-1 ]$.
\item $I [ \exists x \; \varphi , \sigma , i ] = 
\proj ( I [ \varphi , {\cal R} , i ] , \{ x \} )$.
\item $I [ \psi \mathrm{\ where\ } r_j ( x_j) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} , {\cal R} , i ] = 
 I [ \psi , {\cal Q} ] $ where ${\cal Q} = {\cal R} \cup
\{ r_j | r_j [ i ] =  I [ 
\varphi_j , {\cal Q} , i ] \}$
\end{itemize}
In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame}.

We describe an algorithm for monitoring 
FLTL properties, first presented in~\cite{HPU} and implemented
in the tool~\dejavu. 
Instead of storing the data values occurring in events, 
we enumerate these data values as soon as we see them and use Boolean
encodings of this enumeration.
For example, if the runtime verifier sees the input 
events 
$\mathit{open}(\text{``a''})$, 
$\mathit{open}(\text{``b''})$, 
$\mathit{open}(\text{``c''})$, 
it will encode them as
$000$, $001$ and $010$ (say, we use 3 bits $b_0$, $b_1$ and $b_2$
to represent each enumeration, with $b_2$ being the most significant bit).

%

Since we want to be able to deal with infinite domains
(where only a finite number of elements may appear in a given observed prefix) and maintain the ability to perform
complementation, unused enumerations represent the
values that have not been seen and their relations
to all other values. The algorithm maintains that these unseen values are correctly represented, as shown in~\cite{HPU}. In fact, it is sufficient to have just one enumeration representing these values per each variable of the LTL formula. However, it is important to guarantee that at least one such enumeration
exists. We guarantee that by preserving the enumeration $11\ldots11$, i.e., the highest Boolean representation.
We present here only the basic algorithm. For versions that
allow extending the number of bits used for enumerations and garbage collection of enumerations, consult~\cite{HP}.

A Boolean representation of the {\em set} of values 
$\{\text{``a''},\text{``c''}\} $ would be equivalent to a Boolean function $(\neg b_0 \wedge \neg b_2)$ that returns 1 for $000$ and $010$ (the value
of $b_1$ can be arbitrary).

    Our preferred c of sets of values and
assignments is as an Ordered Binary Decision
Diagram (OBDD, although we write simply BDD)~\cite{Bryant}.
A BDD is essentially a compact representation 
of a Boolean tree, where compaction glues together isomorphic subtrees. Each non-leaf node is labeled with one of the
Boolean variables $b_0,\ldots,b_{k-1}$. A non-leaf node $b_i$ 
is the source of two 
arrows leading to other nodes. 
%A dotted-line arrow represents that $b_i$ has the Boolean value $0$, while a thick-line 
%arrow represents that it has the value $1$. 
The nodes in the DAG have the
same order along all paths from the root. However, some of the nodes may be
absent along some paths, when the result of the Boolean function does not 
depend on the value of the corresponding Boolean variable. Each path leads 
to a leaf node that is marked by either a $0$ or a $1$, representing the 
Boolean value returned by the function for the Boolean values on the path.

The use of BDDs can be replaced by other representations that
can compactly and efficiently represent sets of values (e.g., ZDDs) and
apply set operations like complementation, intersection,union (which are simply negation, conjunction and disjunction
in BDDs) and projection (for the quantification).
The marriage of BDDs and Boolean enumeration is in particular
efficient, since collections of adjacent Boolean enumerations tend to compact well.

\iffalse %%%%%
The definition of the {\em since} operator $S$ can be rewritten in a standard
way such that it refers only to the positions $i$ and $i-1$ in
the sequence $\sigma$. 
This is based on the
fact that according to the semantics of {\em since},
$(\varphi \, \since \, \psi) = ( \psi \vee (\varphi \wedge \ominus 
(\varphi \, \since \, \psi )))$.
This will serve in the
implementation to work with only two versions of
the sets of assignments, for the current and previous state:

\begin{itemize}
\item $( \gamma , \sigma , i ) \models ( \varphi \, \since \, \psi )$ if
$( \gamma |_{\vars ( \psi )} , \sigma , i ) \models \psi$ or $i>1$,
$( \gamma |_{\vars ( \varphi )} , \sigma , i ) \models \varphi$, and 
$( \gamma , \sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}
\fi %%%%%

%The rest of the operators are defined as follows:
%$\false = \neg \true$, $\forall x \; \varphi = \neg \exists x \neg \varphi$,
%$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$.
%We can also define the following useful operators:
%$P \varphi = (\true \, \since \, \varphi)$ (for ``previously''),
%$( \varphi \, R \, \psi ) = \neg ( \neg \varphi \, \since \, \neg \psi )$ 
%(the dual of the Since operator), and
%$H \varphi = ( \false \, R \, \varphi )$ (for ``always in the past'').

%\noindent {\bf Set Semantics.}
%It helps to represent the algorithm to 
%redefine the semantics of
%the logic in terms of sets of assignments satisfying a formula.
%Then $I [ \varphi , \sigma, i ]$ is a function that returns
%a set of assignments such that $\gamma \in I [ \varphi , \sigma, i ]$ 
%%Then a BDD for a set containing the values ``de'' and ``af''
%%(2nd and 3rd values)
%will return $1$ for $01$ and $10$. If the Boolean function is over
%$b_0$ (for most significant bit) and $b_1$ (for least significant), 
%then this is the Boolean function 
%$(\neg b_0 \wedge b_1) \vee (b_0 \wedge \neg b_1)$.

Given some ground predicate $p ( a )$ observed in the execution 
matching with $p ( x )$ in the monitored property,
let $\lookup ( x, a )$ be the enumeration of $a$. If this
is $a$'s first occurrence, then it will be assigned a new enumeration.
Otherwise, $\lookup$ returns the enumeration that $a$ received before. We can use a counter, for each variable $x$, counting the number of different values appearing so far for $x$. When a new value appears, this counter is incremented, and the value is converted to
a Boolean representation. Enumerations that were not yet used represent
the values not seen yet. 
%In particular, we always leave at least
%one enumeration, $11\ldots 11$, for this purpose.
In the next section we introduce data reclaiming, which allows 
reusing enumerations for values that no longer affect the checked property. 
This involves a more
complicated enumeration mechanism.

The function $\bddset (x, A)$ returns
a BDD that represents the set of assignments where $x$ is mapped to 
(the enumeration of) $v$ for
$v \in A$. This BDD is independent
of the values assigned to any variable
other than $x$, i.e., they can have any value.
For example, assume that we use three Boolean variables (bits) $x_0$, $x_1$ and $x_2$
for representing enumerations over $x$ (with $x_0$ being the least significant bit), and
assume that $A = \{ a , b \}$, 
$\lookup ( x, a ) = 011$, and $\lookup ( x , b ) = 001$.
Then $\bddset ( x , A )$ is a BDD representation of the Boolean function 
$x_0 \wedge \neg x_2$. 
%But the BDD function will be independent of the values of Boolean variables
%representing other variables.
%Notice that $p ( x )$ is
%equivalent, and is represented as $\exists y \, \exists z \, p ( x )$
%when the variables used in the formula are $x$, $y$ and $z$.

Intersection and union of sets of assignments are translated simply
to conjunction and disjunction of their BDD representation,
respectively, and complementation
becomes BDD negation. We will denote
the Boolean BDD operators as $\bddwedge$, $\bddvee$ and $\bddneg$.
To implement the existential (universal, respectively) operators, 
%as in the interpretation of $\exists x \; \varphi$, 
we use the BDD existential (universal, respectively) operators over
the Boolean variables that represent (the enumerations of) the values of $x$. 
%Thus, $x$ is translated into
%is represented using the Boolean variables $x_1$, $x_1, \ldots x_{k}$ into 
Thus, if $B_{\varphi}$ is the BDD representing
the assignments satisfying $\varphi$ in
the current state of the monitor, then 
$\bddexists (\langle x_0,\ldots,x_{k-1} \rangle, B_{\varphi})$
is the BDD that represents the assignments satisfying $\exists x \; \varphi$ in the current
state.
Finally, $\bfalse$ and $\btrue$  are the BDDs that return always
$0$ or $1$, respectively.

%The case where the number of bits used to represent values for
%some domains is eventually insufficient is dealt with by
%expanding the number of bits for the BDD. This will be explained later,
%but is much harder to implement, and may require a lot of overhead
%to update the BDDs while online computation is monitored. This
%may be unnecessarily, as we can assign a large number of bits
%a priory, where these bits represent a set of values that is
%exponentially larger.

The algorithm shown below works similarly to the algorithm for the 
propositional case shown in
Section~\ref{LTLruntime}. That is,
it operates on two vectors (arrays) of values indexed by subformulas: 
$\old$ for the state before that event, and
$\current$ for the current
state (after the last seen event).
However, while in the propositional case the vectors contain Boolean values, here
they contain BDDs. The temporal relations are
represented by
sets of events, each being a tuple of one of the relations. Thus, $R_k [ i ]$ refers to the set
of tuples of the relation $R_k$ at time instant $i$. Often (and in our implementation), the set of events consists of a singleton tuple of a single relation.
The algorithm follows.

\begin{enumerate}
\item Initially, for each subformula $\varphi$,
%$B ( \varphi , \current ) = \bfalse$.
$\current ( \varphi ) := \bfalse$.
\item Observe a new state (as a set of ground predicates) $s_i$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.
\begin{itemize}
  \item $\current ( \true ) := \btrue$.
  \item $\current ( p_k ( a ) )$ := if $R_k [ i ] ( a )$ then
  $\btrue$ else $\bfalse$.
  \item $\current ( p_k ( x ) )$ :=
     $\bddset ( x ,  \{ a | R_k[i] ( a  ) \} )$.
  \item $\current ( ( \varphi \wedge \psi ) ) := 
  \bddwedge ( \current ( \varphi ) , \current ( \psi ))$.
  \item $\current ( \neg \varphi  ) := \bddneg ( \current ( \varphi ))$.
  \item $\current ( ( \varphi\ \since\ \psi ) ) := \\ 
  \bddvee ( \current ( \psi  ) , \bddwedge ( \current ( \varphi ) ,
      \old ( ( \varphi \since  \psi ))))$.
%DAP: I removed the \; around \since, since it penetrated the 2nd column!
  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
  \item $\current ( \exists x \; \varphi ) := 
  \bddexists (\langle x_0,\ldots,x_{k-1} \rangle, \current ( \varphi ))$.
\end{itemize}
\item Goto step~2.
\end{enumerate}

We extend now the algorithm to capture EPFLTL.
The auxiliary relations $r_j$ extend the model, and we need to keep 
a BDDs representing $\current ( r_j )$ and
$\old (r_j)$ for each relation $r_j$. We also need to calculate the subformulas $\varphi_i$ that appear in
a specification, as part of the runtime verification,
as per the above FPLTL algorithm. The auxiliary relations $r_j$ are calculated with respect to
a variable $x_j$, which is the one appearing
free in $\varphi_j$(this can be generalized to any number of variables). However, $r_j$ can be used
as a subformula with other parameters in other rules or in the statement. Thus,
$r_j ( y )$ constructing a relation
that is the same as $r_j$, but over the variable $y$. This can be achieved by BDD renaming function $\rename ( r_j (x_j ) , y)$, which is easy to implement in BDDs.

We then have to add
following updates between steps 3 and 4 of the above algorithm, when performing runtime verification of
$\psi$.

\noindent \begin{description}
\item For each $r_j ( x_j ) = \varphi_j ( x_j )$:
\item{\ \ }  calculate $\current ( \varphi_j )$.
\item{\ \ }  $\current ( r_j )  := \current (\varphi_j)$
\item{\ \ } $\current ( r_j ( y ) )$ := $\rename ( r_j(x_j) ,  y )$.
\item{\ \ } $\current ( r_j ( a ) )$ := if $\current ( r_j ) ( a )$
then $\btrue$ 
\item{\ \ \ \ \ \ \ \ \ } else $\bfalse$.
\end{description}

As in the propositional case, the evaluation order cannot be
simply top down or bottom up, since relations can appear
both on the left and the right of a definition
such as $r ( x ) := p ( x ) \vee \ominus r ( x )$. For that,
we need to use the mixed evaluation order, described in Section~\ref{LTLruntime}.

### 2019-02-02 rule-ltl.tex
fixing problems
Updates from ShareLaTeX

@@
 execution traces against formal specifications. 
-These include e.g. \cite{Meredith2011} \cite{larva} \cite{Reger2015} \cite{halle-beepbeep-ieee-12} \cite{LOLA}.
+These include e.g. \cite{Meredith2011,larva,Reger2015,halle-beepbeep-ieee-12,LOLA}.
 
@@
 \item{F} allows first-order quantification over
-variables and relations over variables and constants.
+variables and relations over variables and constants. This is referred to as {\em static} quantification.
 \item{Q} adds quantification over predicates (for the propositional versions)
-or over relations (for the first-order version). These objects can have different values at different states.
+or over relations (for the first-order version). These objects can have different values at different states. This is referred to as {\em dynamic} quantification.
 \item{E} allows extending the objects that occur in the input with
@@
 \end{itemize}
+
+\klaus{You above introduce some concepts before they really are explained. I wonder whether the reader will understand. Specifically the E, but also the Q. Why not just introduce as we go along?}
 

### 2019-02-02 rule-ltl.tex
fixing problems

@@
 
-\noindent {\bf Related work.}
-%
+\subsubsection{Related Work}
+
 For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
@@
 
-\klaus{You above introduce some concepts before they really are explained. I wonder whether the reader will understand. Specifically the E, but also the Q. Why not just introduce as we go along?}
+\klaus{You above introduce some concepts before they really are explained.
+Q and E.
+I understand why you do this (to have it in one place), but it may be too difficult for the reader to grasp if not "initiated".
+Especially the E.}
+
+\klaus{We use ELTL (etc) to denote Extended LTL. 
+But QLTL is also an extension of LTL. These logics 
+are all extensions of LTL. Perhaps RLTL (for Rule-based 
+LTL) would have been better ... or something. Not urgent, 
+we can always change that if we want to.}
 
@@
 
+\subsubsection{Quantified LTL}
+
 Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
@@
 
-\noindent {\bf Evaluation order.} The order of calculating $\current$ is subtle here, because
+\subsubsection{Evaluation Order}
+
+The order of calculating $\current$ is subtle here, because
 auxiliary variables can appear both on the left and right of their
@@
 
-\noindent {\bf Syntax.} 
+\subsubsection{Syntax}
+
 The formulas of the core FLTL logic are 
@@
 
-\noindent {\bf Semantics.} 
+\subsubsection{Semantics}
+
 A model is a set of
  relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$, where each of these relations has a last parameter $i$ that is a natural number. 
-
+% 
 It it standard, and more intuitive (in particular for runtime verification) to cast models of temporal logic in terms of sequences states rather than temporal relations. To do that,

### 2019-02-02 rule-ltl.tex
fixing problems

@@
 
-\section{Propositional Linear Temporal Logic}
+\section{Propositional LTL}
 
@@
 
-\section{Making propositional LTL more Expressive}
+\section{Extending Propositional LTL}
 
@@
 
+\klaus{Are the last two lines above needed?}
+
 This kind of quantification is considered to be ``dynamic'', since the quantified variables can have different truth values depending on the states, or, "second order", the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
@@
 
+\subsubsection{Rule-based LTL}
+
 We present now an extension of LTL, which we call ELTL. It has the same
@@
 It has a form that puts it, conceptually, in between temporal
-logic and automata: on the one hand, it is possible to simply represent
-a pure PLTL formula, and on the other hand, one can represent a specification that resembles a finite automaton. As will be showed later, this extension is
+logic and automata: on the one hand, it is possible to simply write
+a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. As will be showed later, this extension is
 is very natural for runtime verification.
@@
 Let $\varphi$ be an ELTL formula as in~(\ref{ELTL}),
-over the auxiliary propositions $A$ and basic propositions $B$ 
-and a model $\sigma$ with states over the $B$, such that
+over the auxiliary propositions $A$ and basic propositions $B$, 
+and let $\sigma$ be a model with states over the $B$, such that
 $\sigma \models \psi$.
@@
 
-%\klaus{I think this lemma needs to tie the trace to the formula. Doron: I do not understand this comment.}
+\klaus{(1) is this lemma saying the right thing? (2) is the proof really a proof of that lemma, (3) 
+should $\varphi$ be $\psi$, (4) now $\psi$ is not the whole spec (it is missing the rules). What do we call the whole thing? It is not a formula like the 
+$\varphi$s and the $\psi$s. Do we need a special greek symbol for it?}
 
@@
 is expressible in QLTL using the following formula:
-\[   \Exists a_1 \ldots \Exists a_n  ( \varphi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
+\[   \Exists a_1 \ldots \Exists a_n  ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
 \]
@@
 
-\section{Runtime Verification for PLTL and its extensions}
+\section{RV for Propositional Past-Time LTL and its Extension}
 \label{LTLruntime}
@@
 
-\section{First-Order Linear Temporal Logic FLTL}
+\section{First-Order LTL}
 
@@
 
-\section{Extending First-Order Temporal Logic}
+\section{Extending First-Order LTL}
 
@@
 
-\section{A Runtime Verification Algorithm for EPFLTL}
+\section{RV for First-Order Past-Time LTL and its Extension}
 \label{EPLTLRV}

### 2019-02-02 rule-ltl.tex
fixing problems

@@
 is expressible in QLTL using the following formula:
+
 \[   \Exists a_1 \ldots \Exists a_n  ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
 \]
-For the other direction, one can translate any QLTL property into an $\omega$ automaton\footnote{The past operators of a QLTL property do not change this fact, since the translation can be done first to second order monadic logic, which can be translated into an $\omega$-automaton}.
-We assume that the property is given as a deterministic Muller automaton $M = (S, i , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $i \in S$ is the initial state, the set $\Sigma$ consists of the inputs, each represented as a Boolean combination of the variables $B$, $\Delta : S \times \Sigma \mapsto S$, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set,
+
+\klaus{How do we know that the ``other direction'' below is possible?}
+
+For the other direction, one can translate any QLTL property into an $\omega$ automaton\footnote{The past operators of a QLTL property do not change this fact, since the translation can be done first to second order monadic logic, which can be translated into an $\omega$-automaton.}.
+We assume that the property is given as a deterministic Muller automaton $M = (S, i , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $i \in S$ is the initial state, $\Sigma$ is the set of inputs, each represented as a Boolean combination of the variables $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set,
 where an accepted input needs to pass infinitely often in exactly
@@
 
+\klaus{The following formulation above is not not clear: ``where an accepted input needs to pass infinitely often in exactly
+one of the specified sets''.}
+
+\noindent
 Then we translate $M$ into an ELTL formula
-of the form \begin{tabbing}
-$\bigvee_{F \in {\cal F}} (\bigwedge_{s \in F} \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
+of the form:
+
+\begin{tabbing}
+\ \ \ \ \ $\bigvee_{F \in {\cal F}} (\bigwedge_{s \in F} \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
 where \\
-\ \ \ \ \ $a_s := \vee _{\Delta ( s'
+\ \ \ \ \ \ \ \ \ \ $a_s := \vee _{\Delta ( s'
 , \alpha ) = s } 
@@
 \end{tabbing}
+
+\klaus{Above: should $s=i \rightarrow \neg \ominus \true$ instead be (note: two changes):
+$\neg \ominus \true \rightarrow a_s=i$
+}
+
+\klaus{Above: can't judge whether the $\bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$ is correct.}
+
+\noindent
 With $(s=i)$ is $\true$ when $s$ is $i$, and $\false$ otherwise.
@@
 $\eta$ can be viewed as representing a
-state machine, in particular, in
-particular when the only modality that is used in the rules part is $\ominus$.
+state machine, in particular when the only modality that is used in the rules part is $\ominus$.
 In that case, we can consider an ELTL specification $\eta$ as consisting with two synchronous parts: an LTL property 
@@
 
+\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}
+
 We can extend PLTL into EPLTL by disallowing the future temporal operators (except a prefixing $\Box$, when interpreted over infinite sequences). We obtain
@@
 with the simpler formula $\vee_{s \in A} a_s$, where $A$ are the accepting states of the finite automaton.
+
+\klaus{Last sentence above: perhaps say with words that we are always in an acceptance state - if that's what the point is.}
 
@@
 axiomatization of QLTL in~\cite{MP}, e.g.,
-$\varphi \since \psi \leftrightarrow ( \psi \vee ( \varphi \wedge \varphi \since \psi) ) )$. \qed
+$\varphi\ \since\ \psi \leftrightarrow \psi \vee ( \varphi \wedge (\varphi\ \since\ \psi))$. \qed
+
+\klaus{In the axiom above, isn't there missing a $\ominus$ before
+the second $(\varphi\ \since\ \psi)$?}
+
+\klaus{Above: is the following really correct?: 
+``with $p$ being one of the auxiliary variables $a_j$''.}
 
@@
 
-Runtime verification of temporal temporal specification often
+Runtime verification of temporal specifications often
 concentrates on the past portion of the logic.
-Past specifications have the important property that one can distinguish when they are violated.
+Past-time  specifications have the important property that one can distinguish when they are violated.
 For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.
@@
 
-The runtime verification exploits the fact that
+\noindent
+Runtime verification exploits the fact that
 the semantic definition is recursive in both the structure of the property, where subformulas
@@
 
+\noindent
 Runtime verification of QPLTL can be done by translating a
-formula into a deterministic finite automaton.
-Then the sequence of states to be checked forms the input for this automaton.
-QPLTL can be extremely compact, in particular when using alternating existential and universal
+formula into a deterministic finite automaton. Then the sequence of states to be checked forms the input for this automaton. 
+However, QPLTL can be extremely compact, in particular when using alternating existential and universal 
 quantifiers, which can result in a huge (non-elementary) explosion of
@@
 In order to extend the algorithm to perform runtime verification of an EPLTL formula $\psi$, as defined in~(\ref{ELTL}),  we need to  add to the above
-algorithm memory for $\current ( a_j )$ and $\old ( a_j )$
-and perform, between steps 3 and 4, the following, per each $a_j$: \\
-
-\ \ \ \ -\ $\current ( a_j ) = \current ( \varphi_j )$ \\
+algorithm memory for $\current ( a_j )$ and $\old ( a_j )$,
+and perform the following between steps 3 and 4 for each $a_j$: \\
+
+\ \ \ \ \ $\current ( a_j ) := \current ( \varphi_j )$ \\
 

### 2019-02-02 rule-ltl.tex
fixing problems

@@
 The order of calculating $\current$ is subtle here, because
-auxiliary variables can appear both on the left and right of their
-definition. For PLTL, one can start calculating the subformulas
+auxiliary variables can appear recursively on the right-hand 
+side of their definition. For PLTL, one can start calculating the subformulas
 bottom up or top down. But for EPLTL, this would not work. To see this, consider, for example, the formula~(\ref{form3}). It contains
 the definition $q := \ominus \neg q$. Now, we cannot calculate this
-bottom up, since the left-hand $q$ is not calculated yet.
+bottom up, since the left-hand $q$ is not calculated yet, and we need its value to compute the right-hand side $q$ leaf node.
 If we calculate it top down, we will encounter the same problem.
@@
 
+\klaus{I don't see that we need to talk about the top-down evaluation.}
+
 Under {\em mixed evaluation order}, one calculates $\current$ in the following order. The calculation within each of the four categories can be performed top down or bottom up.
-\begin{itemize}
+
+\begin{enumerate}
 \item Calculate 
@@
 $\current ( a_j )$ to $\current ( \varphi_j )$ for each $j$.
-\item Calculate $\current ( \delta )$ for each subformulas $\delta$
-that appear in $\varphi_j$ of a rule
+\item Calculate $\current ( \delta )$ for each subformula $\delta$
+that appears in $\varphi_j$ of a rule
 $a_j = \varphi_j$ {\em  within} the scope of
 a $\ominus$ operator.
-\item Calculate $now ( \psi )$ (using $\current ( a_j )$) for
-the statement $\psi$.
-\end{itemize}
+\item Calculate $\current ( \delta )$ for each subformula $\delta$
+that appears in the statement $\psi$ - using $\current ( a_j )$.
+\end{enumerate}
 
@@
 the values $5$ to $x$ and the value ``abc'' to $y$.
-Let $T$ a set of {\em relation} 
+Let $T$ be a set of {\em relations} 
 with typical instances $R$.
@@
 
-We defined models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is,
+\klaus{Above: let T be ... Is T used later? The relations are used in the next paragraph, why not introduce them there? Also, relations are not defined really.}
+
+\klaus{I don't quite understand the motivation below for this way of modeling events: relations with a time stamp, and the whole story of the time stamp not being mentioned etc.
+Why not just the usual models: traces?
+Too late to change now for CAV though.}
+
+We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is,
 their last parameter

### 2019-02-02 rule-ltl.tex
fixing problems

@@
 \section{Extending Propositional LTL}
-
-\subsubsection{Quantified LTL}
+\label{sec:extending-prop-ltl}
+
+\subsubsection{Extending with Dynamic Quantification}
 
@@
 
-\subsubsection{Rule-based LTL}
+\subsubsection{Extending with Rules}
 
@@
 restricting it to the value $i$ in the last parameter. For simplicity, we will describe henceforth
-the logic with relations $R$ that have exactly two parameters
-, hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$; this domain is fixed over all the value of the second parameter $i$.
-The the definition of the logic over other number of parameters is quite straightforward, and our implementation of runtime verification fully supports it.
+the logic with relations $R$ that have exactly two parameters, 
+hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$; this domain is fixed over all the values of the second parameter $i$.
+The definition of the logic over other numbers of parameters is quite straightforward, and our implementation of runtime verification fully supports it.
+
+\klaus{Above, what about instead of writing: ``this domain is fixed over all the values of the second parameter $i$'' to write:
+``this domain is independent of the values of the second parameter $i$'' - if that is what you mean to say.}
+
+\klaus{Above: we need to say explicitly what $i$ intuitively is meant to represent: positions, time, ....}
 
@@
 defined by the following grammar,
-where $p$ denotes to a relation $R$,
+where $p$ denotes a relation $R$,
 $a$ is a constant representing a value in $d ( R )$ and $x$ is a variable with $\domain ( x ) = d ( p )$.
@@
    \bigcirc \varphi \; | \; 
-   ( \varphi \; \until \; \varphi ) \; | $ \\
+   ( \varphi \; \until \; \varphi ) \; | $ %\\
    $ \ominus \; \varphi \; | \;
@@
 forms the logic PFLTL.
-The quantification over values of variables here is `static' in the sense that
-the assigned value is independent of the state in the execution. We
+The quantification over values of variables here is `static' in the sense that values quantified over are independent of the state in the execution. We
 denote static quantification with $\exists$ and $\forall$, while
-dynamic quantification uses $\Exists$ and $\Forall$.
+dynamic quantification is expressed with $\Exists$ and $\Forall$
+(see e.g. Section \ref{sec:extending-prop-ltl}).
 
@@
 % 
-It it standard, and more intuitive (in particular for runtime verification) to cast models of temporal logic in terms of sequences states rather than temporal relations. To do that,
+It is standard, and more intuitive (in particular for runtime verification) to cast models of temporal logic in terms of sequences states rather than temporal relations. To do that,
  we denote by $R_k [ i ]$ the restriction of
  the relation $R_k$ to the value $i$ of the last parameter. A model ${\cal R}$ is then a sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$, where
- each {\em state} ${\cal R} [ i ] = {\cal R} [i] = \{ R_1 [i], \ldots , R_m [ i ] \}$  (a set of relations).
+ each {\em state} ${\cal R} [i] = \{ R_1 [i], \ldots , R_m [ i ] \}$  (a set of relations).
  The denotation $p_k$ will correspond to the relation $R_k$.
+
+\klaus{Above, did you not already say the following before:
+``we denote by $R_k [ i ]$ the restriction of
+ the relation $R_k$ to the value $i$ of the last parameter.''.}
 
@@
 
-\noindent {\bf Example 1.}
-Let $p$ and $q$ be temporal reations, where their
-restrictions to the state $s_i$ is $p[i]$ and $q[i]$, respectively, are unary relations $q(x)$ and $p(x)$. The
+\noindent {\bf Example.}
+Let $p$ and $q$ be temporal relations, where their
+restrictions to the state $s_i$, $p[i]$ and $q[i]$, respectively, are unary relations $p(x)$ and $q(x)$. The
 specification that we want to monitor is that for each value
@@
 such a formula cannot express the said property.
-
 This example demonstrates that the so called ``non-counting''~\cite{Thomas} property of LTL carries over
-to FLTL. This lack of expressiveness can be resolved by
+to FLTL.
+
+\subsubsection{Extending with Quantification}
+
+This lack of expressiveness can be resolved by
 adding propositional variables (i.e., predicates that are
@@
 Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015} can also express this property.
-
-\noindent {\bf Example 2.}
-%%%%
 
@@
 where $p$ denotes an auxiliary relation.
-We also allow $\Forall p \varphi = \neg \Exists \neg \varphi$. The semantics is as follows.
+We also allow $\Forall p\ \varphi = \neg \Exists \neg \varphi$. The semantics is as follows.
 
@@
 Note that quantification here is dynmaic (as in QLTL and QPLTL).
+
+\subsubsection{Extending with Rules}
 

### 2019-02-03 rule-ltl.tex
fixing problems

@@
 Finally, we provide some experimental results.
+
+\klaus{I think that the structure of the paper should be
+crystal clear so that the reader does not get lost. A lost reader leads to rejection. If a reader understand the structure, but perhaps not the details, the reader may still be happy.}
 
@@
 
+\klaus{Below, has the notation ${\cal R}' |_{\{ q \} }$ been defined?}
+
 \begin{itemize}
@@
 
-We now extend FLTL into EFLTL in a way that was motivated by the the propositional extension from LTL (PLTL) to ELTL (PELTL). We thus allow the following formula:
+We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We thus allow the following formula:
+
 \begin{equation}
@@
 \psi \mathit{\ where\ } r_j  ( x_j ) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n \end{equation}
+\varphi_j (x_j) : j \in \{ 1 , \ldots , n 
+\end{equation}
+
 \begin{enumerate}
@@
 no free variables, 
-\item $\varphi_j$ are PFLTL formula with a single
+\item $\varphi_j$ are PFLTL formulas with a single
 free variable $x_j$,
-\item $r_j$ is an auxiliary temporal relations with $2$ parameters; the first parameter is of the same 
-type as $x$ and the second one is, as usual in this paper,
+\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
+type as $x_j$ and the second one is, as usual,
 a natural number that is omitted in the temporal formulas. 
-An auxiliary relations $r_j$ can appear within $\psi$. It can also appears in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, only within the
-scope\footnote{This restricted has the same motivation as
+An auxiliary relation $r_j$ can appear within $\psi$. It can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
+scope\footnote{This restriction has the same motivation as in
 the propositional version ELTL (and EPLTL): to avoid making the
@@
 requires multi-indexing notation that is trivial and adds no further insight. Our implementation indeed uses
-this more general capabilities.}.
-The logic EPFLTL is obtain by restricting the temporal modalities of EFLTL to the past ones:
+this more general capability.}.
+The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
 $\since$ and $\ominus$, and those obtained from them.
@@
 
+\noindent
 We define\footnote{This is done for lack of space. Formal semantics 
-can be given by constructing a set of temporal relation extended
+can be given by constructing a set of temporal relations extended
 with the auxiliary ones
-inductively, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL specification by using the following equivalent
+inductively, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL 
+specification (\ref{EFLTL}) 
+by using the following equivalent
 QFLTL formula:
+
 \begin{equation} \label{equiv}
@@
 
+\noindent 
 The analogous of Lemma~\ref{simplecase} shows that EPFLTL does not
 need the past temporal operators besides $\ominus$.
+
 \begin{lemma}
@@
 \end{lemma}
+
+\klaus{Something seems wrong in the above formulation.
+E.g.: Is the $\Box$ a restriction?, what is ``fully quantified single occurrence of some auxiliary relation $r_j$''?}
+
 \noindent {\bf Proof.} Basically, the proof is the same as Lemma~\ref{simplecase}, based on the axiomatization of the
@@
 
-%\klaus{Something seems wrong above.}
-
 %\klaus{Is the formula above really correct?}
 
-We showed in Theorem~\ref{theo1} that although ELTL
-can be translated into QLTL, both have the same
+We showed in Theorem~\ref{theo1} that ELTL
+and QLTL have the same
 expressive power. We study now the relationship between
@@
 
+\klaus{In the text above there are some imprecisions, such as
+(1) ``tuples of relations'' - relations are sets of tuples, we don't have tuples of relations. 
+(2) ``${\cal R}[i]$ consists of some relations'': it is a set of
+relations (I think).
+(3) ``to have, per each state just some finite tuples'': tuples are finite. You mean: to contain a finite set of tuples?
+(4) what is the story about a set of assignments also being a relation? Needed?
+}
+
+\subsubsection{Syntax}
+
+\klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}
+
 Consider again the EPFLTL form
-of specification, described in~(\ref{EFLTL}), restricted to the past operators. As syntax and semantics of various versions
-of LTL are given in this paper, we repeat now the concrete definitions for EPFLTL. The syntax is as follows:
-
+of specification, described in~(\ref{EFLTL}), restricted to the past operators. We now present the definition of EPFLTL, the final main logic proposed in this paper. 
+The syntax is as follows:
+
+\[ 
+\psi \mathrm{\ where\ } r_j ( x_j) := 
+\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
+\]
 \begin{center}
@@
 \end{center}
+
+\klaus{Note that $\psi$ above is not defined as a syntactic category. It should really be $\varphi$. Likewise in other parts of the paper. Perhaps too big a change now? Also, there should be
+a greek letter for a specification. E.g.:
+
 \[ 
-\psi \mathrm{\ where\ } r_j ( x_j) := 
+\pi ::= \varphi \mathrm{\ where\ } r_j ( x_j) := 
 \varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
 \]
+\[
+\varphi ::= ...
+\]
+}
 
@@
 
+\klaus{One may of course ask whether the following section on the 
+set semantics is needed. Maybe it is needed, since we probably need a semantics. Something to consider.}
+
+\subsubsection{Semantics}
+
 We denote by $A_{\vars ( \varphi )}$ the set of all possible assignments
@@
 The set semantics is defined as follows, where $i \ge 1$.
+
+\klaus{Above: assignments are introduced without any gentle introduction.
+Remember: you are not communicating with Moshe Vardi.
+You are communicating with John Fingelstein who had three glasses of wine at dinner.}
+
+\klaus{Above: $I [ \varphi, {\cal R} , i]$ is introduced out of the blue without any explanation.}
 
@@
 \end{itemize}
+
+\klaus{Above: as the semantics is presented, it looks as if 
+the `$\psi\ where ...$' is just another subformula.}
+
+\klaus{Above: hide is not defined.}
+
+\klaus{Above: the semantics of $p(a)$ and $p(x)$ do not refer to 
+${\cal R} $.}
+
 In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame}.
 
-We describe an algorithm for monitoring 
-FLTL properties, first presented in~\cite{HPU} and implemented
+\subsubsection{Runtime Verification Algorithm for PFLTL}
+
+We start by describing an algorithm for monitoring 
+PFLTL properties, first presented in~\cite{HPU} and implemented
 in the tool~\dejavu. 
@@
 we enumerate these data values as soon as we see them and use Boolean
-encodings of this enumeration.
+encodings of this enumeration, represented as BDDs.
 For example, if the runtime verifier sees the input 
@@
 to represent each enumeration, with $b_2$ being the most significant bit).
+%
+A Boolean representation of the {\em set} of values 
+$\{\text{``a''},\text{``b''}\} $ would be equivalent to a Boolean function $(\neg b_1 \wedge \neg b_2)$ that returns 1 for $000$ and $001$ (the value
+of $b_0$ can be arbitrary).
 
@@
 However, it is important to guarantee that at least one such enumeration
-exists. We guarantee that by preserving the enumeration $11\ldots11$, i.e., the highest Boolean representation.
+exists. We guarantee that by preserving the enumeration $11\ldots11$, i.e., the highest Boolean representation,
+to represent values not yet seen.
 We present here only the basic algorithm. For versions that
@@
 
-A Boolean representation of the {\em set} of values 
-$\{\text{``a''},\text{``c''}\} $ would be equivalent to a Boolean function $(\neg b_0 \wedge \neg b_2)$ that returns 1 for $000$ and $010$ (the value
-of $b_1$ can be arbitrary).
-
-    Our preferred c of sets of values and
-assignments is as an Ordered Binary Decision
+Our preferred representation of a set of values 
+(assignments) is as an Ordered Binary Decision
 Diagram (OBDD, although we write simply BDD)~\cite{Bryant}.
@@
 of a Boolean tree, where compaction glues together isomorphic subtrees. Each non-leaf node is labeled with one of the
-Boolean variables $b_0,\ldots,b_{k-1}$. A non-leaf node $b_i$ 
+Boolean variables $b_0,\ldots,b_{k-1}$ (where $k$ is the number of bits used to represent the values of a variable). A non-leaf node $b_i$ 
 is the source of two 
-arrows leading to other nodes. 
+arrows leading to other nodes, representing respectively that the node has the value 0 (false) or 1 (true).
 %A dotted-line arrow represents that $b_i$ has the Boolean value $0$, while a thick-line 
@@
 
+\subsubsection{Runtime Verification Algorithm for EPFLTL}
+
 We extend now the algorithm to capture EPFLTL.

### 2019-02-03 implementation.tex
fixing problems

@@
 
-\noindent {\bf Example properties.}
-%
+\subsubsection{Example Properties}
+
 Figure \ref{fig:properties} shows three properties in the 
-input format of the tool, which are not expressible in the original first-order logic of \dejavu{}. The first property \idsl{telemetry1} is a variant of
+input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The first property \idsl{telemetry1} is a variant of
 formula \ref{eq:wolper-first-order}, illustrating the use of a 
-rule to express Wolper's example (\cite{Wolper}) that all the states with even indexes of a sequence satisfy a property.
+rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
 In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
-Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when the radio is toggled on.
+Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.
 
@@
 (the only state where \idsl{@true} is false), and there is no \idsl{toggle(x)} event. The next alternative states that
-if \idsl{closed(x)} was true in the previous state and there is
-no \idsl{toggle(x)} event, we stay in the \idsl{closed(x)} state.
-The third alternative states that if we in the previous state were
-in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} event, we transition to the \idsl{closed(x)} state. Similarly for the \idsl{open(x)} rule.
+\idsl{closed(x)} was true in the previous state and there is
+no \idsl{toggle(x)} event.
+The third alternative states that we in the previous state were
+in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} 
+event. Similarly for the \idsl{open(x)} rule.
 
@@
 
-\noindent {\bf Monitor synthesis.}
-%
+\subsubsection{Monitor Synthesis}
+
 Consider the first \idsl{telemetry1} property. The property-specific 
-part\footnote{An additional 600+ lines of property independent 
-boilerplate code is generated.} of the synthesized monitor, is shown 
-in Figure \ref{fig:monitor}.
+part of the synthesized monitor, is shown 
+in Figure \ref{fig:monitor}\footnote{An additional 600+ lines of property independent 
+boilerplate code is generated.}.
 This function is called for each new event.
@@
 override def evaluate(): Boolean = {
-  // event nodes:
+  // -- 1. rule nodes not below @:
+  now(7) = pre(8)
+  now(6) = build("toggle")(V("x"))
+  now(5) = now(6).biimp(now(7))
+  // -- 2. rule calls:
+  now(9) = now(5)
+  now(2) = now(5)
+  // -- 3. rest of rules below @:
+  now(8) = now(9).not()
+  // -- 4. rest of main formula:
   now(4) = build("telem")(V("x"))
-  now(6) = build("toggle")(V("x"))
-  // rule nodes not below @:
-  now(7) = pre(8)
-  now(5) = now(6).biimp(now(7))
-  // rule calls:
-  now(2) = now(5)
-  now(9) = now(5)
-  // rest of rules below @:
-  now(8) = now(9).not()
-  // rest of main formula:
   now(3) = now(4).not()
@@
 The indexing scheme is guided by an enumeration of the annotated abstract syntax tree of the formula, shown in Figure \ref{fig:ast}.
-The tree has been annotated by pointing out (blue node 5) the rule body and calls to the rule (purple nodes 2 and 9).
-The assignments are evaluated in five steps. First the leaf nodes corresponding to events (4 and 6). Then rule nodes not below the previous-operator \idsl{@} (7 and 5). Then calls of rules (2 and 9)
+The figure (generated by \dejavu) points out (blue node 5) the top node for the rule \idsl{closed(x)}, and calls to the rule (purple nodes 2 and 9).
+The subformulas are evaluated in four steps, as outlined at the end of Section \ref{LTLruntime}. First rule nodes not below the previous-operator \idsl{@} (7, 6, 5). Then calls of rules (9, 2)
 which can now be calculated since the rule bodies have been calculated.
 Then the rest of the rules below \idsl{@}  (8).
-Finally the rest of the main formula (3, 1, 0).
+Finally the rest of the main formula (4, 3, 1, 0).
 At composite subformula nodes, BDD operators are applied.
@@
 
-\noindent {\bf Example execution.}
-%
+\subsubsection{Example Execution}
+
 As an example of how a trace is processed consider the following simple (correct) trace containing two events, turning on low and high frequency channels:
-$\mathit{toggle}(``low").\mathit{toggle}(``high")$.
-The value $``low"$ is assigned the binary enumeration $110$ in node 6, shown in Figure \ref{fig:bdd1}.
+$\mathit{toggle}(``low").\mathit{toggle}(``high")$. The value $``low"$ is assigned the binary enumeration $110$ in node 6, shown 
+as a BDD in Figure \ref{fig:bdd1}.
 A dashed line leaving a node means the node has value 0, and a fully drawn line means it has value 1.
@@
 Since channels $``low"$ and $``high"$, corresponding to enumerations 110 and 101, are now open, this BDD represents all other enumerations.
-That is, all enumerations except 110 and 110 lead to 1 in this BDD.
+That is, all enumerations except 110 and 101 lead to 1 in this BDD.
 
@@
     \centering   
-    \begin{subfigure}[b]{0.15\textwidth}
+    \begin{subfigure}[b]{0.3\textwidth}
         \includegraphics[width=\textwidth]{figures/bdd1.pdf}
@@
     ~ 
-    \begin{subfigure}[b]{0.15\textwidth}
+    \begin{subfigure}[b]{0.3\textwidth}
         \includegraphics[width=\textwidth]{figures/bdd2.pdf}
@@
     ~ 
-    \begin{subfigure}[b]{0.15\textwidth}
+    \begin{subfigure}[b]{0.3\textwidth}
         \includegraphics[width=\textwidth]{figures/bdd3.pdf}
@@
 
-\noindent {\bf Evaluation}
-%
+\subsubsection{Evaluation}
+
 We evaluated the three properties in Figure 
-\ref{fig:properties}. Table \ref{tab:evaluation} shows the  
+\ref{fig:properties} on a collection of traces. Table \ref{tab:evaluation} shows the  
 analysis time (excluding time to compile the generated 
@@
 \end{tabular}
-\caption{Evaluation}
+\caption{Evaluation - trace lengths and analysis time in seconds}
 \label{tab:evaluation}

### 2019-02-03 rule-ltl.tex
fixing problems

@@
 The use of BDDs can be replaced by other representations that
-can compactly and efficiently represent sets of values (e.g., ZDDs) and
-apply set operations like complementation, intersection,union (which are simply negation, conjunction and disjunction
+can compactly and efficiently represent sets of values (e.g., ZDDs), and to which one can apply set operations like complementation, intersection, union (which are simply negation, conjunction and disjunction
 in BDDs) and projection (for the quantification).
@@
 %one enumeration, $11\ldots 11$, for this purpose.
-In the next section we introduce data reclaiming, which allows 
-reusing enumerations for values that no longer affect the checked property. 
-This involves a more
-complicated enumeration mechanism.
+%In the next section we introduce data reclaiming, which allows 
+%reusing enumerations for values that no longer affect the checked 
+%property. 
+%This involves a more
+%complicated enumeration mechanism.
 
@@
 other than $x$, i.e., they can have any value.
-For example, assume that we use three Boolean variables (bits) $x_0$, $x_1$ and $x_2$
+For example, assume that we use the three Boolean variables (bits) $x_0$, $x_1$ and $x_2$
 for representing enumerations over $x$ (with $x_0$ being the least significant bit), and
 assume that $A = \{ a , b \}$, 
-$\lookup ( x, a ) = 011$, and $\lookup ( x , b ) = 001$.
+$\lookup ( x, a ) = 000$, and $\lookup ( x , b ) = 001$.
 Then $\bddset ( x , A )$ is a BDD representation of the Boolean function 
-$x_0 \wedge \neg x_2$. 
+$(\neg x_1 \wedge \neg x_2)$. 
 %But the BDD function will be independent of the values of Boolean variables
@@
   \item $\current ( p_k ( x ) )$ :=
-     $\bddset ( x ,  \{ a | R_k[i] ( a  ) \} )$.
+     $\bddset ( x ,  \{ a \ | \ R_k[i] ( a  ) \} )$.
   \item $\current ( ( \varphi \wedge \psi ) ) := 
@@
   \bddvee ( \current ( \psi  ) , \bddwedge ( \current ( \varphi ) ,
-      \old ( ( \varphi \since  \psi ))))$.
+      \old ( ( \varphi \ \since \ \psi ))))$.
 %DAP: I removed the \; around \since, since it penetrated the 2nd column!
@@
 a variable $x_j$, which is the one appearing
-free in $\varphi_j$(this can be generalized to any number of variables). However, $r_j$ can be used
+free in $\varphi_j$ (this can be generalized to any number of variables). However, $r_j$ can be used
 as a subformula with other parameters in other rules or in the statement. Thus,
 $r_j ( y )$ constructing a relation
-that is the same as $r_j$, but over the variable $y$. This can be achieved by BDD renaming function $\rename ( r_j (x_j ) , y)$, which is easy to implement in BDDs.
+that is the same as $r_j$, but over the variable $y$. This can be achieved by the BDD renaming function $\rename ( r_j (x_j ) , y)$, which is easy to implement in BDDs.
 
@@
 \item For each $r_j ( x_j ) = \varphi_j ( x_j )$:
-\item{\ \ }  calculate $\current ( \varphi_j )$.
-\item{\ \ }  $\current ( r_j )  := \current (\varphi_j)$
-\item{\ \ } $\current ( r_j ( y ) )$ := $\rename ( r_j(x_j) ,  y )$.
+\item{\ \ }  calculate $\current ( \varphi_j )$;
+\item{\ \ }  $\current ( r_j )  := \current (\varphi_j)$;
+\item{\ \ } $\current ( r_j ( y ) )$ := $\rename ( r_j(x_j) ,  y )$;
 \item{\ \ } $\current ( r_j ( a ) )$ := if $\current ( r_j ) ( a )$
-then $\btrue$ 
-\item{\ \ \ \ \ \ \ \ \ } else $\bfalse$.
+then $\btrue$ else $\bfalse$; 
 \end{description}
 
+\klaus{Above: $\current ( r_j ) ( a )$ has not been defined I think.}
+
+\noindent
 As in the propositional case, the evaluation order cannot be

### 2019-02-03 main.tex
fixing problems

@@
 
-\title{An Extension of LTL with Rules 
+\title{An Extension of LTL with Rules\\ 
 and its Application to Runtime Verification}
@@
 
+\klaus{In the text below mention what $i$ is (time point, position, ...}
+
 \noindent {\bf Sketch of proof of Theorem~\ref{theo2}.}
@@
 
+\klaus{Below: isn't it the other way around, the first is not a model and the second is a model?}
+
 Hence, the sequence of states 
 \[ \{ r \} \{ v(1) \}  \{ v(2) \} \{ v(3) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(3) \} \{  r \} \] is a model of this formula, while \[ \{ r \} \{ v(1) \}  \{ v(2) \}  \{ v(2) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(5) \}  \{  r \}\] is not.
-The formula is a bit involved, hence given in several parts (macros).
+The QFLTL formula is a bit involved, hence given in several parts (macros).
+
 \begin{itemize}
@@
 \item $collect(P, \varphi, \psi) ::=
-       q \wedge same(P) \wedge \varphi \wedge 
-       \ominus ((add(P)\wedge \neg q \wedge \varphi) \since (q \wedge empty(P) \wedge \psi))$  [Collect elements $a$ in $v(a)$, while $\varphi$ holds, 
-       since the previous state where $q$ 
-       and $\psi$ held. ]
-\item $init ::= r \wedge \Box (r \leftrightarrow \neg \exists x \, v(x))$ [$r$ holds in the first state, and $v(a)$ with some value $a$ holds in states where $q$ does not hold.]
+       r \wedge same(P) \wedge \varphi \wedge 
+       \ominus ((add(P)\wedge \neg r \wedge \varphi) \ \since \ (r \wedge empty(P) \wedge \psi))$  [Collect elements $a$ in $v(a)$, while $\varphi$ holds, 
+       since the previous state where $r$ 
+       and $\psi$ held.]
+\item $init ::= r \wedge \Box (r \leftrightarrow \neg \exists x \, v(x))$ [$r$ holds in the first state, and $v(a)$ with some value $a$ holds in states where $r$ does not hold.]
 \item $init \wedge \neg \Exists A \Exists B (equal (A, B) \wedge \\
-collect(B, same(A),\, same(A) \since collect(A, true, true))$
+collect(B, same(A),\, same(A)) \ \since \ collect(A, true, true))$
 [There are no two sequences of states where $v(a)$ holds
@@
 \end{itemize}
+
+\klaus{Above: maybe need to point out what the top formula is.}
+
+\klaus{Above: isn't there missing a $\Box$ in the main formula?}
+
 We prove that the above property cannot be expressed in EFLTL.

### 2019-02-03 implementation.tex
fixing problems

@@
 Figure \ref{fig:properties} shows three properties in the 
-input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The first property \idsl{telemetry1} is a variant of
+input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 
+%
+The ASCII version of the logic here uses 
+\idsl{@} for $\ominus$,
+\idsl{|} for $\vee$,
+\idsl{&} for $\wedge$, and
+\idsl{!} for $\neg$.
+%
+The first property \idsl{telemetry1} is a variant of
 formula \ref{eq:wolper-first-order}, illustrating the use of a 
@@
 prop telemetry1: 
-  Forall x . closed(x) -> !telem(x)
-  where closed(x) := toggle(x) <-> @!closed(x)
+  Forall x . closed(x) -> !telem(x) where closed(x) := toggle(x) <-> @!closed(x)
 
 prop telemetry2: 
-  Forall x . closed(x) -> !telem(x)
-  where
-    closed(x) :=
-        (!@true & !toggle(x))
-      | (@closed(x) & !toggle(x))
-      | (@open(x) & toggle(x)),
-    open(x) :=
-        (@open(x) & !toggle(x))
-      | (@closed(x) & toggle(x))
+  Forall x . closed(x) -> !telem(x) where
+    closed(x) := (!@true & !toggle(x)) | (@closed(x) & !toggle(x)) | 
+                 (@open(x) & toggle(x)),
+    open(x) := (@open(x) & !toggle(x)) | (@closed(x) & toggle(x))
 
@@
   Forall x . Forall y . 
-  spawn(x,y) -> 
-    (P main(x) | Exists z . (P main(z) & leadsto(z,y)))
-  where
-    leadsto(z,y) := 
-        spawn(z,y) 
-      | @ leadsto(z,y) 
-      | Exists x . (@leadsto(z,x) & spawn(x,y))
+    spawn(x,y) ->  (P main(x) | Exists z . (P main(z) & leadsto(z,y))) 
+    where leadsto(z,y) := 
+      spawn(z,y) | @ leadsto(z,y) | Exists x . (@leadsto(z,x) & spawn(x,y))
 \end{lstlisting}
@@
 part of the synthesized monitor, is shown 
-in Figure \ref{fig:monitor}\footnote{An additional 600+ lines of property independent 
+in Figure \ref{fig:monitor} (left)\footnote{An additional 600+ lines of property independent 
 boilerplate code is generated.}.
@@
 
-\begin{figure}
-%\centering
-\begin{center}
-\begin{tabular}{c}
-\begin{lstlisting}[language=scala,basicstyle=\small\sffamily]
+\begin{figure}[htb]
+\centering
+\begin{minipage}[b]{.51\textwidth}
+\begin{lstlisting}[language=scala,basicstyle=\sffamily]
 override def evaluate(): Boolean = {
@@
 \end{lstlisting}
-\end{tabular}
-\end{center}
-\caption{Monitor evaluation function for telemetry1}
+\end{minipage}
+\qquad
+\begin{minipage}[b]{.43\textwidth}
+\includegraphics[width=7cm]{figures/ast.pdf}
+\end{minipage}
+\caption{Monitor (left) and subformula enumeration (right) for property telemetry1}
 \label{fig:monitor}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=7cm]{figures/ast.pdf}
-\caption{Annotated abstract syntax tree for telemetry1}
-\label{fig:ast}
-\end{figure}
-
-The indexing scheme is guided by an enumeration of the annotated abstract syntax tree of the formula, shown in Figure \ref{fig:ast}.
+The indexing scheme is guided by an enumeration of the annotated abstract syntax tree of the formula, shown in Figure \ref{fig:monitor} (right).
 The figure (generated by \dejavu) points out (blue node 5) the top node for the rule \idsl{closed(x)}, and calls to the rule (purple nodes 2 and 9).
@@
     \centering   
-    \begin{subfigure}[b]{0.3\textwidth}
+    \begin{subfigure}[b]{0.25\textwidth}
         \includegraphics[width=\textwidth]{figures/bdd1.pdf}
@@
     ~ 
-    \begin{subfigure}[b]{0.3\textwidth}
+    \begin{subfigure}[b]{0.25\textwidth}
         \includegraphics[width=\textwidth]{figures/bdd2.pdf}
@@
     ~ 
-    \begin{subfigure}[b]{0.3\textwidth}
+    \begin{subfigure}[b]{0.25\textwidth}
         \includegraphics[width=\textwidth]{figures/bdd3.pdf}

### 2019-02-03 main.tex
fixing problems
Updates from ShareLaTeX

@@
 Runtime Verification (RV) consists of analyzing execution 
-traces using formal techniques. RV is scalable and 
-therefore has a big potential for infusion in software 
-engineering practices. It includes e.g. monitoring the 
+traces using formal techniques. 
+%RV is scalable and 
+%therefore has a big potential for infusion in software 
+%engineering practices. 
+It includes monitoring the 
 execution of a system against properties formulated in 
@@
 properties. Finally, we expand the BDD-based runtime verification 
-tool \dejavu{}  to support our extension and we perform some 
+tool \dejavu{}  to support our extension and perform some 
 experiments.
@@
 
+\klaus{Above: why set of subsets?}
+
 \iffalse

### 2019-02-03 rule-ltl.tex
Updates from ShareLaTeX

@@
 
-We present first an extension to the propositional case that is based on using additional (auxiliary) propositions, not appearing in the execution, which summarize the past 
+We present first an extension of the propositional LTL that is based on using additional (auxiliary) propositions, not appearing in the execution, which summarize the past 
 of the current execution.
@@
 This can be seen as adding
-summary states to the temporal specification, resulting in a specification that is between logic and finite automata representation. This extension, which we believe is conceptually simpler
+summary states to the temporal specification, resulting in a specification that is in between logic and finite automata representation. This extension, which we believe is conceptually simpler
 and more intuitive than other suggested extensions, is shown to be
@@
 order case to motivate the need for
-extending first-order LTL.
-We present such an extension where additional (auxiliary) relations are
-added. We show that this extension is more expressive than
-first-order LTL, e.g., by repeating Wolper's classical algorithm
-with data. On the other hand, we show that for the first-order version, this has less 
-expressive power than adding quantification over relations, in distinction from the propositional case.
+extending first-order LTL. Another example is based on the inexpressiveness of transitive closure of a relation in first order logic.
+We present an extension where auxiliary relations, not appearing in the model, are
+added. While this extension
+is capable of expressing the above
+properties, we show that for the first-order version, this has less 
+expressive power than adding quantification over relations, in distinction from the corresponding propositional case.
 
@@
 on calculating and updating a summary of the prefix of the execution.
-We provide runtime verification algorithms for the extended
-logics and present an extension of the 
+Runtime verification often concentrates on a past  version of temporal logic~\cite{HR} mainly because for these properties we can decide when the monitored property is violated immediately when this happen, while for future properties we may never know for sure.
+Consistent with that,
+we provide runtime verification algorithms for the past-time versions of the extended propositional and first-order
+logics. We further present an extension of the 
 \dejavu{} tool (https://github.com/havelund/dejavu) that
 realizes the extended first-order past temporal logic verification.
+The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first order temporal logic over infinite domains (e.g., the integers, strings).
+It achieves efficiency by BDD representation of enumerations of data (rather than over the data values themselves) and using
+a garbage collection algorithm. 
 Finally, we provide some experimental results.
 
-\klaus{I think that the structure of the paper should be
-crystal clear so that the reader does not get lost. A lost reader leads to rejection. If a reader understand the structure, but perhaps not the details, the reader may still be happy.}
-
-\subsubsection{Related Work}
-
+The contributions of this paper are both theoretical and practical.
+On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions.
+On the practical side, we present algorithms for the extended logic, and 
+show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.
+
+The structure of this paper is as follows.
+In Section~2 we first define the syntax and semantics of propositional temporal logic.
+We then review some of its classical extensions and present our own extension.
+In Section~3 we present a runtime verification algorithm for the extended past LTL.
+In Section~4 we review first order linear temporal logic, present our extension and study the relationships between the presented versions of first-order LTL.
+IIn Section~5 we present a runtime verification algorithm for the past version of the extended first-order LTL.
+Section~6 describes the implementation and provides experimental results.
+Finally, Section~7 concludes the paper.
+
+%\klaus{I think that the structure of the paper should be
+%crystal clear so that the reader does not get lost. A lost reader leads to rejection. If a reader understand the structure, but perhaps not the details, the reader may still be happy.}
+
+\vspace{1ex}
+\noindent {\bf Related Work}
 For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
@@
 \fi
-
-\section{Preliminaries}
+\vspace{1ex}
+\noindent {\bf conventions}
 \label{sec:prelim}
-
-In this paper, we treat several versions of the logic LTL. To simplify
+In this paper, we treat and present several versions of the logic LTL. To simplify
 the presentation, we prefix LTL with letters as follows:
@@
 
-\section{Extending Propositional LTL}
+\section*{Extending Propositional LTL}
 \label{sec:extending-prop-ltl}
@@
 
-\section{Extending First-Order LTL}
+\section*{Extending First-Order LTL}
 
@@
 \psi \mathit{\ where\ } r_j  ( x_j ) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n 
+\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} 
 \end{equation}

### 2019-02-04 main.tex
Updates from ShareLaTeX

@@
 \and
-Doron Peled~\thanks{The research performed by this author was partially funded by Israeli Science Foundation grant 2239/15: 
-``Runtime Measuring and Checking of Cyber Physical Systems''.
+Doron Peled~\thanks{The research performed by this author was partially funded by Israeli Science Foundation grant 1464/18: 
+``Efficient Runtime Verification for Systems with Lots of Data and its Applications''.
 }

### 2019-02-04 rule-ltl.tex
Updates from ShareLaTeX

@@
 \vspace{1ex}
-\noindent {\bf Related Work}
+\noindent {\bf Related Work.}
 For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
@@
 \vspace{1ex}
-\noindent {\bf conventions}
+\noindent {\bf Conventions.}
 \label{sec:prelim}
@@
 \begin{itemize}
-\item{P} restricts to {\em past} temporal operators.
-\item{F} allows first-order quantification over
-variables and relations over variables and constants. This is referred to as {\em static} quantification.
-\item{Q} adds quantification over predicates (for the propositional versions)
-or over relations (for the first-order version). These objects can have different values at different states. This is referred to as {\em dynamic} quantification.
-\item{E} allows extending the objects that occur in the input with
-new auxiliary predicates (for the propositional versions) or relations (for the first-order cases).
+\item{P} restricts to P{\em ast} temporal operators.
+\item{F} allows F{\em irst-order} quantification over
+data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
+\item{Q} adds (second order) Q{\em uantification} that is state-dependent. This is referred to as {\em dynamic} quantification.
+\item{E} allows E{\em xtending} the objects that occur in the input with
+new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) and past temporal logic rules that define them.
 \end{itemize}
 
-\klaus{You above introduce some concepts before they really are explained.
+%\klaus{You above introduce some concepts before they really are explained.
 Q and E.
 I understand why you do this (to have it in one place), but it may be too difficult for the reader to grasp if not "initiated".
-Especially the E.}
-
-\klaus{We use ELTL (etc) to denote Extended LTL. 
-But QLTL is also an extension of LTL. These logics 
-are all extensions of LTL. Perhaps RLTL (for Rule-based 
-LTL) would have been better ... or something. Not urgent, 
-we can always change that if we want to.}
+%Especially the E.}
+
+%\klaus{We use ELTL (etc) to denote Extended LTL. 
+%But QLTL is also an extension of LTL. These logics 
+%are all extensions of LTL. Perhaps RLTL (for Rule-based 
+%LTL) would have been better ... or something. Not urgent, 
+%we can always change that if we want to.}
 
 %\klaus{The explanation of Q and E could be improved.}
+
+\doron{Did my best to make the letters more self explained.
+I think R is usually reserved in temporal logic for realtime.
+But if you insist, one thing that I cannot perform from here is changing the two figures where these letters occur.}
 
@@
 The models for LTL formulas 
-are infinite sequence of states.
+are infinite sequence of states, of the form
 $\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
@@
 \item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
-\item $(\sigma, {i}) \models ( \varphi\ \cal{U}\ \psi )$ iff for some 
+\item $(\sigma, {i}) \models ( \varphi \cal{U} \psi )$ iff for some $j$,
 $j\geq i$, $(\sigma, {j}) \models \psi$,
-and for all $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
+and for each $k$, $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
 \item $(\sigma , i) \models \ominus \varphi$ iff $| \sigma | > 1$ and $(\sigma, i-1) \models \varphi$.
-\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $1 \leq k \leq i$ such that
-$(\sigma , i) \models \psi$  and for each
-$j < k \leq i$, 
-$(\sigma , k)  \models \psi$.
+\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $j$, $1 \leq j \leq i$, such that
+$(\sigma , j)  \models \psi$ and for each
+$k$, $j < k \leq i$, $(\sigma , k ) \models \varphi$.
 \end{itemize}
@@
 $\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
-${\bf P}\  \varphi = (\true \; \since \; \varphi)$, and 
+${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}), and 
 ${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$.
@@
 The expressive power of propositional LTL is often compared to regular expressions
-over the alphabet $\Sigma = 2^P$ and to  {\em monadic} logics where the temporal propositions
-correspond to unary predicates (over time). A full overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
-LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions with the following operators: constant letters, concatenation and complementation, but no star (or $\omega$-star for infinite words).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer n such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}.
+over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second order logics where the temporal propositions
+correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
+LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star (or $\omega$-star for infinite words).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}.
 In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
@@
 operators: $\since$, $\ominus$ and the ones derived
-from them $\mathbf{H}$ and $\mathbf{P}$. The logic is often
+from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is often
 (and also here) interpreted over finite sequences, 
@@
 Alternatively,
-it is also a common practice to prefix the property with a single $\Box$ (always) operator, where each of the prefixes has to satisfy $\varphi$. This later
-form expresses {\em safety} properties~\cite{AS} over infinite sequences.
+it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; there, each of the prefixes has to satisfy $\varphi$. This later
+form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences.
 
@@
 
-\section*{Extending Propositional LTL}
+%\section*{Extending Propositional LTL}
 \label{sec:extending-prop-ltl}
-
-\subsubsection{Extending with Dynamic Quantification}
 
@@
 to some important properties. As a canonical example, he demonstrated that it is impossible to
-express that all the states with even indexes of a sequence
+express that all the states with even indexes in a sequence
 satisfy $p$ (this is
 different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states).
+Wolper's example property cannot be expressed in PLTL either.
 In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
-propositions.
-
+propositions, as shown below.
+
+\subsubsection{Extending LTL with Dynamic Quantification.}
 Adding quantification over propositions allows writing a formula of the form
-$\Exists q \, \varphi$, where $q$ is a proposition that can appear in $\varphi$.
+$\Exists q \, \varphi$, where
+$\Exists q$ repesents {\em existential} quantification over a propositiona $q$ that can appear in $\varphi$.
 Let the propositions $P$ be partitioned into two disjoint sets: 
@@
 %has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.
-We also allow $\Forall q\, \varphi =
-\neg \Exists q\, \neg \varphi$.
-To define the semantics, let $X$ be a subset of 
-the propositional variables appearing in the states of the sequence $\sigma = s_1 s_2 \ldots$ and 
-denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)
+
+To define the semantics, let $X \subseteq P$ 
+%be a subset of 
+%the propositional variables appearing in the %states of the sequence $\sigma = s_1 s_2 %\ldots$ 
+and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)
 
@@
 \end{itemize}
-Since all the auxiliary variables appear in the scope of
-quantifiers, it holds that $\sigma \models \varphi$
-(in $\varphi$, auxiliary variables appear only within
-the scope of quantifiers)
-iff $\sigma |_A \models \varphi$.
-
-\klaus{Are the last two lines above needed?}
-
-This kind of quantification is considered to be ``dynamic'', since the quantified variables can have different truth values depending on the states, or, "second order", the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
-Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
+%Since all the auxiliary variables appear in the scope of
+%quantifiers, it holds that $\sigma \models \varphi$
+%(in $\varphi$, auxiliary variables appear only within
+%the scope of quantifiers)
+%iff $\sigma |_A \models \varphi$.
+%
+%\klaus{Are the last two lines above needed?}
+%
+Universal quantification is also allowed, where $\Forall q\, \varphi =
+\neg \Exists q\, \neg \varphi$.
+This kind of quantification is considered to be ``dynamic'', since the quantified variables can have different truth values depending on the states It is also callsed  "second order" quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
+Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions (over infinite words), full B\"{u}chi
 automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
@@
 ( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}
-Wolper's example property cannot be expressed in PLTL either.
 
@@
 
-\subsubsection{Extending with Rules}
+\subsubsection{Extending LTL with Rules.}
 
@@
 \item $( \gamma , {\cal R} , i ) \models ( \varphi \; \until \; \psi )$ if
-for some $j > i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
-\models \psi $ and for all $i \leq k < j $,
+for some $j$, $j > i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
+\models \psi $ and for each $k$,  $i \leq k < j $,
 $( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.
@@
 \item $( \gamma , {\cal R} , i ) \models ( \varphi \; \since \; \psi )$ if
-for some $1 \le j \leq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
-\models \psi $ and for all $j < k \leq i$,
+for some $j$, $1 \le j \leq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
+\models \psi $ and for each $k$, $j < k \leq i$,
 $( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.
@@
 
-\section*{Extending First-Order LTL}
+%\section*{Extending First-Order LTL}
 
@@
 
-\subsubsection{Extending with Quantification}
+\vspace{1ex}
+\noindent {\bf Example 2.}
+Consider the property that asserts that
+a main process $y$ (where
+$mathit{main} (x)$ appears
+in the sequence), is a descendant of a process $x$ if $x$ spawns a chain of processes spawning (of
+the form $\mathit{spawn} (z, w$ that leadsfrom $x$ to $y$. This property cannot be expressed in FLTL. The reason
+is that the required property needs to
+assert about the transitive closure
+of the relation $\mathit{spawn}$. 
+FLTL can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}.
+Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
+E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, (p(x, i) \rightarrow \exists t' p(x, j) )$
+\footnote{The obtain model will have, in addition
+to the aforementioned relations, also
+the linear order $<$ over the naturals.}
+It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in this first order setting. The proof is based on the
+compactness theory of first order logic~\cite{Flum}.
+
+\subsubsection{Extending FLTL with Quantification.}
 
@@
 
-\subsubsection{Extending with Rules}
+\subsubsection{ExtendingFLTL  with Rules.}
 

### 2019-02-05 rule-ltl.tex
Updates from Overleaf

@@
 
-\doron{Did my best to make the letters more self explained.
+%\doron{Did my best to make the letters more self explained.
 I think R is usually reserved in temporal logic for realtime.
@@
 $\Exists q$ repesents {\em existential} quantification over a propositiona $q$ that can appear in $\varphi$.
-Let the propositions $P$ be partitioned into two disjoint sets: 
-{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
-and {\em basic propositions} $B$. 
+
 %A model for such a formula 
 %has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.
-
 To define the semantics, let $X \subseteq P$ 
@@
 
-We partition again, as in QLTL, the propositions $P$ into
+We partition the propositions $P$ into
 {\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
@@
 $a_j  := \varphi_j \, : \, 
- {j \in \{1, \ldots , n\}}$ as the rules. In text, rules
+ {j \in \{1, \ldots , n\}}$ as the {\em rules}. In text, rules
  will be separated from each other by a comma (",").
@@
 ELTL extends the set of propositional variables with new variables whose values at a state are
-based on the prefix of the model. We say that
-the extension of the states are {\em well founded},
-as the added propositions are dependent and effectively summarize the past execution prefix up to and
+based on the prefix of the model. The added propositions are dependent and effectively summarize the past execution prefix up to and
 including the current state.
@@
 must occur within the scope of a $\ominus$ operator
-is required to guarantee that the values
+is required to help preventing that the values
 of the auxiliary variables are not conflicting, as in
-$a_1 = \neg a_2$ and $a_2 = a_1$, which makes a formula
-that uses $a_1$ and $a_2$ uniformly $\false$.
+$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
+that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write conflicting rules, e.g., by simlpy 
+adding the rule $a: = \false$.
 
 \begin{lemma} \label{fourone}
-Let $\varphi$ be an ELTL formula as in~(\ref{ELTL}),
-over the auxiliary propositions $A$ and basic propositions $B$, 
-and let $\sigma$ be a model with states over the $B$, such that
-$\sigma \models \psi$.
-Then there is a unique
-model $\sigma'$ such that $\sigma' |_A = \sigma$.
+Let $\varphi$ be an ELTL formula
+over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
+a set of rules $a_j  := \varphi_j \, : \, 
+ {j \in \{1, \ldots , n\}}$. 
+Let $\sigma$ be a model with states over the $B$. 
+Then there is at most one
+model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $
 
@@
 
-\klaus{(1) is this lemma saying the right thing? (2) is the proof really a proof of that lemma, (3) 
-should $\varphi$ be $\psi$, (4) now $\psi$ is not the whole spec (it is missing the rules). What do we call the whole thing? It is not a formula like the 
-$\varphi$s and the $\psi$s. Do we need a special greek symbol for it?}
+%\klaus{(1) is this lemma saying the right thing? (2) is the proof really a proof of that lemma, (3) 
+%$\varphi$s and the $\psi$s. Do we need a special greek symbol for it?}
 
@@
 is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states. A PLTL
-property depends on a state only on the prefix
-of the model that precedes and includes that state.
-Restricting the auxiliary variables to appear only within the scope of a previous-time operator guarantees that
-their values at the current state cannot be conflicting. \qed
+property depends in a state only on the prefix
+of the model that precedes and includes that state. \qed
 
@@
 
-\klaus{How do we know that the ``other direction'' below is possible?}
-
-For the other direction, one can translate any QLTL property into an $\omega$ automaton\footnote{The past operators of a QLTL property do not change this fact, since the translation can be done first to second order monadic logic, which can be translated into an $\omega$-automaton.}.
-We assume that the property is given as a deterministic Muller automaton $M = (S, i , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $i \in S$ is the initial state, $\Sigma$ is the set of inputs, each represented as a Boolean combination of the variables $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set,
-where an accepted input needs to pass infinitely often in exactly
-one of the specified sets. To formulate an ELTL formula,
-he auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
-
-\klaus{The following formulation above is not not clear: ``where an accepted input needs to pass infinitely often in exactly
-one of the specified sets''.}
-
-\noindent
-Then we translate $M$ into an ELTL formula
-of the form:
-
-\begin{tabbing}
-\ \ \ \ \ $\bigvee_{F \in {\cal F}} (\bigwedge_{s \in F} \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
-where \\
-\ \ \ \ \ \ \ \ \ \ $a_s := \vee _{\Delta ( s'
-, \alpha ) = s } 
-\alpha \wedge \ominus a_{s'} )
-\wedge (s=i \rightarrow \neg \ominus \true)
-\, :\,  s \in S
-$
-\end{tabbing}
-
-\klaus{Above: should $s=i \rightarrow \neg \ominus \true$ instead be (note: two changes):
-$\neg \ominus \true \rightarrow a_s=i$
-}
-
-\klaus{Above: can't judge whether the $\bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$ is correct.}
-
-\noindent
-With $(s=i)$ is $\true$ when $s$ is $i$, and $\false$ otherwise.
-Basically, the assignments to the auxiliary variables are used to simulate the change of states\footnote{It is possible to use only $\lceil \log ( | S | ) \rceil$
-variables, at the expense of longer and more complicated formulas.}, and the main formula expresses the Muller acceptance condition. \qed
-
-The use of the auxiliary variables $a_1 \ldots a_n$ in the
+%\klaus{How do we know that the ``other direction'' below is possible?}
+
+For the other direction, one can translate the QLTL property first into
+a second order monadic logic formula, then to a deterministic Muller automata and then constructs an ELTL formula that holds for the accepting executions of this automaton. The details of the
+constructions are given in the appendix. \qed
+
+%EPLTL, where the Muller automaton can be
+%replaced with a B\"{u}chi automaton,
+%where all of its states, except one, sink
+%state, are accepting. \qed
+
+\iffalse
+The use of the auxiliary variables $a_1 \ldots a_n$ and the
 rules of an ELTL specification 
-$\eta$ can be viewed as representing a
+can be viewed as representing a
 state machine, in particular when the only modality that is used in the rules part is $\ominus$.
-In that case, we can consider an ELTL specification $\eta$ as consisting with two synchronous parts: an LTL property 
-(the statement) $\psi$
+In that case, we can consider an ELTL specification $\eta$ as consisting of two synchronous parts: an LTL property 
+(the statement) 
 and a finite automaton, whose states correspond to combinations
@@
 
-\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}
-
-We can extend PLTL into EPLTL by disallowing the future temporal operators (except a prefixing $\Box$, when interpreted over infinite sequences). We obtain
-a formalism that is equivalent to finite automata. In this case, 
+%\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}
+\fi
+
+We extend PLTL into EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$). This results in
+a formalism that is equivalent to finite automata, where all the states except one are accepting and where the nonaccepting state is a sink. 
+We can use related, but simpler construction than in Theorem~\ref{theo1} to prove the following:
+
+\iffalse
+In this case, 
 we replace the first part of the construction in the proof of Theorem~\ref{theo1}
-with the simpler formula $\vee_{s \in A} a_s$, where $A$ are the accepting states of the finite automaton.
+with the simpler formula $\vee_{s \in A} a_s$, where all the states but one are
+accepting.$A$ are the accepting states of the finite automaton.
 
 \klaus{Last sentence above: perhaps say with words that we are always in an acceptance state - if that's what the point is.}
-
-For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.
+\fi
+
+%For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.
+
 \begin{lemma}
 \label{simplecase}
-Any safety QPLTL property can be expressed using the following restricted
-ELTL form:
+The expressive power of 
+ELTL is the same as QLTL.
+%Any safety QPLTL property can be expressed using the following restricted
+%ELTL form:
+%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
+% {j \in \{1, \ldots , n\}}  \] 
+% with $p$ being one of the auxiliary variables %$a_j$.
+\end{lemma}
+In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
 \[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}  \] 
- with $p$ being one of the auxiliary variables $a_j$.
-\end{lemma}
+with $p$ being one of the auxiliary variables $a_j$. The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.
+
+\iffalse
 \noindent {\bf Proof.} The proof of the Lemma follows from the
@@
 While this Lemma makes the use of the temporal operators
-besides $\ominus$ redundant, we believe that it is still
-more convenient to use a combination of
+besides $\ominus$ in EPLTL redundant, it is still
+convenient to use a combination of
 temporal operators when possible, and
 auxiliary variables definitions.
-The relative expressive power between the propositional temporal logics discussed here
-appear in Figure~\ref{prop}.
+\fi
 
@@
 
-Runtime verification of temporal specifications often
+Runtime verification of temporal specification often
 concentrates on the past portion of the logic.
@@
 of $\since$ in an equivalent form that is
-more directly amenable for runtime verification.
-
+more directly applicable for runtime verification.
 \begin{itemize}
@@
 \end{itemize}
-
-\noindent
 Runtime verification exploits the fact that
@@
 The algorithm, shown below, operates on two vectors (arrays) of values indexed by subformulas:  $\old$ that summarizes the truth values of the
-subformulas just {\em before} the new state, and $\current$ for the {\em current} state.
-
+subformulas just {\em before} the new state, and $\current$ for the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
 \begin{enumerate}
@@
 
-\noindent
-Runtime verification of QPLTL can be done by translating a
+\subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
 formula into a deterministic finite automaton. Then the sequence of states to be checked forms the input for this automaton. 
@@
 
-The well foundedness of the formulas defining the new auxiliary variables in EPLTL makes the extension of the above
-runtime algorithm simple.
-In order to extend the algorithm to perform runtime verification of an EPLTL formula $\psi$, as defined in~(\ref{ELTL}),  we need to  add to the above
-algorithm memory for $\current ( a_j )$ and $\old ( a_j )$,
-and perform the following between steps 3 and 4 for each $a_j$: \\
-
-\ \ \ \ \ $\current ( a_j ) := \current ( \varphi_j )$ \\
-
-\subsubsection{Evaluation Order}
-
-The order of calculating $\current$ is subtle here, because
-auxiliary variables can appear recursively on the right-hand 
-side of their definition. For PLTL, one can start calculating the subformulas
-bottom up or top down. But for EPLTL, this would not work. To see this, consider, for example, the formula~(\ref{form3}). It contains
+\subsubsection*{Runtime verification for EPLTL.} For EPLTL,  we need to  add to the above
+algorithm calculations for the rules of
+the form $a_j := \varphi_j$ (calculations
+for the statement part $\psi$ remain the same).
+$\current ( a_j )$ and $\old ( a_j )$,
+and perform the following between steps~3 and~4 for each $a_j$: $\current ( a_j ) := \current ( \varphi_j )$.
+
+For PLTL, we can calculate  $\current$, for subformulas,
+bottom up. But
+because
+auxiliary variables can appear recursively in EPLTL rules, i.e., both on the left and the right sides,
+this would not work correctly. To see this, consider, for example, the formula~(\ref{form3}). It contains
 the definition $q := \ominus \neg q$. Now, we cannot calculate this
@@
 
-\klaus{I don't see that we need to talk about the top-down evaluation.}
+%\klaus{I don't see that we need to talk about the top-down evaluation.}
 
@@
 $\current ( \delta )$ for each subformula $\delta$
-that appears in $\varphi_j$ of rule
-$a_j := \varphi_j$ but not within the scope of
+that appears in $\varphi_j$ of a rule
+$a_j := \varphi_j$, but {\em not} within the scope of
 a $\ominus$ operator. (This is not
-a problem since for $\delta = \ominus \gamma$,
-$\current ( \delta )$ is set to $\old ( \gamma )$.)
+a problem, since
+$\current ( \ominus \gamma )$ is set to $\old ( \gamma )$.)
 \item Set 
@@
 \item Calculate $\current ( \delta )$ for each subformula $\delta$
-that appears in the statement $\psi$ - using $\current ( a_j )$.
+that appears in the statement $\psi$, using the calculated $\current ( a_j )$.
 \end{enumerate}
@@
 the values $5$ to $x$ and the value ``abc'' to $y$.
-Let $T$ be a set of {\em relations} 
-with typical instances $R$.
-Each relation is associated with some position-dependent domains.
-
-\klaus{Above: let T be ... Is T used later? The relations are used in the next paragraph, why not introduce them there? Also, relations are not defined really.}
-
-\klaus{I don't quite understand the motivation below for this way of modeling events: relations with a time stamp, and the whole story of the time stamp not being mentioned etc.
+%Let ${\cal R}$ be a set of {\em relations} 
+%with typical instances $R$.
+%Each relation is associated with some position-dependent domains.
+
+%\klaus{Above: let T be ... Is T used later? The relations are used in the next paragraph, why not introduce them there? Also, relations are not defined really.}
+%\klaus{I don't quite understand the motivation below for this way of modeling events: relations with a time stamp, and the whole story of the time stamp not being mentioned etc.
 Why not just the usual models: traces?
@@
 
-We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is,
-their last parameter
-is a natural number. For a relation $R$, $R [ i ]$ is the relation obtained from $R$ by
-restricting it to the value $i$ in the last parameter. For simplicity, we will describe henceforth
+We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is relations with
+last parameter
+that is a natural number represented a time instance. So a tuple of
+a relation $R$ can be $(``a'', 5, ``cbb'', 3)$, where
+the $3$ is the time parameter. The last parameter $i$
+does not correspond to modeling physical time. This
+is rather used to allow the relations to have
+different tuples in different instances of $i$,
+corresponding to states, in the propositional temporal
+logics, so that we can define the temporal operators
+such as $\bigcirc$ or $\since$.
+
+For a relation $R$, $R [ i ]$ is the relation obtained from $R$ by
+restricting it to the value $i$ in the last parameter, and removing that last $i$ from the tuples. For simplicity, we will describe henceforth
 the logic with relations $R$ that have exactly two parameters, 
-hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$; this domain is fixed over all the values of the second parameter $i$.
-The definition of the logic over other numbers of parameters is quite straightforward, and our implementation of runtime verification fully supports it.
-
-\klaus{Above, what about instead of writing: ``this domain is fixed over all the values of the second parameter $i$'' to write:
-``this domain is independent of the values of the second parameter $i$'' - if that is what you mean to say.}
-
-\klaus{Above: we need to say explicitly what $i$ intuitively is meant to represent: positions, time, ....}
+hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$.
+The definition of the logic over other numbers of parameters is quite straightforward.
+e.g., when $R$ has one parameter,
+$R [ i ]$ is a Boolean value. Our implementation of runtime verification described later fully supports relations with zero or more parameter.
+
+%\klaus{Above, what about instead of writing: ``this domain is fixed over all the values of the second parameter $i$'' to write:
+%``this domain is independent of the values of the second %parameter $i$'' - if that is what you mean to say.}
+
+%\klaus{Above: we need to say explicitly what $i$ intuitively is meant to represent: positions, time, ....}
 
@@
 
-Additional operators are defined as in the propositional logic. In addition we define
+Additional operators are defined as in the propositional logic. We שךדם define
 $\forall x \; \varphi = \neg \exists x \neg \varphi$.
-Restricting the modal operators to the past ones
+Restricting the modal operators to the past operators
 ($\since$, $\ominus$ and the ones obtained from them) 
 forms the logic PFLTL.
-The quantification over values of variables here is `static' in the sense that values quantified over are independent of the state in the execution. We
-denote static quantification with $\exists$ and $\forall$, while
-dynamic quantification is expressed with $\Exists$ and $\Forall$
-(see e.g. Section \ref{sec:extending-prop-ltl}).
+The quantification over values of variables here is `static' in the sense that they are independent of the state in the execution. We
+denote static quantification with $\exists$ and
+$\forall$.
+Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
+and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.
 
@@
 
-A model is a set of
- relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$, where each of these relations has a last parameter $i$ that is a natural number. 
-% 
-It is standard, and more intuitive (in particular for runtime verification) to cast models of temporal logic in terms of sequences states rather than temporal relations. To do that,
- we denote by $R_k [ i ]$ the restriction of
- the relation $R_k$ to the value $i$ of the last parameter. A model ${\cal R}$ is then a sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$, where
- each {\em state} ${\cal R} [i] = \{ R_1 [i], \ldots , R_m [ i ] \}$  (a set of relations).
- The denotation $p_k$ will correspond to the relation $R_k$.
-
-\klaus{Above, did you not already say the following before:
-``we denote by $R_k [ i ]$ the restriction of
- the relation $R_k$ to the value $i$ of the last parameter.''.}
+A model is a set of temporal
+ relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$.
+ Since the standard definition of temporal logic is
+ over a sequence (``the execution''). let
+ ${\cal R} [ i ] = \{ R_1 [ i ] \, \ldots , R_m [ i ] \}$. A model ${\cal R}$ can thus be
+seen as the sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$.
+
+%\klaus{Above, did you not already say the following before:
+%``we denote by $R_k [ i ]$ the restriction of
+% the relation $R_k$ to the value $i$ of the last parameter.''.}
 
@@
 We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We thus allow the following formula:
-
 \begin{equation}
@@
 \end{equation}
-
 \begin{enumerate}

### 2019-02-05 main.tex

@@
 
+\newpage
 \appendix
+
+\section{Appendix}
 
@@
 
-\noindent {\bf Sketch of proof of Theorem~\ref{theo2}.}
+\noisubsubsection{Sketch of proof of Theorem~\ref{theo2}.}
 As we defined the semantics of
@@
 
+%%%M
+\subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
+One can translate any QLTL property into a second order monadic logic in a standard
+way~\cite{Thomas}, and then from
+second order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
+each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set.
+A muller automaton
+accepts a run if it 
+passes infinitely often in exactly
+one of the specified accepting sets. To formulate an equivalent ELTL formula,
+the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
+
+%\klaus{The following formulation above is not not clear: ``where an accepted input needs to pass infinitely often in exactly
+%one of the specified sets''.}
+
+\noindent
+Then we translate $M$ into an ELTL formula
+of the form:
+
+\begin{tabbing}
+\ \ \ \ \ $\bigvee_{F \in {\cal F}} \bigwedge_{s \in F} ( \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
+where \\
+\ \ \ \ \ \ \ \ \ \ $a_s :=   ( \alpha_{\iota} \wedge \ominus \false ) \vee \bigvee_{\Delta ( s'
+, \alpha ) = s } 
+(\alpha \wedge \ominus a_{s'}) 
+\, :\,  s \in S$
+\end{tabbing}
+
+%\klaus{Above: should $s=i \rightarrow \neg \ominus \true$ instead be (note: two changes):
+%$\neg \ominus \true \rightarrow a_s=i$
+%}
+%\klaus{Above: can't judge whether the $\bigwedge_{s \not\in F} \Diamond \Box %\neg a_s)$ is correct.}
+\noindent
+Basically, the assignments to the auxiliary variables are used to simulate the change of states\footnote{It is possible to use only $\lceil \log ( | S | ) \rceil$
+variables, at the expense of longer and more complicated formulas.}, and the main formula expresses the Muller acceptance condition. \qed
+
 \end{document}

### 2019-02-05 main.tex
fixed figures

@@
 
-\noisubsubsection{Sketch of proof of Theorem~\ref{theo2}.}
+\subsubsection{Sketch of proof of Theorem~\ref{theo2}.}
 As we defined the semantics of

### 2019-02-05 rule-ltl.tex
fixed figures
fixed figs
Updates from Overleaf

@@
 \begin{center}
-\includegraphics[height=1.1in,width=2.7in]{PROP.jpg}
+%\includegraphics[height=0.6in,width=2in]{PROP.jpg}
+\includegraphics[width=3.5in]{PROP.jpg}
 \caption{\label{prop} An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness}
@@
 defined by the following grammar,
-where $p$ denotes a relation $R$,
-$a$ is a constant representing a value in $d ( R )$ and $x$ is a variable with $\domain ( x ) = d ( p )$.
+where $p$ is a relation,
+$a$ is a constant and $x$ is a variable. 
 \begin{center}
@@
 
-Additional operators are defined as in the propositional logic. We שךדם define
+Additional operators are defined as in the propositional logic. We define
 $\forall x \; \varphi = \neg \exists x \neg \varphi$.
@@
 forms the logic PFLTL.
-The quantification over values of variables here is `static' in the sense that they are independent of the state in the execution. We
-denote static quantification with $\exists$ and
-$\forall$.
-Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
-and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.
-
-\subsubsection{Semantics}
-
+
+\subsubsection*{Semantics.}
 A model is a set of temporal
@@
  Since the standard definition of temporal logic is
- over a sequence (``the execution''). let
- ${\cal R} [ i ] = \{ R_1 [ i ] \, \ldots , R_m [ i ] \}$. A model ${\cal R}$ can thus be
-seen as the sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$.
+ over a sequence (``the execution''), let
+ ${\cal R} [ i ] = \{ R_1 [ i ] \, \ldots , R_m [ i ] \}$. ${\cal R} [i]$ represents a {\em state}.
+ A model ${\cal R}$ can thus be seen as a sequence
+ of states ${\cal R}[1] {\cal R}[2] \ldots$. Let 
+ $m$ be a bijection
+from relation names (syntax)
+onto the relations ${\cal R}$ (semantics).
+%After the formal semantical definition, we will abuse notation
+%and use the same symbol for the representation of the relation in the logic and for the relation as mathematical object.
 
@@
 %
-Let $\epsilon$ be the empty assignment. 
+Let $\epsilon$ be the empty assignment (with no variables). 
 %
@@
 \item $( \epsilon ,  {\cal R} , i )
- \models p_k ( a ) $ if $R_k ( a, i )$, where $a$ denotes a constant from  $d(p)$.
-
-\item $( [ x \mapsto a ] , {\cal R} , i ) \models p_k ( x )$ if $R_k ( a , i )$.
+ \models p ( a ) $ if $m(p) ( a, i )$, where $a$ denotes a constant from  $d(m(p))$.
+
+\item $( [ x \mapsto a ] , {\cal R} , i ) \models p ( x )$ if $m(p)  ( a , i )$, where $\domain (x) = d ( m ( p ))$.
 
@@
 \item $( \gamma , {\cal R} , i ) \models ( \varphi \; \until \; \psi )$ if
-for some $j$, $j > i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
+for some $j$, $j \geq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
 \models \psi $ and for each $k$,  $i \leq k < j $,
@@
 
-Then for an FLTL (PFLTL) formula with no free variables, ${\cal R} \models \varphi$ iff $( \epsilon , {\cal R} , 1 ) \models \varphi$.
+For an FLTL (PFLTL) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
 We will henceforth sometimes abuse notation, as is very often done in logic, and use the same
@@
 the logic (syntax).
+The quantification over values of variables here is `static' in the sense that they are independent of the state in the execution. We
+denote static quantification with $\exists$ and
+$\forall$.
+Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
+and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.
 
@@
 
-We first demonstrate that the lack of expressiveness
+We demonstrate that the lack of expressiveness
 carries over from LTL (PLTL) to FLTL (PFLTL).
 
-\noindent {\bf Example.}
+\vspace{1ex}
+\noindent {\bf Example 1.}
 Let $p$ and $q$ be temporal relations, where their
-restrictions to the state $s_i$, $p[i]$ and $q[i]$, respectively, are unary relations $p(x)$ and $q(x)$. The
+restrictions to the state $s_i$, $p[i]$ and $q[i]$, respectively, are unary relations $p$ and $q$. The
 specification that we want to monitor is that for each value
@@
 Wolper's example over LTL, carried over to the first-order case. To show that this is not expressible in FLTL (and PFLTL),
-consider models (executions) where only one data element $a$ appears. We can prove that in this case, by a simple structural induction on (sub)formulas, that for each property $\varphi$
+consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
+that in this case, for each property $\varphi$
 we can replace occurrences of variables within
@@
 the occurrences of $p(a)$ and $q(a)$ by propositions
-$p_a$ and $p_b$, respectively. However, by~\cite{Wolper},
-such a formula cannot express the said property.
-This example demonstrates that the so called ``non-counting''~\cite{Thomas} property of LTL carries over
-to FLTL.
+$p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.
 
@@
 \noindent {\bf Example 2.}
-Consider the property that asserts that
-a main process $y$ (where
-$mathit{main} (x)$ appears
-in the sequence), is a descendant of a process $x$ if $x$ spawns a chain of processes spawning (of
-the form $\mathit{spawn} (z, w$ that leadsfrom $x$ to $y$. This property cannot be expressed in FLTL. The reason
+Consider the property that asserts that when
+$\mathit{inherent}(x , y)$ appears
+in a state, denoting that a process 
+$y$ wants to inherent variables from a process $x$,
+there were a chain of states with process spawning
+$\mathit{spawn}(x, x_1)$, $\mathit{spawn} (x_1, x_2)$ $\ldots$
+$\mathit{spawn}(x_l , y)$. The reason
 is that the required property needs to
@@
 Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
-E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, (p(x, i) \rightarrow \exists t' p(x, j) )$
-\footnote{The obtain model will have, in addition
+E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, ((p(x, i) \rightarrow \exists t' p(x, j) )$
+\footnote{The obtained model will have, in addition
 to the aforementioned relations, also
 the linear order $<$ over the naturals.}
-It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in this first order setting. The proof is based on the
+It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in first order setting. The proof is based on the
 compactness theory of first order logic~\cite{Flum}.
@@
 
-Extending FLTL (PFLTL) with quantifiers over relations, we obtain QFLTL
+Extending FLTL (PFLTL) with quantifiers over temporal relations, we obtain QFLTL
 (and the past-restricted version QPFLTL, respectively). 
@@
 where $p$ denotes an auxiliary relation.
-We also allow $\Forall p\ \varphi = \neg \Exists \neg \varphi$. The semantics is as follows.
-
-\klaus{Below, has the notation ${\cal R}' |_{\{ q \} }$ been defined?}
+We also allow $\Forall p\ \varphi = \neg \Exists \neg \varphi$.
+The semantics is as follows.
+%We denote by ${\cal R} |_Q$ the set of temporal relations obtained from ${\cal R}$ by removing the set of relations $Q$.
+
+%\klaus{Below, has the notation ${\cal R}' |_{\{ q \} }$ been defined?}
 
@@
 Consequently, quantification over relations effectively extends the model ${\cal R}$ into a model ${\cal R}'$ within the scope of the quantifier.
-Note that quantification here is dynmaic (as in QLTL and QPLTL).
-
-\subsubsection{ExtendingFLTL  with Rules.}
-
-We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We thus allow the following formula:
+Note that quantification here is dynmaic (as in QLTL and QPLTL), since the relations are temporal and can have different sets of tuples in different states.
+
+\subsubsection{Extending FLTL  with Rules.}
+
+We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We allow the following formula:
 \begin{equation}
 \label{EFLTL}
-\psi \mathit{\ where\ } r_j  ( x_j ) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} 
+\ \ \ \ \ \psi \mathit{\ where\ } r_j  ( x_j ) := 
+\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \mathrm{\ \ \ \     such\ that,}
 \end{equation}
 \begin{enumerate}
-\item $\psi$, the {\em statement}, is a FLTL formula with
+\item $\psi$, the {\em statement}, is an FLTL formula with
 no free variables, 
@@
 a natural number that is omitted in the temporal formulas. 
-An auxiliary relation $r_j$ can appear within $\psi$. It can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
-scope\footnote{This restriction has the same motivation as in
-the propositional version ELTL (and EPLTL): to avoid making the
-formula $\false$ just by means of the definition of the auxiliary relations.}
+An auxiliary relations $r_j$ can appear within $\psi$. They can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
+scope
 of a previous-time operator $\ominus$. 
@@
 relations $r_j$ with different number of arguments, and, correspondingly, formulas $\varphi_j$ with corresponding number of free
-variables\footnote{We chose not to do that here (and in the definition of the semantics of FLTL above), because it
-requires multi-indexing notation that is trivial and adds no further insight. Our implementation indeed uses
-this more general capability.}.
+variables.
 The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
@@
 \end{lemma}
-\noindent {\bf Proof.} by a simple induction, as in Lemma~\ref{fourone}. \qed
-
+\noindent {\bf Proof.} By a simple induction, similar to Lemma~\ref{fourone}. \qed
+\vspace{1ex}
 The following formula expresses the property that was shown to be 
@@
 with the auxiliary ones
-inductively, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL 
+inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL 
 specification (\ref{EFLTL}) 
@@
 \noindent 
-The analogous of Lemma~\ref{simplecase} shows that EPFLTL does not
+Analogously to the propositional case, it is easy to show that EPFLTL does not
 need the past temporal operators besides $\ominus$.
 
-\begin{lemma}
-The full expressive power of EPFLTL can be obtained by allowing only 
-specifications of the form
-\[ \Box \psi \mathrm{\ where\ } r_j  ( x_j ) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \]
-where $\psi$ consists of a fully quantified single occurrence of some
-auxiliary relation $r_j$.
-\end{lemma}
-
-\klaus{Something seems wrong in the above formulation.
-E.g.: Is the $\Box$ a restriction?, what is ``fully quantified single occurrence of some auxiliary relation $r_j$''?}
+%\klaus{Something seems wrong in the above formulation.
+%E.g.: Is the $\Box$ a restriction?, what is ``fully quantified single occurrence of some auxiliary relation $r_j$''?}
 
@@
 \end{theorem}
-The proof of this theorem, which includes the encoding of a nontrivial property, and a combinatorial argument about the amount of information
-needed after a given prefix, appears in an appendix.
-The relative expressive power between the first-order temporal logics discussed here
+The proof of this theorem, which includes the encoding of a nontrivial property, and a combinatorial argument, appears in an appendix.
+The relative expressive power between the first-order temporal logics presented in this Section
 appear in Figure~\ref{fsfr}.
@@
 \begin{center}
-\includegraphics[height=1.1in,width=2.7in]{FIRST.jpg}
+%\includegraphics[height=1.5in,width=3in]{FIRST.jpg}
+\includegraphics[width=3in]{FIRST.jpg}
 \caption{\label{fsfr} An arrow from A to B means that logic  B is more expressive than A}
@@
 input that consists of {\em events} in the form of
-tuples of relations. In our notation, the input
+tuples. In our notation, the input
 consists of a sequence ${\cal R}[1] \, {\cal R}[2]  \ldots$,
 which we earlier identified with states, where each
-${\cal R}[i]$ consists of some relations. A typical use
-of runtime verification restricts these relations to have,
-per each state just some finite tuples, and, quite commonly, have a single tuple from a single relation per each state.
-A set of assignments over the same variables, which has some fixed order
-(say alphabetical) also forms a relation.
-
+${\cal R}[i]$ consists of the relations in
+${\cal R}$,
+where the last parameter is restricted to  $i$. A typical use
+of runtime verification restricts the
+events (tuples) for
+each state to a finite number, and often even to a single event.
+The RV algorithm will also make
+use of sets of assignments over a set of variables for satisfying a subformula at
+some state (corresponding to $\old$ and $\current$, which were used with propositional values in Section~\ref{LTLruntime});  these also form a relation over the domains of these variables (having some fixed order).
+
+\ignore{
 \klaus{In the text above there are some imprecisions, such as
@@
 (4) what is the story about a set of assignments also being a relation? Needed?
-}
-
-\subsubsection{Syntax}
-
-\klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}
+}}
+
+%\subsubsection{Syntax}
+
+\ignore{
+\klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}}
 
 Consider again the EPFLTL form
-of specification, described in~(\ref{EFLTL}), restricted to the past operators. We now present the definition of EPFLTL, the final main logic proposed in this paper. 
+of specification, described in~(\ref{EFLTL}).
+
+\iffalse
+, restricted to the past operators. We now present the definition of EPFLTL, the final main logic proposed in this paper. 
 The syntax is as follows:
@@
 \varphi ::= ...
-\]
-}
+\]}
+\fi
 
@@
 %union. Accordingly, $\mathbf{T}$ is idempotent to intersection.
-\fi
 
@@
 set semantics is needed. Maybe it is needed, since we probably need a semantics. Something to consider.}
+\fi
 

### 2019-02-05 implementation.tex
Updates from Overleaf
fixed third property

@@
 prop spawning :
-  Forall x . Forall y . 
-    spawn(x,y) ->  (P main(x) | Exists z . (P main(z) & leadsto(z,y))) 
-    where leadsto(z,y) := 
-      spawn(z,y) | @ leadsto(z,y) | Exists x . (@leadsto(z,x) & spawn(x,y))
+  Forall x . Forall y . Forall d . report(y,x,d) -> spawned(x,y) where
+    spawned(x,y) := @ spawned(x,y) | spawn(x,y) | 
+                    Exists z . (@spawned(x,z) & spawn(z,y))
 \end{lstlisting}
@@
 about threads being spawned in an operating system. We want to 
-ensure that all threads are spawned by the main program or are spawned by a thread that transitively can be traced back to a thread being spawned by the main program. The events are
-\idsl{main(x)} (the main program spawns thread \idsl{x}) and
-\idsl{spawn(x,y)} (thread \idsl{x} spawns thread \idsl{y}).
-For this we need to compute a transitive closure of spawning relationships, here expressed with the rule \idsl{leadsto(z,y)}.
+ensure that when a thread \idsl{y} reports some data \idsl{d} 
+back to another thread \idsl{x}, then thread \idsl{y} has been spawned by thread \idsl{x} either directly, or transitively 
+via a sequence of spawn events. The events are
+\idsl{spawn(x,y)} (thread \idsl{x} spawns thread \idsl{y})
+and \idsl{report(y,x,d)} (thread \idsl{y} reports data \idsl{d}
+back to thread \idsl{x}).
+For this we need to compute a transitive closure of spawning relationships, here expressed with the rule \idsl{spawning(x,y)}.
 
@@
 analysis time (excluding time to compile the generated 
-monitor) for different traces (format is `trace length : seconds').
-The telemetry traces will repeatedly open hundreds of channels and send telemetry over them, and the spawning traces will spawn a new thread for each event in the trace.
-The transitive closure of the spawning property needs to process these, therefore the more extensive processing time.
+monitor) for different traces (format is `trace length : seconds'). The telemetry traces will repeatedly open hundreds of channels and send telemetry over them. 
+%
+The spawning traces consist of 50\% spawning events and 50\% reporting events.
+The spawning property  requires larger processing time compared to the telemetry properties since more data (the transitive closure) has to be computed and stored.
 
@@
 \hline
-spawning   & 4,950\,:\,9.4s & 10,000\,:\,88.4s & 19,900\,:\,488.8s \\
+spawning   & 9,899\,:\,42.7s & 19,999\,:\,127.6s & 39,799\,:\,572.4s \\
 \hline

### 2019-02-07 implementation.tex

@@
 
-\subsubsection{Example Properties}
-
+\vspace{1ex}\noindent {\bf Example Properties.}
 Figure \ref{fig:properties} shows three properties in the 
 input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}.
+The properties are related to the examples in Section~\ref{sec:syntax-semantics}, which
+are not expressible in (P)FLTL.
 %
@@
 
-\begin{center}
-\begin{figure}
-\begin{lstlisting}[language=dsl,frame=single,linewidth=0.95\textwidth,backgroundcolor=\color{white},linewidth=\columnwidth,breaklines=true,basicstyle=\small]
-prop telemetry1: 
-  Forall x . closed(x) -> !telem(x) where closed(x) := toggle(x) <-> @!closed(x)
-
-prop telemetry2: 
-  Forall x . closed(x) -> !telem(x) where
-    closed(x) := (!@true & !toggle(x)) | (@closed(x) & !toggle(x)) | 
-                 (@open(x) & toggle(x)),
-    open(x) := (@open(x) & !toggle(x)) | (@closed(x) & toggle(x))
-
-prop spawning :
-  Forall x . Forall y . Forall d . report(y,x,d) -> spawned(x,y) where
-    spawned(x,y) := @ spawned(x,y) | spawn(x,y) | 
-                    Exists z . (@spawned(x,z) & spawn(z,y))
-\end{lstlisting}
-\caption{Three properties stated in \dejavu's logic}
-\label{fig:properties}
-\end{figure}
-\end{center}
+%%%%
 
@@
 
+\begin{figure}
+\begin{center}
+\begin{lstlisting}[language=dsl,frame=single,linewidth=0.95\textwidth,backgroundcolor=\color{white},linewidth=\columnwidth,breaklines=true,basicstyle=\small]
+prop telemetry1: 
+  Forall x . closed(x) -> !telem(x) where closed(x) := toggle(x) <-> @!closed(x)
+
+prop telemetry2: 
+  Forall x . closed(x) -> !telem(x) where
+    closed(x) := (!@true & !toggle(x)) | (@closed(x) & !toggle(x)) | 
+                 (@open(x) & toggle(x)),
+    open(x) := (@open(x) & !toggle(x)) | (@closed(x) & toggle(x))
+
+prop spawning :
+  Forall x . Forall y . Forall d . report(y,x,d) -> spawned(x,y) where
+    spawned(x,y) := @ spawned(x,y) | spawn(x,y) | 
+                    Exists z . (@spawned(x,z) & spawn(z,y))
+\end{lstlisting}
+\caption{Three properties stated in \dejavu's logic}
+\label{fig:properties}
+\end{center}
+\end{figure}
+
+\iffalse
 \subsubsection{Monitor Synthesis}
@@
 
-\subsubsection{Evaluation}
+\fi
 
+\vspace{1ex}\noindent {\bf Evaluation.}
 We evaluated the three properties in Figure 

### 2019-02-07 conclusion.tex
Updates from Overleaf

@@
 
-We then demonstrated that first-order linear temporal logic (FLTL), which
+We demonstrated that first-order linear temporal logic (FLTL), which
 can be used to assert properties about systems with data, also has expressiveness deficiencies, and extended it with relations that summarize prefixes.

### 2019-02-07 main.tex
Updates from Overleaf

@@
 
+%\author{Submitted Anonymosly}
+
 %\thanks{Copyright 2018. All rights reserved.}
 
+\iffalse
 \author{ 
@@
 }
+
+\fi
 
@@
 expressiveness in the propositional case, and several 
-theoretic (non-operational) extensions have therefore been proposed. Furthermore, for many practical cases, there is a need to monitor properties 
+theoretic extensions have therefore been proposed. Furthermore, for many practical cases, there is a need to monitor properties 
 that carry data, where one can use formalisms like 
 first-order LTL. We show that 
-first-order LTL has similar expressive
+first-order LTL has similar expressiveness
 limitations as the propositional version. We suggest here 
@@
 properties. Finally, we expand the BDD-based runtime verification 
-tool \dejavu{}  to support our extension and perform some 
+tool \dejavu{}\footnote{The actual tool name is not given due to anonymous reviewing.}  to support our extension and perform some 
 experiments.
@@
 
-\klaus{In the text below mention what $i$ is (time point, position, ...}
+%\klaus{In the text below mention what $i$ is (time point, position, ...}
 
-\subsubsection{Sketch of proof of Theorem~\ref{theo2}.}
-As we defined the semantics of
-EFLTL by translating into QFLTL, QFLTL is at least as
-expressive as EFLTL. We show here that 
-the expressive power of EFLTL is strictly smaller than that of
-QFLTL through the following example. It consists of
-a temporal relation\footnote{We abuse notation and
-use the same name for the relation and for its
-denotation in the logic. Also, $r [ i ]$ has zero
-parameters; as noted before, this is a natural extension to the syntax and semantics presented above.} $r$ with only one parameter, so $r [ i ]$
-has a Boolean value. We will henceforth say that $r$ occurs
+\subsubsection{Construction of a property
+for Theorem~\ref{theo2}.}
+The model consists of
+a temporal relation.
+%\footnote{We abuse notation and
+%use the same name for the relation and for its
+%denotation in the logic. Also, $r [ i ]$ has zero
+%parameters; as noted before, this is a natural %extension to the syntax and semantics presented %above.} 
+$r$ with only the time parameter, so $r [ i ]$
+has a Boolean value at the $i$th state. We will henceforth say that $r$ occurs
 in state $i$. Another relation is $v$, with
@@
 
-\klaus{Below: isn't it the other way around, the first is not a model and the second is a model?}
+%\klaus{Below: isn't it the other way around, the first is not a model and the second is a model?}
 
 Hence, the sequence of states 
-\[ \{ r \} \{ v(1) \}  \{ v(2) \} \{ v(3) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(3) \} \{  r \} \] is a model of this formula, while \[ \{ r \} \{ v(1) \}  \{ v(2) \}  \{ v(2) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(5) \}  \{  r \}\] is not.
-The QFLTL formula is a bit involved, hence given in several parts (macros).
+\[ \{ r \} \{ v(1) \}  \{ v(2) \}  \{ v(2) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(5) \}  \{  r \}\]
+is a model of this formula, while 
+\[ \{ r \} \{ v(1) \}  \{ v(2) \} \{ v(3) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(3) \} \{  r \} \]
+is not.
+The QFLTL formula is a bit involved, hence given in several parts (macros), where the main formula appears last.
 
@@
 \item $init ::= r \wedge \Box (r \leftrightarrow \neg \exists x \, v(x))$ [$r$ holds in the first state, and $v(a)$ with some value $a$ holds in states where $r$ does not hold.]
-\item $init \wedge \neg \Exists A \Exists B (equal (A, B) \wedge \\
+\item $init \wedge \Box \neg \Exists A \Exists B (equal (A, B) \wedge \\
 collect(B, same(A),\, same(A)) \ \since \ collect(A, true, true))$
@@
 
-\klaus{Above: maybe need to point out what the top formula is.}
+%\klaus{Above: maybe need to point out what the top formula is.}
 
-\klaus{Above: isn't there missing a $\Box$ in the main formula?}
+%\klaus{Above: isn't there missing a $\Box$ in the main formula?}
 
-We prove that the above property cannot be expressed in EFLTL.
-We use a combinatorial argument.
-In EFLTL, after every prefix of the input, it is
+\iffalse
+We prove that the above property cannot be expressed in EPFLTL.
+We use a combinatorial argument, showing that
+the information that is available in any EPFLTL
+property
+after a prefix is not sufficient to distinguish between two models (executions), one satisfying and one not satisfying it.
+
+In EPFLTL, after every prefix of the input, it is
 sufficient to keep a finite set of (non temporal) relations over the
@@
 of a property a prefix of length $i$ only depends on
-these relations. Now, the number of such possible relations
-for a fixed number of data elements $N$ is polynomial
-in $N$. However, the above property checks whether
-a set of values that appear within the relation $v$, between two consecutive appearances
-of $r$'s, are the same. For that, one needs to keep a set of
-subsets of the elements that appeared in the input so far. 
+these relations. Now, the amount of information
+that summarizes an observed prefix for
+EPFLTL is is polynomial
+in $N$, consisting of a finite number
+of relations over
+$N$ elements. The number of different tuples
+for such relations can be shown, inductively, to
+be polynomial in $N$ (beach parameter
+of each relation can be one out of the $N$ possibilities, or a representation the values not seen yet).
+However, the above property depends on whether
+a set of values that appear as parameters to
+the relation $v$, between two consecutive appearances
+of $r$'s, are the same. For that, one needs to keep some representation of all the
+subsets of the elements that appeared in the input so far; by contradiction, any summary representation of the observed prefix that
+would identify two prefixes with some different sets, would not be able to distinguish between
+sequences that extend this prefix with the set
+of elements that appears in one prefix but not the other.
 But the number of such subsets is exponential in $N$. \qed
+\fi
 
-\klaus{Above: why set of subsets?}
+%\klaus{Above: why set of subsets?}
 
@@
 
+\subsubsection*{Set Semantics.}
+
+Let $\Gamma$ be
+a set of assignments over the variables $W$, and
+$U \subseteq W$.
+Then $\proj ( \Gamma , U )$ (for ``projecting {\em out}'' or  ``hiding''
+the variables $U$)
+is the largest set of assignments over 
+$W \setminus U$, 
+each agreeing with some
+assignment of $\Gamma$ on all the variables in $W \setminus U$.
+Let $U \cap W = \emptyset$, then
+$\ext ( \Gamma , U )$ is the largest set of assignments
+over $W \cup U$, where each such assignment agrees 
+with some assignment in $\Gamma$ on the
+values assigned to the variables $W$. 
+\iffalse
+This means that we extend $\Gamma$ by
+adding arbitrary values to the variables in $U$ from their domains.
+Then $\proj ( \ext ( \Gamma , U ) , U ) = \Gamma$ holds.
+We define the union and intersection operators on sets of
+assignments, even if they are defined over non identical
+sets of variables. 
+In this case, the assignments are extended
+over the union of the variables. 
+\fi
+Thus, if $\Gamma$ is a
+set of assignments over $W$ and $\Gamma'$ is
+a set of assignments over $W'$, then
+$\Gamma \; \bigcup \;  \Gamma'$ is defined as $\ext (\Gamma , W' \setminus W ) \cup
+\ext ( \Gamma' , W \setminus W' )$ and
+$\Gamma \; \bigcap \; \Gamma'$ is $\ext (\Gamma , W' \setminus W ) \cap
+\ext ( \Gamma' , W \setminus W' )$.  Hence, both are defined
+over the set of variables $W \cup W'$.
+
+We denote by $A_{\vars ( \varphi )}$ the set of all possible assignments
+of values to the variables that appear free
+in $\varphi$. Thus,
+$I [ \varphi, {\cal R} , i] \subseteq A_{\vars ( \varphi )}$.
+To simplify definitions, we add
+a dummy position $0$ for sequence $\sigma$ (which starts with $s_1$), 
+where every formula is interpreted as an empty set.
+Observe that the value $\emptyset$ and $\{ \epsilon \}$, behave
+as the Boolean constants $0$ and $1$, respectively.
+The set semantics is defined as follows, where $i \ge 1$.
+
+%\klaus{Above: assignments are introduced without any gentle introduction.
+Remember: you are not communicating with Moshe Vardi.
+You are communicating with John Fingelstein who had three glasses of wine at dinner.}
+
+%\klaus{Above: $I [ \varphi, {\cal R} , i]$ is introduced out of the blue without any explanation.}
+
+\begin{itemize}
+\item $I [ \varphi , {\cal R} , 0 ] = \emptyset$.
+\item $I [ \true , {\cal R} , i ] = \{ \epsilon \}$.
+\item $I [ p ( a ) , {\cal R} , i ] =$ if $m (p )  ( a , i ) $ then
+$\{ \epsilon \}$ else $\emptyset$.
+\item $I [ p ( x ) , {\cal R} , i ] = \{ [ x \mapsto a ] | m (p) ( a , i)  \}$.
+%\item $I [ \true , \sigma , i ] = D^n$.
+%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
+%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
+\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
+I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
+\item $I [ \neg \varphi , {\cal R} , i ] = 
+A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , {\cal R} , i ]$.
+\item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
+I [ \psi , {\cal R}, i ] \; \bigcup \;
+( I [ \varphi , {\cal R} , i ] \bigcap 
+I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
+\item $I [ \ominus \varphi , {\cal R} , i ] = I [ \varphi , {\cal R} , i-1 ]$.
+\item $I [ \exists x \; \varphi , \sigma , i ] = 
+\proj ( I [ \varphi , {\cal R} , i ] , \{ x \} )$.
+\item $I [ \psi \mathrm{\ where\ } r_j ( x_j) := 
+\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} , {\cal R} , i ] = 
+ I [ \psi , {\cal Q} ] $ where ${\cal Q} = {\cal R} \cup
+\{ r_j | r_j [ i ] =  I [ 
+\varphi_j , {\cal Q} , i ] \}$
+\end{itemize}
+Note that the last time interprets a formula of
+the form defined in~(\ref{EFLTL}).
+
+%\klaus{Above: as the semantics is presented, it looks as if 
+%the `$\psi\ where ...$' is just another subformula.}
+
+%\klaus{Above: hide is not defined.}
+
+%\klaus{Above: the semantics of $p(a)$ and $p(x)$ do not refer to 
+%${\cal R} $.}
+
+In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame}.
 \end{document}

### 2019-02-07 rule-ltl.tex
Updates from Overleaf

@@
 in various ways.
-We are mainly motivated by runtime verification of executions with data, for which a first-order temporal logic is
+We are mainly motivated by runtime verification of executions {\em with data}, for which a first-order temporal logic is
 appropriate. There, too, we show that expressiveness can be extended.
 
-We present first an extension of the propositional LTL that is based on using additional (auxiliary) propositions, not appearing in the execution, which summarize the past 
+We present first an extension of  propositional LTL that is based on using additional (auxiliary) propositions, not appearing in the execution, which are dependent on the past 
 of the current execution.
@@
 order case to motivate the need for
-extending first-order LTL. Another example is based on the inexpressiveness of transitive closure of a relation in first order logic.
-We present an extension where auxiliary relations, not appearing in the model, are
+extending first-order LTL.
+Another example is based on the in-expressiveness of transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
+We present an extension where auxiliary relations, not appearing in the execution, are
 added. While this extension
@@
 properties, we show that for the first-order version, this has less 
-expressive power than adding quantification over relations, in distinction from the corresponding propositional case.
-
-These extensions to the logics are in particular natural for
+expressive power than adding alternating quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
+provides the full expressive power of, say B\"{u}chi automata.
+
+The extensions we proposed to the logics are in particular natural for
 runtime verification of past (safety) temporal properties, since
@@
 on calculating and updating a summary of the prefix of the execution.
-Runtime verification often concentrates on a past  version of temporal logic~\cite{HR} mainly because for these properties we can decide when the monitored property is violated immediately when this happen, while for future properties we may never know for sure.
+Runtime verification is often restricted to a {\em past}  version of temporal logic~\cite{HR}, mainly because for past properties we can decide when the monitored property is violated immediately when this happens, while for future properties we may never know for sure.
 Consistent with that,
@@
 logics. We further present an extension of the 
-\dejavu{} tool (https://github.com/havelund/dejavu) that
+\dejavu{} tool (publicly available on {\sf github},
+hidden here because of anonymous reviewing)
+\ignore{(https://github.com/havelund/dejavu)} that
 realizes the extended first-order past temporal logic verification.
-The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first order temporal logic over infinite domains (e.g., the integers, strings).
+The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first-order temporal logic over infinite domains (e.g., the integers, strings).
 It achieves efficiency by BDD representation of enumerations of data (rather than over the data values themselves) and using
 a garbage collection algorithm.
-Finally, we provide some experimental results.
 
@@
 On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions.
-On the practical side, we present algorithms for the extended logic, and 
+On the practical side, we also present RV algorithms for the extended logic, and 
 show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.
@@
 In Section~4 we review first order linear temporal logic, present our extension and study the relationships between the presented versions of first-order LTL.
-IIn Section~5 we present a runtime verification algorithm for the past version of the extended first-order LTL.
+In Section~5 we present a runtime verification algorithm for the past version of the extended first-order LTL.
 Section~6 describes the implementation and provides experimental results.
@@
 where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
-in~\cite{Wolper}, adds
-state-dependent quantification over propositions to LTL.
+in~\cite{Wolper}, adds dynamic, i.e.,
+state-dependent, quantification over propositions to LTL.
 
@@
 data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
-\item{Q} adds (second order) Q{\em uantification} that is state-dependent. This is referred to as {\em dynamic} quantification.
-\item{E} allows E{\em xtending} the objects that occur in the input with
-new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) and past temporal logic rules that define them.
+\item{Q} adds (second order) Q{\em uantification} that is dynamic (state-dependent). This is referred to as {\em dynamic} quantification.
+\item{E} allows E{\em xtending} the 
+set of objects that occur in the input with
+new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
 \end{itemize}
@@
 \section{Propositional LTL}
+\label{propLTL}
 
@@
    \, ( \varphi \since \psi ) \,  \]
-where $p$ is a proposition from a finite set of propositions $P$, and with $\bigcirc$, $\until$, $\ominus$, $\since$ standing for {\em next-time}, 
+where $p$ is a proposition from a finite set of propositions $P$, and $\bigcirc$, $\until$, $\ominus$, $\since$ stand for {\em next-time}, 
 {\em until}, {\em previous-time} and {\em since}, respectively.
@@
 $\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
-where $s_i \subseteq P$ for each $i\geq 0$. These are the propositions that {\em hold} in that state.
-We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
+where $s_i \subseteq P$ for each $i\geq 1$. These are the propositions that {\em hold} in that state.
+%We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
 LTL's semantics is defined as follows:
@@
 and for each $k$, $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
-\item $(\sigma , i) \models \ominus \varphi$ iff $| \sigma | > 1$ and $(\sigma, i-1) \models \varphi$.
+\item $(\sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
 \item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $j$, $1 \leq j \leq i$, such that
@@
 \end{itemize}
-Then $\sigma \models \varphi$ when $( \sigma ,0 ) \models \varphi$.
+Then $\sigma \models \varphi$ when $( \sigma ,1 ) \models \varphi$.
 We can use the following abbreviations:
@@
 ${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}), and 
-${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$.
+${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$ ({\bf H} stands for 
+H{\em istory}).
 
 %%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
-The expressive power of propositional LTL is often compared to regular expressions
+The expressive power of different versions of propositional LTL is often compared to regular expressions
 over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second order logics where the temporal propositions
 correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
-LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star (or $\omega$-star for infinite words).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}.
+LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star operator (or $\omega$).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}.
 In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
@@
 An important subset of LTL, called here PLTL, allows only past temporal
-operators: $\since$, $\ominus$ and the ones derived
-from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is often
-(and also here) interpreted over finite sequences, 
+operators: $\since$, $\ominus$ and the operators derived
+from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is sometimes interpreted over finite sequences, 
 where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
 Alternatively,
-it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; there, each of the prefixes has to satisfy $\varphi$. This later
+it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; in this case, each of the prefixes has to satisfy $\varphi$. This later
 form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences. 
-
-When PLTL is interpreted over finite sequences, its
-expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.
+(When PLTL is interpreted over finite sequences, its
+expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.)
 
@@
 \subsubsection{Extending LTL with Dynamic Quantification.}
-Adding quantification over propositions allows writing a formula of the form
+Adding quantification over propositions, as done in~\cite{Wolper}, allows writing a formula of the form
 $\Exists q \, \varphi$, where
-$\Exists q$ repesents {\em existential} quantification over a propositiona $q$ that can appear in $\varphi$.
+$\Exists q$ represents {\em existential} quantification over a proposition $q$ that can appear in $\varphi$.
 
@@
 \neg \Exists q\, \neg \varphi$.
-This kind of quantification is considered to be ``dynamic'', since the quantified variables can have different truth values depending on the states It is also callsed  "second order" quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
-Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions (over infinite words), full B\"{u}chi
+This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states.
+It is also called  {\em second order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
+Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
 automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
@@
 
-\noindent
 Wolper's property can be expressed in QLTL as follows:
@@
 
-If we allow
-quantification over propositions, we obtain the logic PQLTL
-(QLTL restricted to the past modalities), with the following formula:
+Restricting QPLTL to the past modalities, one obtain the logic PQLTL. PQLTL has the same expressive
+power as regular expressions and finite automata.
+Equation~(\ref{form0}) can
+be rewritten in PQLTL as:
 \begin{equation} \label{form1} \Exists q \,
@@
 %In fact, it can be shown that a single existential quantifier is sufficient. 
-PQLTL has the same expressive
-power as regular expressions and finite automata.
-Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
-PQLTL property into a deterministic
-automaton.
+Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
+$q$ is $\false$ in the initial state. Then $q$ alternates between even and odd states.
+
+%Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
+%PQLTL property into a deterministic
+%automaton.
 
@@
 
-We present now an extension of LTL, which we call ELTL. It has the same
-expressive power as  QLTL.
-It has a form that puts it, conceptually, in between temporal
-logic and automata: on the one hand, it is possible to simply write
-a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. As will be showed later, this extension is
+We introduce now another extension of LTL, which we call ELTL.
+%It has a form that puts it, conceptually, in between temporal
+%logic and automata: on the one hand, it is possible to simply write
+%a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. 
+As will be showed later, this extension is
 is very natural for runtime verification.
@@
 {\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
-and {\em basic propositions} $B$. A model for an ELTL formula has truth assignments only over the propositions in $B$.
+and {\em basic propositions} $B$.
+%A model for an ELTL formula has truth assignments only over the propositions in $B$.
 An ELTL property $\eta$ has the following form: 
@@
  {j \in \{1, \ldots , n\}} \end{equation}
-where each $a_i$ is a distinct auxiliary proposition from $A$,
+where each $a_j$ is a distinct auxiliary proposition from $A$,
 $\psi$ is an LTL property and each $\varphi_i$ 
@@
 as the {\em statement} of $\eta$ and to 
-$a_j  := \varphi_j \, : \, 
- {j \in \{1, \ldots , n\}}$ as the {\em rules}. In text, rules
- will be separated from each other by a comma (",").
-
+$a_j  := \varphi_j$ as a {\em rule}. In text, rules
+ will be separated from each other by commas. 
 \begin{tabbing}
@@
 The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
-and each {\em even} state will satisfy $p$.  As $\ominus \varphi \models \false$ in the initial state, regardless of $\varphi$, then
-$q$ is $\false$ in the initial state, and then it subsequently alternates.
-
-ELTL extends the set of propositional variables with new variables whose values at a state are
-based on the prefix of the model. The added propositions are dependent and effectively summarize the past execution prefix up to and
+and each {\em even} state will satisfy $p$.
+
+ELTL extends the set of propositional variables with new variables, whose values at a state are
+functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
 including the current state.
 This differs from 
-the use of auxiliary variables in QLTL, where the auxiliary variables do not have to extend the execution
-states in a fixed way throughout the interpretation of 
+the use of auxiliary variables in QLTL, where the values assigned to the
+auxiliary variables do not have to extend the 
+states of the model in a unique way throughout the interpretation of 
 the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
@@
 is required to help preventing that the values
-of the auxiliary variables are not conflicting, as in
+of the auxiliary variables are conflicting, as in
 $a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
-that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write conflicting rules, e.g., by simlpy 
-adding the rule $a: = \false$.
+that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write conflicting rules, e.g., by using the rule $a_1 : = \false$.
 
@@
 \noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
-is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states. A PLTL
-property depends in a state only on the prefix
-of the model that precedes and includes that state. \qed
+is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
+%A PLTL property depends in a state only on the prefix of the model that precedes and includes that state.
 
@@
 Each ELTL formula $\eta$, as defined above,
-is expressible in QLTL using the following formula:
+is expressible using the following equivalent QLTL formula:
 
@@
 
+\noindent
 For the other direction, one can translate the QLTL property first into
-a second order monadic logic formula, then to a deterministic Muller automata and then constructs an ELTL formula that holds for the accepting executions of this automaton. The details of the
+a second order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton.
+The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
+The details of the
 constructions are given in the appendix. \qed
@@
 
-We extend PLTL into EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$). This results in
-a formalism that is equivalent to finite automata, where all the states except one are accepting and where the nonaccepting state is a sink. 
-We can use related, but simpler construction than in Theorem~\ref{theo1} to prove the following:
+We define EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$) in ELTL. This results in
+a formalism that is equivalent to finite automata, where all the states except one are accepting and where the non-accepting state is a sink. 
+We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:
 
@@
 The expressive power of 
-ELTL is the same as QLTL.
+EPLTL is the same as QPLTL.
 %Any safety QPLTL property can be expressed using the following restricted
@@
 {j \in \{1, \ldots , n\}}  \] 
-with $p$ being one of the auxiliary variables $a_j$. The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.
+with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
+of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part, may provide a more intuitive specification.
+The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.
+An arrow from A to B means that logic  B is more expressive than A.
+A double headed arrow shows same expressiveness
 
@@
 %\includegraphics[height=0.6in,width=2in]{PROP.jpg}
-\includegraphics[width=3.5in]{PROP.jpg}
-\caption{\label{prop} An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness}
+\includegraphics[width=2.8in]{PROP.jpg}
+\caption{\label{prop}
+Comparing the expressive power of the temporal logics in Section~\ref{propLTL}
+}
 \end{center}
@@
 
-Runtime verification of temporal specification often
-concentrates on the past portion of the logic.
-Past-time  specifications have the important property that one can distinguish when they are violated.
-For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.
+Runtime verification of temporal specifications often
+concentrates on the past portion of the logic.  Past-time  specifications have the important property that one can distinguish when they are violated after observing a finite prefix of
+an execution. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.
 
@@
 \end{itemize}
-Runtime verification exploits the fact that
-the semantic definition is recursive in both the structure of the property, where subformulas
+Runtime verification for a PLTL formula $\eta$ exploits the fact that
+the semantic definition is recursive in both the structure of the property. Thus, subformulas
 are evaluated based on smaller subformulas, and the
 evaluation of subformulas in the previous state.
-The algorithm, shown below, operates on two vectors (arrays) of values indexed by subformulas:  $\old$ that summarizes the truth values of the
-subformulas just {\em before} the new state, and $\current$ for the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
+The algorithm, shown below, uses two vectors (arrays) of values indexed by subformulas:  $\old$, which summarizes the truth values of the
+subformulas for the execution prefix that
+ends just {\em before} the new state, and $\current$, for the execution prefix that
+ends with the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
 \begin{enumerate}
-\item Initially, for each subformula $\varphi$,
+\item Initially, for each subformula $\varphi$
+of $\eta$,
 %$B ( \varphi , \current ) = \bfalse $.
 $\current ( \varphi ) := \false$.
-\item Observe a new event (as a set of ground predicates) $s$ as input. 
+
+\item Observe a new event (as a set propositions) $s$ as input. 
 \item Let $\old := \current$.
@@
       $\current ( \psi )$.
-
 \begin{itemize}
@@
   \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
-
 \end{itemize}
+\item If $\current (\eta) = \false$ then
+report a violation and terminate.
 \item Goto step~2.
@@
 \subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
-formula into a deterministic finite automaton. Then the sequence of states to be checked forms the input for this automaton. 
-However, QPLTL can be extremely compact, in particular when using alternating existential and universal 
+formula into a deterministic finite automaton, where one distinguished states, $s$, in a sink state.
+Then the sequence of states to be checked forms the input for this automaton, and violation is
+announced if $s$ is reached.
+QPLTL can be extremely compact, in particular when using alternating existential and universal 
 quantifiers, which can result in a huge (non-elementary) explosion of
@@
 \subsubsection*{Runtime verification for EPLTL.} For EPLTL,  we need to  add to the above
-algorithm calculations for the rules of
-the form $a_j := \varphi_j$ (calculations
-for the statement part $\psi$ remain the same).
-$\current ( a_j )$ and $\old ( a_j )$,
-and perform the following between steps~3 and~4 for each $a_j$: $\current ( a_j ) := \current ( \varphi_j )$.
-
-For PLTL, we can calculate  $\current$, for subformulas,
-bottom up. But
-because
-auxiliary variables can appear recursively in EPLTL rules, i.e., both on the left and the right sides,
-this would not work correctly. To see this, consider, for example, the formula~(\ref{form3}). It contains
+algorithm calculations 
+of $\current ( a_j )$ and
+$\current ( \varphi_j )$, for each of the rules of
+the form $a_j := \varphi_j$ (the corresponding
+$\old$ entries will be updated as in line~3 in the above algorithm).
+
+%(calculations
+%for the statement part $\psi$ remain the same).
+%$\current ( a_j )$ and $\old ( a_j )$,
+%and perform the following between steps~3 and~4 %for each $a_j$: $\current ( a_j ) := \current ( %\varphi_j )$.
+
+%For PLTL, we can calculate  $\current$, for subformulas,
+%bottom up. But
+Because the
+auxiliary variables can appear recursively in EPLTL rules, the order of calculation is
+subtle. To see this, consider, for example, the formula~(\ref{form3}). It contains
 the definition $q := \ominus \neg q$. Now, we cannot calculate this
-bottom up, since the left-hand $q$ is not calculated yet, and we need its value to compute the right-hand side $q$ leaf node.
-If we calculate it top down, we will encounter the same problem.
-However, notice that the top down calculation does not need the
-value of $q$ to calculate $\ominus \neg q$; this value should already be available as $\old ( \neg q )$.
+bottom up, as we did for PLTL, since $\current ( q )$ is not computed yet, and we need 
+to calculate $\current ( \ominus \neg q )$ in order 
+to compute $\current ( q )$.
+However, notice that the calculation is not dependent on the
+value of $q$ to calculate $\ominus \neg q$; 
+in Step~4 above, we have that
+$\current ( \ominus \; \varphi ) := \old ( \varphi )$
+so  $\current ( \ominus \neg q ) := \old ( \neg q )$.
 
@@
 
-Under {\em mixed evaluation order}, one calculates $\current$ in the following order. The calculation within each of the four categories can be performed top down or bottom up.
-
+Under {\em mixed evaluation order}, one calculates $\current$ in Step~4 in the following order. The first~4 steps are for the rules, and step~5 is for the statement $\eta$.
 \begin{enumerate}
@@
 that appears in $\varphi_j$ of a rule
-$a_j = \varphi_j$ {\em  within} the scope of
+$a_j := \varphi_j$ {\em  within} the scope of
 a $\ominus$ operator.
@@
 that appears in the statement $\psi$, using the calculated $\current ( a_j )$.
+\item Calculate $\current ( \eta )$ for the statement
+$\eta$.
 \end{enumerate}
@@
 that is a natural number represented a time instance. So a tuple of
-a relation $R$ can be $(``a'', 5, ``cbb'', 3)$, where
+a relation $R$ can be $(``a", 5, ``cbb", 3)$, where
 the $3$ is the time parameter. The last parameter $i$
@@
 corresponding to states, in the propositional temporal
-logics, so that we can define the temporal operators
-such as $\bigcirc$ or $\since$.
+logics.
 
@@
 The definition of the logic over other numbers of parameters is quite straightforward.
-e.g., when $R$ has one parameter,
-$R [ i ]$ is a Boolean value. Our implementation of runtime verification described later fully supports relations with zero or more parameter.
+e.g., when $R$ has one parameter, $i$,
+$R [ i ]$ is a Boolean value.
+Our implementation of runtime verification described later fully supports relations with zero or more parameter.
+This also applies to the extensions of the logic in the rest of this section, and to some of the examples used.
 
@@
 The formulas of the core FLTL logic are 
-defined by the following grammar,
-where $p$ is a relation,
-$a$ is a constant and $x$ is a variable. 
+defined by the following grammar,where $p$ denotes a relation,
+$a$ denotes a constant and $x$ denotes a variable. 
 \begin{center}
@@
 \noindent 
-
 Additional operators are defined as in the propositional logic. We define
@@
 Restricting the modal operators to the past operators
-($\since$, $\ominus$ and the ones obtained from them) 
+($\since$, $\ominus$ and the ones derived from them) 
 forms the logic PFLTL.
@@
 For an FLTL (PFLTL) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
-We will henceforth sometimes abuse notation, as is very often done in logic, and use the same
+We will henceforth sometimes abuse notation, and use the same
 symbols both for the relations (semantics) and their representation in
 the logic (syntax).
-The quantification over values of variables here is `static' in the sense that they are independent of the state in the execution. We
+The quantification over values of variables here is {\em static} in the sense that they are independent of the state in the execution. We
 denote static quantification with $\exists$ and
 $\forall$.
-Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
-and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.
+%Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
+%and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.
 
@@
 Let $p$ and $q$ be temporal relations, where their
-restrictions to the state $s_i$, $p[i]$ and $q[i]$, respectively, are unary relations $p$ and $q$. The
+restrictions to the state $s_i$, $p[i]$ and $q[i]$, are unary relations $p$ and $q$. The
 specification that we want to monitor is that for each value
-$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). This is an extension of
-Wolper's example over LTL, carried over to the first-order case. To show that this is not expressible in FLTL (and PFLTL),
+$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to).
+To show that this is not expressible in FLTL (and PFLTL),
 consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
@@
 $p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.
+Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015} can express this property, 
+where for each value $a$ there is a separate automaton that counts the number of times that $q(a)$ has occurred.
 
@@
 Consider the property that asserts that when
-$\mathit{inherent}(x , y)$ appears
-in a state, denoting that a process 
-$y$ wants to inherent variables from a process $x$,
-there were a chain of states with process spawning
+$\mathit{report}(y , x, d)$ appears
+in a state, denoting that  process 
+$y$ sends some data $d$ to a process $x$,
+there was a chain of states with process spawns
 $\mathit{spawn}(x, x_1)$, $\mathit{spawn} (x_1, x_2)$ $\ldots$
-$\mathit{spawn}(x_l , y)$. The reason
+$\mathit{spawn}(x_l , y)$. i.,e., $y$ is a descendent process of $x$. The reason
 is that the required property needs to
@@
 Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
-E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, ((p(x, i) \rightarrow \exists t' p(x, j) )$
-\footnote{The obtained model will have, in addition
+E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, ((p(x, t) \rightarrow \exists t' p(x, t') )$\footnote{The obtained model will have, in addition
 to the aforementioned relations, also
-the linear order $<$ over the naturals.}
+the linear order $<$ over the naturals.}.
 It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in first order setting. The proof is based on the
@@
 
-\subsubsection{Extending FLTL with Quantification.}
-
-This lack of expressiveness can be resolved by
-adding propositional variables (i.e., predicates that are
-only state dependent) with dynamic quantification over them.
-Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015} can also express this property.
+\subsubsection{Extending FLTL with Dynamic Quantification.}
 
 Extending FLTL (PFLTL) with quantifiers over temporal relations, we obtain QFLTL
-(and the past-restricted version QPFLTL, respectively). 
+(and the past-restricted version QPFLTL). 
 The syntax includes $\Exists p \, \varphi$, 
 where $p$ denotes an auxiliary relation.
-We also allow $\Forall p\ \varphi = \neg \Exists \neg \varphi$.
+We also allow $\Forall p\ \varphi = \neg \Exists \neg p \varphi$.
 The semantics is as follows.
@@
 \item $(\gamma ,  {\cal R}, i) \models \Exists q \, \varphi$ iff there exists
-${\cal R}'$ such that ${\cal R}' |_{\{ q \} } = {\cal R}$ and 
+${\cal R}'$ such that ${\cal R}' \setminus {\{ q \} } = {\cal R}$ and 
 $( \gamma, {\cal R}' , i) \models \varphi$.
@@
 Consequently, quantification over relations effectively extends the model ${\cal R}$ into a model ${\cal R}'$ within the scope of the quantifier.
-Note that quantification here is dynmaic (as in QLTL and QPLTL), since the relations are temporal and can have different sets of tuples in different states.
+Note that quantification here is dynamic (as in QLTL and QPLTL), since the relations are temporal and can have different sets of tuples in different states.
 
@@
 
-Again, it
-is simple to extend the syntax and semantics to include
-relations $r_j$ with different number of arguments, and, correspondingly, formulas $\varphi_j$ with corresponding number of free
-variables.
 The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
-$\since$ and $\ominus$, and those obtained from them.
+$\since$ and $\ominus$, and those derived from them.
 
@@
 The auxiliary temporal relations $r_1, \ldots r_n$ are
-uniquely defined from ${\cal R}$.
+uniquely defined by the rules for each model ${\cal R}$.
 \end{lemma}
@@
 \noindent
-We define\footnote{This is done for lack of space. Formal semantics 
-can be given by constructing a set of temporal relations extended
+We define\footnote{ Formal semantics 
+can also be given by constructing a set of temporal relations extended
 with the auxiliary ones
@@
 
+\ignore{
 \noindent {\bf Proof.} Basically, the proof is the same as Lemma~\ref{simplecase}, based on the axiomatization of the
-past temporal operators. \qed
+past temporal operators. \qed}
 
@@
 
-We showed in Theorem~\ref{theo1} that ELTL
-and QLTL have the same
+%We showed in Theorem~\ref{theo1} that ELTL (EPLTL)
+%and QLTL (PQLTL, respectively) have the same
 expressive power. We study now the relationship between
-EFLTL and QFLTL.
+%EFLTL (EPFLTL) and QFLTL (QPFLTL, respectively).
 
@@
 \label{theo2}
-The expressive power of EFLTL is stricktly weaker than that of QFLTL.
+The expressive power of EPFLTL is strictly weaker than that of QFLTL.
 \end{theorem}
-The proof of this theorem, which includes the encoding of a nontrivial property, and a combinatorial argument, appears in an appendix.
-The relative expressive power between the first-order temporal logics presented in this Section
-appear in Figure~\ref{fsfr}.
-
-It follows from~(\ref{equiv}) that EFLTL (EPFLTL, respectively) has the same
-expressive power as QFLTL (EPFLTL, respectively) restricted to having its
-dynamic 
-quantification to being existential, and prefixing the
-rest of the formula. Thus, we immediately obtain from Theorem~\ref{theo2} the following:
+
+\noindent {\bf Sketch of Proof.}
+The proof of this theorem, which includes encoding of a nontrivial property, where the property
+observes the data elements (appearing as $v (a)$)
+in between states where $r$ appears (a relation with
+one parameter, the time or index parameter $i$).
+The property asserts that there are no two sets of events that are equivalent.
+This property can be expressed in QPFLTL, and hence
+also in QFLTL. The details appear in the appendix.
+
+We use a combinatorial argument to show by contradiction that one cannot express this property
+using EFLTL, and also using EPFLTL. The reason is that every prefix of an EFLTL model, extended uniquely with the
+auxiliary relations, according to Lemma~\ref{samesame}, can be summarized with a finite number of relations, the ones in
+the model, the auxiliary relations and the
+assignments satisfying the subformulas. The size of
+each such relation is bounde by ${\cal O} (  m ^ N )$
+where $m$ is the number of values appearing in the
+prefix, and $N$ is the sum of the number of parameters on all the above mentioned relations.
+However, the number of different subsets over $m$ values is
+$2^m$. This means that with large enough number of different values, EPFLTL can have two prefixes with the same summary. It would then not be able to distinguish between two executions that extend
+this prefix, where one of them is extended with a set of values that appeared before and the other with 
+a set that already appeared. \qed
+
+%The relative expressive power between the first-order %temporal logics presented in this Section
+%appear in Figure~\ref{fsfr}.
+
+%It follows from~(\ref{equiv}) that EFLTL (EPFLTL, %respectively) has the same
+%expressive power as QFLTL (EPFLTL, respectively) %restricted to having its
+%dynamic 
+%quantification to being existential, and prefixing %the
+%rest of the formula.
+The following follows immediately from Theorem~\ref{theo2} the following:
 
@@
 \end{corollary}
+
+\ignore{
 
@@
 \end{figure}
-
-\section{RV for First-Order Past-Time LTL and its Extension}
+}
+
+\section{RV for past-time
+first-order  LTL and its Extension}
 \label{EPLTLRV}
@@
 ${\cal R}[i]$ consists of the relations in
-${\cal R}$,
-where the last parameter is restricted to  $i$. A typical use
+${\cal R}$ with
+the last parameter is restricted to  $i$. A typical use
 of runtime verification restricts the
-events (tuples) for
+events (tuples from the relations in ${\cal R}$) for
 each state to a finite number, and often even to a single event.
@@
 use of sets of assignments over a set of variables for satisfying a subformula at
-some state (corresponding to $\old$ and $\current$, which were used with propositional values in Section~\ref{LTLruntime});  these also form a relation over the domains of these variables (having some fixed order).
+some state, corresponding to $\old$ and $\current$, which had Boolean values in the 
+algorithm for propositional LTL in Section~\ref{LTLruntime}.
 
@@
 \klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}}
-
-Consider again the EPFLTL form
-of specification, described in~(\ref{EFLTL}).
 
@@
 
-\klaus{One may of course ask whether the following section on the 
-set semantics is needed. Maybe it is needed, since we probably need a semantics. Something to consider.}
+%\klaus{One may of course ask whether the following section on the 
+%set semantics is needed. Maybe it is needed, since we probably need a semantics. Something to consider.}
 \fi
 
-\subsubsection{Semantics}
-
-We denote by $A_{\vars ( \varphi )}$ the set of all possible assignments
-of values to the variables that appear free
-in $\varphi$. Thus,
-$I [ \varphi, {\cal R} , i] \subseteq A_{\vars ( \varphi )}$.
-To simplify definitions, we add
-a dummy position $0$ for sequence $\sigma$ (which starts with $s_1$), 
-where every formula is interpreted as an empty set.
-Observe that the value $\emptyset$ and $\{ \epsilon \}$, behave
-as the Boolean constants $0$ and $1$, respectively.
-The set semantics is defined as follows, where $i \ge 1$.
-
-\klaus{Above: assignments are introduced without any gentle introduction.
-Remember: you are not communicating with Moshe Vardi.
-You are communicating with John Fingelstein who had three glasses of wine at dinner.}
-
-\klaus{Above: $I [ \varphi, {\cal R} , i]$ is introduced out of the blue without any explanation.}
-
+\subsubsection{Set Semantics}
+
+The RV algorithm for
+(E)PLTL, presented in~Section \ref{LTLruntime}
+calculates
+$\current ( \eta )$, for $\eta$ a
+subformula of the monitored property, to be the
+truth value of $\varphi$ over the
+prefix inspected by the RV algorithm so far. For (E)PFLTL,
+$\current (\eta )$ consists of the
+sets of assignments (in the form of relations
+over the free variables in the subformula),
+rather than Boolean variables.
+
+We redefine the semantics for EPFLTL in
+equivalent way, which will be more directly related to the calculation of values
+in $\current$ by the
+RV algorithm that will be presented below. 
+Under the  {\em set semantics}
+(introduced in~\cite{HPU} for PFLTL, and extended here for EPFLTL),
+$I [ \varphi , \sigma, i ]$ is a function that returns
+a set of assignments such that $\gamma \in I [ \varphi , \sigma, i ]$ 
+iff $( \gamma , \sigma , i ) \models \varphi$.
+
+We present here only two simple cases of the set semantics. The full set semantics appears in the appendix. 
 \begin{itemize}
-\item $I [ \varphi , {\cal R} , 0 ] = \emptyset$.
-\item $I [ \true , {\cal R} , i ] = \{ \epsilon \}$.
-\item $I [ p ( a ) , {\cal R} , i ] =$ if $p ( a ) \in \sigma [ i ]$ then
-$\{ \epsilon \}$ else $\emptyset$.
-\item $I [ p ( x ) , {\cal R} , i ] = \{ [ x \mapsto a ] | p ( a ) \in
-\sigma [ i ] \}$.
-%\item $I [ \true , \sigma , i ] = D^n$.
-%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
-%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
 \item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
 I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
-\item $I [ \neg \varphi , {\cal R} , i ] = 
-A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , {\cal R} , i ]$.
 \item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
 I [ \psi , {\cal R}, i ] \; \bigcup \;
-( I [ \varphi , {\cal R} , i ] \bigcap 
+( I [ \varphi , {\cal R} , i ] \ \bigcap \ 
 I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
-\item $I [ \ominus \varphi , {\cal R} , i ] = I [ \varphi , {\cal R} , i-1 ]$.
-\item $I [ \exists x \; \varphi , \sigma , i ] = 
-\proj ( I [ \varphi , {\cal R} , i ] , \{ x \} )$.
-\item $I [ \psi \mathrm{\ where\ } r_j ( x_j) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} , {\cal R} , i ] = 
- I [ \psi , {\cal Q} ] $ where ${\cal Q} = {\cal R} \cup
-\{ r_j | r_j [ i ] =  I [ 
-\varphi_j , {\cal Q} , i ] \}$
 \end{itemize}
 
-\klaus{Above: as the semantics is presented, it looks as if 
-the `$\psi\ where ...$' is just another subformula.}
-
-\klaus{Above: hide is not defined.}
-
-\klaus{Above: the semantics of $p(a)$ and $p(x)$ do not refer to 
-${\cal R} $.}
-
-In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame}.
-
-\subsubsection{Runtime Verification Algorithm for PFLTL}
+\subsubsection{Runtime Verification Algorithm for PFLTL.}
 
@@
 in the tool~\dejavu. 
-Instead of storing the data values occurring in events, 
-we enumerate these data values as soon as we see them and use Boolean
-encodings of this enumeration, represented as BDDs.
+We enumerate data values appearing in
+monitored events, as soon as we first see them. We
+represent relations over the Boolean
+encodings of these enumeration, rather than over the data values themselves. A hash function is used to connect the data values to their enumerations to maintain consistency
+between these two representations.
+The relations are
+represented as BDDs~\cite{Bryant}. 
 For example, if the runtime verifier sees the input 
@@
 to all other values. 
-The algorithm maintains that these unseen values are correctly represented, as shown in~\cite{HPU}.
+%The algorithm maintains that these unseen values are correctly represented, as shown in~\cite{HPU}. 
 In fact, it is sufficient to have just one enumeration representing these values per each variable of the LTL formula. 
-However, it is important to guarantee that at least one such enumeration
-exists. We guarantee that by preserving the enumeration $11\ldots11$, i.e., the highest Boolean representation,
-to represent values not yet seen.
+%However, it is important to guarantee that at least one such enumeration exists. 
+We guarantee that at least one such enumeration exists by preserving for that purpose the enumeration $11\ldots11$.
+%, i.e., the highest Boolean representation,
+%to represent values not yet seen.
 We present here only the basic algorithm. For versions that
@@
 
-Our preferred representation of a set of values 
-(assignments) is as an Ordered Binary Decision
+\iffalse
+Our preferred representation of a set of values (assignments) is as an Ordered Binary Decision
 Diagram (OBDD, although we write simply BDD)~\cite{Bryant}.
-A BDD is essentially a compact representation 
+A BDD~\cite{Bryant} is essentially a compact representation 
 of a Boolean tree, where compaction glues together isomorphic subtrees. Each non-leaf node is labeled with one of the
@@
 Boolean value returned by the function for the Boolean values on the path.
+\fi
 
@@
 
-Given some ground predicate $p ( a )$ observed in the execution 
+Given some ground predicate $p ( a )$, observed in the monitored execution, 
 matching with $p ( x )$ in the monitored property,
-let $\lookup ( x, a )$ be the enumeration of $a$. If this
+let $\lookup ( x, a )$ be the enumeration of $a$ (a lookup in the hash table). If this
 is $a$'s first occurrence, then it will be assigned a new enumeration.
@@
 When a new value appears, this counter is incremented, and the value is converted to
-a Boolean representation. Enumerations that were not yet used represent
-the values not seen yet. 
+a Boolean representation. 
+%Enumerations that were not yet used represent
+%the values not seen yet. 
 %In particular, we always leave at least
@@
 
-The algorithm shown below works similarly to the algorithm for the 
-propositional case shown in
-Section~\ref{LTLruntime}. That is,
-it operates on two vectors (arrays) of values indexed by subformulas: 
+\iffalse
+The algorithm shown below uses two vectors (arrays) of values indexed by subformulas: 
 $\old$ for the state before that event, and
@@
 state (after the last seen event).
-However, while in the propositional case the vectors contain Boolean values, here
-they contain BDDs. The temporal relations are
-represented by
-sets of events, each being a tuple of one of the relations. Thus, $R_k [ i ]$ refers to the set
-of tuples of the relation $R_k$ at time instant $i$. Often (and in our implementation), the set of events consists of a singleton tuple of a single relation.
+While in the propositional case, in Section~\ref{LTLruntime}, the vectors contain Boolean values, here
+they contain BDDs. The temporal relations,
+restricted to each state, are
+represented also as BDDs.
 The algorithm follows.
+\fi
 
@@
   \item $\current ( \neg \varphi  ) := \bddneg ( \current ( \varphi ))$.
-  \item $\current ( ( \varphi\ \since\ \psi ) ) := \\ 
+  \item $\current ( ( \varphi\ \since\ \psi ) ) :=  
   \bddvee ( \current ( \psi  ) , \bddwedge ( \current ( \varphi ) ,
@@
 The auxiliary relations $r_j$ extend the model, and we need to keep 
-a BDDs representing $\current ( r_j )$ and
+BDDs representing $\current ( r_j )$ and
 $\old (r_j)$ for each relation $r_j$. We also need to calculate the subformulas $\varphi_i$ that appear in
 a specification, as part of the runtime verification,
-as per the above FPLTL algorithm. The auxiliary relations $r_j$ are calculated with respect to
-a variable $x_j$, which is the one appearing
-free in $\varphi_j$ (this can be generalized to any number of variables). However, $r_j$ can be used
-as a subformula with other parameters in other rules or in the statement. Thus,
-$r_j ( y )$ constructing a relation
-that is the same as $r_j$, but over the variable $y$. This can be achieved by the BDD renaming function $\rename ( r_j (x_j ) , y)$, which is easy to implement in BDDs.
+as per the above FPLTL algorithm. One subtle point is that the auxiliary relations $r_j$ may be defined in a rule with respect to
+a variable $x_j$ (as $r_j (x_j ):= \varphi_j (x_j)$ (this can be generalized to any number of variables) נוא $r_j$ can be used
+as a subformula with other parameters in other rules or in the statement e.g., as $r_j ( y )$. This can be resolved by a BDD renaming function $\rename ( r_j (x_j ) , y)$.
 
@@
 \item{\ \ } $\current ( r_j ( a ) )$ := if $\current ( r_j ) ( a )$
-then $\btrue$ else $\bfalse$; 
+then $\btrue$ else $\bfalse$
 \end{description}
 
-\klaus{Above: $\current ( r_j ) ( a )$ has not been defined I think.}
+%\klaus{Above: $\current ( r_j ) ( a )$ has not been defined I think.}
 

### 2019-02-07 implementation.tex
fixing bugs

@@
 
-\vspace{1ex}\noindent {\bf Example Properties.}
+\vspace{1ex}\noindent {\bf Example properties.}
 Figure \ref{fig:properties} shows three properties in the 

### 2019-02-07 main.tex
fixing bugs

@@
 \subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
-One can translate any QLTL property into a second order monadic logic in a standard
+One can translate any QLTL property into a second-order monadic logic in a standard
 way~\cite{Thomas}, and then from
-second order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
+second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
 each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set.

### 2019-02-07 rule-ltl.tex
fixing bugs

@@
 \vspace{1ex}
-\noindent {\bf Related Work.}
+\noindent {\bf Related work.}
 For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
@@
 data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
-\item{Q} adds (second order) Q{\em uantification} that is dynamic (state-dependent). This is referred to as {\em dynamic} quantification.
+\item{Q} adds (second-order) Q{\em uantification} that is dynamic (state-dependent). This is referred to as {\em dynamic} quantification.
 \item{E} allows E{\em xtending} the 
@@
 \, ( \varphi  \until  \varphi ) \,|  \, \ominus \varphi |
-   \, ( \varphi \since \psi ) \,  \]
+   \, ( \varphi \ \since \ \psi ) \,  \]
 where $p$ is a proposition from a finite set of propositions $P$, and $\bigcirc$, $\until$, $\ominus$, $\since$ stand for {\em next-time}, 
@@
 The expressive power of different versions of propositional LTL is often compared to regular expressions
-over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second order logics where the temporal propositions
+over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second-order logics where the temporal propositions
 correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
@@
 
-\subsubsection{Extending LTL with Dynamic Quantification.}
+\subsubsection{Extending LTL with dynamic quantification.}
 Adding quantification over propositions, as done in~\cite{Wolper}, allows writing a formula of the form
@@
 This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states.
-It is also called  {\em second order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
+It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
 Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
-automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
+automata, or monadic second-order logic with unary predicates over the naturals, see again~\cite{Thomas}.
 In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.
@@
 
-\subsubsection{Extending LTL with Rules.}
+\subsubsection{Extending LTL with rules.}
 
@@
 For the other direction, one can translate the QLTL property first into
-a second order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton.
+a second-order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton.
 The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
@@
 
-\subsubsection{Syntax}
+\subsubsection{Syntax.}
 
@@
 
-\subsubsection{Extending FLTL with Dynamic Quantification.}
+\subsubsection{Extending FLTL with dynamic quantification.}
 
@@
 
-\subsubsection{Extending FLTL  with Rules.}
+\subsubsection{Extending FLTL  with rules.}
 
@@
 
-\subsubsection{Set Semantics}
+\subsubsection{Set Semantics.}
 
@@
 
-\subsubsection{Runtime Verification Algorithm for PFLTL.}
+\subsubsection{Runtime verification algorithm for PFLTL.}
 
@@
 
-\subsubsection{Runtime Verification Algorithm for EPFLTL}
+\subsubsection{Runtime verification algorithm for EPFLTL}
 
@@
 BDDs representing $\current ( r_j )$ and
-$\old (r_j)$ for each relation $r_j$. We also need to calculate the subformulas $\varphi_i$ that appear in
+$\old (r_j)$ for each relation $r_j$. We also need to calculate the 
+subformulas $\varphi_i$ that appear in
 a specification, as part of the runtime verification,
-as per the above FPLTL algorithm. One subtle point is that the auxiliary relations $r_j$ may be defined in a rule with respect to
-a variable $x_j$ (as $r_j (x_j ):= \varphi_j (x_j)$ (this can be generalized to any number of variables) נוא $r_j$ can be used
-as a subformula with other parameters in other rules or in the statement e.g., as $r_j ( y )$. This can be resolved by a BDD renaming function $\rename ( r_j (x_j ) , y)$.
+as per the above FPLTL algorithm. One subtle point is that the auxiliary 
+relations $r_j$ may be defined in a rule with respect to
+a variable $x_j$ (as $r_j (x_j ):= \varphi_j (x_j)$ (this can be generalized 
+to any number of variables), but $r_j$ can be used
+as a subformula with other parameters in other rules or in the statement e.g., 
+as $r_j ( y )$. This can be resolved by a BDD renaming function 
+$\rename ( r_j (x_j ) , y)$.
 

### 2019-02-07 implementation.tex
Updates from Overleaf

@@
 Figure \ref{fig:properties} shows three properties in the 
-input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}.
-The properties are related to the examples in Section~\ref{sec:syntax-semantics}, which
+input format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
 are not expressible in (P)FLTL.
+These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The properties are 
 %

### 2019-02-07 rule-ltl.tex
Updates from Overleaf

@@
 \item If $\current (\eta) = \false$ then
-report a violation and terminate.
-\item Goto step~2.
+report a violation, otherwise goto step~2.
 \end{enumerate}
@@
 with the auxiliary ones
-inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL 
+inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL (EPFLTL)
 specification (\ref{EFLTL}) 
 by using the following equivalent
-QFLTL formula:
+QFLTL (QPFLTL, respectively) formula:
 
@@
 \noindent {\bf Sketch of Proof.}
-The proof of this theorem, which includes encoding of a nontrivial property, where the property
-observes the data elements (appearing as $v (a)$)
-in between states where $r$ appears (a relation with
-one parameter, the time or index parameter $i$).
-The property asserts that there are no two sets of events that are equivalent.
-This property can be expressed in QPFLTL, and hence
-also in QFLTL. The details appear in the appendix.
+The proof of this theorem includes encoding of a  property that
+observes sets of data elements $a$, appearing separately, one per state as $v (a)$,
+in between states where $r$ appears. The domain of data elements is unbounded.
+%(i.e., $r$ is a relation with
+%one parameter: the time/state-index parameter $i$).
+The set of $a$-values observed in between two consecutive $r$'s is called a {\em data set}.
+The property asserts that there are no two consecutive data sets that are equivalent.
+This property can be expressed in QPFLTL. The details appear in the appendix.
 
 We use a combinatorial argument to show by contradiction that one cannot express this property
-using EFLTL, and also using EPFLTL. The reason is that every prefix of an EFLTL model, extended uniquely with the
-auxiliary relations, according to Lemma~\ref{samesame}, can be summarized with a finite number of relations, the ones in
+in EPFLTL. The reason is that every prefix of an a model for an EPFLTL property is extended uniquely with
+auxiliary relations, according to Lemma~\ref{samesame}. Each prefix can be summarized by a finite number of relations: the ones in
 the model, the auxiliary relations and the
 assignments satisfying the subformulas. The size of
-each such relation is bounde by ${\cal O} (  m ^ N )$
+each such relation is bounded by ${\cal O} (  m ^ N )$
 where $m$ is the number of values appearing in the
-prefix, and $N$ is the sum of the number of parameters on all the above mentioned relations.
-However, the number of different subsets over $m$ values is
-$2^m$. This means that with large enough number of different values, EPFLTL can have two prefixes with the same summary. It would then not be able to distinguish between two executions that extend
-this prefix, where one of them is extended with a set of values that appeared before and the other with 
-a set that already appeared. \qed
+prefix, and $N$ is the number of parameters of the relations.
+However, the number of different data sets over $m$ values is
+$2^m$.
+This means that with large enough number of different values, each EPFLTL formula over the models of this property can have two prefixes with the same summary where at least one of them has a data set that the other one does not.
+The semantics of EPFLTL implies that extending
+two models with the same summary in the same way would have the same truth value. So, we can extend
+the two models with a data set that appears in one of the
+prefixes but not  in the other, and the EPFLTL property will not be able to distinguish between them. \qed
 
@@
 %rest of the formula.
-The following follows immediately from Theorem~\ref{theo2} the following:
-
+
+\vspace{0.7ex} From Theorem~\ref{theo2} and Equation~(\ref{equiv}) we immediately obtain:
 \begin{corollary}
-Constraining the dynamic quantification of QFLTL (QPFLTL) to existential quantification prefixing the
-formula strictly weakens its expressive power.
+Restricting the quantification of QPFLTL to existential quantification, strictly weakens its expressive power.
 \end{corollary}
@@
 \begin{enumerate}
-\item Initially, for each subformula $\varphi$,
+\item Initially, for each subformula $\varphi$ of $\eta$,
 %$B ( \varphi , \current ) = \bfalse$.
@@
 \end{itemize}
-\item Goto step~2.
+\item If $\current (\eta) = \false$ then
+report a violation, otherwise goto step~2.
 \end{enumerate}

### 2019-02-07 rule-ltl.tex
Updates from Overleaf

@@
 
-\section{RV for past-time
-first-order  LTL and its Extension}
+\section{RV for Past-Time
+First-Order  LTL and its Extension}
 \label{EPLTLRV}
@@
 We then have to add
-following updates between steps 3 and 4 of the above algorithm, when performing runtime verification of
+following updates to step~4 of the above algorithm, when performing runtime verification of
 $\psi$.
@@
 \noindent \begin{description}
-\item For each $r_j ( x_j ) = \varphi_j ( x_j )$:
+\item For each rule 
+$r_j ( x_j ) := \varphi_j ( x_j )$:
 \item{\ \ }  calculate $\current ( \varphi_j )$;
@@
 such as $r ( x ) := p ( x ) \vee \ominus r ( x )$. For that,
-we need to use the mixed evaluation order, described in Section~\ref{LTLruntime}.
+we need to use the {\em mixed} evaluation order, described in Section~\ref{LTLruntime}.

### 2019-03-24 main.tex
Updates from Overleaf

@@
 the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
-
-%\klaus{The following formulation above is not not clear: ``where an accepted input needs to pass infinitely often in exactly
-%one of the specified sets''.}
-
-\noindent
 Then we translate $M$ into an ELTL formula
@@
 the form defined in~(\ref{EFLTL}).
-
 %\klaus{Above: as the semantics is presented, it looks as if 
 %the `$\psi\ where ...$' is just another subformula.}
-
 %\klaus{Above: hide is not defined.}
-
 %\klaus{Above: the semantics of $p(a)$ and $p(x)$ do not refer to 
 %${\cal R} $.}
-
-In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign. It is well defined because of Lemma~\ref{samesame}.
+In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign.
+It is well defined because of Lemma~\ref{samesame} (equivalently, a more verbose definition can include a stepwise construction of ${\cal Q}$ from ${\cal R}$).
 \end{document}

### 2019-03-24 rule-ltl.tex

@@
 extending first-order LTL.
-Another example is based on the in-expressiveness of transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
+Another example is based on the unexpressiveness of the transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
 We present an extension where auxiliary relations, not appearing in the execution, are
@@
 properties, we show that for the first-order version, this has less 
-expressive power than adding alternating quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
-provides the full expressive power of, say B\"{u}chi automata.
+expressive power than adding full quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
+provides the full expressive power of, say, B\"{u}chi automata.
 
@@
 on calculating and updating a summary of the prefix of the execution.
-Runtime verification is often restricted to a {\em past}  version of temporal logic~\cite{HR}, mainly because for past properties we can decide when the monitored property is violated immediately when this happens, while for future properties we may never know for sure.
+Runtime verification is often restricted to a {\em past}  version of temporal logic~\cite{HR}, mainly because for past properties we can decide when the monitored property is violated after monitoring a finite prefix of it, while for future properties we may never know for sure.
 Consistent with that,
@@
 The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first-order temporal logic over infinite domains (e.g., the integers, strings).
-It achieves efficiency by BDD representation of enumerations of data (rather than over the data values themselves) and using
+It achieves efficiency by using BDD representation of enumerations of data, rather than over the data values themselves, and by using
 a garbage collection algorithm.
@@
 On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions.
-On the practical side, we also present RV algorithms for the extended logic, and 
+On the practical side, we present RV algorithms for the extended logic, and 
 show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.
@@
 \noindent {\bf Related work.}
-For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
+For propositional LTL, the logic ETL~\cite{Wolper,WVS} extends
 the set of temporal operators 
-by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
-where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
-where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
+by using automata (or, equivalently a right linear grammar).
+Each automaton defines a temporal operator, where the input letters of that automaton correspond to future temporal subformulas.
+Each accepted run of such an automaton defines then the places
+where the subformulas need to hold. Another expressive-equivalent extension, 
 in~\cite{Wolper}, adds dynamic, i.e.,
@@
 Numerous other systems have been produced for monitoring
-execution traces against formal specifications. 
-These include e.g. \cite{Meredith2011,larva,Reger2015,halle-beepbeep-ieee-12,LOLA}.
+execution traces eith data against formal specifications. 
+These include e.g. \cite{LOLA,larva,halle-beepbeep-ieee-12,Meredith2011,Reger2015}.
 
@@
 data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
-\item{Q} adds (second-order) Q{\em uantification} that is dynamic (state-dependent). This is referred to as {\em dynamic} quantification.
+\item{Q} adds (second-order) Q{\em uantification} that is state-dependent. This is referred to as {\em dynamic} quantification.
 \item{E} allows E{\em xtending} the 
@@
 Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
-$q$ is $\false$ in the initial state. Then $q$ alternates between even and odd states.
+$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.
 
@@
 $a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
-that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write conflicting rules, e.g., by using the rule $a_1 : = \false$.
+that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.
 
@@
 We define EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$) in ELTL. This results in
-a formalism that is equivalent to finite automata, where all the states except one are accepting and where the non-accepting state is a sink. 
+a formalism that is equivalent to finite (B\"{u}chi) automata, where all the states except one are accepting and where the non-accepting state is a sink. 
 We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:
@@
 with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
-of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part, may provide a more intuitive specification.
+of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a more intuitive specification.
 The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.
@@
 
-The algorithm for PLTL, first presented in \cite{HR}, 
+The algorithm for PLTL first presented in \cite{HR} 
 is based on the observation that the semantics of the 
@@
 
-Under {\em mixed evaluation order}, one calculates $\current$ in Step~4 in the following order. The first~4 steps are for the rules, and step~5 is for the statement $\eta$.
-\begin{enumerate}
-\item Calculate 
+Under {\em mixed evaluation order}, one calculates $\current$ as part of Step~4 of the above algorithm in the following order.
+The first~four steps, $a$-$d$, are for the rules, and the fifth step $e$ is for the statement $\eta$.
+\begin{description}
+\item{\em a.} Calculate 
 $\current ( \delta )$ for each subformula $\delta$
@@
 $\current ( \ominus \gamma )$ is set to $\old ( \gamma )$.)
-\item Set 
+\item{\em b.} Set 
 $\current ( a_j )$ to $\current ( \varphi_j )$ for each $j$.
-\item Calculate $\current ( \delta )$ for each subformula $\delta$
+\item{\em c.} Calculate $\current ( \delta )$ for each subformula $\delta$
 that appears in $\varphi_j$ of a rule
@@
 a $\ominus$ operator.
-\item Calculate $\current ( \delta )$ for each subformula $\delta$
+\item{\em d.} Calculate $\current ( \delta )$ for each subformula $\delta$
 that appears in the statement $\psi$, using the calculated $\current ( a_j )$.
-\item Calculate $\current ( \eta )$ for the statement
+\item{\em e.} Calculate $\current ( \eta )$ for the statement
 $\eta$.
-\end{enumerate}
+\end{description}
 
@@
 a relation $R$ can be $(``a", 5, ``cbb", 3)$, where
-the $3$ is the time parameter. The last parameter $i$
+the  value $3$ is the time parameter. The last parameter $i$
 does not correspond to modeling physical time. This
@@
 $p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.
-Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015} can express this property, 
+Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015}, can express this property, 
 where for each value $a$ there is a separate automaton that counts the number of times that $q(a)$ has occurred.
@@
 
-The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
-$\since$ and $\ominus$, and those derived from them.
-
-\begin{lemma} \label{samesame}
-The auxiliary temporal relations $r_1, \ldots r_n$ are
-uniquely defined by the rules for each model ${\cal R}$.
-\end{lemma}
-\noindent {\bf Proof.} By a simple induction, similar to Lemma~\ref{fourone}. \qed
-\vspace{1ex}
-The following formula expresses the property that was shown to be 
-not expressible using FLTL, at the beginning of this section.
-\begin{equation}
-\Box \forall x \, (r(x)\rightarrow p(x)) \mathrm{\ where\ }
-r(x) = ( q(x) \leftrightarrow \ominus \neg  r(x)) 
-\label{eq:wolper-first-order}
-\end{equation}
-
 \noindent
@@
 
-\noindent 
+The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
+$\since$ and $\ominus$, and those derived from them.
+
+\begin{lemma} \label{samesame}
+The auxiliary temporal relations $r_1, \ldots r_n$ are
+uniquely defined by the rules for each model ${\cal R}$.
+\end{lemma}
+\noindent {\bf Proof.} By a simple induction, similar to Lemma~\ref{fourone}. \qed
+\vspace{1ex}
+The following formula expresses the property described in Example~1, which was shown to be 
+not expressible using FLTL.
+\begin{equation}
+\Box \forall x \, (r(x)\rightarrow p(x)) \mathrm{\ where\ }
+r(x) = ( q(x) \leftrightarrow \ominus \neg  r(x)) 
+\label{eq:wolper-first-order}
+\end{equation}
+The property that corresponds to Example~2 appears as the third checked example in the implementation section~\ref{implement}.
+
 Analogously to the propositional case, it is easy to show that EPFLTL does not
@@
 %and QLTL (PQLTL, respectively) have the same
-expressive power. We study now the relationship between
+%expressive power. We study now the relationship between
 %EFLTL (EPFLTL) and QFLTL (QPFLTL, respectively).
@@
 %(i.e., $r$ is a relation with
-%one parameter: the time/state-index parameter $i$).
+%one parameter: the time parameter $i$).
 The set of $a$-values observed in between two consecutive $r$'s is called a {\em data set}.

### 2019-03-27 RESPONSE.txt
added response

General comment: Two of the reviews seem to ask (more or less
directly) what the contribution of the paper is.  We admit that the
introduction may not have provided a sufficiently clear overview over
this work, since many logics are introduced, but in the following we
try to briefly outline this (and we will of course later improve the
introduction).

In previous work [23,24] we presented a BDD-based monitoring algorithm
for:

  PFLTL : Past-time First-order LTL

as defined on page 9. This logic has limitations in expressiveness as
shown with two examples on page 10. As main contribution, we
consequently present an extension of this logic, called:

  EPFLTL: Extended Past-time First-ordrer LTL

See page 11 equation (5). The extension coniststs of adding rules to
temporal properties.  Example properties (not expresssible in PFLTL)
are shown in Figure 2 page 16.  This logic is strictly more expressive
than PFLTL by allowing e.g. Wolper's in-every-even-state property and
generally transitive closures over data in the trace to be
expressed. The algorithm is presented on page 14.  The corresponding
tool has been extended as well (and is availbale open source).

The paper is structured in such a way that it gradually builds up this
result in steps, with some intermediate results, first making the
argument for the propositional LTL case in Sections 2 and 3, and then
for the first-order LTL case in Sections 4 and 5, as follows.

For the propositional case:

- LTL is introduced and its limitation to past time operartors PLTL
  (known results)

- QLTL (and its past time version QPLTL) is LTL (PLTL) with (so-called
  dynamic) quantification over states. This is a known extension
  described by Wopler in [21]. This extends the expressiveness of LTL
  to allow formulation of the property: "every even state satisfies
  p", not expressible in LTL. However it is not obvious how to design
  an algorithm for efficiently monitoring QLTL properties. Therefore
  we alternatively propose (and this is the main theme in the paper):

- ELTL (and its past time version EPLTL) that instead extends LTL
  (PLTL) with rules, achieving the same expressivness as QLTL (QPLTL)
  but with a well defined monitoring algorithm. This is a new result.

For the first-order case we repeat the above story, ending with the final
contribution:

- FLTL and specifically PFLTL are introduced. The logic PFLTL and its
  monitoring algorithm was introduced in previous work by us [23,24].

- QFLTL (and its past time version QPFLTL) is introduced, showing how
  to extend FLTL with quantification over states as in the
  propositional case. This is a new theoretic result. However, as in
  the propositional case, it is not clear how to construct an
  efficient monitoring algorithm. This therefore leads us to the final
  result and main contribution:

- EFLTL and specifically the past time version EPFLTL are introduced,
  and a monitoring algorithm for EPFLTL is presented.

----------------------- REVIEW 1 ---------------------

> The authors claim that these extensions are useful for runtime
> verification and present some minimal experimental results from
> running a BDD-based implementation of the extensions.

As for experiments with BDDs, we have performed many experiments, they
are mentioned in our past papers, and are publically available with
our tool. But due to anonymous reviewing, we cannot reveal it. We only
here presented experiments to show the strength of the extension of
the logic.

> I am not quite on board with the motivation of the paper. In runtime
> verification, the truth values of formulas are determined on a single
> trace, hence, in particular, LTL and CTL are equivalent, and the
> verification is hence polynomial. It is also quite standard in
> model-checking to augment the intermediate states on the path with the
> truth values of subformulas (this is how CTL model checking is done,
> and hence also LTL for a single trace). Wouldn't it address at least
> some of the concerns of the authors (that is, which subformulas are
> satisfied in which states)?

*** TODO ***

We do not quite understand this comment.

> Also, I see no reference to PSL, which is the IEEE standard for
> property specification languages, or to ForSpec, which is Intel's
> specification language. I am wondering whether these would already
> provide the necessary functionality.

We should definitely cite the excellent well known work on
PSL/ForSpec. However, PSL and other standards are limited in the way
they handle data, and are basically strong propositional logics. They
are equivalent to logics that we cited, e.g. regular expressions and
2nd order monadic logic (this is not a logic with data!). We
introduced the propositional extension with rules (ELTL/EPLTL) to
demonstrate how simple (not more expressive than regular expressions)
our extension is, and as an easy to read motivation for a related
extension for first-order LTL (EFLTL/EPFLTL).  The subject of our
paper is not about expressiveness of propositional LTL, but a
practical extension to first-order LTL. As a side effect, the
propositional introduction also makes a new nice style of expressive
LTL properties.

> Adding quantification is solvable, in general, with a QBF solver, and
> in some cases even without it. In the Intel's paper about enhanced
> vacuity from the beginning of 2000's, they studied the problem of
> model-checking a quantified formula and demonstrated that for the
> types of formulas they constructed it is no harder than
> model-checking.

We assume you are referring to the paper:

  Enhanced Vacuity Detection in Linear Temporal Logic
  R. Armoni, L. Fix, A. Flaisher, O. Grumberg, N. Piterman, A. Tiemeyer, M. Vardi
  CAV 2003

The main difference is the first-order nature of our logic, in
contrast to the logic presented in that paper, which is propositional
(and by the way limited in where quantifiers can appear). However,
thanks for the reference, we will explore this work.

> Finally, I might be wrong, but it seems to me that BDDs are no longer
> widely used in model-checking, replaced by SAT and SMT solvers. I am
> wondering whether the implementation with BDDs is interesting enough.

The main problem studied in this paper is that of a linear temporal
logic with data, i.e., (past) first order linear temporal logic. For
this reason, a representation that can efficiently handle relations
over (basically unbounded) data is important. The BDD representation
is used to represent such big relations, where the BDDs hold
enumerations of the seen data values rather than that (hashed) data
values themselves. Consequently, we can use efficient standard Boolean
operators on BDDs to implement intersection, union and complementation
of sets (since the sets are only semi infinite, there is a special
encoding value for "all values not seen"). This is a surprisingly
efficient representation which works with first-order LTL.  BDDs are
nowdays less frequently used for model checking, but (a) this is not
model checking but runtime verification (b) they are used for the
specific and ingenious way of representing relations, and (c) SAT
solvers are not needed here. We are aware of (and will cite) work on
using SMT for runtime verification:

  Monitoring Modulo Theories
  Normann Decker, Martin Leucker, and Daniel Thoma

but it is not particularly efficient.

----------------------- REVIEW 2 ---------------------

> While the paper explores a potentially useful abstraction for runtime
> verification, it was hard to fully understand the motivation for this
> work. Some ways that the work could be strengthened are listed below.

Sure, see also leading comment above.

> Thm. 1 and Lemma 2 show that the introduced logics (ELTL and EPLTL)
> have the same expressive power as Q(P)LTL. The authors should make the
> benefit of introducing these new logics more clear.
> 
> Thm. 2 shows that
> QPFLTL is more expressive than EPFLTL, but the authors do not provide
> any complexity results on the decidability of these logics - for what
> benefit are we trading the expressivity?

As explained in the general introduction of this response, the main
reason for moving from 'Q' to 'E', e.g. from QPFLTL to EPFLTL, is that
we can show how to construct an efficient monitoring algorithm for 'E'
(e.g. EPFLTL), whereas we do not see how to do this for more theoretic 'Q'
(e.g. QPFLTL). That is the main message of the paper.

> Thm 2 also opens the question
> of whether EQPFLTL makes sense and if so how it relates to the other
permutations of (E/Q/F/PLTL).

There is no need to extend QPFLTL with rules since the rule extension 'E'
is an easier monitorable *alternative* to the dynamic quantification 'Q'.

> The limitations of existing runtime verification systems for data
> intensive systems could be made more clear through an evaluation that
> compares the use of EFPLTL for runtime verification with the
> alternative existing approaches. At the moment the evaluation only
> presents benchmarks for the introduced tool, but does not situate this
> evaluation in the context of existing runtime verification tools.

That is a fair comment.

> Furthermore, the evaluation includes a 'spawning' trace where analysis
> time appears to scale exponentially in the size of the trace. The
> authors mention this is possibly due to the increasingly large amount
> of data that must be stored as the trace grows. As the system is
> intended to be used for runtime verification, what practical
> consideration must be made given the exponentially scaling with
> respect to the lifecycle of the task being monitored?

It is certainly true that dealing with data introduces the problem
that monitoring becomes trace length dependent.  To tackle this
problem, the original version of the tool performs a clever form of
garbage collection, described in [24], where data no longer needed are
dropped from the BDDs.  In the here (artificially created) property
and trace, there is indeed a problem if too many threads are spawned.
We have not in this work studied how to limit monitoring to become
less trace length dependent. We consider this as an orthogonal,
although very important, line of work in the field of runtime
verification. Only little work has been done in this direction so far.

> In addition, a more thorough analysis of memory usage should be
> included in the evaluation. No information is provided on the system
> that ran these evaluation results. Do these benchmarks' memory usage
> exceed the available RAM and trigger disk operations? As the temporal
> relations of (P)FLTL require increasingly large hash tables to save
> their evaluations, what impact will this have on memory limited
> systems? This is especially relevant for the many use cases of runtime
> verification on embedded systems.

Indeed, memory analysis should be included. This information was not
directly visible when running our scripts (not a good excuse,
admitted). Memory is of course critical. As mentioned above, we do
perform garbage collection on unused values as described in [24]. For
example for the property

  forall x . p(x) -> !q(x) S r(x)

(meaning: forall x . if p(x) then in the past r(x) and since then no
q(x)), the monitor will drop any value x for which we in the past saw
an r(x) and since then saw! a q(x). But the general problem of
monitoring embedded systems is challenging. Not all problems can be
solved at once.

> Another issue is that it was hard to identify the paper's
> contribution. The section on 'Runtime verification algorithm for
> PFLTL' has already been described in [24]. To handle EPFLTL, Sec 5
> introduces a single 'rename' function which seems a relatively small
> delta with respect to [24]. The authors point us back to section 3
> (mixed evaluation order) for further contributions with respect to
> [24] to handle EPFLTL, but it is then not clear why Section 3 is
> separate from Section 5.

See general introduction to this response for a more general outline
of the work. The handling of EPFLTL, compared to PFLTL, is not trivial
and was quite difficult to implement. We do not consider it a
relatively small delta, although the theory perhaps is small (and
elegant). Concering the question why Section 3 is separate from
Section 5: as outlined in the introductory text above, Section 3 deals
with the propositional case whereas Section 5 deals with the
first-order case.

> The paper should better position its contributions wrt. related work.
> Throughout the paper, the authors should make it more obvious to the
> reader that only logics prefixed with an E are new logics introduced
> in this work (if I understood the paper correctly).

Well, QFLTL and QPFLTL are new (first-order LTL extended with dynamic
quantification).  But yes, the comment is largely correct. Our
introduction did not clarify the structure well enough.

> Additionally, the use of predicates to encode information about the
> prefix of the state (system inputs seen so far), and thereby allowing
> arbitrary data over infinite domains, was explored in 'Synthesizing
> Functional Reactive Programs' by Finkbeiner et. al
> https://arxiv.org/abs/1712.00246.

*** TODO ***

> Finally, the paper contains typos that create problematic ambiguities.
> As an example, the top of page 5 notes that the logic PQLTL is
> obtained by restricting QPLTL to past modalities. From context, it
> appears this should read as QLTL, not QPLTL.

Indeed.

> Furthermore, by page 6
> the authors are proving properties of QPLTL, but PQLTL is no longer
> referenced. Although from context the reader can infer that QPLTL and
> PQLTL these should be taken as the same logic (except on page 5 when
> QPLTL refers to QLTL), these inferences should not be left to the
> reader. I also believe that there is an issue with Theorem 2 - when
> comparing the expressive power of EPFLTL and QFPLTL, the statement
> instead reads as comparing EPFLTL and QFLTL.

Indeed to all observations here, we should have used/should use latex
macros for this!

----------------------- REVIEW 3 ---------------------

> The experimental results seem
> preliminary: apart from the fact that the examples are made up, it's
> unclear how the collection of traces was generated.

The artificial traces were generated by a trace generating program
parameterized with various constants, indicating how many events of
various kinds should be generated, and the patterns with which they
are generated. The objectieve was to generate as many data values in
the trace as possible, while at the same time also cause repetition of
patterns.

> It would be more
> convincing to see this applied to more realistic examples (esp the
> transitive closure property surely has many applications).  However,
> the tool is apparently available publicly (links hidden due to blind
> review).

Yes, it would be better with more realistic examples. The tool is
publicly available.

------------------------------------------------------

### 2019-03-28 RESPONSE-V2.txt
added response version 2

Contribution: in previous work [23,24] we presented a BDD-based
monitoring algorithm for PFLTL: Past-time First-order LTL. This logic
has limitations in expressiveness exemplified on page 10. As main
contribution, we consequently present an extension of this logic (page
11):

  EPFLTL: Extended Past-time First-ordrer LTL

The extension adds rules to temporal properties. Example properties
(not expressible in PFLTL) are shown (Figure 2).

We build this result in steps, making the argument initially for
propositional LTL (Sections 2,3) and then for first-order LTL
(Sections 4,5).

For each case (propositional and first-order) we present three
logic-classes.  Let X stand for LTL/PLTL (propositional case) or
FLTL/PFLTL (first-order case):

  core form       :  X
  quantified form : QX (extends expressiveness but no monitoring algorithm known)
  extended form   : EX (extends expressiveness with shown monitoring algorithm)

----------[REVIEW-1]----------

> and present some _minimal_ experimental results

We have presented many experiments in previous papers, which are
publicly available with our tool. We here added experiments to show
the strength of the extension.

> verification is hence polynomial ...

For unbounded data and unbounded sequences, the size of information
that needs to be stored can grow unboundedly. Our BDD representation,
however, shows impressive results, handling large traces.

> Also, I see no reference to PSL, ... ForSpec, ...

We will cite this work. However, PSL is limited in the way it handles
data, and is basically a strong propositional logic. It is equivalent
to some propositional logics we cite.  We introduced the propositional
extension with rules to demonstrate how simple (not more expressive!)
our extension is, that a simple extension to a classical RV algorithm
handles it, and to motivate the related extension for first-order LTL.

> In the Intel's paper about enhanced
> vacuity from the beginning of 2000's, they studied the problem of
> model-checking a quantified formula

There are two types of quantifiers: propositional quantifiers, as
QLTL, as dealt with in the paper you mention, and first-order
quantification (in our logics named with letter 'F'). Our contribution
is w.r.t. handling the first-order version, so the work mentioned is
not connected.

> ... BDDs are no longer widely used in model-checking, replaced by SAT and SMT ..

The main problem studied in handling RV for first-order LTL with data
is the representation of relations over data, which are accumulated
throughout the execution. In our BDD encoding of relations, we can use
efficient Boolean BDD operators to implement intersection, union and
complementation of relations. This is surprisingly efficient allowing
handling of very large traces with lots of data.

----------[REVIEW-2]----------

> have the same expressive power as Q(P)LTL. The authors should make the
> benefit of introducing these new logics more clear.
> Thm. 2 shows that
> QPFLTL is more expressive than EPFLTL...for what benefit are we trading the expressivity?

The main reason for moving from 'Q' to 'E', e.g. from QPFLTL to
EPFLTL, is that we can show how to construct an efficient monitoring
algorithm for 'E' (e.g. EPFLTL), whereas we do not see how to do this
for 'Q' (e.g. QPFLTL). We do not believe there is a reasonable
algorithm for the 'Q' version and the example in the proof gives some
insight of what is involved.

> the evaluation only presents benchmarks for the introduced tool

In [23] we compared with MonPoly, which is closest, since it includes
FPLTL.

> time appears to scale exponentially in the size of the trace.

The BDD representation of relations showed on many examples (open
source) to be quite compact. The original version of the tool also
performs a clever form of garbage collection, described in [24].
Indeed, BDDs are not a magic fix-all tool, but we obtain impressive
results.

> a more thorough analysis of memory usage should be included

Memory analysis is missing, agreed.

> The paper should better position its contributions wrt. related work.

Yes, although see also introduction to this response above.

> Additionally, the use of predicates ... was explored in 'Synthesizing Functional Reactive Programs'.

We were not aware of this paper (it appears as an archived paper).  It
seems that the TSL assignments only assign one value to a
variable. Our approach maintains sets of such values, needed for
monitoring. But we do not claim to have introduced predicates to
encode information about the prefix of the state, we refer to paper
[4].

----------------------- REVIEW 3 ---------------------

> unclear how the collection of traces was generated.

The artificial traces were generated by a trace generating program
parameterized with various constants, indicating how many events of
various kinds should be generated, and the patterns with which they
are generated. The objective was to generate as many data values in
the trace as possible, while at the same time also cause repetition of
patterns.

> more convincing to see this applied to more realistic examples

Agreed!

### 2019-04-18 Response/RESPONSE-V1.txt
uploaded reviews

--- 2019-03-27 RESPONSE.txt
@@
 
-*** TODO ***
+The paper referred to indeed introduces a concept of assignments in
+the temporal logic TSL.  An interesting read (nice paper)! we were not
+aware of this work. There are some essential differences though.
+
+First of all, assignments in TSL are to simple variables that along
+one trace only can hold one value at a time. This in contrast to our
+solution, where rules (and quantified variables in general) can keep
+track of multiple values (sets of values, they are relations in
+general), which is necessary for expressing the kind of properties we
+are interested in. E.g. it is as far as we understand (we may be
+wrong) not possible in TSL to express the following property, assuming
+that file identifiers range over the infinite domain of strings:
+
+  "whenever a file x is opened, file x should not be zip'ed
+   until file x has been closed"
+
+Assume that numerous files can be open at the same time, an unknown
+number up front.  If in TSL one uses a variable that stores a single
+file name this will not work.  To handle such a property, a variable
+storing file identifiers would need to be able to store all files
+opened so far and not yet closed. This of course can be programmed by
+letting a variable store a set of file names. However, this is like
+programming, and it becomes more complicated with more complicated
+properties. Another way of phrasing this is that TSL is closer to the
+concept of extended state machines, where there is one state machine
+with one state of data values that is updated in addition to the
+control states changing. In our framework we can say that we have a
+state machine for each value x.
+
+Second, TSL does not support the transitive closure of relations,
+whereas our logic does.  This is non-trivial as well.
+
+Finally, TSL is a future time logic, whereas our logic is a past time
+logic. There is a substantial difference in how algorithms in these
+two different spaces work.
 

### 2019-04-18 implementation.tex
started editing small things according to reviews

@@
 \section{Implementation}
+\label{implement}
 
@@
 Figure \ref{fig:properties} shows three properties in the 
-input format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
+input ASCII format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
 are not expressible in (P)FLTL.
-These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The properties are 
+These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 
 %
-The ASCII version of the logic here uses 
+The ASCII version of the logic uses 
 \idsl{@} for $\ominus$,
@@
 no \idsl{toggle(x)} event.
-The third alternative states that we in the previous state were
+The third alternative states that in the previous state we were
 in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} 

### 2019-04-18 rule-ltl.tex
started editing small things according to reviews

@@
 is usually performed by first translating the specification into a B\"{u}chi automaton.
-The automata formalisms are more expressive than LTL, with a classical example By Wolper~\cite{Wolper} showing that
+The automata formalisms are more expressive than LTL, with a classical example by Wolper~\cite{Wolper} showing that
 it is not possible to express in LTL that every even state in
@@
 extending first-order LTL.
-Another example is based on the unexpressiveness of the transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
+Another example is based on the inexpressiveness of the transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
 We present an extension where auxiliary relations, not appearing in the execution, are
@@
 
-Restricting QPLTL to the past modalities, one obtain the logic PQLTL. PQLTL has the same expressive
+Restricting QLTL to the past modalities, one obtains the logic QPLTL. QPLTL has the same expressive
 power as regular expressions and finite automata.
 Equation~(\ref{form0}) can
-be rewritten in PQLTL as:
+be rewritten in QPLTL as:
 \begin{equation} \label{form1} \Exists q \,
@@
 %Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
-%PQLTL property into a deterministic
+%QPLTL property into a deterministic
 %automaton.
@@
 $\psi$ is an LTL property and each $\varphi_i$ 
-is a PLTL property where propositions from $B$ can
+is a PLTL property where propositions from $A$ can
 only occur within the scope of a $\ominus$ operator.
@@
  {j \in \{1, \ldots , n\}}$. 
-Let $\sigma$ be a model with states over the $B$. 
+Let $\sigma$ be a model with states over $B$. 
 Then there is at most one
@@
 \subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
-formula into a deterministic finite automaton, where one distinguished states, $s$, in a sink state.
+formula into a deterministic finite automaton, where one distinguished state, $s$, is a sink state.
 Then the sequence of states to be checked forms the input for this automaton, and violation is
@@
 a natural number that is omitted in the temporal formulas. 
-An auxiliary relations $r_j$ can appear within $\psi$. They can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
+An auxiliary relation $r_j$ can appear within $\psi$. They can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
 scope
@@
 %We showed in Theorem~\ref{theo1} that ELTL (EPLTL)
-%and QLTL (PQLTL, respectively) have the same
+%and QLTL (QPLTL, respectively) have the same
 %expressive power. We study now the relationship between
@@
 \label{theo2}
-The expressive power of EPFLTL is strictly weaker than that of QFLTL.
+The expressive power of EPFLTL is strictly weaker than that of QPFLTL.
 \end{theorem}
@@
 We use a combinatorial argument to show by contradiction that one cannot express this property
-in EPFLTL. The reason is that every prefix of an a model for an EPFLTL property is extended uniquely with
+in EPFLTL. The reason is that every prefix of a model for an EPFLTL property is extended uniquely with
 auxiliary relations, according to Lemma~\ref{samesame}. Each prefix can be summarized by a finite number of relations: the ones in

### 2019-04-19 conclusion.tex
Updates from Overleaf
introduced logic macros

@@
 observed prefixes.
-
-We demonstrated that first-order linear temporal logic (FLTL), which
+%
+We demonstrated that first-order linear temporal logic (\FLTL{}), which
 can be used to assert properties about systems with data, also has expressiveness deficiencies, and extended it with relations that summarize prefixes.
@@
 this extension is not identical to the addition of dynamic (i.e., state dependent) quantification.
-
+%
 We presented a monitoring
@@
 Finally we described the implementation of this extension in the \dejavu{} tool and provided experimental results.
+%
+Future work includes performing additional experiments, and making further comparisons to other formalisms.
+Finally we would like to study further extensions, exploring the space between logic and programming.

### 2019-04-19 implementation.tex
introduced logic macros

@@
 input ASCII format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
-are not expressible in (P)FLTL.
+are not expressible in (P)\FLTL{}.
 These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 

### 2019-04-19 main.tex
introduced logic macros

@@
 is not.
-The QFLTL formula is a bit involved, hence given in several parts (macros), where the main formula appears last.
+The \QFLTL{} formula is a bit involved, hence given in several parts (macros), where the main formula appears last.
 
@@
 \iffalse
-We prove that the above property cannot be expressed in EPFLTL.
+We prove that the above property cannot be expressed in \EPFLTL{}.
 We use a combinatorial argument, showing that
-the information that is available in any EPFLTL
+the information that is available in any \EPFLTL{}
 property
@@
 
-In EPFLTL, after every prefix of the input, it is
+In \EPFLTL{}, after every prefix of the input, it is
 sufficient to keep a finite set of (non temporal) relations over the
@@
 that summarizes an observed prefix for
-EPFLTL is is polynomial
+\EPFLTL{} is is polynomial
 in $N$, consisting of a finite number
@@
 \iffalse
-{\bf Comments.} First, note that the reason that QFLTL {\em can} express the property in Theorem~\ref{theo2}, is that it allows to augment the input relations with auxiliary relations that are not fixed.
+{\bf Comments.} First, note that the reason that \QFLTL{} {\em can} express the property in Theorem~\ref{theo2}, is that it allows to augment the input relations with auxiliary relations that are not fixed.
 That is, for an auxiliary relation $R$,
@@
 prefixes (of length at least $i$), or even in different
-subformulas. Second, given a simple extension to EFLTL, where 
+subformulas. Second, given a simple extension to \EFLTL{}, where 
 one can use a variable that represents an updatable
@@
 \subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
-One can translate any QLTL property into a second-order monadic logic in a standard
+One can translate any \QLTL{} property into a second-order monadic logic in a standard
 way~\cite{Thomas}, and then from
@@
 passes infinitely often in exactly
-one of the specified accepting sets. To formulate an equivalent ELTL formula,
+one of the specified accepting sets. To formulate an equivalent \ELTL{} formula,
 the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
-Then we translate $M$ into an ELTL formula
+Then we translate $M$ into an \ELTL{} formula
 of the form:

### 2019-04-19 rule-ltl.tex
introduced logic macros

@@
 
-%\klaus{We use ELTL (etc) to denote Extended LTL. 
-%But QLTL is also an extension of LTL. These logics 
+%\klaus{We use \ELTL{} (etc) to denote Extended LTL. 
+%But \QLTL{} is also an extension of LTL. These logics 
 %are all extensions of LTL. Perhaps RLTL (for Rule-based 
@@
 
-An important subset of LTL, called here PLTL, allows only past temporal
+An important subset of LTL, called here \PLTL{}, allows only past temporal
 operators: $\since$, $\ominus$ and the operators derived
@@
 form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences. 
-(When PLTL is interpreted over finite sequences, its
+(When \PLTL{} is interpreted over finite sequences, its
 expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.)
@@
 different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states).
-Wolper's example property cannot be expressed in PLTL either.
+Wolper's example property cannot be expressed in \PLTL{} either.
 In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
@@
 It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
-Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
+Extending LTL with such quantification, the logic \QLTL{} has the same expressive power as regular expressions, full B\"{u}chi
 automata, or monadic second-order logic with unary predicates over the naturals, see again~\cite{Thomas}.
-In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.
-
-Wolper's property can be expressed in QLTL as follows:
+In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of \QLTL{}~\cite{Thomas}.
+
+Wolper's property can be expressed in \QLTL{} as follows:
 \begin{equation} \label{form0} \Exists q \,
@@
 
-Restricting QLTL to the past modalities, one obtains the logic QPLTL. QPLTL has the same expressive
+Restricting \QLTL{} to the past modalities, one obtains the logic \QPLTL{}. \QPLTL{} has the same expressive
 power as regular expressions and finite automata.
 Equation~(\ref{form0}) can
-be rewritten in QPLTL as:
+be rewritten in \QPLTL{} as:
 \begin{equation} \label{form1} \Exists q \,
@@
 %Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
-%QPLTL property into a deterministic
+%\QPLTL{} property into a deterministic
 %automaton.
@@
 
-We introduce now another extension of LTL, which we call ELTL.
+We introduce now another extension of LTL, which we call \ELTL{}.
 %It has a form that puts it, conceptually, in between temporal
@@
 and {\em basic propositions} $B$.
-%A model for an ELTL formula has truth assignments only over the propositions in $B$.
-An ELTL property $\eta$ has the following form: 
+%A model for an \ELTL{} formula has truth assignments only over the propositions in $B$.
+An \ELTL{} property $\eta$ has the following form: 
 \begin{equation} \label{ELTL}
@@
 $\psi$ is an LTL property and each $\varphi_i$ 
-is a PLTL property where propositions from $A$ can
+is a \PLTL{} property where propositions from $A$ can
 only occur within the scope of a $\ominus$ operator.
@@
 \noindent
-Wolper's example can be written in ELTL
+Wolper's example can be written in \ELTL{}
 as follows:
@@
 
-ELTL extends the set of propositional variables with new variables, whose values at a state are
+\ELTL{} extends the set of propositional variables with new variables, whose values at a state are
 functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
@@
 This differs from 
-the use of auxiliary variables in QLTL, where the values assigned to the
+the use of auxiliary variables in \QLTL{}, where the values assigned to the
 auxiliary variables do not have to extend the 
@@
 \begin{lemma} \label{fourone}
-Let $\varphi$ be an ELTL formula
+Let $\varphi$ be an \ELTL{} formula
 over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
@@
 \noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
-is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
-%A PLTL property depends in a state only on the prefix of the model that precedes and includes that state.
+is defined, via a \PLTL{} formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
+%A \PLTL{} property depends in a state only on the prefix of the model that precedes and includes that state.
 
@@
 The expressive power of 
-ELTL is the same as QLTL.
+\ELTL{} is the same as \QLTL{}.
 \end{theorem}
 \noindent {\bf Proof.}
-Each ELTL formula $\eta$, as defined above,
-is expressible using the following equivalent QLTL formula:
+Each \ELTL{} formula $\eta$, as defined above,
+is expressible using the following equivalent \QLTL{} formula:
 
@@
 \noindent
-For the other direction, one can translate the QLTL property first into
-a second-order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton.
+For the other direction, one can translate the \QLTL{} property first into
+a second-order monadic logic formula, then to a deterministic Muller automata and then construct an \ELTL{} formula that holds for the accepting executions of this automaton.
 The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
@@
 
-%EPLTL, where the Muller automaton can be
+%\EPLTL{}, where the Muller automaton can be
 %replaced with a B\"{u}chi automaton,
@@
 The use of the auxiliary variables $a_1 \ldots a_n$ and the
-rules of an ELTL specification 
+rules of an \ELTL{} specification 
 can be viewed as representing a
 state machine, in particular when the only modality that is used in the rules part is $\ominus$.
-In that case, we can consider an ELTL specification $\eta$ as consisting of two synchronous parts: an LTL property 
+In that case, we can consider an \ELTL{} specification $\eta$ as consisting of two synchronous parts: an LTL property 
 (the statement) 
@@
 within $\psi$. As a corollary of Theorem~\ref{theo1}, 
-this restricted representation is also equivalent to QLTL.
+this restricted representation is also equivalent to \QLTL{}.
 This representation can be useful, for RV, as explored further in this paper, because of the possibility to
@@
 
-%\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}
+%\klaus{First sentence below sounds wrong: (1) \PLTL{} already does not have future time operators, and (2) why mention infinite sequences?}
 \fi
 
-We define EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$) in ELTL. This results in
+We define \EPLTL{} by disallowing the future temporal operators (except an implied prefixing $\Box$) in \ELTL{}. This results in
 a formalism that is equivalent to finite (B\"{u}chi) automata, where all the states except one are accepting and where the non-accepting state is a sink. 
@@
 
-%For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.
+%For representing safety properties that can be expressed in \QPLTL{}, we can use a restricted version as shown in the following simple Lemma.
 
@@
 The expressive power of 
-EPLTL is the same as QPLTL.
-%Any safety QPLTL property can be expressed using the following restricted
-%ELTL form:
+\EPLTL{} is the same as \QPLTL{}.
+%Any safety \QPLTL{} property can be expressed using the following restricted
+%\ELTL{} form:
 %\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
@@
 \end{lemma}
-In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
+In fact, it is easy to show that \EPLTL{} can be restricted to the following form, while retaining its expressive power.
 \[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
@@
 with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
-of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a more intuitive specification.
+of an \EPLTL{} property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a more intuitive specification.
 The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.
@@
 \noindent {\bf Proof.} The proof of the Lemma follows from the
-axiomatization of QLTL in~\cite{MP}, e.g.,
+axiomatization of \QLTL{} in~\cite{MP}, e.g.,
 $\varphi\ \since\ \psi \leftrightarrow \psi \vee ( \varphi \wedge (\varphi\ \since\ \psi))$. \qed
@@
 While this Lemma makes the use of the temporal operators
-besides $\ominus$ in EPLTL redundant, it is still
+besides $\ominus$ in \EPLTL{} redundant, it is still
 convenient to use a combination of
@@
 
-The algorithm for PLTL first presented in \cite{HR} 
+The algorithm for \PLTL{} first presented in \cite{HR} 
 is based on the observation that the semantics of the 
@@
 \end{itemize}
-Runtime verification for a PLTL formula $\eta$ exploits the fact that
+Runtime verification for a \PLTL{} formula $\eta$ exploits the fact that
 the semantic definition is recursive in both the structure of the property. Thus, subformulas
@@
 
-\subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
+\subsubsection*{Runtime verification for \QPLTL{}.}  This can be performed by translating a
 formula into a deterministic finite automaton, where one distinguished state, $s$, is a sink state.
@@
 announced if $s$ is reached.
-QPLTL can be extremely compact, in particular when using alternating existential and universal 
+\QPLTL{} can be extremely compact, in particular when using alternating existential and universal 
 quantifiers, which can result in a huge (non-elementary) explosion of
@@
 
-\subsubsection*{Runtime verification for EPLTL.} For EPLTL,  we need to  add to the above
+\subsubsection*{Runtime verification for \EPLTL{}.} For \EPLTL{},  we need to  add to the above
 algorithm calculations 
@@
 
-%For PLTL, we can calculate  $\current$, for subformulas,
+%For \PLTL{}, we can calculate  $\current$, for subformulas,
 %bottom up. But
 Because the
-auxiliary variables can appear recursively in EPLTL rules, the order of calculation is
+auxiliary variables can appear recursively in \EPLTL{} rules, the order of calculation is
 subtle. To see this, consider, for example, the formula~(\ref{form3}). It contains
 the definition $q := \ominus \neg q$. Now, we cannot calculate this
-bottom up, as we did for PLTL, since $\current ( q )$ is not computed yet, and we need 
+bottom up, as we did for \PLTL{}, since $\current ( q )$ is not computed yet, and we need 
 to calculate $\current ( \ominus \neg q )$ in order 
@@
 
-We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is relations with
+We define models for \FLTL{} based on {\em temporal} relations~\cite{Chomicki}. That is relations with
 last parameter
@@
 
-The formulas of the core FLTL logic are 
+The formulas of the core \FLTL{} logic are 
 defined by the following grammar,where $p$ denotes a relation,
@@
 ($\since$, $\ominus$ and the ones derived from them) 
-forms the logic PFLTL.
+forms the logic \PFLTL{}.
 
@@
 
-For an FLTL (PFLTL) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
+For an \FLTL{} (\PFLTL{}) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
 We will henceforth sometimes abuse notation, and use the same
@@
 %Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
-%and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.
+%and will extend \FLTL{} below, is expressed with $\Exists$ and $\Forall$.
 
@@
 We demonstrate that the lack of expressiveness
-carries over from LTL (PLTL) to FLTL (PFLTL).
+carries over from LTL (\PLTL{}) to \FLTL{} (\PFLTL{}).
 
@@
 $a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to).
-To show that this is not expressible in FLTL (and PFLTL),
+To show that this is not expressible in \FLTL{} (and \PFLTL{}),
 consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
@@
 of the relation $\mathit{spawn}$. 
-FLTL can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}.
+\FLTL{} can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}.
 Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
@@
 
-\subsubsection{Extending FLTL with dynamic quantification.}
-
-Extending FLTL (PFLTL) with quantifiers over temporal relations, we obtain QFLTL
-(and the past-restricted version QPFLTL). 
+\subsubsection{Extending \FLTL{} with dynamic quantification.}
+
+Extending \FLTL{} (\PFLTL{}) with quantifiers over temporal relations, we obtain \QFLTL{}
+(and the past-restricted version \QPFLTL{}). 
 The syntax includes $\Exists p \, \varphi$, 
@@
 Consequently, quantification over relations effectively extends the model ${\cal R}$ into a model ${\cal R}'$ within the scope of the quantifier.
-Note that quantification here is dynamic (as in QLTL and QPLTL), since the relations are temporal and can have different sets of tuples in different states.
-
-\subsubsection{Extending FLTL  with rules.}
-
-We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We allow the following formula:
+Note that quantification here is dynamic (as in \QLTL{} and \QPLTL{}), since the relations are temporal and can have different sets of tuples in different states.
+
+\subsubsection{Extending \FLTL{}  with rules.}
+
+We now extend \FLTL{} into \EFLTL{} in a way that is motivated by the  propositional extension from LTL (\PLTL{}) to \ELTL{} (\EPLTL{}). We allow the following formula:
 \begin{equation}
@@
 \begin{enumerate}
-\item $\psi$, the {\em statement}, is an FLTL formula with
+\item $\psi$, the {\em statement}, is an \FLTL{} formula with
 no free variables, 
-\item $\varphi_j$ are PFLTL formulas with a single
+\item $\varphi_j$ are \PFLTL{} formulas with a single
 free variable $x_j$,
@@
 with the auxiliary ones
-inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL (EPFLTL)
+inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the \EFLTL{} (\EPFLTL{})
 specification (\ref{EFLTL}) 
 by using the following equivalent
-QFLTL (QPFLTL, respectively) formula:
+\QFLTL{} (\QPFLTL{}, respectively) formula:
 
@@
 
-The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
+The logic \EPFLTL{} is obtained by restricting the temporal modalities of \EFLTL{} to the past ones:
 $\since$ and $\ominus$, and those derived from them.
@@
 The following formula expresses the property described in Example~1, which was shown to be 
-not expressible using FLTL.
+not expressible using \FLTL{}.
 \begin{equation}
@@
 
-Analogously to the propositional case, it is easy to show that EPFLTL does not
+Analogously to the propositional case, it is easy to show that \EPFLTL{} does not
 need the past temporal operators besides $\ominus$.
@@
 
-%We showed in Theorem~\ref{theo1} that ELTL (EPLTL)
-%and QLTL (QPLTL, respectively) have the same
+%We showed in Theorem~\ref{theo1} that \ELTL{} (\EPLTL{})
+%and \QLTL{} (\QPLTL{}, respectively) have the same
 %expressive power. We study now the relationship between
-%EFLTL (EPFLTL) and QFLTL (QPFLTL, respectively).
+%\EFLTL{} (\EPFLTL{}) and \QFLTL{} (\QPFLTL{}, respectively).
 
@@
 \label{theo2}
-The expressive power of EPFLTL is strictly weaker than that of QPFLTL.
+The expressive power of \EPFLTL{} is strictly weaker than that of \QPFLTL{}.
 \end{theorem}
@@
 The property asserts that there are no two consecutive data sets that are equivalent.
-This property can be expressed in QPFLTL. The details appear in the appendix.
+This property can be expressed in \QPFLTL{}. The details appear in the appendix.
 
 We use a combinatorial argument to show by contradiction that one cannot express this property
-in EPFLTL. The reason is that every prefix of a model for an EPFLTL property is extended uniquely with
+in \EPFLTL{}. The reason is that every prefix of a model for an \EPFLTL{} property is extended uniquely with
 auxiliary relations, according to Lemma~\ref{samesame}. Each prefix can be summarized by a finite number of relations: the ones in
@@
 $2^m$.
-This means that with large enough number of different values, each EPFLTL formula over the models of this property can have two prefixes with the same summary where at least one of them has a data set that the other one does not.
-The semantics of EPFLTL implies that extending
+This means that with large enough number of different values, each \EPFLTL{} formula over the models of this property can have two prefixes with the same summary where at least one of them has a data set that the other one does not.
+The semantics of \EPFLTL{} implies that extending
 two models with the same summary in the same way would have the same truth value. So, we can extend
 the two models with a data set that appears in one of the
-prefixes but not  in the other, and the EPFLTL property will not be able to distinguish between them. \qed
+prefixes but not  in the other, and the \EPFLTL{} property will not be able to distinguish between them. \qed
 
@@
 
-%It follows from~(\ref{equiv}) that EFLTL (EPFLTL, %respectively) has the same
-%expressive power as QFLTL (EPFLTL, respectively) %restricted to having its
+%It follows from~(\ref{equiv}) that \EFLTL{} (\EPFLTL{}, %respectively) has the same
+%expressive power as \QFLTL{} (\EPFLTL{}, respectively) %restricted to having its
 %dynamic 
@@
 \begin{corollary}
-Restricting the quantification of QPFLTL to existential quantification, strictly weakens its expressive power.
+Restricting the quantification of \QPFLTL{} to existential quantification, strictly weakens its expressive power.
 \end{corollary}
@@
 
-Runtime verification of FLTL is performed on an
+Runtime verification of \FLTL{} is performed on an
 input that consists of {\em events} in the form of
@@
 \ignore{
-\klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}}
+\klaus{Below: (\ref{EFLTL}) was not a \EPFLTL{} form of specification.}}
 
 \iffalse
-, restricted to the past operators. We now present the definition of EPFLTL, the final main logic proposed in this paper. 
+, restricted to the past operators. We now present the definition of \EPFLTL{}, the final main logic proposed in this paper. 
 The syntax is as follows:
@@
 The RV algorithm for
-(E)PLTL, presented in~Section \ref{LTLruntime}
+(E)\PLTL{}, presented in~Section \ref{LTLruntime}
 calculates
@@
 truth value of $\varphi$ over the
-prefix inspected by the RV algorithm so far. For (E)PFLTL,
+prefix inspected by the RV algorithm so far. For (E)\PFLTL{},
 $\current (\eta )$ consists of the
@@
 
-We redefine the semantics for EPFLTL in
+We redefine the semantics for \EPFLTL{} in
 equivalent way, which will be more directly related to the calculation of values
@@
 Under the  {\em set semantics}
-(introduced in~\cite{HPU} for PFLTL, and extended here for EPFLTL),
+(introduced in~\cite{HPU} for \PFLTL{}, and extended here for \EPFLTL{}),
 $I [ \varphi , \sigma, i ]$ is a function that returns
@@
 
-\subsubsection{Runtime verification algorithm for PFLTL.}
+\subsubsection{Runtime verification algorithm for \PFLTL{}.}
 
 We start by describing an algorithm for monitoring 
-PFLTL properties, first presented in~\cite{HPU} and implemented
+\PFLTL{} properties, first presented in~\cite{HPU} and implemented
 in the tool~\dejavu. 
@@
 
-\subsubsection{Runtime verification algorithm for EPFLTL}
-
-We extend now the algorithm to capture EPFLTL.
+\subsubsection{Runtime verification algorithm for \EPFLTL{}}
+
+We extend now the algorithm to capture \EPFLTL{}.
 The auxiliary relations $r_j$ extend the model, and we need to keep 
@@
 a specification, as part of the runtime verification,
-as per the above FPLTL algorithm. One subtle point is that the auxiliary 
+as per the above \PFLTL{} algorithm. One subtle point is that the auxiliary 
 relations $r_j$ may be defined in a rule with respect to

### 2019-04-19 main.tex
Updates from Overleaf

@@
 
-\title{An Extension of LTL with Rules\\ 
-and its Application to Runtime Verification}
+\title{An Extension of LTL with Rules and its application to Runtime Verification}
 
-\ignore{\author{
+\author{
 \begin{tabular}{ccc}
@@
 \end{tabular}
-}
 }
@@
 
-\iffalse
 \author{ 
@@
 }
-
-\fi
 
@@
 properties. Finally, we expand the BDD-based runtime verification 
-tool \dejavu{}\footnote{The actual tool name is not given due to anonymous reviewing.}  to support our extension and perform some 
+tool \dejavu{} to support our extension and perform some 
 experiments.

### 2019-07-15 abstract-old.tex
Updates from Overleaf

--- 2019-04-19 main.tex
@@
-\documentclass{llncs}
-%\documentclass[converence,10pt]{IEEEtran}
-%\IEEEoverridecommandlockouts
-%\usepackage{cite}
-%\usepackage{amsmath,amssymb,amsfonts}
-%\usepackage{algorithmic}
-%\usepackage{graphicx}
-\usepackage{textcomp}
-\usepackage{bm}
-%\usepackage{xcolor}
-%\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\newcommand{\Exists}{\bm{\exists}\kern-0.6em\bm{\exists}}
-\newcommand{\Forall}{\bm{\forall}\kern-0.6em\bm{\forall}}
-
-\input{macros}
-
-\begin{document}
-
-\title{An Extension of LTL with Rules and its application to Runtime Verification}
-
-\author{
-\begin{tabular}{ccc}
-Klaus Havelund & \vspace{0.2cm} & 
-Doron Peled \\
-NASA Jet Propulsion Laboratory, Pasadena, USA & & 
-Bar Ilan University, Ramat Gan, Israel \\
-klaus.havelund@jpl.nasa.gov & & doron.peled@gmail.com
-\end{tabular}
-}
-
-\ignore{
-\author{\IEEEauthorblockN{Klaus Havelund} 
-\IEEEauthorblockA{\textit{Laboratory for Reliable 
-Software} \\
-\textit{NASA Jet Propulsion Laboratory}\\
-Pasadena, USA \\
-klaus.havelund@jpl.nasa.gov}\\\mbox{}\\
-\and
-\IEEEauthorblockN{Doron Peled} 
-\IEEEauthorblockA{\textit{Dep. of Computer Science} \\
-\textit{Bar Ilan University}\\
-Ramat Gan, Israel \\
-doron.peled@gmail.com}
-}
-}
-
-%\author{Submitted Anonymosly}
-
-%\thanks{Copyright 2018. All rights reserved.}
-
-\author{ 
-Klaus Havelund~\thanks{The research performed by this author was carried out at Jet Propulsion Laboratory, California Institute of Technology, under a contract with the National Aeronautics and Space Administration.}
-\and
-Doron Peled~\thanks{The research performed by this author was partially funded by Israeli Science Foundation grant 1464/18: 
-``Efficient Runtime Verification for Systems with Lots of Data and its Applications''.
-}
-}
-
-\institute{
-Jet Propulsion Laboratory, \\
-California Institute of Technology, USA 
-\and
-Department of Computer Science \\
-Bar Ilan University, Israel
-}
-
-%\hfill
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
-
-\input{rule-ltl}
-
-\input{implementation}
-
-\input{conclusion}
-
-\input{prop.bib}
-
-\newpage
-\appendix
-
-\section{Appendix}
-
-%\klaus{In the text below mention what $i$ is (time point, position, ...}
-
-\subsubsection{Construction of a property
-for Theorem~\ref{theo2}.}
-The model consists of
-a temporal relation.
-%\footnote{We abuse notation and
-%use the same name for the relation and for its
-%denotation in the logic. Also, $r [ i ]$ has zero
-%parameters; as noted before, this is a natural %extension to the syntax and semantics presented %above.} 
-$r$ with only the time parameter, so $r [ i ]$
-has a Boolean value at the $i$th state. We will henceforth say that $r$ occurs
-in state $i$. Another relation is $v$, with
-two parameters. For each $i$, if $r$ does not appear in the
-state, 
-then there is a value $a$ from some (infinite) domain,
-such that $v ( a, i )$, 
-i.e., $v(a)$ appears in $i$.
-The property requires that there are no two sets of values appearing within the relation $v$ (as $v(a)$) between successive occurrences of $r$ that are the same.
-
-%\klaus{Below: isn't it the other way around, the first is not a model and the second is a model?}
-
-Hence, the sequence of states 
-\[ \{ r \} \{ v(1) \}  \{ v(2) \}  \{ v(2) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(5) \}  \{  r \}\]
-is a model of this formula, while 
-\[ \{ r \} \{ v(1) \}  \{ v(2) \} \{ v(3) \}  \{ r\}  \{ v(2) \}  \{  v(1) \}  \{  v(3) \} \{  r \} \]
-is not.
-The \QFLTL{} formula is a bit involved, hence given in several parts (macros), where the main formula appears last.
-
-\begin{itemize}
-\item $add(P) ::= \forall x \, (P(x) \leftrightarrow (v(x) \vee \ominus P(x))$ [The set $P$ differs in the current state from the previous one only by values $a$ that appear in $v(a)$.]
-\item $same(P) ::= \forall x \, (P(x) \leftrightarrow \ominus P(x))$ 
-[The set of elements $x$ for which
-$P (x)$ holds
-contains the same elements as in the previous state.]
-\item $empty (P) ::= \forall x \neg P(x)$ [The set $P$ is empty.]
-\item $equal (P,Q) ::= \forall x (P(x) \leftrightarrow Q(x))$ 
-[The sets represented by the relations
-$P$ and $Q$ contain the same elements {\em in the current state}.]
-\item $collect(P, \varphi, \psi) ::=
-       r \wedge same(P) \wedge \varphi \wedge 
-       \ominus ((add(P)\wedge \neg r \wedge \varphi) \ \since \ (r \wedge empty(P) \wedge \psi))$  [Collect elements $a$ in $v(a)$, while $\varphi$ holds, 
-       since the previous state where $r$ 
-       and $\psi$ held.]
-\item $init ::= r \wedge \Box (r \leftrightarrow \neg \exists x \, v(x))$ [$r$ holds in the first state, and $v(a)$ with some value $a$ holds in states where $r$ does not hold.]
-\item $init \wedge \Box \neg \Exists A \Exists B (equal (A, B) \wedge \\
-collect(B, same(A),\, same(A)) \ \since \ collect(A, true, true))$
-[There are no two sequences of states where $v(a)$ holds
-with various values $a$, between a pair of states satisfying $r$ that contain exactly the same elements (regardless of repetition).]
-\end{itemize}
-
-%\klaus{Above: maybe need to point out what the top formula is.}
-
-%\klaus{Above: isn't there missing a $\Box$ in the main formula?}
-
-\iffalse
-We prove that the above property cannot be expressed in \EPFLTL{}.
-We use a combinatorial argument, showing that
-the information that is available in any \EPFLTL{}
-property
-after a prefix is not sufficient to distinguish between two models (executions), one satisfying and one not satisfying it.
-
-In \EPFLTL{}, after every prefix of the input, it is
-sufficient to keep a finite set of (non temporal) relations over the
-data elements seen in the input, which
-summarizes the prefix. This follows from the set
-semantics, which appears in Section~\ref{EPLTLRV}, 
-where the interpretation of
-each subformula, and the auxiliary relations,
-at a particular state is a relation, and the interpretation 
-of a property a prefix of length $i$ only depends on
-these relations. Now, the amount of information
-that summarizes an observed prefix for
-\EPFLTL{} is is polynomial
-in $N$, consisting of a finite number
-of relations over
-$N$ elements. The number of different tuples
-for such relations can be shown, inductively, to
-be polynomial in $N$ (beach parameter
-of each relation can be one out of the $N$ possibilities, or a representation the values not seen yet).
-However, the above property depends on whether
-a set of values that appear as parameters to
-the relation $v$, between two consecutive appearances
-of $r$'s, are the same. For that, one needs to keep some representation of all the
-subsets of the elements that appeared in the input so far; by contradiction, any summary representation of the observed prefix that
-would identify two prefixes with some different sets, would not be able to distinguish between
-sequences that extend this prefix with the set
-of elements that appears in one prefix but not the other.
-But the number of such subsets is exponential in $N$. \qed
-\fi
-
-%\klaus{Above: why set of subsets?}
-
-\iffalse
-{\bf Comments.} First, note that the reason that \QFLTL{} {\em can} express the property in Theorem~\ref{theo2}, is that it allows to augment the input relations with auxiliary relations that are not fixed.
-That is, for an auxiliary relation $R$,
-$R [ i ]$ is not fixed, and can change per different
-prefixes (of length at least $i$), or even in different
-subformulas. Second, given a simple extension to \EFLTL{}, where 
-one can use a variable that represents an updatable
-counter (or a generator of distinct values), one can express the property used in~\ref{theo2}, by
-keeping a relation where values between different occurrences of the event $r$ are paired with distinct values, hence, representing an unbounded set of sets.
-\fi
-
-%%%M
-\subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
-One can translate any \QLTL{} property into a second-order monadic logic in a standard
-way~\cite{Thomas}, and then from
-second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
-each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set.
-A muller automaton
-accepts a run if it 
-passes infinitely often in exactly
-one of the specified accepting sets. To formulate an equivalent \ELTL{} formula,
-the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.
-Then we translate $M$ into an \ELTL{} formula
-of the form:
-
-\begin{tabbing}
-\ \ \ \ \ $\bigvee_{F \in {\cal F}} \bigwedge_{s \in F} ( \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
-where \\
-\ \ \ \ \ \ \ \ \ \ $a_s :=   ( \alpha_{\iota} \wedge \ominus \false ) \vee \bigvee_{\Delta ( s'
-, \alpha ) = s } 
-(\alpha \wedge \ominus a_{s'}) 
-\, :\,  s \in S$
-\end{tabbing}
-
-%\klaus{Above: should $s=i \rightarrow \neg \ominus \true$ instead be (note: two changes):
-%$\neg \ominus \true \rightarrow a_s=i$
-%}
-%\klaus{Above: can't judge whether the $\bigwedge_{s \not\in F} \Diamond \Box %\neg a_s)$ is correct.}
-\noindent
-Basically, the assignments to the auxiliary variables are used to simulate the change of states\footnote{It is possible to use only $\lceil \log ( | S | ) \rceil$
-variables, at the expense of longer and more complicated formulas.}, and the main formula expresses the Muller acceptance condition. \qed
-
-\subsubsection*{Set Semantics.}
-
-Let $\Gamma$ be
-a set of assignments over the variables $W$, and
-$U \subseteq W$.
-Then $\proj ( \Gamma , U )$ (for ``projecting {\em out}'' or  ``hiding''
-the variables $U$)
-is the largest set of assignments over 
-$W \setminus U$, 
-each agreeing with some
-assignment of $\Gamma$ on all the variables in $W \setminus U$.
-Let $U \cap W = \emptyset$, then
-$\ext ( \Gamma , U )$ is the largest set of assignments
-over $W \cup U$, where each such assignment agrees 
-with some assignment in $\Gamma$ on the
-values assigned to the variables $W$. 
-\iffalse
-This means that we extend $\Gamma$ by
-adding arbitrary values to the variables in $U$ from their domains.
-Then $\proj ( \ext ( \Gamma , U ) , U ) = \Gamma$ holds.
-We define the union and intersection operators on sets of
-assignments, even if they are defined over non identical
-sets of variables. 
-In this case, the assignments are extended
-over the union of the variables. 
-\fi
-Thus, if $\Gamma$ is a
-set of assignments over $W$ and $\Gamma'$ is
-a set of assignments over $W'$, then
-$\Gamma \; \bigcup \;  \Gamma'$ is defined as $\ext (\Gamma , W' \setminus W ) \cup
-\ext ( \Gamma' , W \setminus W' )$ and
-$\Gamma \; \bigcap \; \Gamma'$ is $\ext (\Gamma , W' \setminus W ) \cap
-\ext ( \Gamma' , W \setminus W' )$.  Hence, both are defined
-over the set of variables $W \cup W'$.
-
-We denote by $A_{\vars ( \varphi )}$ the set of all possible assignments
-of values to the variables that appear free
-in $\varphi$. Thus,
-$I [ \varphi, {\cal R} , i] \subseteq A_{\vars ( \varphi )}$.
-To simplify definitions, we add
-a dummy position $0$ for sequence $\sigma$ (which starts with $s_1$), 
-where every formula is interpreted as an empty set.
-Observe that the value $\emptyset$ and $\{ \epsilon \}$, behave
-as the Boolean constants $0$ and $1$, respectively.
-The set semantics is defined as follows, where $i \ge 1$.
-
-%\klaus{Above: assignments are introduced without any gentle introduction.
-Remember: you are not communicating with Moshe Vardi.
-You are communicating with John Fingelstein who had three glasses of wine at dinner.}
-
-%\klaus{Above: $I [ \varphi, {\cal R} , i]$ is introduced out of the blue without any explanation.}
-
-\begin{itemize}
-\item $I [ \varphi , {\cal R} , 0 ] = \emptyset$.
-\item $I [ \true , {\cal R} , i ] = \{ \epsilon \}$.
-\item $I [ p ( a ) , {\cal R} , i ] =$ if $m (p )  ( a , i ) $ then
-$\{ \epsilon \}$ else $\emptyset$.
-\item $I [ p ( x ) , {\cal R} , i ] = \{ [ x \mapsto a ] | m (p) ( a , i)  \}$.
-%\item $I [ \true , \sigma , i ] = D^n$.
-%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
-%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
-\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
-I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
-\item $I [ \neg \varphi , {\cal R} , i ] = 
-A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , {\cal R} , i ]$.
-\item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
-I [ \psi , {\cal R}, i ] \; \bigcup \;
-( I [ \varphi , {\cal R} , i ] \bigcap 
-I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
-\item $I [ \ominus \varphi , {\cal R} , i ] = I [ \varphi , {\cal R} , i-1 ]$.
-\item $I [ \exists x \; \varphi , \sigma , i ] = 
-\proj ( I [ \varphi , {\cal R} , i ] , \{ x \} )$.
-\item $I [ \psi \mathrm{\ where\ } r_j ( x_j) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} , {\cal R} , i ] = 
- I [ \psi , {\cal Q} ] $ where ${\cal Q} = {\cal R} \cup
-\{ r_j | r_j [ i ] =  I [ 
-\varphi_j , {\cal Q} , i ] \}$
-\end{itemize}
-Note that the last time interprets a formula of
-the form defined in~(\ref{EFLTL}).
-%\klaus{Above: as the semantics is presented, it looks as if 
-%the `$\psi\ where ...$' is just another subformula.}
-%\klaus{Above: hide is not defined.}
-%\klaus{Above: the semantics of $p(a)$ and $p(x)$ do not refer to 
-%${\cal R} $.}
-In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign.
-It is well defined because of Lemma~\ref{samesame} (equivalently, a more verbose definition can include a stepwise construction of ${\cal Q}$ from ${\cal R}$).
-\end{document}

### 2019-07-15 abstract.tex
Updates from Overleaf

\begin{abstract}

Runtime Verification (RV) consists of analyzing execution 
traces using formal techniques, e.g., monitoring executions against 
Linear Temporal Logic (LTL) properties. Propositional LTL is, however, 
limited in expressiveness, as first shown by Wolper \cite{Wolper}.
%, e.g. not express that every even state in a sequence satisfies a given proposition. 
Several extensions to propositional LTL, which promote the
expressive power to that of regular expressions,
have therefore been proposed; however, none of which was, by and large, adopted for RV.
In addition, for many practical cases, there is a need in RV to monitor properties 
that carry data. This problem has been addressed by numerous authors, and 
in previous work we addressed this by providing an algorithm that uses BDDs to represent relations over data elements.
We show expressiveness deficiencies of first-order LTL and suggest an extension of (propositional as well as first-order) LTL with rules to address these 
limitations.  We describe how the \dejavu{} tool is correspondingly extended and provide some experimental results. 
%It has an incremental operational 
%semantics that suggests a simple extension for RV algorithms. 
%We demonstrate that by expanding the \dejavu{} tool to 
%support our extension, and performing some experiments.

\end{abstract}

### 2019-07-15 conclusion.tex
Updates from Overleaf

@@
+\vspace{-1.7cm} \quad
+
 \section{Conclusions}
@@
 progresses. There has been several proposed extensions to LTL that extend
-its expressive power to that of related automata formalisms. We proposed here
-a simple extension for propositional LTL that is based on adding propositions that summarize the prefix of an execution.
+its expressive power to that of related automata formalisms.
+
+We proposed here
+a simple extension for propositional LTL that adds auxiliary propositions that summarize the prefix of the execution, based on rules written as past formulas.
 Conceptually, this extension puts the specification in between propositional LTL and automata, as the additional
@@
 that is synchronized with the temporal property. It is shown to have
-the same expressive power as automata models, and is in particular
+the same expressive power as B\"{u}chi automata, and is in particular
 appealing for runtime verification of past (i.e., safety) temporal properties,
@@
 observed prefixes.
+
 %
 We demonstrated that first-order linear temporal logic (\FLTL{}), which
-can be used to assert properties about systems with data, also has expressiveness deficiencies, and extended it with relations that summarize prefixes.
-We showed that for the first-order case, unlike the propositional case,
+can be used to assert properties about systems with data, also has expressiveness deficiencies, and similarly
+extended it with rules
+that define {\em relations} that summarize prefixes of the execution.
+We proved that for the first-order case, unlike the propositional case,
 this extension is not identical to the addition of dynamic (i.e., state dependent) quantification.
@@
 We presented a monitoring
-algorithm for propositional past time temporal logic, extending a classical algorithm, and similarly presented an algorithm for 
-first-order past temporal logic.
-Finally we described the implementation of this extension in the \dejavu{} tool and provided experimental results.
+algorithm for propositional past-time temporal logic with rules, extending a classical algorithm, and similarly presented an algorithm for 
+first-order past temporal logic with rules.
+Finally we described the implementation of this extension in the \dejavu{} tool and provided experimental results. The
+code and many more examples appear on:
+\begin{center}
+\url{https://github.com/havelund/dejavu}
+\end{center}
+
 %
-Future work includes performing additional experiments, and making further comparisons to other formalisms.
-Finally we would like to study further extensions, exploring the space between logic and programming.
+Future work includes performing additional experiments, and making further comparisons to other formalisms. We intend to study further extensions, exploring the space between logic and programming.

### 2019-07-15 implementation.tex
Updates from Overleaf

--- 2019-04-18 implementation.tex
@@
 
+\begin{figure}
+\begin{center}
+\begin{lstlisting}[language=dsl,frame=single,linewidth=0.95\textwidth,backgroundcolor=\color{white},linewidth=\columnwidth,breaklines=true,basicstyle=\small]
+prop telemetry1: Forall x . closed(x) -> !telem(x) where 
+  closed(x) := toggle(x) <-> @!closed(x)
+
+prop telemetry2: Forall x . closed(x) -> !telem(x) where
+  closed(x) := (!@true & !toggle(x)) | (@closed(x) & !toggle(x)) | 
+               (@open(x) & toggle(x)),
+  open(x) := (@open(x) & !toggle(x)) | (@closed(x) & toggle(x))
+
+prop spawning : Forall x . Forall y . Forall d . report(y,x,d) -> spawned(x,y) where
+  spawned(x,y) := @spawned(x,y) | spawn(x,y) | 
+                  Exists z . (@spawned(x,z) & spawn(z,y))
+
+prop commands : Forall c . dispatch(c) -> ! already_dispatched(c) where
+  already_dispatched(c) := @ [dispatch(c) , complete(c)),
+  dispatch(c) := Exists t . CMD_DISPATCH(c,t),
+  complete(c) := Exists t . CMD_COMPLETE(c,t)
+\end{lstlisting}
+\caption{Properties stated in \dejavu's logic}
+\label{fig:properties}
+\end{center}
+\end{figure}
+
 \vspace{1ex}\noindent {\bf Example properties.}
-Figure \ref{fig:properties} shows three properties in the 
-input ASCII format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
-are not expressible in (P)FLTL.
-These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 
+Figure \ref{fig:properties} shows four properties in the 
+input ASCII format of the tool, the first three of which are related 
+to the examples in Section~\ref{sec:syntax-semantics}, which
+are not expressible in (P)\FLTL{}. That is,
+these properties are not expressible in the original first-order 
+logic of \dejavu{} presented in \cite{HPU}. The last property illustrates the use of rules to perform conceptual abstraction.
 %
@@
 rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
-In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
+In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula) that can be turned on and off with a \idsl{toggle(x)}; they are initially off.
 Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.
-
-%%%%
 
@@
 property as \idsl{telemetry1}, but in this case using two 
-rules, more closely reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
+rules, reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
 \idsl{closed(x)} and \idsl{open(x)}. The rule \idsl{closed(x)} is defined as a disjunction between three alternatives. The first states that this predicate is true if we are in the initial state 
@@
 
-\begin{figure}
-\begin{center}
-\begin{lstlisting}[language=dsl,frame=single,linewidth=0.95\textwidth,backgroundcolor=\color{white},linewidth=\columnwidth,breaklines=true,basicstyle=\small]
-prop telemetry1: 
-  Forall x . closed(x) -> !telem(x) where closed(x) := toggle(x) <-> @!closed(x)
+The fourth property, \idsl{commands}, 
+concerns a realistic log from from the Mars rover Curiosity 
+\cite{msl}. The log consists of events 
+(here renamed) \idsl{CMD_DISPATCH(c,t)} and \idsl{CMD_COMPLETE(c,t)}, 
+representing the dispatch and subsequent completion  of a command 
+\idsl{c} at time \idsl{t}. The property to be verified is that a command, once 
+dispatched, is not dispatched again before completed. Rules 
+are used to break down the formula to conceptually simpler pieces. 
+%The example
+%illustrates how rules elegantly replace the role of macros, a non-trivial extension 
+%introduced in an earlier version of \dejavu.
 
-prop telemetry2: 
-  Forall x . closed(x) -> !telem(x) where
-    closed(x) := (!@true & !toggle(x)) | (@closed(x) & !toggle(x)) | 
-                 (@open(x) & toggle(x)),
-    open(x) := (@open(x) & !toggle(x)) | (@closed(x) & toggle(x))
-
-prop spawning :
-  Forall x . Forall y . Forall d . report(y,x,d) -> spawned(x,y) where
-    spawned(x,y) := @ spawned(x,y) | spawn(x,y) | 
-                    Exists z . (@spawned(x,z) & spawn(z,y))
-\end{lstlisting}
-\caption{Three properties stated in \dejavu's logic}
-\label{fig:properties}
-\end{center}
-\end{figure}
+% Out of 25,463 dispatches and 23,740 completions, there were 2,466
+% violations of this property.
+%
+% A formulation of this property without using rules is:
+%
+% prop commands :
+%   Forall c .
+%     (Exists t1 . CMD_DISPATCH(c,t1)) -> 
+%       ! @ [
+%             Exists t2 . CMD_DISPATCH(c,t2), 
+%             Exists t3 . CMD_COMPLETE(c,t3)
+%           )
 
@@
 
-\vspace{1ex}\noindent {\bf Evaluation.}
-We evaluated the three properties in Figure 
-\ref{fig:properties} on a collection of traces. Table \ref{tab:evaluation} shows the  
-analysis time (excluding time to compile the generated 
-monitor) for different traces (format is `trace length : seconds'). The telemetry traces will repeatedly open hundreds of channels and send telemetry over them. 
+\vspace{1.5ex}\noindent {\bf Evaluation.}
+In \cite{HPU,HPU-FMSD} we evaluated \dejavu{}
+without the rule extension against the \monpoly{} tool \cite{agrebasin}, which supports a logic close to \dejavu's.
+In \cite{HP} we evaluated \dejavu's garbage collection capability.
+In this section we evaluate the rule extension for the properties in Figure
+\ref{fig:properties} on a collection of 
+traces. Table \ref{tab:evaluation} shows the  
+analysis time (excluding time to compile the 
+generated monitor) and maximal memory usage in 
+MB for different 
+% ---
+traces
+%\footnote{
+%For the first two properties the same traces were used, generated as follows for varying constants $P$, $Q$ and $R$.
+%Repeat $P$ times the following: toggle $Q$ 
+%channels; then for each channel send telemetry $R$ %times; then toggle the channels again.
+%For the third property, traces were 
+%generated as follows for varying constants $S$ 
+%and $T$: spawn $S$ threads from the main thread 0, and %let each report back to thread 0; then repeat $T$ times %the following: each newly spawned thread in previous %step
+%spawns a new thread, which reports back to the main %thread 0.
+%}
+% ---
+(format is `trace length : time / memory'). 
 %
-The spawning traces consist of 50\% spawning events and 50\% reporting events.
-The spawning property  requires larger processing time compared to the telemetry properties since more data (the transitive closure) has to be computed and stored.
+%The telemetry traces will repeatedly open 
+%hundreds of channels and send telemetry over 
+%them. 
+%
+%The spawning traces consist of 50\% spawning 
+%events and 50\% reporting events. 
+%
+The processing time is generally
+very reasonable for very large traces. However,
+the spawning property  requires considerably larger 
+processing time and memory compared to the other 
+properties since more data (the transitive 
+closure) has to be computed and stored.
+%
+The evaluation was performed on a Mac laptop, with the Mac OS X 10.10.5 operating 
+system, on a 2.8 GHz Intel Core i7 with 16 GB of memory.
 
@@
 }
-
 \begin{table}[h!]
@@
 \hline
-telemetry1 & 1,200,001\,:\,3.3s & 5,200,001\,:\,7.1s & 10,200,001\,:\,12.8s \\
+telemetry1 & 1,200,001\;:\;2.6s / 194 MB & 5,200,001\;:\;5.9s / 210 MB & 10,200,001\;:\;10.7s / 239 MB \\
 \hline
-telemetry2 & 1,200,001\,:\,4.4s & 5,200,001\,:\,9.9s & 10,200,001\,:\,17.9s \\
+telemetry2 & 1,200,001\;:\;3.8s / 225 MB & 5,200,001\;:\;8.7s / 218 MB & 10,200,001\;:\;16.6s / 214 MB \\
 \hline
-spawning   & 9,899\,:\,42.7s & 19,999\,:\,127.6s & 39,799\,:\,572.4s \\
+spawning   & 9,899\;:\;29.5s / 737 MB & 19,999\;:\;117.3s / 1,153 MB & 39,799\;:\;512.5s / 3,513 MB \\
+\hline
+commands   & 49,999\;:\;1.5s / 169 MB & N/A & N/A \\
 \hline
 \end{tabular}
-\caption{Evaluation - trace lengths and analysis time in seconds}
+\caption{  Evaluation - trace lengths, analysis time in seconds, and maximal memory use}
 \label{tab:evaluation}
 \end{table}
+
+%\begin{table}[h!]
+%\centering
+%\begin{tabular}{|l|c|c|c|}
+%\hline
+%Property   & Trace 1 & Trace 2 & Trace 3 \\
+%\hline
+%telemetry1 & 1,200,001\,:\,3.3s & 5,200,001\,:\,7.1s & 10,200,001\,:\,12.8s \\
+%\hline
+%telemetry2 & 1,200,001\,:\,4.4s & 5,200,001\,:\,9.9s & 10,200,001\,:\,17.9s \\
+%\hline
+%spawning   & 9,899\,:\,42.7s & 19,999\,:\,127.6s & 39,799\,:\,572.4s \\
+%\hline
+%\end{tabular}
+%\caption{Evaluation - trace lengths and analysis time in seconds}
+%\label{tab:evaluation}
+%\end{table}

### 2019-07-15 introduction-old.tex
Updates from Overleaf

--- 2019-04-18 rule-ltl.tex
@@
 But if you insist, one thing that I cannot perform from here is changing the two figures where these letters occur.}
-
-\newcommand\eventty{\mathbb{E}}
-\newcommand\setof[1]{\mathcal{P}(#1)}
-
-\section{Propositional LTL}
-\label{propLTL}
-
-The classical definition of linear temporal logic~\cite{MP} has the following syntax:
-\[ \varphi ::= \true \, | \,  p \, | \, ( \varphi \wedge \varphi ) \, |  \, \neg  \varphi \, |   \, \bigcirc \varphi \, |  
-\, ( \varphi  \until  \varphi ) \,|  \, \ominus \varphi |
-   \, ( \varphi \ \since \ \psi ) \,  \]
-where $p$ is a proposition from a finite set of propositions $P$, and $\bigcirc$, $\until$, $\ominus$, $\since$ stand for {\em next-time}, 
-{\em until}, {\em previous-time} and {\em since}, respectively.
-
-The models for LTL formulas 
-are infinite sequence of states, of the form
-$\sigma =  s_{1}\, s_{2}\, s_3 \ldots$,
-where $s_i \subseteq P$ for each $i\geq 1$. These are the propositions that {\em hold} in that state.
-%We denote by $\xi_{i}$ the suffix $s_{i} s_{i+1} s_{i+2}\ldots$ of $\sigma$. 
-LTL's semantics is defined as follows:
-\begin{itemize}
-\item $( \sigma, {i}) \models \mathit{true}$.
-\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$.
-\item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
-\item $(\sigma, {i}) \models ( \varphi \land \psi)$ iff $(\sigma, {i}) \models \varphi$ and $(\sigma, {i} ) \models \psi$.
-\item $(\sigma, {i}) \models \bigcirc \varphi$ iff $(\sigma, {i+1}) \models \varphi$.
-\item $(\sigma, {i}) \models ( \varphi  \cal{U}  \psi )$ iff for some $j$,
-$j\geq i$, $(\sigma, {j}) \models \psi$,
-and for each $k$, $i \leq k < j$, $(\sigma, {k}) \models \varphi$.
-\item $(\sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
-\item $(\sigma , i) \models (\varphi \since \psi)$ iff there exists $j$, $1 \leq j \leq i$, such that
-$(\sigma , j)  \models \psi$ and for each
-$k$, $j < k \leq i$, $(\sigma , k ) \models \varphi$.
-\end{itemize}
-Then $\sigma \models \varphi$ when $( \sigma ,1 ) \models \varphi$.
-We can use the following abbreviations:
-$\false = \neg \true$, 
-$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$, 
-$(\varphi \rightarrow \psi) = ( \neg \varphi \vee \psi)$,
-$(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
-$\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
-${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}), and 
-${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$ ({\bf H} stands for 
-H{\em istory}).
-
-%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
-The expressive power of different versions of propositional LTL is often compared to regular expressions
-over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second-order logics where the temporal propositions
-correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
-LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star operator (or $\omega$).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}.
-In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
-(and the ones derived from them $\Box$ and $\Diamond$)
-leaves the same expressive power and is commonly used.
-
-%\klaus{Is that also the true for the first-order case? Doron: I have no idea. But at this point, I would not ask to ask Gabbay.}
-
-An important subset of LTL, called here PLTL, allows only past temporal
-operators: $\since$, $\ominus$ and the operators derived
-from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is sometimes interpreted over finite sequences, 
-where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
-Alternatively,
-it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; in this case, each of the prefixes has to satisfy $\varphi$. This later
-form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences. 
-(When PLTL is interpreted over finite sequences, its
-expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.)
-
-%\section*{Extending Propositional LTL}
-\label{sec:extending-prop-ltl}
-
-Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
-to some important properties. As a canonical example, he demonstrated that it is impossible to
-express that all the states with even indexes in a sequence
-satisfy $p$ (this is
-different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states).
-Wolper's example property cannot be expressed in PLTL either.
-In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
-propositions, as shown below.
-
-\subsubsection{Extending LTL with dynamic quantification.}
-Adding quantification over propositions, as done in~\cite{Wolper}, allows writing a formula of the form
-$\Exists q \, \varphi$, where
-$\Exists q$ represents {\em existential} quantification over a proposition $q$ that can appear in $\varphi$.
-
-%A model for such a formula 
-%has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.
-To define the semantics, let $X \subseteq P$ 
-%be a subset of 
-%the propositional variables appearing in the %states of the sequence $\sigma = s_1 s_2 %\ldots$
-and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)
-
-\begin{itemize}
-\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
-$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
-$( \sigma', i) \models \varphi$.
-\end{itemize}
-%Since all the auxiliary variables appear in the scope of
-%quantifiers, it holds that $\sigma \models \varphi$
-%(in $\varphi$, auxiliary variables appear only within
-%the scope of quantifiers)
-%iff $\sigma |_A \models \varphi$.
-%
-%\klaus{Are the last two lines above needed?}
-%
-Universal quantification is also allowed, where $\Forall q\, \varphi =
-\neg \Exists q\, \neg \varphi$.
-This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states.
-It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
-Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
-automata, or monadic second-order logic with unary predicates over the naturals, see again~\cite{Thomas}.
-In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.
-
-Wolper's property can be expressed in QLTL as follows:
-\begin{equation} \label{form0} \Exists q \,
-( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}
-
-Restricting QLTL to the past modalities, one obtains the logic QPLTL. QPLTL has the same expressive
-power as regular expressions and finite automata.
-Equation~(\ref{form0}) can
-be rewritten in QPLTL as:
-\begin{equation} \label{form1} \Exists q \,
-{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
-%In fact, it can be shown that a single existential quantifier is sufficient. 
-Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
-$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.
-
-%Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
-%QPLTL property into a deterministic
-%automaton.
-
-\subsubsection{Extending LTL with rules.}
-
-We introduce now another extension of LTL, which we call ELTL.
-%It has a form that puts it, conceptually, in between temporal
-%logic and automata: on the one hand, it is possible to simply write
-%a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. 
-As will be showed later, this extension is
-is very natural for runtime verification.
-
-We partition the propositions $P$ into
-{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
-and {\em basic propositions} $B$.
-%A model for an ELTL formula has truth assignments only over the propositions in $B$.
-An ELTL property $\eta$ has the following form: 
-\begin{equation} \label{ELTL}
-\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
- {j \in \{1, \ldots , n\}} \end{equation}
-where each $a_j$ is a distinct auxiliary proposition from $A$,
-$\psi$ is an LTL property and each $\varphi_i$ 
-is a PLTL property where propositions from $A$ can
-only occur within the scope of a $\ominus$ operator.
-The semantics can be defined as follows. We refer to $\psi$
-as the {\em statement} of $\eta$ and to 
-$a_j  := \varphi_j$ as a {\em rule}. In text, rules
- will be separated from each other by commas. 
-\begin{tabbing}
-\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
-$\sigma' |_A = \sigma$ such that \\
- \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
-\end{tabbing}
-
-\noindent
-Wolper's example can be written in ELTL
-as follows:
-\begin{equation} 
-\label{form3}
-\Box ( q \rightarrow p ) \, \mathrm{where} \, 
- ( q := \ominus \neg q)
-\end{equation}
-where $A = \{ q\}$ and $B = \{ p \}$.
-The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
-and each {\em even} state will satisfy $p$.
-
-ELTL extends the set of propositional variables with new variables, whose values at a state are
-functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
-including the current state.
-This differs from 
-the use of auxiliary variables in QLTL, where the values assigned to the
-auxiliary variables do not have to extend the 
-states of the model in a unique way throughout the interpretation of 
-the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
-must occur within the scope of a $\ominus$ operator
-is required to help preventing that the values
-of the auxiliary variables are conflicting, as in
-$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
-that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.
-
-\begin{lemma} \label{fourone}
-Let $\varphi$ be an ELTL formula
-over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
-a set of rules $a_j  := \varphi_j \, : \, 
- {j \in \{1, \ldots , n\}}$. 
-Let $\sigma$ be a model with states over $B$. 
-Then there is at most one
-model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $
-
-\end{lemma}
-
-%\klaus{(1) is this lemma saying the right thing? (2) is the proof really a proof of that lemma, (3) 
-%$\varphi$s and the $\psi$s. Do we need a special greek symbol for it?}
-
-\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
-is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
-%A PLTL property depends in a state only on the prefix of the model that precedes and includes that state.
-
-\begin{theorem}
-\label{theo1}
-The expressive power of 
-ELTL is the same as QLTL.
-\end{theorem}
-\noindent {\bf Proof.}
-Each ELTL formula $\eta$, as defined above,
-is expressible using the following equivalent QLTL formula:
-
-\[   \Exists a_1 \ldots \Exists a_n  ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
-\]
-
-%\klaus{How do we know that the ``other direction'' below is possible?}
-
-\noindent
-For the other direction, one can translate the QLTL property first into
-a second-order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton.
-The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
-The details of the
-constructions are given in the appendix. \qed
-
-%EPLTL, where the Muller automaton can be
-%replaced with a B\"{u}chi automaton,
-%where all of its states, except one, sink
-%state, are accepting. \qed
-
-\iffalse
-The use of the auxiliary variables $a_1 \ldots a_n$ and the
-rules of an ELTL specification 
-can be viewed as representing a
-state machine, in particular when the only modality that is used in the rules part is $\ominus$.
-In that case, we can consider an ELTL specification $\eta$ as consisting of two synchronous parts: an LTL property 
-(the statement) 
-and a finite automaton, whose states correspond to combinations
-of truth values of the auxiliary variables, and can be referred to
-within $\psi$. As a corollary of Theorem~\ref{theo1}, 
-this restricted representation is also equivalent to QLTL.
-This representation can be useful, for RV, as explored further in this paper, because of the possibility to
-incrementally calculate the values of the auxiliary variables. 
-Similarly, it can be
-used for extending the expressiveness of LTL model checking.
-
-%\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}
-\fi
-
-We define EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$) in ELTL. This results in
-a formalism that is equivalent to finite (B\"{u}chi) automata, where all the states except one are accepting and where the non-accepting state is a sink. 
-We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:
-
-\iffalse
-In this case, 
-we replace the first part of the construction in the proof of Theorem~\ref{theo1}
-with the simpler formula $\vee_{s \in A} a_s$, where all the states but one are
-accepting.$A$ are the accepting states of the finite automaton.
-
-\klaus{Last sentence above: perhaps say with words that we are always in an acceptance state - if that's what the point is.}
-\fi
-
-%For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.
-
-\begin{lemma}
-\label{simplecase}
-The expressive power of 
-EPLTL is the same as QPLTL.
-%Any safety QPLTL property can be expressed using the following restricted
-%ELTL form:
-%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
-% {j \in \{1, \ldots , n\}}  \] 
-% with $p$ being one of the auxiliary variables %$a_j$.
-\end{lemma}
-In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
-\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
-{j \in \{1, \ldots , n\}}  \] 
-with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
-of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a more intuitive specification.
-The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.
-An arrow from A to B means that logic  B is more expressive than A.
-A double headed arrow shows same expressiveness
-
-\iffalse
-\noindent {\bf Proof.} The proof of the Lemma follows from the
-axiomatization of QLTL in~\cite{MP}, e.g.,
-$\varphi\ \since\ \psi \leftrightarrow \psi \vee ( \varphi \wedge (\varphi\ \since\ \psi))$. \qed
-
-\klaus{In the axiom above, isn't there missing a $\ominus$ before
-the second $(\varphi\ \since\ \psi)$?}
-
-\klaus{Above: is the following really correct?: 
-``with $p$ being one of the auxiliary variables $a_j$''.}
-
-While this Lemma makes the use of the temporal operators
-besides $\ominus$ in EPLTL redundant, it is still
-convenient to use a combination of
-temporal operators when possible, and
-auxiliary variables definitions.
-\fi
-
-\begin{figure}
-\begin{center}
-%\includegraphics[height=0.6in,width=2in]{PROP.jpg}
-\includegraphics[width=2.8in]{PROP.jpg}
-\caption{\label{prop}
-Comparing the expressive power of the temporal logics in Section~\ref{propLTL}
-}
-\end{center}
-\end{figure}
-
-\section{RV for Propositional Past-Time LTL and its Extension}
-\label{LTLruntime}
-
-Runtime verification of temporal specifications often
-concentrates on the past portion of the logic.  Past-time  specifications have the important property that one can distinguish when they are violated after observing a finite prefix of
-an execution. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.
-
-The algorithm for PLTL first presented in \cite{HR} 
-is based on the observation that the semantics of the 
-past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
-in the previous step $i - 1$ of a subformula.
-To demonstrate this, we repeat here the definition of the previous-time
-operator $\ominus$, and rewrite the definition
-of $\since$ in an equivalent form that is
-more directly applicable for runtime verification.
-\begin{itemize}
-\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
-\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
-$( \sigma , i)  \models \varphi$ and 
-$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
-\end{itemize}
-Runtime verification for a PLTL formula $\eta$ exploits the fact that
-the semantic definition is recursive in both the structure of the property. Thus, subformulas
-are evaluated based on smaller subformulas, and the
-evaluation of subformulas in the previous state.
-The algorithm, shown below, uses two vectors (arrays) of values indexed by subformulas:  $\old$, which summarizes the truth values of the
-subformulas for the execution prefix that
-ends just {\em before} the new state, and $\current$, for the execution prefix that
-ends with the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
-\begin{enumerate}
-\item Initially, for each subformula $\varphi$
-of $\eta$,
-%$B ( \varphi , \current ) = \bfalse $.
-$\current ( \varphi ) := \false$.
-
-\item Observe a new event (as a set propositions) $s$ as input. 
-\item Let $\old := \current$.
-\item Make the following updates for each subformula. If $\varphi$ is
-      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
-      $\current ( \psi )$.
-\begin{itemize}
-  \item $\current ( \true ) := \true$.
-  \item $\current (  \varphi \wedge \psi  ) := 
-  \current ( \varphi )\  and\ \current ( \psi )$.
-  \item $\current ( \neg \varphi  ) := not\ \current ( \varphi )$.
-  \item $\current (  \varphi \since \psi  ) :=  
-  \current ( \psi  )\ or\ ( \current ( \varphi ) \ and\ 
-      \old ( ( \varphi \since  \psi )))$.
-%DAP: I removed the \; around \since, since it penetrated the 2nd column!
-  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
-\end{itemize}
-\item If $\current (\eta) = \false$ then
-report a violation, otherwise goto step~2.
-\end{enumerate}
-
-\subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
-formula into a deterministic finite automaton, where one distinguished state, $s$, is a sink state.
-Then the sequence of states to be checked forms the input for this automaton, and violation is
-announced if $s$ is reached.
-QPLTL can be extremely compact, in particular when using alternating existential and universal 
-quantifiers, which can result in a huge (non-elementary) explosion of
-the automaton representing the same
-specification~\cite{Thomas}.
-
-\subsubsection*{Runtime verification for EPLTL.} For EPLTL,  we need to  add to the above
-algorithm calculations 
-of $\current ( a_j )$ and
-$\current ( \varphi_j )$, for each of the rules of
-the form $a_j := \varphi_j$ (the corresponding
-$\old$ entries will be updated as in line~3 in the above algorithm).
-
-%(calculations
-%for the statement part $\psi$ remain the same).
-%$\current ( a_j )$ and $\old ( a_j )$,
-%and perform the following between steps~3 and~4 %for each $a_j$: $\current ( a_j ) := \current ( %\varphi_j )$.
-
-%For PLTL, we can calculate  $\current$, for subformulas,
-%bottom up. But
-Because the
-auxiliary variables can appear recursively in EPLTL rules, the order of calculation is
-subtle. To see this, consider, for example, the formula~(\ref{form3}). It contains
-the definition $q := \ominus \neg q$. Now, we cannot calculate this
-bottom up, as we did for PLTL, since $\current ( q )$ is not computed yet, and we need 
-to calculate $\current ( \ominus \neg q )$ in order 
-to compute $\current ( q )$.
-However, notice that the calculation is not dependent on the
-value of $q$ to calculate $\ominus \neg q$; 
-in Step~4 above, we have that
-$\current ( \ominus \; \varphi ) := \old ( \varphi )$
-so  $\current ( \ominus \neg q ) := \old ( \neg q )$.
-
-%\klaus{I don't see that we need to talk about the top-down evaluation.}
-
-Under {\em mixed evaluation order}, one calculates $\current$ as part of Step~4 of the above algorithm in the following order.
-The first~four steps, $a$-$d$, are for the rules, and the fifth step $e$ is for the statement $\eta$.
-\begin{description}
-\item{\em a.} Calculate 
-$\current ( \delta )$ for each subformula $\delta$
-that appears in $\varphi_j$ of a rule
-$a_j := \varphi_j$, but {\em not} within the scope of
-a $\ominus$ operator. (This is not
-a problem, since
-$\current ( \ominus \gamma )$ is set to $\old ( \gamma )$.)
-\item{\em b.} Set 
-$\current ( a_j )$ to $\current ( \varphi_j )$ for each $j$.
-\item{\em c.} Calculate $\current ( \delta )$ for each subformula $\delta$
-that appears in $\varphi_j$ of a rule
-$a_j := \varphi_j$ {\em  within} the scope of
-a $\ominus$ operator.
-\item{\em d.} Calculate $\current ( \delta )$ for each subformula $\delta$
-that appears in the statement $\psi$, using the calculated $\current ( a_j )$.
-\item{\em e.} Calculate $\current ( \eta )$ for the statement
-$\eta$.
-\end{description}
-
-\section{First-Order LTL}
-
-%section*{Syntax and Semantics}
-\label{sec:syntax-semantics}
-
-Assume a finite set of infinite domains\footnote{Finite domains are handled with some minor changes, see~\cite{HPU}.}
-$D_1, D_2, \ldots$,
-e.g., integers or strings. 
-Let $V$ be a finite set of {\em variables}, 
-with typical instances $x$, $y$, $z$.
-%We denote by $x: D$ the fact that the variable $x$ has the domain $D$.
-An {\em assignment} over a set of variables $V$
-maps each variable $x \in V$ to a value from
-its associated domain $\domain ( x )$, where multiple variables (or all of them)
-can be related to the same domain. For example
-$[ x \rightarrow 5 , y \rightarrow \text{``abc''} ]$ assigns
-the values $5$ to $x$ and the value ``abc'' to $y$.
-%Let ${\cal R}$ be a set of {\em relations} 
-%with typical instances $R$.
-%Each relation is associated with some position-dependent domains.
-
-%\klaus{Above: let T be ... Is T used later? The relations are used in the next paragraph, why not introduce them there? Also, relations are not defined really.}
-%\klaus{I don't quite understand the motivation below for this way of modeling events: relations with a time stamp, and the whole story of the time stamp not being mentioned etc.
-Why not just the usual models: traces?
-Too late to change now for CAV though.}
-
-We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is relations with
-last parameter
-that is a natural number represented a time instance. So a tuple of
-a relation $R$ can be $(``a", 5, ``cbb", 3)$, where
-the  value $3$ is the time parameter. The last parameter $i$
-does not correspond to modeling physical time. This
-is rather used to allow the relations to have
-different tuples in different instances of $i$,
-corresponding to states, in the propositional temporal
-logics.
-
-For a relation $R$, $R [ i ]$ is the relation obtained from $R$ by
-restricting it to the value $i$ in the last parameter, and removing that last $i$ from the tuples. For simplicity, we will describe henceforth
-the logic with relations $R$ that have exactly two parameters, 
-hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$.
-The definition of the logic over other numbers of parameters is quite straightforward.
-e.g., when $R$ has one parameter, $i$,
-$R [ i ]$ is a Boolean value.
-Our implementation of runtime verification described later fully supports relations with zero or more parameter.
-This also applies to the extensions of the logic in the rest of this section, and to some of the examples used.
-
-%\klaus{Above, what about instead of writing: ``this domain is fixed over all the values of the second parameter $i$'' to write:
-%``this domain is independent of the values of the second %parameter $i$'' - if that is what you mean to say.}
-
-%\klaus{Above: we need to say explicitly what $i$ intuitively is meant to represent: positions, time, ....}
-
-\subsubsection{Syntax.}
-
-The formulas of the core FLTL logic are 
-defined by the following grammar,where $p$ denotes a relation,
-$a$ denotes a constant and $x$ denotes a variable. 
-\begin{center}
-$\varphi ::= \true  \; | \;
-    p ( a ) \; | \;
-    p ( x ) \; | \;
-    ( \varphi \wedge \varphi ) \;  |   \;
-   \neg \varphi \; | \;
-   \bigcirc \varphi \; | \; 
-   ( \varphi \; \until \; \varphi ) \; | $ %\\
-   $ \ominus \; \varphi \; | \;
-    ( \varphi  \; \since  \; \varphi ) \; | \;
-    \exists x \; \varphi$
-\end{center}
-
-%We interpret a predicate $p(a)$ as a subformula that
-%represents the occurrence of some
-%For example, if we open
-%a file named $\text{``xyz''}$, we may have $\mathit{open}
-%(\text{``xyz''})$, where the
-%domain of the predicate ${\mathit{open}}$ is over strings representing
-%file names.
-
-\noindent 
-Additional operators are defined as in the propositional logic. We define
-$\forall x \; \varphi = \neg \exists x \neg \varphi$.
-Restricting the modal operators to the past operators
-($\since$, $\ominus$ and the ones derived from them) 
-forms the logic PFLTL.
-
-\subsubsection*{Semantics.}
-A model is a set of temporal
- relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$.
- Since the standard definition of temporal logic is
- over a sequence (``the execution''), let
- ${\cal R} [ i ] = \{ R_1 [ i ] \, \ldots , R_m [ i ] \}$. ${\cal R} [i]$ represents a {\em state}.
- A model ${\cal R}$ can thus be seen as a sequence
- of states ${\cal R}[1] {\cal R}[2] \ldots$. Let 
- $m$ be a bijection
-from relation names (syntax)
-onto the relations ${\cal R}$ (semantics).
-%After the formal semantical definition, we will abuse notation
-%and use the same symbol for the representation of the relation in the logic and for the relation as mathematical object.
-
-%\klaus{Above, did you not already say the following before:
-%``we denote by $R_k [ i ]$ the restriction of
-% the relation $R_k$ to the value $i$ of the last parameter.''.}
-
-\iffalse
-At a given state the formula
-$p(\text{``a''})$ means that $p (\text{``a''} )$ holds
-in the current state,
-that is, $p (\text{``a''} )$ is among 
-the ground predicates of the state.
-Consider now the formula $p ( x )$, for a variable $x \in V$.
-We interpret it such that $x$ is assigned any value ``$a$'' where
-$p ( \text{``a''} )$ appears in the current state. 
-%DP: Added parentheses on the outside level.
-Thus, for interpreting $(p ( x ) \wedge q ( y ))$ in a state that
-has the predicates
-$p ( \text{``a''} )$ and $q ( 3 )$,
-we have the assignment $[ x \mapsto \text{``a''} , y \mapsto 3 ]$.
-%The past time temporal operators have the following intuitive meaning.
-The formula $(\varphi_1\; \since\; \varphi_2)$ 
-(reads $\varphi_1$ {\em since} $\varphi_2$)
-means that $\varphi_2$ occurred in the past (including now)
-and since then (beyond that state) $\varphi_1$ has been true. This is the 
-past dual of the commonly used %DP: changed "common" to "commonly used".
-future time  {\em until} modality~\cite{MP}. 
-The property $\ominus \; \varphi$ means that $\varphi$ is true 
-in the previous state.
-This is the past dual of the %DP: removed "common".
-future time {\em next} modality.
-We can also define the following additional temporal operators:
-$P \; \varphi = (\true \, \since \, \varphi)$ (``previously''),
-and $H \varphi = \neg P \neg \varphi$ (``always in the past'' or ``historically'').
-The operator $[\varphi_1,\varphi_2)$, borrowed from \cite{MaC}, 
-has the same meaning as $(\neg \varphi_2\; \since\; \varphi_1)$, but reads more naturally as
-a semi-open interval.
-
-%Consider now the formula $P \neg p ( v )$, meaning that
-%sometimes in the past $\neg p ( v )$. The semantics of $P$
-%will be the union on the sets of values in the previous sets,
-%thus, we need to take the union of $\neg p ( v )$ interpreted
-%over all past states (including the current one).
-%
-%Suppose that we have the predicate $p ( i )$ at state $s_i$.
-%There are several ways to interpret that.
-%\begin{enumerate}
-%\item We complement with respect to
-%all the values in $domain ( p )$.
-%Thus, the complementation will be $N \setminus \{ i \}$ at state $i$,
-%where $N$ is the domain natural numbers, and the 
-%$N \setminus \{ i \}$ from $1$ to $n$ is $N$.
-%
-%\item We complement with respect to
-%the values in $domain ( p )$ that appeared so far in the sequence.
-%Thus, at state $s_i$ we already saw $\{1 , \ldots , i \}$
-%and as we see $p ( i )$ in the current state, the complementation 
-%at state $s_i$ is $\{ 1, \ldots , i-1 \}$. The union 
-%up to the $n$th state will give $\{1, \ldots , n-1 \}$.
-%
-%\item When new values appear later in the sequence, we 
-%update the complementation related to previous states to include
-%newly appearing values.
-%Then we have to take the union of $n$ sets
-%that do not include one distinct element each. This
-%gives us at the $n$th state $\{ 1 , \ldots , n \}$.
-%\end{enumerate}
-
-%In our implementation we will take the first interpretation.
-%However, we will assume that we have a reasonable
-%limit on the number of different values that can appear in
-%each domain. As it will cost us very little to assume quite
-%a large size of values (the size of memory and amount of
-%time of the implementation will be propositional to the
-%{\em actual} number of values seen and not to this limit), we deem
-%this assumption quite reasonable. %We will explain how to 
-%remove this constraint (however, this requires quite a
-%complicated implementation). It is also quite easy to see
-%how to change the semantics to conform with case $1$ and
-%how to implement it.
-
-Let $\gamma$ be an assignment to the variables that appear
-free in a formula $\varphi$.
-Then $( \gamma ,  \sigma, i ) \models \varphi$
-if $\varphi$ holds for the prefix $s_1 s_2 \ldots s_i$ of 
-the trace $\sigma$
-with the assignment $\gamma$. This is a standard definition,
-agreeing, e.g., with~\cite{Basin}. Note that by using past %DP: appearing in, instead of appearing with.
-operators, the semantics is not affected by
-states $s_j$ for $j>i$.
-%
-\fi
-
-Let $\vars ( \varphi )$ be the set of free (i.e., unquantified) variables of a
-subformula $\varphi$. 
-%The interpretation of a subformula
-%$\varphi$ over some finite sequence is as a set of assignments
-%to the variables $\vars ( \varphi)$.
-We denote by $\gamma |_{\vars ( \varphi )}$ the restriction (projection) of
-an assignment $\gamma$ to the free variables appearing in $\varphi$.
-%We are careful in combining the sets of assignments
-%between different
-%subformulas, e.g., through union and intersection (corresponding
-%to disjunction and conjunction, respectively) and the quantification.
-%This is done using the projection and extension functions over
-%assignments defined above.
-%
-Let $\epsilon$ be the empty assignment (with no variables). 
-%
-In any of the following cases, $( \gamma,  {\cal R} , i ) \models \varphi$
-is defined when $\gamma$ is an
-assignment over $\vars ( \varphi )$, and $i\ge 1$.
-
-\begin{itemize}
-
-\item $( \epsilon , {\cal R} , i ) \models \true$.
-
-\item $( \epsilon ,  {\cal R} , i )
- \models p ( a ) $ if $m(p) ( a, i )$, where $a$ denotes a constant from  $d(m(p))$.
-
-\item $( [ x \mapsto a ] , {\cal R} , i ) \models p ( x )$ if $m(p)  ( a , i )$, where $\domain (x) = d ( m ( p ))$.
-
-\item $( \gamma,  {\cal R} , i ) \models ( \varphi \wedge \psi ) $~if~$(
-\gamma |_{\vars  ( \varphi )} , {\cal R} , i ) \models \varphi$~and  \\
-$( \gamma |_{\vars ( \psi ) } , {\cal R} , i ) \models \psi$.
-
-\item $( \gamma , {\cal R} , i ) \models \neg \varphi$ if not $( \gamma , {\cal R} , i ) 
-\models \varphi$.
-
-\item $( \gamma , {\cal R} , i) \models \bigcirc \varphi$ if 
-$(\gamma , {\cal R} , i + 1) \models
-\varphi$.
-
-\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \until \; \psi )$ if
-for some $j$, $j \geq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
-\models \psi $ and for each $k$,  $i \leq k < j $,
-$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.
-
-\item $( \gamma , {\cal R} , i) \models \ominus \varphi$ if $i>1$ and
-$(\gamma , {\cal R} , i-1) \models
-\varphi$.
-
-\item $( \gamma , {\cal R} , i ) \models ( \varphi \; \since \; \psi )$ if
-for some $j$, $1 \le j \leq i$, $( \gamma |_{\vars ( \psi )} , {\cal R} , j) 
-\models \psi $ and for each $k$, $j < k \leq i$,
-$( \gamma |_{\vars ( \varphi )} , {\cal R} , k) \models \varphi$.
-
-\item $( \gamma , {\cal R} , i) \models \exists x \; \varphi$ if
-there exists $a \in \domain ( x )$ such that\footnote{$\gamma \, [ x \mapsto a ]$
-is the overriding of $\gamma$ with the binding $[ x \mapsto a ]$.}
-$( \gamma \, [ x \mapsto a ], \sigma , i) \models \varphi$.
-
-\end{itemize}
-
-\iffalse
-The definition of the {\em since} operator $S$ can be simplified in a standard
-way such that it refers only to the positions $i$ and $i-1$ in
-the sequence $\sigma$. This is based on the
-fact that according to the semantics of {\em since},
-$(\varphi \, \since \, \psi) = ( \psi \vee (\varphi \wedge \ominus 
-(\varphi \, \since \, \psi )))$.
-This will serve in the
-implementation to work with only two versions of
-the sets of assignments, for the current and previous state:
-
-\begin{itemize}
-\item $( \gamma , {\cal R} , i ) \models ( \varphi \, \since \, \psi )$ if
-$( \gamma |_{\vars ( \psi )} , {\cal R} , i ) \models \psi$ or $i>1$,
-$( \gamma |_{\vars ( \varphi )} , {\cal R} , i ) \models \varphi$, and \\
-$( \gamma , {\cal R} , i-1 ) \models ( \varphi \, \since \, \psi )$.
-\end{itemize}
-\fi
-
-%We can also define the following useful operators:
-%$P \varphi = (\true \, \since \, \varphi)$ (for ``previously''),
-%$( \varphi \, R \, \psi ) = \neg ( \neg \varphi \, \since \, \neg \psi )$ 
-%(the dual of the Since operator), and
-%$H \varphi = ( \false \, R \, \varphi )$ (for ``always in the past'').
-
-%We define $\mathbf{F}$ and $\mathbf{T}$ as special constants
-%in order to interpret formulas without free variables.
-%They are the complement of each other.
-%When combined with set operators (complementation, union, intersection),
-%$\mathbf{F}$ behaves as the empty set, hence is idempotent to
-%union. Accordingly, $\mathbf{T}$ is idempotent to intersection.
-
-\iffalse
-We denote by $A_{\vars ( \varphi )}$ the set of all possible assignments
-of values to the variables that appear free
-in $\varphi$. Thus,
-$I [ \varphi, \sigma , i] \subseteq A_{\vars ( \varphi )}$.
-To simplify definitions, we add
-a dummy position $0$ for sequence $\sigma$ (which starts with $s_1$), 
-where every formula is interpreted as an empty set.
-Observe that the value $\emptyset$ and $\{ \epsilon \}$, behave
-as the Boolean constants $0$ and $1$, respectively.
-The set semantics is defined as follows, where $i \ge 1$.
-
-\begin{itemize}
-\item $I [ \varphi , \sigma , 0 ] = \emptyset$.
-\item $I [ \true , \sigma , i ] = \{ \epsilon \}$.
-\item $I [ p ( a ) , \sigma , i ] =$ if $p ( a ) \in \sigma [ i ]$ then
-$\{ \epsilon \}$ else $\emptyset$.
-\item $I [ p ( v ) , \sigma , i ] = \{ [ v \mapsto a ] \; | \; p ( a ) \in
-\sigma [ i ] \}$.
-%\item $I [ \true , \sigma , i ] = D^n$.
-%\item $I [ ( \varphi \vee \psi ) , \sigma , i ] = 
-%I [ \varphi , \sigma , i ] \cup I [ \psi , \sigma , i ]$.
-\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
-I [ \varphi , \sigma , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
-\item $I [ \neg \varphi , \sigma , i ] = 
-A_{\vars ( \varphi )} \; \setminus \; I [ \varphi , \sigma , i ]$.
-\item $I [ ( \varphi \; \since \; \psi ) , \sigma , i ] = 
-I [ \psi , \sigma , i ] \; \bigcup \;
-( I [ \varphi , \sigma , i ] \; \bigcap \; 
-I [ (\varphi \since \psi ) , \sigma , i - 1 ] )$.
-\item $I [ \ominus \varphi , \sigma , i ] = I [ \varphi , \sigma , i-1 ]$.
-\item $I [ \exists x \; \varphi , \sigma , i ] = 
-\proj ( I [ \varphi , \sigma , i ] , \{ x \} )$.
-\end{itemize}
-
-\noindent
-As before, the interpretation for the rest of the operators can
-be obtained from the above using the connections between the operators,
-e.g., $I [ P \, \varphi  , \sigma , i ] = 
-I [ ( \true \, \since \, \varphi ) , \sigma , i ]$.
-The correspondence between this set based semantics 
-and the previous semantics, namely that
-$\gamma \in I [ \varphi , \sigma, i ]$ iff
-$( \gamma , \sigma , i ) \models \varphi$
-can be proved by a simple structural induction on
-the size of the formulas.
- \fi
-
-For an FLTL (PFLTL) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
-We will henceforth sometimes abuse notation, and use the same
-symbols both for the relations (semantics) and their representation in
-the logic (syntax).
-The quantification over values of variables here is {\em static} in the sense that they are independent of the state in the execution. We
-denote static quantification with $\exists$ and
-$\forall$.
-%Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
-%and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.
-
-%\section*{Extending First-Order LTL}
-
-%%this is specification formalism that is very expressive and yet monitorable under run time verification~\cite{HPU}.
-
-We demonstrate that the lack of expressiveness
-carries over from LTL (PLTL) to FLTL (PFLTL).
-
-\vspace{1ex}
-\noindent {\bf Example 1.}
-Let $p$ and $q$ be temporal relations, where their
-restrictions to the state $s_i$, $p[i]$ and $q[i]$, are unary relations $p$ and $q$. The
-specification that we want to monitor is that for each value
-$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to).
-To show that this is not expressible in FLTL (and PFLTL),
-consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
-that in this case, for each property $\varphi$
-we can replace occurrences of variables within
-predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
-and $q  (a )$, respectively, and that we can throw away
-the quantification ($\forall x$, $\exists x$), obtaining
-an equivalent formula. This is then equivalent to
-an LTL formula that is obtained by replacing
-the occurrences of $p(a)$ and $q(a)$ by propositions
-$p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.
-Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015}, can express this property, 
-where for each value $a$ there is a separate automaton that counts the number of times that $q(a)$ has occurred.
-
-\vspace{1ex}
-\noindent {\bf Example 2.}
-Consider the property that asserts that when
-$\mathit{report}(y , x, d)$ appears
-in a state, denoting that  process 
-$y$ sends some data $d$ to a process $x$,
-there was a chain of states with process spawns
-$\mathit{spawn}(x, x_1)$, $\mathit{spawn} (x_1, x_2)$ $\ldots$
-$\mathit{spawn}(x_l , y)$. i.,e., $y$ is a descendent process of $x$. The reason
-is that the required property needs to
-assert about the transitive closure
-of the relation $\mathit{spawn}$. 
-FLTL can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}.
-Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
-E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, ((p(x, t) \rightarrow \exists t' p(x, t') )$\footnote{The obtained model will have, in addition
-to the aforementioned relations, also
-the linear order $<$ over the naturals.}.
-It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in first order setting. The proof is based on the
-compactness theory of first order logic~\cite{Flum}.
-
-\subsubsection{Extending FLTL with dynamic quantification.}
-
-Extending FLTL (PFLTL) with quantifiers over temporal relations, we obtain QFLTL
-(and the past-restricted version QPFLTL). 
-The syntax includes $\Exists p \, \varphi$, 
-where $p$ denotes an auxiliary relation.
-We also allow $\Forall p\ \varphi = \neg \Exists \neg p \varphi$.
-The semantics is as follows.
-%We denote by ${\cal R} |_Q$ the set of temporal relations obtained from ${\cal R}$ by removing the set of relations $Q$.
-
-%\klaus{Below, has the notation ${\cal R}' |_{\{ q \} }$ been defined?}
-
-\begin{itemize}
-\item $(\gamma ,  {\cal R}, i) \models \Exists q \, \varphi$ iff there exists
-${\cal R}'$ such that ${\cal R}' \setminus {\{ q \} } = {\cal R}$ and 
-$( \gamma, {\cal R}' , i) \models \varphi$.
-\end{itemize}
-Consequently, quantification over relations effectively extends the model ${\cal R}$ into a model ${\cal R}'$ within the scope of the quantifier.
-Note that quantification here is dynamic (as in QLTL and QPLTL), since the relations are temporal and can have different sets of tuples in different states.
-
-\subsubsection{Extending FLTL  with rules.}
-
-We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We allow the following formula:
-\begin{equation}
-\label{EFLTL}
-\ \ \ \ \ \psi \mathit{\ where\ } r_j  ( x_j ) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \mathrm{\ \ \ \     such\ that,}
-\end{equation}
-\begin{enumerate}
-\item $\psi$, the {\em statement}, is an FLTL formula with
-no free variables, 
-\item $\varphi_j$ are PFLTL formulas with a single
-free variable $x_j$,
-\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
-type as $x_j$ and the second one is, as usual,
-a natural number that is omitted in the temporal formulas. 
-An auxiliary relation $r_j$ can appear within $\psi$. They can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
-scope
-of a previous-time operator $\ominus$. 
-\end{enumerate}
-
-\noindent
-We define\footnote{ Formal semantics 
-can also be given by constructing a set of temporal relations extended
-with the auxiliary ones
-inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL (EPFLTL)
-specification (\ref{EFLTL}) 
-by using the following equivalent
-QFLTL (QPFLTL, respectively) formula:
-
-\begin{equation} \label{equiv}
-\Exists r_1  \, \ldots \, \Exists r_n \,  ( \psi \wedge \Box \bigwedge_{j \in \{ 1, \ldots , n\}} (r_j ( x_j )  \leftrightarrow 
-\varphi_i ( x_j) )
-\end{equation}
-
-The logic EPFLTL is obtained by restricting the temporal modalities of EFLTL to the past ones:
-$\since$ and $\ominus$, and those derived from them.
-
-\begin{lemma} \label{samesame}
-The auxiliary temporal relations $r_1, \ldots r_n$ are
-uniquely defined by the rules for each model ${\cal R}$.
-\end{lemma}
-\noindent {\bf Proof.} By a simple induction, similar to Lemma~\ref{fourone}. \qed
-\vspace{1ex}
-The following formula expresses the property described in Example~1, which was shown to be 
-not expressible using FLTL.
-\begin{equation}
-\Box \forall x \, (r(x)\rightarrow p(x)) \mathrm{\ where\ }
-r(x) = ( q(x) \leftrightarrow \ominus \neg  r(x)) 
-\label{eq:wolper-first-order}
-\end{equation}
-The property that corresponds to Example~2 appears as the third checked example in the implementation section~\ref{implement}.
-
-Analogously to the propositional case, it is easy to show that EPFLTL does not
-need the past temporal operators besides $\ominus$.
-
-%\klaus{Something seems wrong in the above formulation.
-%E.g.: Is the $\Box$ a restriction?, what is ``fully quantified single occurrence of some auxiliary relation $r_j$''?}
-
-\ignore{
-\noindent {\bf Proof.} Basically, the proof is the same as Lemma~\ref{simplecase}, based on the axiomatization of the
-past temporal operators. \qed}
-
-%\klaus{Is the formula above really correct?}
-
-%We showed in Theorem~\ref{theo1} that ELTL (EPLTL)
-%and QLTL (QPLTL, respectively) have the same
-%expressive power. We study now the relationship between
-%EFLTL (EPFLTL) and QFLTL (QPFLTL, respectively).
-
-\begin{theorem}
-\label{theo2}
-The expressive power of EPFLTL is strictly weaker than that of QPFLTL.
-\end{theorem}
-
-\noindent {\bf Sketch of Proof.}
-The proof of this theorem includes encoding of a  property that
-observes sets of data elements $a$, appearing separately, one per state as $v (a)$,
-in between states where $r$ appears. The domain of data elements is unbounded.
-%(i.e., $r$ is a relation with
-%one parameter: the time parameter $i$).
-The set of $a$-values observed in between two consecutive $r$'s is called a {\em data set}.
-The property asserts that there are no two consecutive data sets that are equivalent.
-This property can be expressed in QPFLTL. The details appear in the appendix.
-
-We use a combinatorial argument to show by contradiction that one cannot express this property
-in EPFLTL. The reason is that every prefix of a model for an EPFLTL property is extended uniquely with
-auxiliary relations, according to Lemma~\ref{samesame}. Each prefix can be summarized by a finite number of relations: the ones in
-the model, the auxiliary relations and the
-assignments satisfying the subformulas. The size of
-each such relation is bounded by ${\cal O} (  m ^ N )$
-where $m$ is the number of values appearing in the
-prefix, and $N$ is the number of parameters of the relations.
-However, the number of different data sets over $m$ values is
-$2^m$.
-This means that with large enough number of different values, each EPFLTL formula over the models of this property can have two prefixes with the same summary where at least one of them has a data set that the other one does not.
-The semantics of EPFLTL implies that extending
-two models with the same summary in the same way would have the same truth value. So, we can extend
-the two models with a data set that appears in one of the
-prefixes but not  in the other, and the EPFLTL property will not be able to distinguish between them. \qed
-
-%The relative expressive power between the first-order %temporal logics presented in this Section
-%appear in Figure~\ref{fsfr}.
-
-%It follows from~(\ref{equiv}) that EFLTL (EPFLTL, %respectively) has the same
-%expressive power as QFLTL (EPFLTL, respectively) %restricted to having its
-%dynamic 
-%quantification to being existential, and prefixing %the
-%rest of the formula.
-
-\vspace{0.7ex} From Theorem~\ref{theo2} and Equation~(\ref{equiv}) we immediately obtain:
-\begin{corollary}
-Restricting the quantification of QPFLTL to existential quantification, strictly weakens its expressive power.
-\end{corollary}
-
-\ignore{
-
-\begin{figure}
-\begin{center}
-%\includegraphics[height=1.5in,width=3in]{FIRST.jpg}
-\includegraphics[width=3in]{FIRST.jpg}
-\caption{\label{fsfr} An arrow from A to B means that logic  B is more expressive than A}
-\end{center}
-\end{figure}
-}
-
-\section{RV for Past-Time
-First-Order  LTL and its Extension}
-\label{EPLTLRV}
-
-Runtime verification of FLTL is performed on an
-input that consists of {\em events} in the form of
-tuples. In our notation, the input
-consists of a sequence ${\cal R}[1] \, {\cal R}[2]  \ldots$,
-which we earlier identified with states, where each
-${\cal R}[i]$ consists of the relations in
-${\cal R}$ with
-the last parameter is restricted to  $i$. A typical use
-of runtime verification restricts the
-events (tuples from the relations in ${\cal R}$) for
-each state to a finite number, and often even to a single event.
-The RV algorithm will also make
-use of sets of assignments over a set of variables for satisfying a subformula at
-some state, corresponding to $\old$ and $\current$, which had Boolean values in the 
-algorithm for propositional LTL in Section~\ref{LTLruntime}.
-
-\ignore{
-\klaus{In the text above there are some imprecisions, such as
-(1) ``tuples of relations'' - relations are sets of tuples, we don't have tuples of relations. 
-(2) ``${\cal R}[i]$ consists of some relations'': it is a set of
-relations (I think).
-(3) ``to have, per each state just some finite tuples'': tuples are finite. You mean: to contain a finite set of tuples?
-(4) what is the story about a set of assignments also being a relation? Needed?
-}}
-
-%\subsubsection{Syntax}
-
-\ignore{
-\klaus{Below: (\ref{EFLTL}) was not a EPFLTL form of specification.}}
-
-\iffalse
-, restricted to the past operators. We now present the definition of EPFLTL, the final main logic proposed in this paper. 
-The syntax is as follows:
-
-\[ 
-\psi \mathrm{\ where\ } r_j ( x_j) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
-\]
-\begin{center}
-$\varphi ::= \true  \; | \;
-    p ( a ) \; | \;
-    p ( x ) \; | \;
-    ( \varphi \wedge \varphi ) \;  |   \;
-   \neg \varphi \; | \;
-    \ominus \varphi \; | \;
-    ( \varphi  \; \since  \; \varphi ) \; | \;
-    \exists x \; \varphi$
-\end{center}
-
-\klaus{Note that $\psi$ above is not defined as a syntactic category. It should really be $\varphi$. Likewise in other parts of the paper. Perhaps too big a change now? Also, there should be
-a greek letter for a specification. E.g.:
-
-\[ 
-\pi ::= \varphi \mathrm{\ where\ } r_j ( x_j) := 
-\varphi_j (x_j) : j \in \{ 1 , \ldots , n \}
-\]
-\[
-\varphi ::= ...
-\]}
-\fi
-
-\iffalse
-\noindent {\bf Set Semantics.}
-We now refine the semantics of the logic. Under the new definition, 
-$I [ \varphi , \sigma, i ]$ is a function that returns
-a set of assignments such that $\gamma \in I [ \varphi , \sigma, i ]$ 
-iff $( \gamma , \sigma , i ) \models \varphi$.
-This redefinition will later lead to a simple implementation 
-using BDDs, where each set of assignments will be represented
-as a BDD, and the Boolean operators will correspond directly 
-to Boolean operators on BDDs.
-
-In order to deal with subformulas with different sets of free
-variables (hence, different domains for assignments), 
-we apply a projection and an extension operator to assignments
-over a subset of the variables. Let $\Gamma$ be
-a set of assignments over the variables $W$, and
-$U \subseteq W$.
-Then $\proj ( \Gamma , U )$ (for ``projecting {\em out}'' or  ``hiding''
-the variables $U$)
-is the largest set of assignments over 
-$W \setminus U$, 
-each agreeing with some
-assignment of $\Gamma$ on all the variables in $W \setminus U$.
-Let $U \cap W = \emptyset$, then
-$\ext ( \Gamma , U )$ is the largest set of assignments
-over $W \cup U$, where each such assignment agrees 
-with some assignment in $\Gamma$ on the
-values assigned to the variables $W$. 
-This means that we extend $\Gamma$ by
-adding arbitrary values to the variables in $U$ from their domains.
-Then $\proj ( \ext ( \Gamma , U ) , U ) = \Gamma$ holds.
-We define the union and intersection operators on sets of
-assignments, even if they are defined over non identical
-sets of variables. 
-In this case, the assignments are extended
-over the union of the variables. Thus, if $\Gamma$ is a
-set of assignments over $W$ and $\Gamma'$ is
-a set of assignments over $W'$, then
-$\Gamma \; \bigcup \;  \Gamma'$ is defined as $\ext (\Gamma , W' \setminus W ) \cup
-\ext ( \Gamma' , W \setminus W' )$ and
-$\Gamma \; \bigcap \; \Gamma'$ is $\ext (\Gamma , W' \setminus W ) \cap
-\ext ( \Gamma' , W \setminus W' )$.  Hence, both are defined
-over the set of variables $W \cup W'$.
-
-%We define $\mathbf{F}$ and $\mathbf{T}$ as special constants
-%in order to interpret formulas without free variables.
-%They are the complement of each other.
-%When combined with set operators (complementation, union, intersection),
-%$\mathbf{F}$ behaves as the empty set, hence is idempotent to
-%union. Accordingly, $\mathbf{T}$ is idempotent to intersection.
-
-%\klaus{One may of course ask whether the following section on the 
-%set semantics is needed. Maybe it is needed, since we probably need a semantics. Something to consider.}
-\fi
-
-\subsubsection{Set Semantics.}
-
-The RV algorithm for
-(E)PLTL, presented in~Section \ref{LTLruntime}
-calculates
-$\current ( \eta )$, for $\eta$ a
-subformula of the monitored property, to be the
-truth value of $\varphi$ over the
-prefix inspected by the RV algorithm so far. For (E)PFLTL,
-$\current (\eta )$ consists of the
-sets of assignments (in the form of relations
-over the free variables in the subformula),
-rather than Boolean variables.
-
-We redefine the semantics for EPFLTL in
-equivalent way, which will be more directly related to the calculation of values
-in $\current$ by the
-RV algorithm that will be presented below. 
-Under the  {\em set semantics}
-(introduced in~\cite{HPU} for PFLTL, and extended here for EPFLTL),
-$I [ \varphi , \sigma, i ]$ is a function that returns
-a set of assignments such that $\gamma \in I [ \varphi , \sigma, i ]$ 
-iff $( \gamma , \sigma , i ) \models \varphi$.
-
-We present here only two simple cases of the set semantics. The full set semantics appears in the appendix. 
-\begin{itemize}
-\item $I [ ( \varphi \wedge \psi ) , \sigma , i ] = 
-I [ \varphi , {\cal R} , i ] \;  \bigcap \; I [ \psi , \sigma , i ]$.
-\item $I [ ( \varphi \; \since \; \psi ) , {\cal R} , i ] = 
-I [ \psi , {\cal R}, i ] \; \bigcup \;
-( I [ \varphi , {\cal R} , i ] \ \bigcap \ 
-I [ (\varphi \since \psi ) , {\cal R} , i - 1 ] )$.
-\end{itemize}
-
-\subsubsection{Runtime verification algorithm for PFLTL.}
-
-We start by describing an algorithm for monitoring 
-PFLTL properties, first presented in~\cite{HPU} and implemented
-in the tool~\dejavu. 
-We enumerate data values appearing in
-monitored events, as soon as we first see them. We
-represent relations over the Boolean
-encodings of these enumeration, rather than over the data values themselves. A hash function is used to connect the data values to their enumerations to maintain consistency
-between these two representations.
-The relations are
-represented as BDDs~\cite{Bryant}. 
-For example, if the runtime verifier sees the input 
-events 
-$\mathit{open}(\text{``a''})$, 
-$\mathit{open}(\text{``b''})$, 
-$\mathit{open}(\text{``c''})$, 
-it will encode them as
-$000$, $001$ and $010$ (say, we use 3 bits $b_0$, $b_1$ and $b_2$
-to represent each enumeration, with $b_2$ being the most significant bit).
-%
-A Boolean representation of the {\em set} of values 
-$\{\text{``a''},\text{``b''}\} $ would be equivalent to a Boolean function $(\neg b_1 \wedge \neg b_2)$ that returns 1 for $000$ and $001$ (the value
-of $b_0$ can be arbitrary).
-
-%
-
-Since we want to be able to deal with infinite domains
-(where only a finite number of elements may appear in a given observed prefix) and maintain the ability to perform
-complementation, unused enumerations represent the
-values that have not been seen and their relations
-to all other values. 
-%The algorithm maintains that these unseen values are correctly represented, as shown in~\cite{HPU}. 
-In fact, it is sufficient to have just one enumeration representing these values per each variable of the LTL formula. 
-%However, it is important to guarantee that at least one such enumeration exists. 
-We guarantee that at least one such enumeration exists by preserving for that purpose the enumeration $11\ldots11$.
-%, i.e., the highest Boolean representation,
-%to represent values not yet seen.
-We present here only the basic algorithm. For versions that
-allow extending the number of bits used for enumerations and garbage collection of enumerations, consult~\cite{HP}.
-
-\iffalse
-Our preferred representation of a set of values (assignments) is as an Ordered Binary Decision
-Diagram (OBDD, although we write simply BDD)~\cite{Bryant}.
-A BDD~\cite{Bryant} is essentially a compact representation 
-of a Boolean tree, where compaction glues together isomorphic subtrees. Each non-leaf node is labeled with one of the
-Boolean variables $b_0,\ldots,b_{k-1}$ (where $k$ is the number of bits used to represent the values of a variable). A non-leaf node $b_i$ 
-is the source of two 
-arrows leading to other nodes, representing respectively that the node has the value 0 (false) or 1 (true).
-%A dotted-line arrow represents that $b_i$ has the Boolean value $0$, while a thick-line 
-%arrow represents that it has the value $1$. 
-The nodes in the DAG have the
-same order along all paths from the root. However, some of the nodes may be
-absent along some paths, when the result of the Boolean function does not 
-depend on the value of the corresponding Boolean variable. Each path leads 
-to a leaf node that is marked by either a $0$ or a $1$, representing the 
-Boolean value returned by the function for the Boolean values on the path.
-\fi
-
-The use of BDDs can be replaced by other representations that
-can compactly and efficiently represent sets of values (e.g., ZDDs), and to which one can apply set operations like complementation, intersection, union (which are simply negation, conjunction and disjunction
-in BDDs) and projection (for the quantification).
-The marriage of BDDs and Boolean enumeration is in particular
-efficient, since collections of adjacent Boolean enumerations tend to compact well.
-
-\iffalse %%%%%
-The definition of the {\em since} operator $S$ can be rewritten in a standard
-way such that it refers only to the positions $i$ and $i-1$ in
-the sequence $\sigma$. 
-This is based on the
-fact that according to the semantics of {\em since},
-$(\varphi \, \since \, \psi) = ( \psi \vee (\varphi \wedge \ominus 
-(\varphi \, \since \, \psi )))$.
-This will serve in the
-implementation to work with only two versions of
-the sets of assignments, for the current and previous state:
-
-\begin{itemize}
-\item $( \gamma , \sigma , i ) \models ( \varphi \, \since \, \psi )$ if
-$( \gamma |_{\vars ( \psi )} , \sigma , i ) \models \psi$ or $i>1$,
-$( \gamma |_{\vars ( \varphi )} , \sigma , i ) \models \varphi$, and 
-$( \gamma , \sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
-\end{itemize}
-\fi %%%%%
-
-%The rest of the operators are defined as follows:
-%$\false = \neg \true$, $\forall x \; \varphi = \neg \exists x \neg \varphi$,
-%$(\varphi \vee \psi) = \neg ( \neg \varphi \wedge \neg \psi )$.
-%We can also define the following useful operators:
-%$P \varphi = (\true \, \since \, \varphi)$ (for ``previously''),
-%$( \varphi \, R \, \psi ) = \neg ( \neg \varphi \, \since \, \neg \psi )$ 
-%(the dual of the Since operator), and
-%$H \varphi = ( \false \, R \, \varphi )$ (for ``always in the past'').
-
-%\noindent {\bf Set Semantics.}
-%It helps to represent the algorithm to 
-%redefine the semantics of
-%the logic in terms of sets of assignments satisfying a formula.
-%Then $I [ \varphi , \sigma, i ]$ is a function that returns
-%a set of assignments such that $\gamma \in I [ \varphi , \sigma, i ]$ 
-%%Then a BDD for a set containing the values ``de'' and ``af''
-%%(2nd and 3rd values)
-%will return $1$ for $01$ and $10$. If the Boolean function is over
-%$b_0$ (for most significant bit) and $b_1$ (for least significant), 
-%then this is the Boolean function 
-%$(\neg b_0 \wedge b_1) \vee (b_0 \wedge \neg b_1)$.
-
-Given some ground predicate $p ( a )$, observed in the monitored execution, 
-matching with $p ( x )$ in the monitored property,
-let $\lookup ( x, a )$ be the enumeration of $a$ (a lookup in the hash table). If this
-is $a$'s first occurrence, then it will be assigned a new enumeration.
-Otherwise, $\lookup$ returns the enumeration that $a$ received before.
-We can use a counter, for each variable $x$, counting the number of different values appearing so far for $x$.
-When a new value appears, this counter is incremented, and the value is converted to
-a Boolean representation. 
-%Enumerations that were not yet used represent
-%the values not seen yet. 
-%In particular, we always leave at least
-%one enumeration, $11\ldots 11$, for this purpose.
-%In the next section we introduce data reclaiming, which allows 
-%reusing enumerations for values that no longer affect the checked 
-%property. 
-%This involves a more
-%complicated enumeration mechanism.
-
-The function $\bddset (x, A)$ returns
-a BDD that represents the set of assignments where $x$ is mapped to 
-(the enumeration of) $v$ for
-$v \in A$. This BDD is independent
-of the values assigned to any variable
-other than $x$, i.e., they can have any value.
-For example, assume that we use the three Boolean variables (bits) $x_0$, $x_1$ and $x_2$
-for representing enumerations over $x$ (with $x_0$ being the least significant bit), and
-assume that $A = \{ a , b \}$, 
-$\lookup ( x, a ) = 000$, and $\lookup ( x , b ) = 001$.
-Then $\bddset ( x , A )$ is a BDD representation of the Boolean function 
-$(\neg x_1 \wedge \neg x_2)$. 
-%But the BDD function will be independent of the values of Boolean variables
-%representing other variables.
-%Notice that $p ( x )$ is
-%equivalent, and is represented as $\exists y \, \exists z \, p ( x )$
-%when the variables used in the formula are $x$, $y$ and $z$.
-
-Intersection and union of sets of assignments are translated simply
-to conjunction and disjunction of their BDD representation,
-respectively, and complementation
-becomes BDD negation. We will denote
-the Boolean BDD operators as $\bddwedge$, $\bddvee$ and $\bddneg$.
-To implement the existential (universal, respectively) operators, 
-%as in the interpretation of $\exists x \; \varphi$, 
-we use the BDD existential (universal, respectively) operators over
-the Boolean variables that represent (the enumerations of) the values of $x$. 
-%Thus, $x$ is translated into
-%is represented using the Boolean variables $x_1$, $x_1, \ldots x_{k}$ into 
-Thus, if $B_{\varphi}$ is the BDD representing
-the assignments satisfying $\varphi$ in
-the current state of the monitor, then 
-$\bddexists (\langle x_0,\ldots,x_{k-1} \rangle, B_{\varphi})$
-is the BDD that represents the assignments satisfying $\exists x \; \varphi$ in the current
-state.
-Finally, $\bfalse$ and $\btrue$  are the BDDs that return always
-$0$ or $1$, respectively.
-
-%The case where the number of bits used to represent values for
-%some domains is eventually insufficient is dealt with by
-%expanding the number of bits for the BDD. This will be explained later,
-%but is much harder to implement, and may require a lot of overhead
-%to update the BDDs while online computation is monitored. This
-%may be unnecessarily, as we can assign a large number of bits
-%a priory, where these bits represent a set of values that is
-%exponentially larger.
-
-\iffalse
-The algorithm shown below uses two vectors (arrays) of values indexed by subformulas: 
-$\old$ for the state before that event, and
-$\current$ for the current
-state (after the last seen event).
-While in the propositional case, in Section~\ref{LTLruntime}, the vectors contain Boolean values, here
-they contain BDDs. The temporal relations,
-restricted to each state, are
-represented also as BDDs.
-The algorithm follows.
-\fi
-
-\begin{enumerate}
-\item Initially, for each subformula $\varphi$ of $\eta$,
-%$B ( \varphi , \current ) = \bfalse$.
-$\current ( \varphi ) := \bfalse$.
-\item Observe a new state (as a set of ground predicates) $s_i$ as input. 
-\item Let $\old := \current$.
-\item Make the following updates for each subformula. If $\varphi$ is
-      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
-      $\current ( \psi )$.
-\begin{itemize}
-  \item $\current ( \true ) := \btrue$.
-  \item $\current ( p_k ( a ) )$ := if $R_k [ i ] ( a )$ then
-  $\btrue$ else $\bfalse$.
-  \item $\current ( p_k ( x ) )$ :=
-     $\bddset ( x ,  \{ a \ | \ R_k[i] ( a  ) \} )$.
-  \item $\current ( ( \varphi \wedge \psi ) ) := 
-  \bddwedge ( \current ( \varphi ) , \current ( \psi ))$.
-  \item $\current ( \neg \varphi  ) := \bddneg ( \current ( \varphi ))$.
-  \item $\current ( ( \varphi\ \since\ \psi ) ) :=  
-  \bddvee ( \current ( \psi  ) , \bddwedge ( \current ( \varphi ) ,
-      \old ( ( \varphi \ \since \ \psi ))))$.
-%DAP: I removed the \; around \since, since it penetrated the 2nd column!
-  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
-  \item $\current ( \exists x \; \varphi ) := 
-  \bddexists (\langle x_0,\ldots,x_{k-1} \rangle, \current ( \varphi ))$.
-\end{itemize}
-\item If $\current (\eta) = \false$ then
-report a violation, otherwise goto step~2.
-\end{enumerate}
-
-\subsubsection{Runtime verification algorithm for EPFLTL}
-
-We extend now the algorithm to capture EPFLTL.
-The auxiliary relations $r_j$ extend the model, and we need to keep 
-BDDs representing $\current ( r_j )$ and
-$\old (r_j)$ for each relation $r_j$. We also need to calculate the 
-subformulas $\varphi_i$ that appear in
-a specification, as part of the runtime verification,
-as per the above FPLTL algorithm. One subtle point is that the auxiliary 
-relations $r_j$ may be defined in a rule with respect to
-a variable $x_j$ (as $r_j (x_j ):= \varphi_j (x_j)$ (this can be generalized 
-to any number of variables), but $r_j$ can be used
-as a subformula with other parameters in other rules or in the statement e.g., 
-as $r_j ( y )$. This can be resolved by a BDD renaming function 
-$\rename ( r_j (x_j ) , y)$.
-
-We then have to add
-following updates to step~4 of the above algorithm, when performing runtime verification of
-$\psi$.
-
-\noindent \begin{description}
-\item For each rule 
-$r_j ( x_j ) := \varphi_j ( x_j )$:
-\item{\ \ }  calculate $\current ( \varphi_j )$;
-\item{\ \ }  $\current ( r_j )  := \current (\varphi_j)$;
-\item{\ \ } $\current ( r_j ( y ) )$ := $\rename ( r_j(x_j) ,  y )$;
-\item{\ \ } $\current ( r_j ( a ) )$ := if $\current ( r_j ) ( a )$
-then $\btrue$ else $\bfalse$
-\end{description}
-
-%\klaus{Above: $\current ( r_j ) ( a )$ has not been defined I think.}
-
-\noindent
-As in the propositional case, the evaluation order cannot be
-simply top down or bottom up, since relations can appear
-both on the left and the right of a definition
-such as $r ( x ) := p ( x ) \vee \ominus r ( x )$. For that,
-we need to use the {\em mixed} evaluation order, described in Section~\ref{LTLruntime}.

### 2019-07-15 introduction.tex
Updates from Overleaf

\section{Introduction}

Runtime verification (RV)  \cite{bartocci18,havelund-rv-data-2018} 
refers to the use of rigorous (formal) 
techniques for {\em processing} execution traces emitted by a system being observed. 
The purpose is typically to evaluate the behavior of the observed system. We focus here on {\em specification-based} runtime verification, 
where an execution trace is checked against a property expressed in a formal logic, 
in our case variants of Linear Temporal Logic (LTL).

LTL is a common specification formalism for reactive and concurrent systems. It is often used in model checking and runtime verification. Another formalism that is used for the same purpose is finite automata, often over infinite words. This includes B\"{u}chi, Rabin, Street, Muller and Parity automata~\cite{Thomas}, all having the same expressive power. In fact, model checking of an LTL specification
is usually performed by first translating the specification into a B\"{u}chi automaton. The automata formalisms are more expressive than LTL, with a classical example by Wolper~\cite{Wolper}, showing that
it is not possible to express in LTL that every even state in
the sequence satisfies some proposition $p$. 
%
This has motivated extending LTL in various ways to achieve the same expressive power as B\"{u}chi automata:
Wolper's ETL \cite{Wolper,WVS} uses right-linear grammars,
%\footnote{ETL allows a formula $\varphi$ to be
%the definition of a right-linear grammar
%(equivalent to a finite automaton) together with a mapping $M$ from the terminals 
%of the grammar to temporal sub-formulas. A trace satisfies $\varphi$ if there is some word $w$ of terminals 
%generated by the grammar, such that the trace in every position $i$ satisfies the sub-formula $M(w_i)$, where 
%$w_i$ is the terminal at position $i$.}.
Sistla's~\QLTL{} extends LTL  
with dynamic (i.e., state-dependent, second-order) {\em quantification over propositions}~\cite{sistla-phd-83} and the PSL standard~\cite{psl-standard-2010} extends LTL with regular expressions.
However, these and other extensions have not been extensively used
for RV (e.g., \QLTL\  can be nonelementary more compact, hence
correspondingly more complex to monitor than LTL).

We therefore first present an alternative 
extension of  propositional LTL with {\em rules},
named \ELTL{}. These rules define and use
auxiliary propositions, not appearing in the execution itself. These propositions obtain their values in a state as a 
function of the prefix of the execution
up to and including that state, expressed as 
a past-time temporal formula. This extension fits
easily and naturally to existing RV algorithms that
use incremental
summaries of prefixes, e.g., the classical algorithm~\cite{HR}
for past time LTL (denoted here~\PLTL{}), maintaining also its linear time complexity (in the length of
the trace and the size of the formula).
In fact, our extension of the logic is inspired by that RV algorithm.
The logic~\ELTL{}
is shown to be equivalent to \QLTL{}
and its restriction to past properties is equivalent to 
B\"{u}chi automata and regular expressions.

Another expressiveness dimension is runtime verification of events that carry 
data, for which a first-order LTL supporting {\em quantification over data}
is appropriate, here referred to as \FLTL.
We demonstrate the weakness of \FLTL{} in
expressing Wolper's example, relativized to the first-order 
case,  and in expressing
the transitive closure of temporal relations over events. 
We therefore introduce two alternative ways 
of extending the expressive power of \FLTL{}, corresponding, respectively, to the propositional logics \QLTL{} and \ELTL{}. 
The first adds 
  quantification over relations of data, obtaining a logic 
referred to as \QFLTL{}.
The second extension adds rules for the first-order case, and is referred to as \EFLTL. Both of these extended logics can express the above examples.

%Adding rules to LTL furthermore interestingly allows to express the 
%transitive closure of a relation. That is, in first-order logic, given a relation 
%representing neighbors in a graph, it is not possible to express that there is a path between nodes. The  addition of rules allow this to be expressed.
We show that for the first-order case, in contrast to the propositional case, the extension of the logic with  quantification is more expressive than the extension with
rules.
\iffalse
While
it is not clear how to construct a monitoring
algorithm for \QFLTL{}, for which we 
can construct a monitoring
algorithm. 
\fi

Runtime verification is commonly restricted to
the {\em past-time} versions of LTL, i.e.,
to {\em safety} properties~\cite{AS}, where
a violation can be detected and demonstrated
after a finite prefix of the execution.
We refer to the logic
\PLTL{} for the propositional case and to \PFLTL{} for the first-order case; 
these logics also enjoy elegant RV algorithms, based on the ability to compute summaries of the observed
prefixes~\cite{HR,HPU}, as opposed to future
temporal logics~\cite{KV}.  The RV algorithm, presented here
for~\EPFLTL{} (the safety part of~\EFLTL{})
naturally
extends the RV algorithm for \PFLTL{} in~\cite{HPU} 
in the same way that the algorithm we present for \EPLTL{}
(the safety part of~\ELTL{}) extends the RV algorithm
in~\cite{HR} for \PLTL{}.

\iffalse
A past-time property has the quality
that given a finite prefix of the execution, it is known whether the property
is violated or not; this in contrast to future time properties.
Consistent with that, we provide runtime verification algorithms for the past-time versions of the rule-based extensions of propositional and first-order LTL. 
\fi
%
We further present a corresponding extension of the 
\dejavu{} tool
%\footnote{\dejavu{} is available at 
%\url{https://github.com/havelund/dejavu}.}
\cite{HPU,HPU-FMSD,HP}, that realizes the extension 
of first-order past-time LTL with rules (\EPFLTL). The \dejavu{} tool allows runtime 
verification on past-time first-order temporal logic over infinite domains (e.g., the 
integers, strings, etc.). It achieves efficiency by using a unique BDD representation of the data part; BDDs correspond to relations over a Boolean enumeration of the
input data (with a hash table representing the
correspondence between the data and the enumerations). This is a very different
use of BDDs from the classical model checking representation of sets of Boolean states\footnote{E.g., in~\cite{Green-berg},
BDDs are used to represent sets of program locations,
and the data
elements are represented
symbolically as a formulas.}.
A garbage collection algorithm tailored for that representation
also assists in obtaining efficiency.

Our main contribution is the LTL logics extended with rules (extensions prefixed with `R'), and in particular the
logic \EPFLTL{} (past-time first-order LTL extended with rules) and its implementation.
 The structure of the paper reflects our 
step-wise approach by 
first exploring the problem in the propositional case to form a basic understanding (but we believe that
our extension of propositional LTL is
interesting by itself), and then by addressing the more interesting first-order case.

Numerous monitoring related expressive logics and systems have been developed over 
the past decades.
In the database community, relations have been added to temporal 
databases for aggregation \cite{Libkin}, calculating functions (sums etc.). 
Aggregations were also used in the runtime verification tool {\sc MonPoly}
\cite{agrebasin}.   
Numerous other systems have been produced for monitoring execution traces with data 
against formal specifications. 
These include e.g. {\sc Mop} \cite{Meredith2011}, {\sc Qea} \cite{Reger2015}, and {\sc Larva} \cite{larva}, which
provide automaton-based data parameterized logics; {\sc Lola}
\cite{LOLA}, which is based on stream processing; {\sc BeepBeep} 
\cite{halle-beepbeep-ieee-12} which is temporal logic-based; and the 
rule-based {\sc LogFire} \cite{havelund-logfire-sttt14}. These systems address the
expressiveness issues discussed in this paper in different ways. Our approach differs
from earlier such work by taking a starting point in LTL and
extending it with rules, implemented using BDDs.
%{\sc PSL} \cite{psl-standard-2010} is designed for model 
%checking and monitoring of hardware systems. By
%combining linear temporal logic and regular expressions 
%t achieves part of the goals of the here presented work. 
%However, it is finitely limited in the quantification it 
%allows over data.

%The contributions of this paper are both theoretical and practical. On the theory 
%side, we present and study extensions for propositional and first order linear-
%temporal logics and show the relations between them and existing versions. 
%On the practical side, we present RV algorithms for the extended logic, and 
%show how to cast them specifically in an efficient BDD-based implementation. We 
%present some experiments performed with our implementation and report on the 
%results.

\iffalse
That is,
in Section~2 we first define the syntax and semantics of propositional LTL. We then review some of its classical extensions,
including \QLTL, and present our own extension \ELTL. In Section~3 we present a runtime verification algorithm for the extended past-time propositional LTL (\EPLTL). In Section~4 we review first-order linear temporal logic (\FLTL), present our extension \EFLTL, and study the relationships between the presented versions of first-order LTL. In Section~5 we present a runtime verification algorithm for the past-time version of the extended first-order LTL (\EPFLTL). Section~6 describes the implementation and provides experimental results. Finally, Section~7 concludes the paper. 
\fi
% ---

% Explanation of letters

\noindent {\bf Conventions.} As already outlined above,  we present several versions of LTL. 
We name the different versions by prefixing LTL with the following letters.
`P' : restricted to {\em P}ast-time temporal operators;
`F' : allowing {\em F}irst-order (static) quantification over data assigned to variables;
`Q' : adding second-order (dynamic) {\em Q}uantification over propositions/predicates; and finally
`R' : adding {\em R}ules, our main contribution.

% ---

% ----------------

%\iffalse
%The contributions of the papers are both theoretical, and
%practical. We study the extensions of propositional temporal logic (both full 
%version and past/safety version) and suggest a restricted (and, to our opinion, an 
%intuitive) extension that is equivalent to the power of quantified LTL or B\"{u}chi/
%finite automata. We then study 
%The corresponding extension to first-order LTL, replacing propositions with 
%relations, and allowing variables and constant values. We propose, for the first-
%order version, an extension similar to the one we propose for the propositional 
%case. But then we show that for the first-order case, this extension has
%less expressive power than adding quantification in the style of propositional 
%temporal logic.
%
%The extensions we suggest are in particular easy to implement, as an extension to 
%existing propositional/first-order runtime verification. We extend our tool \dejavu, 
%experiment with new examples that take advantage of the extension.
%\fi

%\noindent {\bf Conventions.}
%\label{sec:prelim}
%In this paper, we treat and present several versions of the logic LTL. To simplify
%the presentation, we prefix LTL with letters as follows:

%\begin{itemize}
%\item{P} restricts to P{\em ast} temporal operators.
%\item{F} allows F{\em irst-order} quantification over
%data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
%\item{Q} adds (second-order) Q{\em uantification} that is state-dependent. This is referred to as {\em } quantification.
%\item{E} allows E{\em xtending} the 
%set of objects that occur in the input with
%new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
%\end{itemize}

### 2019-07-15 main.tex
Updates from Overleaf

--- 2019-04-19 main.tex
@@
 \documentclass{llncs}
+\usepackage{authblk}
 %\documentclass[converence,10pt]{IEEEtran}
@@
 
-\title{An Extension of LTL with Rules\\ 
-and its Application to Runtime Verification}
+\title{An Extension of LTL with Rules\\ and its Application to Runtime Verification}
 
-\ignore{\author{
+%\ignore{
+%\author{\IEEEauthorblockN{Klaus Havelund} 
+%\IEEEauthorblockA{\textit{Laboratory for Reliable 
+%Software} \\
+%\textit{NASA Jet Propulsion Laboratory}\\
+%Pasadena, USA \\
+%klaus.havelund@jpl.nasa.gov}\\\mbox{}\\
+%\and
+%\IEEEauthorblockN{Doron Peled} 
+%\IEEEauthorblockA{\textit{Dep. of Computer Science} \\
+%\textit{Bar Ilan University}\\
+%Ramat Gan, Israel \\
+%doron.peled@gmail.com}
+%}
+
+%\author{Submitted Anonymosly}
+
+%\thanks{Copyright 2018. All rights reserved.}
+
+\author{Klaus Havelund$^1$ \and Doron Peled$^2$}
+
+%\ignore{
+%Klaus Havelund~\thanks{The research performed by this %author was carried out at Jet Propulsion Laboratory, %California Institute of Technology, under a contract %with the National Aeronautics and Space %Administration.}
+%\and
+%Doron Peled
+%~\thanks{The research performed by this author was %partially funded by Israeli Science Foundation grant %1464/18: 
+%``Efficient Runtime Verification for Systems with Lots %of Data and its Applications''.
+
+\institute{
+Jet Propulsion Laboratory, 
+California Institute of Technology, USA 
+\and
+Department of Computer Science,
+Bar Ilan University, Israel
+}
+
+\iffalse
+\author{
 \begin{tabular}{ccc}
@@
 }
-}
-
-\ignore{
-\author{\IEEEauthorblockN{Klaus Havelund} 
-\IEEEauthorblockA{\textit{Laboratory for Reliable 
-Software} \\
-\textit{NASA Jet Propulsion Laboratory}\\
-Pasadena, USA \\
-klaus.havelund@jpl.nasa.gov}\\\mbox{}\\
-\and
-\IEEEauthorblockN{Doron Peled} 
-\IEEEauthorblockA{\textit{Dep. of Computer Science} \\
-\textit{Bar Ilan University}\\
-Ramat Gan, Israel \\
-doron.peled@gmail.com}
-}
-}
-
-%\author{Submitted Anonymosly}
-
-%\thanks{Copyright 2018. All rights reserved.}
-
-\iffalse
-\author{ 
-Klaus Havelund~\thanks{The research performed by this author was carried out at Jet Propulsion Laboratory, California Institute of Technology, under a contract with the National Aeronautics and Space Administration.}
-\and
-Doron Peled~\thanks{The research performed by this author was partially funded by Israeli Science Foundation grant 1464/18: 
-``Efficient Runtime Verification for Systems with Lots of Data and its Applications''.
-}
-}
-
-\institute{
-Jet Propulsion Laboratory, \\
-California Institute of Technology, USA 
-\and
-Department of Computer Science \\
-Bar Ilan University, Israel
-}
-
 \fi
@@
 
-\begin{abstract}
+\input{abstract}
 
-Runtime Verification (RV) consists of analyzing execution 
-traces using formal techniques. 
-%RV is scalable and 
-%therefore has a big potential for infusion in software 
-%engineering practices. 
-It includes monitoring the 
-execution of a system against properties formulated in 
-Linear Temporal Logic (LTL). 
-LTL offers a succinct notation for writing useful 
-specification properties. However, it is limited in 
-expressiveness in the propositional case, and several 
-theoretic extensions have therefore been proposed. Furthermore, for many practical cases, there is a need to monitor properties 
-that carry data, where one can use formalisms like 
-first-order LTL. We show that 
-first-order LTL has similar expressiveness
-limitations as the propositional version. We suggest here 
-two related extensions for increasing the expressive power: 
-one for propositional LTL and one for 
-first-order LTL. These extensions have a simple incremental operational 
-semantics that is more suitable for RV than previously 
-suggested extensions. We show that the 
-propositional extension has the same expressiveness as the 
-classical extensions, and demonstrate its
-adoption for runtime verification of first-order safety 
-properties. Finally, we expand the BDD-based runtime verification 
-tool \dejavu{}\footnote{The actual tool name is not given due to anonymous reviewing.}  to support our extension and perform some 
-experiments.
-
-\end{abstract}
+\input{introduction}
 
@@
 \input{conclusion}
+
+%\newpage
 
@@
 way~\cite{Thomas}, and then from
-second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
+second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a deterministic B\"{u}chi automaton may not always be possible), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
 each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set.
@@
 In the last item, the extended model ${\cal Q}$ appears both on the left and the right of the equality sign.
-It is well defined because of Lemma~\ref{samesame} (equivalently, a more verbose definition can include a stepwise construction of ${\cal Q}$ from ${\cal R}$).
+It is well defined because of Lemma~\ref{samesame} (equivalently, a more verbose definition can include a prefix-inductive construction of ${\cal Q}$ from ${\cal R}$).
 \end{document}

### 2019-07-15 rule-ltl.tex
Updates from Overleaf

@@
-\section{Introduction}
-
-Runtime verification (RV)  \cite{bartocci18,havelund-rv-data-2018} very generally refers to the use of rigorous (formal) 
-techniques for {\em processing} execution traces emitted by a system being observed. The purpose is typically
-%, again generally viewed, 
-to evaluate the state of the observed system. Processing can take numerous
-forms.
-We focus here on {\em specification-based} runtime verification, where an execution trace is checked against a property expressed in a formal logic, in our case variants of linear temporal logic.
-
-Linear Temporal Logic (LTL) is a common specification formalism for reactive and concurrent systems.
-It is often used in e.g.
-model checking and runtime verification.
-Another formalism that is used for the same purpose is finite automata, often over infinite words.
-This includes B\"{u}chi, Rabin, Street, Muller and Parity automata~\cite{Thomas}, all having the same expressive power.
-In fact, model checking of an LTL specification
-is usually performed by first translating the specification into a B\"{u}chi automaton.
-The automata formalisms are more expressive than LTL, with a classical example by Wolper~\cite{Wolper} showing that
-it is not possible to express in LTL that every even state in
-the sequence satisfies $p$. This has motivated extending LTL
-in various ways.
-We are mainly motivated by runtime verification of executions {\em with data}, for which a first-order temporal logic is
-appropriate. There, too, we show that expressiveness can be extended.
-
-We present first an extension of  propositional LTL that is based on using additional (auxiliary) propositions, not appearing in the execution, which are dependent on the past 
-of the current execution.
-These propositions obtain their value in a state as a function of the prefix of the execution up and including that state through a past temporal formula.
-This can be seen as adding
-summary states to the temporal specification, resulting in a specification that is in between logic and finite automata representation. This extension, which we believe is conceptually simpler
-and more intuitive than other suggested extensions, is shown to be
-equivalent to the other common specification formalism:
-B\"{u}chi automata, regular expressions and quantified LTL.
-
-We then proceed to first-order LTL. We relativise Wolper's example to the first
-order case to motivate the need for
-extending first-order LTL.
-Another example is based on the inexpressiveness of the transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
-We present an extension where auxiliary relations, not appearing in the execution, are
-added. While this extension
-is capable of expressing the above
-properties, we show that for the first-order version, this has less 
-expressive power than adding full quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
-provides the full expressive power of, say, B\"{u}chi automata.
-
-The extensions we proposed to the logics are in particular natural for
-runtime verification of past (safety) temporal properties, since
-the values of propositional and first-order LTL are based
-on calculating and updating a summary of the prefix of the execution.
-Runtime verification is often restricted to a {\em past}  version of temporal logic~\cite{HR}, mainly because for past properties we can decide when the monitored property is violated after monitoring a finite prefix of it, while for future properties we may never know for sure.
-Consistent with that,
-we provide runtime verification algorithms for the past-time versions of the extended propositional and first-order
-logics. We further present an extension of the 
-\dejavu{} tool (publicly available on {\sf github},
-hidden here because of anonymous reviewing)
-\ignore{(https://github.com/havelund/dejavu)} that
-realizes the extended first-order past temporal logic verification.
-The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first-order temporal logic over infinite domains (e.g., the integers, strings).
-It achieves efficiency by using BDD representation of enumerations of data, rather than over the data values themselves, and by using
-a garbage collection algorithm.
-
-The contributions of this paper are both theoretical and practical.
-On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions.
-On the practical side, we present RV algorithms for the extended logic, and 
-show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.
-
-The structure of this paper is as follows.
-In Section~2 we first define the syntax and semantics of propositional temporal logic.
-We then review some of its classical extensions and present our own extension.
-In Section~3 we present a runtime verification algorithm for the extended past LTL.
-In Section~4 we review first order linear temporal logic, present our extension and study the relationships between the presented versions of first-order LTL.
-In Section~5 we present a runtime verification algorithm for the past version of the extended first-order LTL.
-Section~6 describes the implementation and provides experimental results.
-Finally, Section~7 concludes the paper.
-
-%\klaus{I think that the structure of the paper should be
-%crystal clear so that the reader does not get lost. A lost reader leads to rejection. If a reader understand the structure, but perhaps not the details, the reader may still be happy.}
-
-\vspace{1ex}
-\noindent {\bf Related work.}
-For propositional LTL, the logic ETL~\cite{Wolper,WVS} extends
-the set of temporal operators 
-by using automata (or, equivalently a right linear grammar).
-Each automaton defines a temporal operator, where the input letters of that automaton correspond to future temporal subformulas.
-Each accepted run of such an automaton defines then the places
-where the subformulas need to hold. Another expressive-equivalent extension, 
-in~\cite{Wolper}, adds dynamic, i.e.,
-state-dependent, quantification over propositions to LTL.
-
-Addition of relations to temporal databases is done
-for aggregation~\cite{Libkin}, where relations are
-calculated, based on the current state, and are used for further calculating
-functions (sums, maxima) over the sequence of states. Aggregations were also used in the runtime verification tool 
-{\sf MonPoly}~\cite{agrebasin}.
-%
-% --- other less directly related work: ---
-%
-Numerous other systems have been produced for monitoring
-execution traces eith data against formal specifications. 
-These include e.g. \cite{LOLA,larva,halle-beepbeep-ieee-12,Meredith2011,Reger2015}.
-
-% ----------------
-
-\iffalse
-The contributions of the papers are both theoretical, and
-practical.
-We study the extensions of propositional temporal logic (both full version and past/safety version) and suggest a restricted (and, to our opinion, an intuitive) extension that is equivalent to the power of quantified LTL or B\"{u}chi/finite automata.
-We then study
-The corresponding extension to first-order LTL, replacing propositions with relations, and allowing variables and constant values.
-We propose, for the first-order version, an extension similar to the one we propose for the propositional case.
-But then we show that for the first-order case, this extension has
-less expressive power than adding quantification in the style of propositional temporal logic.
-
-The extensions we suggest are in particular easy to implement, as an extension to existing propositional/first-order runtime verification.
-We extend our tool \dejavu, experiment with new examples that take advantage of the extension.
-\fi
-\vspace{1ex}
-\noindent {\bf Conventions.}
-\label{sec:prelim}
-In this paper, we treat and present several versions of the logic LTL. To simplify
-the presentation, we prefix LTL with letters as follows:
-
-\begin{itemize}
-\item{P} restricts to P{\em ast} temporal operators.
-\item{F} allows F{\em irst-order} quantification over
-data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
-\item{Q} adds (second-order) Q{\em uantification} that is state-dependent. This is referred to as {\em dynamic} quantification.
-\item{E} allows E{\em xtending} the 
-set of objects that occur in the input with
-new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
-\end{itemize}
-
-%\klaus{You above introduce some concepts before they really are explained.
-Q and E.
-I understand why you do this (to have it in one place), but it may be too difficult for the reader to grasp if not "initiated".
-%Especially the E.}
-
-%\klaus{We use \ELTL{} (etc) to denote Extended LTL. 
-%But \QLTL{} is also an extension of LTL. These logics 
-%are all extensions of LTL. Perhaps RLTL (for Rule-based 
-%LTL) would have been better ... or something. Not urgent, 
-%we can always change that if we want to.}
-
-%\klaus{The explanation of Q and E could be improved.}
-
-%\doron{Did my best to make the letters more self explained.
-I think R is usually reserved in temporal logic for realtime.
-But if you insist, one thing that I cannot perform from here is changing the two figures where these letters occur.}
-
 \newcommand\eventty{\mathbb{E}}
@@
 {\em until}, {\em previous-time} and {\em since}, respectively.
-
+%
 The models for LTL formulas 
@@
 \begin{itemize}
-\item $( \sigma, {i}) \models \mathit{true}$.
-\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$.
+\item $( \sigma, {i}) \models \mathit{true}$. \ \ \ \ \ 
+\item $(\sigma, {i}) \models p$ iff $p \in s_{i}$. \ 
 \item $(\sigma, {i}) \models \neg\varphi$ iff  $(\sigma, {i}) \not\models \varphi$.
@@
 $(\varphi \leftrightarrow \psi) = ((\neg \varphi \wedge \neg \psi) \vee (\varphi \wedge \psi ))$,
-$\Diamond \varphi = (\true \until \varphi)$ and $\Box \varphi = \neg \Diamond \neg \varphi$,
-${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}), and 
+$\Diamond \varphi = (\true \until \varphi)$, $\Box \varphi = \neg \Diamond \neg \varphi$,
+${\bf P}\  \varphi = (\true \; \since \; \varphi)$ ({\bf P} stands for P{\em reviously}) and 
 ${\bf H}\  \varphi = \neg {\bf P}\ \neg \varphi$ ({\bf H} stands for 
@@
 The expressive power of different versions of propositional LTL is often compared to regular expressions
-over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second-order logics where the temporal propositions
-correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
-LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star operator (or $\omega$).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}.
-In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
+over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second-order logic.
+%, where the  propositions in the LTL formula become
+%unary predicates over the naturals, and with the $<$
+%relation. 
+ Accordingly, we have the following characterizations:
+LTL is equivalent to monadic first-order logic,  star-free regular expressions\footnote{Regular expressions without the star operator (or $\omega$).} and counter-free B\"{u}chi automata.
+%\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}.
+For an overview of logic and automata see~\cite{Thomas}.
+Restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
 (and the ones derived from them $\Box$ and $\Diamond$)
-leaves the same expressive power and is commonly used.
-
-%\klaus{Is that also the true for the first-order case? Doron: I have no idea. But at this point, I would not ask to ask Gabbay.}
+maintains the same expressive power.
 
@@
 operators: $\since$, $\ominus$ and the operators derived
-from them, $\mathbf{H}$ and $\mathbf{P}$. The past time logic is sometimes interpreted over finite sequences, 
+from them, $\mathbf{H}$ and $\mathbf{P}$. The past-time logic is sometimes interpreted over finite sequences, 
 where $\sigma \models \varphi$ when $( \sigma , | \sigma | ) \models \varphi$.
-Alternatively,
-it is also a common practice to use the past time restricted version, prefixed with  a single $\Box$ (always) operator; in this case, each of the prefixes has to satisfy $\varphi$. This later
-form expresses {\em safety} LTL properties~\cite{AS} over infinite sequences. 
-(When \PLTL{} is interpreted over finite sequences, its
-expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.)
+It is also a common practice to use a \PLTL{} formula, prefixed with  a single $\Box$ (always) operator; in this case, {\em each of the prefixes} has to satisfy $\varphi$. This later
+form expresses {\em safety} LTL properties~\cite{AS}.
+When \PLTL{} is interpreted over finite sequences, its
+expressive power is the same as star-free regular expressions, first-order monadic logic over finite sequences and counting-free automata.
 
@@
 
-Wolper~\cite{Wolper} showed that the expressive power of LTL is lacking with respect
-to some important properties. As a canonical example, he demonstrated that it is impossible to
-express that all the states with even indexes in a sequence
-satisfy $p$ (this is
-different than saying that $p$ alternates between $\true$ and $\false$ on consecutive states).
-Wolper's example property cannot be expressed in \PLTL{} either.
-In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
-propositions, as shown below.
-
-\subsubsection{Extending LTL with dynamic quantification.}
-Adding quantification over propositions, as done in~\cite{Wolper}, allows writing a formula of the form
+Wolper~\cite{Wolper} demonstrated that the expressive power of LTL is lacking using
+the property that all the states with even\footnote{This is
+different than stating that $p$ alternates between $\true$ and $\false$ on consecutive states.} indexes in a sequence
+satisfy some proposition $p$.
+
+%Wolper's example property cannot be expressed in \PLTL{} either.
+%In order to extend the expressiveness of LTL, Wolper suggested to use linear grammars, or, alternatively, the ability to quantify over
+%propositions, as shown below.
+
+\vspace{1.2ex}
+\noindent {\bf Extending LTL with dynamic quantification.}
+Adding quantification over propositions, suggested by Sistla in~\cite{sistla-phd-83}, allows writing a formula of the form
 $\Exists q \, \varphi$, where
 $\Exists q$ represents {\em existential} quantification over a proposition $q$ that can appear in $\varphi$.
-
+%
 %A model for such a formula 
@@
 %the propositional variables appearing in the %states of the sequence $\sigma = s_1 s_2 %\ldots$
-and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)
+and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the propositions in $X$.) The semantics is defined as follows:
 
@@
 \end{itemize}
-%Since all the auxiliary variables appear in the scope of
+%Since all the auxiliary propositions appear in the scope of
 %quantifiers, it holds that $\sigma \models \varphi$
@@
 \neg \Exists q\, \neg \varphi$.
-This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states.
-It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
+This kind of quantification is considered to be {\em dynamic}, since the quantified propositions can have different truth values depending on the states.
+It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a proposition has the value $\true$.
 Extending LTL with such quantification, the logic \QLTL{} has the same expressive power as regular expressions, full B\"{u}chi
-automata, or monadic second-order logic with unary predicates over the naturals, see again~\cite{Thomas}.
+automata, or monadic second-order logic with unary predicates over the naturals (see again~\cite{Thomas}).
 In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of \QLTL{}~\cite{Thomas}. 
-
+\iffalse
 Wolper's property can be expressed in \QLTL{} as follows:
@@
 ( (  \neg q \wedge \Box (q \leftrightarrow \bigcirc \neg q )  )  \wedge \Box (q \rightarrow p ) )\end{equation}
-
+\fi
 Restricting \QLTL{} to the past modalities, one obtains the logic \QPLTL{}. \QPLTL{} has the same expressive
 power as regular expressions and finite automata.
-Equation~(\ref{form0}) can
+Wolper's property can
 be rewritten in \QPLTL{} as:
@@
 
-\subsubsection{Extending LTL with rules.}
-
-We introduce now another extension of LTL, which we call \ELTL{}.
+\vspace{1.5ex}
+\noindent {\bf Extending LTL with rules.}
+We introduce another extension of LTL, which we call \ELTL{}.
 %It has a form that puts it, conceptually, in between temporal
@@
 %a pure LTL formula, and on the other hand, one can write a specification that resembles a finite automaton. 
-As will be showed later, this extension is
+As will be showed later, this extension
 is very natural for runtime verification.
-
+%
 We partition the propositions $P$ into
@@
 as the {\em statement} of $\eta$ and to 
-$a_j  := \varphi_j$ as a {\em rule}. In text, rules
- will be separated from each other by commas. 
-\begin{tabbing}
-\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
-$\sigma' |_A = \sigma$ such that \\
- \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
-\end{tabbing}
+$a_j  := \varphi_j$ as a {\em rule} (in text, rules
+ will be separated  by commas).
+
+\begin{center}
+$\sigma \models \eta$ if there exists $\sigma'$, where
+$\sigma' |_A = \sigma$ s.t.  
+ $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
+ \end{center}
 
@@
 \Box ( q \rightarrow p ) \, \mathrm{where} \, 
- ( q := \ominus \neg q)
+  q := \ominus \neg q
 \end{equation}
 where $A = \{ q\}$ and $B = \{ p \}$.
-The auxiliary variable $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
+The auxiliary proposition $q$ is used to augment the input sequence such that each {\em odd} state will satisfy $\neg q$
 and each {\em even} state will satisfy $p$.  
-
-\ELTL{} extends the set of propositional variables with new variables, whose values at a state are
-functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
+%
+\ELTL{} extends the set of propositions with new propositions, whose values at a state are
+functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions summarize the past execution prefix up to and
 including the current state.
 This differs from 
-the use of auxiliary variables in \QLTL{}, where the values assigned to the
-auxiliary variables do not have to extend the 
+the use of auxiliary propositions in \QLTL{}, where the values assigned to the
+auxiliary propositions do not have to extend the 
 states of the model in a unique way throughout the interpretation of 
-the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
+the property over a model. The constraint that auxiliary propositions appearing in the formulas $\varphi_i$ 
 must occur within the scope of a $\ominus$ operator
-is required to help preventing that the values
-of the auxiliary variables are conflicting, as in
-$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
-that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.
-
-\begin{lemma} \label{fourone}
-Let $\varphi$ be an \ELTL{} formula
-over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
-a set of rules $a_j  := \varphi_j \, : \, 
- {j \in \{1, \ldots , n\}}$. 
-Let $\sigma$ be a model with states over $B$. 
-Then there is at most one
-model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $
+is required to prevent conflicting rules, as in
+$a_1 := \neg a_2$ and $a_2 := a_1$.
+%, which makes a formula
+%that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.
+
+\begin{lemma}[Well foundedness of auxiliary propositions] \label{fourone}
+The values of the
+auxiliary propositions
+of an \ELTL{} formula $\eta$
+are uniquely defined in a state of an execution
+by the prefix of the execution upto and including that state.
 
@@
 
-\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
-is defined, via a \PLTL{} formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states.  \qed
+\noindent {\bf Proof.}
+Let $\eta$ be a formula over auxiliary propositions $A$ and basic propositions $B$,
+with rules
+$a_j  := \varphi_j \, : \, 
+ {j \in \{1, \ldots , n\}}$.
+%, with
+%a set of rules $a_j  := \varphi_j \, : \, 
+% {j \in \{1, \ldots , n\}}$. 
+Let $\sigma$ be a model with states over $B$. 
+Then there is a unique
+model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $: inductively,
+the value of each auxiliary proposition $a_j$ at the $i$th state
+of $\sigma'$
+is defined, via a rule $ a_j := \varphi_j$, where $\varphi_j$ is a \PLTL{} formula; 
+hence it depends on
+the values of the propositions $B$
+in the $i$th state of $\sigma$, and on the values of $A \cup B$ in the previous states
+of $\sigma'$.  \qed
 %A \PLTL{} property depends in a state only on the prefix of the model that precedes and includes that state.
@@
 \end{theorem}
-\noindent {\bf Proof.}
-Each \ELTL{} formula $\eta$, as defined above,
+\noindent {\bf Sketch of proof.}
+Each \ELTL{} formula $\eta$, as defined in~(\ref{ELTL}),
 is expressible using the following equivalent \QLTL{} formula:
-
 \[   \Exists a_1 \ldots \Exists a_n  ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j))
 \]
-
 %\klaus{How do we know that the ``other direction'' below is possible?}
-
 \noindent
-For the other direction, one can translate the \QLTL{} property first into
+For the other direction, one can  first translate the \QLTL{} property  into
 a second-order monadic logic formula, then to a deterministic Muller automata and then construct an \ELTL{} formula that holds for the accepting executions of this automaton.
 The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
-The details of the
-constructions are given in the appendix. \qed
+Details are provided in an appendix. \qed
 
@@
 \iffalse
-The use of the auxiliary variables $a_1 \ldots a_n$ and the
+The use of the auxiliary propositions $a_1 \ldots a_n$ and the
 rules of an \ELTL{} specification 
@@
 and a finite automaton, whose states correspond to combinations
-of truth values of the auxiliary variables, and can be referred to
+of truth values of the auxiliary propositions, and can be referred to
 within $\psi$. As a corollary of Theorem~\ref{theo1}, 
@@
 This representation can be useful, for RV, as explored further in this paper, because of the possibility to
-incrementally calculate the values of the auxiliary variables. 
+incrementally calculate the values of the auxiliary propositions. 
 Similarly, it can be
@@
 We define \EPLTL{} by disallowing the future temporal operators (except an implied prefixing $\Box$) in \ELTL{}. This results in
-a formalism that is equivalent to finite (B\"{u}chi) automata, where all the states except one are accepting and where the non-accepting state is a sink. 
+a formalism that is equivalent to a B\"{u}chi automata, where all the states except one are accepting and where the non-accepting state is a sink. 
 We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:
@@
 \end{lemma}
-In fact, it is easy to show that \EPLTL{} can be restricted to the following form, while retaining its expressive power.
+
+\begin{lemma}
+\EPLTL{} can with no loss of expressive power be restricted to the form:
 \[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}  \] 
-with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
-of an \EPLTL{} property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a more intuitive specification.
-The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.
-An arrow from A to B means that logic  B is more expressive than A.
-A double headed arrow shows same expressiveness
+with $p$ being one of the auxiliary propositions $a_j$
+and $\varphi_j$ contains only a single
+occurrence of the $\ominus$ temporal operator (and the Boolean operators). 
+\end{lemma}
+In this form, the value of the Boolean variables $a_j$ encodes the states of an automaton, and the rules encode the transitions.
+%
+Yet, this may not be the preferred form
+of an \EPLTL{} property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a much more intuitive specification.
+%
+%The relative expressive power between the propositional temporal logics discussed 
+%in this section is summarized in Figure~\ref{prop}. An arrow from A to B means 
+%that logic  B is more expressive than A. A double headed arrow shows same 
+%expressiveness
 
@@
 
-\begin{figure}
-\begin{center}
-%\includegraphics[height=0.6in,width=2in]{PROP.jpg}
-\includegraphics[width=2.8in]{PROP.jpg}
-\caption{\label{prop}
-Comparing the expressive power of the temporal logics in Section~\ref{propLTL}
-}
-\end{center}
-\end{figure}
+%\begin{figure}
+%\begin{center}
+%\includegraphics[width=2.8in]{PROP.jpg}
+%\caption{\label{prop}
+%Comparing the expressive power of the temporal logics in Section~\ref{propLTL}
+%}
+%\end{center}
+%\end{figure}
 
@@
 
-The algorithm for \PLTL{} first presented in \cite{HR} 
+The RV algorithm for \PLTL{}, presented in \cite{HR},
 is based on the observation that the semantics of the 
-past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
-in the previous step $i - 1$ of a subformula.
-To demonstrate this, we repeat here the definition of the previous-time
-operator $\ominus$, and rewrite the definition
-of $\since$ in an equivalent form that is
+past-time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current state $i$ is defined in terms of the semantics of its subformula(s)
+in the previous state $i - 1$.
+To demonstrate this, we rewrite here the semantic definition of the 
+%previous-time
+%operator $\ominus$, and rewrite the definition
+ $\since$ operator in an equivalent form that is
 more directly applicable for runtime verification.
 \begin{itemize}
-\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
-\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
+%\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
+\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or: $i>1$ and
 $( \sigma , i)  \models \varphi$ and 
@@
 \end{itemize}
-Runtime verification for a \PLTL{} formula $\eta$ exploits the fact that
-the semantic definition is recursive in both the structure of the property. Thus, subformulas
+The semantic definition is recursive in both the length of the prefix and the
+structure of the property. Thus, subformulas
 are evaluated based on smaller subformulas, and the
 evaluation of subformulas in the previous state.
-The algorithm, shown below, uses two vectors (arrays) of values indexed by subformulas:  $\old$, which summarizes the truth values of the
+The algorithm, shown below, uses two vectors of values indexed by subformulas:  $\old$, which summarizes the truth values of the
 subformulas for the execution prefix that
-ends just {\em before} the new state, and $\current$, for the execution prefix that
-ends with the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
+ends just {\em before} the current state, and $\current$, for the execution prefix that
+ends with the current state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
 \begin{enumerate}
@@
 
-\item Observe a new event (as a set propositions) $s$ as input. 
+\item Observe a new event (as a set of propositions) $s$ as input. 
 \item Let $\old := \current$.
@@
 
-\subsubsection*{Runtime verification for \QPLTL{}.}  This can be performed by translating a
+\iffalse
+\subsubsection*{Extending~\cite{HR}: Runtime verification for \QPLTL{}.}  This can be performed by translating a
 formula into a deterministic finite automaton, where one distinguished state, $s$, is a sink state.
@@
 specification~\cite{Thomas}.
-
-\subsubsection*{Runtime verification for \EPLTL{}.} For \EPLTL{},  we need to  add to the above
+\fi
+
+\noindent {\bf Runtime verification for \EPLTL{}.} For \EPLTL{},  we need to  add to the above
 algorithm calculations 
 of $\current ( a_j )$ and
-$\current ( \varphi_j )$, for each of the rules of
+$\current ( \varphi_j )$ for each rule of
 the form $a_j := \varphi_j$ (the corresponding
 $\old$ entries will be updated as in line~3 in the above algorithm).
-
+%
 %(calculations
@@
 %and perform the following between steps~3 and~4 %for each $a_j$: $\current ( a_j ) := \current ( %\varphi_j )$.
-
+%
 %For \PLTL{}, we can calculate  $\current$, for subformulas,
@@
 Because the
-auxiliary variables can appear recursively in \EPLTL{} rules, the order of calculation is
-subtle. To see this, consider, for example, the formula~(\ref{form3}). It contains
-the definition $q := \ominus \neg q$. Now, we cannot calculate this
+auxiliary propositions can appear recursively in \EPLTL{} rules, the order of calculation is
+subtle. To see this, consider, for example, Formula~(\ref{form3}). It contains
+the definition $q := \ominus \neg q$. We cannot calculate this
 bottom up, as we did for \PLTL{}, since $\current ( q )$ is not computed yet, and we need 
@@
 so  $\current ( \ominus \neg q ) := \old ( \neg q )$.
-
-%\klaus{I don't see that we need to talk about the top-down evaluation.}
-
-Under {\em mixed evaluation order}, one calculates $\current$ as part of Step~4 of the above algorithm in the following order.
-The first~four steps, $a$-$d$, are for the rules, and the fifth step $e$ is for the statement $\eta$.
+%
+
+{\em Mixed evaluation order.}
+Under mixed evaluation order, one calculates $\current$ as part of Step~4 of the above algorithm in the following order. 
+%The first~four steps, $a$-$d$, are for the rules of the specification $\eta$, and the fifth step $e$ is for the statement $\psi$ of  $\eta$.
 \begin{description}
@@
 $a_j := \varphi_j$, but {\em not} within the scope of
-a $\ominus$ operator. (This is not
-a problem, since
-$\current ( \ominus \gamma )$ is set to $\old ( \gamma )$.)
+a $\ominus$ operator (observe that
+$\current ( \ominus \gamma )$ is set to $\old ( \gamma )$).
 \item{\em b.} Set 
@@
 that appears in the statement $\psi$, using the calculated $\current ( a_j )$.
-\item{\em e.} Calculate $\current ( \eta )$ for the statement
-$\eta$.
+%\item{\em e.} Calculate $\current ( \psi )$ for the statement
+%$\psi$.
 \end{description}
@@
 
-We define models for \FLTL{} based on {\em temporal} relations~\cite{Chomicki}. That is relations with
+We define models for \FLTL{} based on {\em temporal} relations~\cite{Chomicki}, that is, relations with
 last parameter
-that is a natural number represented a time instance. So a tuple of
-a relation $R$ can be $(``a", 5, ``cbb", 3)$, where
-the  value $3$ is the time parameter. The last parameter $i$
-does not correspond to modeling physical time. This
-is rather used to allow the relations to have
+that is a natural number, representing a time instance
+in the execution. So a tuple of
+a relation $R$ can be $(``a", 5, ``cbb", 3)$, where $3$ is
+the value of the time parameter. The last parameter $i$ represents a discrete progress of time
+rather than modeling {\em physical real time}. It
+is used to allow the relations to have
 different tuples in different instances of $i$,
-corresponding to states, in the propositional temporal
+corresponding to states in the propositional temporal
 logics.
@@
 restricting it to the value $i$ in the last parameter, and removing that last $i$ from the tuples. For simplicity, we will describe henceforth
-the logic with relations $R$ that have exactly two parameters, 
-hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$.
+the logic with relations $R$ that have exactly two parameters, the second of which is the time instance. 
+Hence $R [ i ]$ is a relation with just one parameter over a domain
+that will be denoted as $dom ( R )$.
 The definition of the logic over other numbers of parameters is quite straightforward.
-e.g., when $R$ has one parameter, $i$,
-$R [ i ]$ is a Boolean value.
-Our implementation of runtime verification described later fully supports relations with zero or more parameter.
-This also applies to the extensions of the logic in the rest of this section, and to some of the examples used.
+Our implementation of runtime verification, and the examples described later, fully support relations with zero or more parameter.
+%This also applies to the extensions of the logic in the rest of this section, and to some of the examples used.
 
@@
 
-\subsubsection{Syntax.}
-
+\noindent{\bf Syntax.} 
 The formulas of the core \FLTL{} logic are 
-defined by the following grammar,where $p$ denotes a relation,
+defined by the following grammar, where $p$ denotes a relation,
 $a$ denotes a constant and $x$ denotes a variable. 
@@
 
-\subsubsection*{Semantics.}
+\noindent{\bf Semantics.}
 A model is a set of temporal
@@
 we have the assignment $[ x \mapsto \text{``a''} , y \mapsto 3 ]$.
-%The past time temporal operators have the following intuitive meaning.
+%The past-time temporal operators have the following intuitive meaning.
 The formula $(\varphi_1\; \since\; \varphi_2)$ 
@@
 In any of the following cases, $( \gamma,  {\cal R} , i ) \models \varphi$
-is defined when $\gamma$ is an
+is defined where $\gamma$ is an
 assignment over $\vars ( \varphi )$, and $i\ge 1$.
@@
 \item $( \epsilon ,  {\cal R} , i )
- \models p ( a ) $ if $m(p) ( a, i )$, where $a$ denotes a constant from  $d(m(p))$.
-
-\item $( [ x \mapsto a ] , {\cal R} , i ) \models p ( x )$ if $m(p)  ( a , i )$, where $\domain (x) = d ( m ( p ))$.
+ \models p ( a ) $ if $m(p) ( a, i )$, where $a$ denotes a constant from  $dom (m(p))$.
+
+\item $( [ x \mapsto a ] , {\cal R} , i ) \models p ( x )$ if $m(p)  ( a , i )$, where $\domain (x) = dom ( m ( p ))$.
 
 \item $( \gamma,  {\cal R} , i ) \models ( \varphi \wedge \psi ) $~if~$(
-\gamma |_{\vars  ( \varphi )} , {\cal R} , i ) \models \varphi$~and  \\
+\gamma |_{\vars  ( \varphi )} , {\cal R} , i ) \models \varphi$~and 
 $( \gamma |_{\vars ( \psi ) } , {\cal R} , i ) \models \psi$.
@@
 For an \FLTL{} (\PFLTL{}) formula with no free variables, denote ${\cal R} \models \varphi$ when $( \epsilon , {\cal R} , 1 ) \models \varphi$.
-We will henceforth sometimes abuse notation, and use the same
+We will henceforce, less formally, use the same
 symbols both for the relations (semantics) and their representation in
 the logic (syntax).
-The quantification over values of variables here is {\em static} in the sense that they are independent of the state in the execution. We
-denote static quantification with $\exists$ and
-$\forall$.
+Note that the letters $p, q, r$, which were
+used for representing propositions in the propositional versions
+of the logic in previous sections, will represent relations in
+the first order versions.
+The quantification over values of variables, 
+denoted with $\exists$ and
+$\forall$,
+here is {\em static} in the sense that they are independent of the state in the execution. 
 %Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
@@
 \noindent {\bf Example 1.}
-Let $p$ and $q$ be temporal relations, where their
-restrictions to the state $s_i$, $p[i]$ and $q[i]$, are unary relations $p$ and $q$. The
+Let $p$ and $q$ be temporal relations (recall that we omit writing the time parameter $i$, denoting $p ( a, i)$ as
+$p(a)$). The
 specification that we want to monitor is that for each value
@@
 To show that this is not expressible in \FLTL{} (and \PFLTL{}),
-consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
+consider models (executions) where only one data element $a$ appears. We can prove, by a structural induction on subformulas,
 that in this case, for each property $\varphi$
@@
 $\mathit{spawn}(x, x_1)$, $\mathit{spawn} (x_1, x_2)$ $\ldots$
-$\mathit{spawn}(x_l , y)$. i.,e., $y$ is a descendent process of $x$. The reason
-is that the required property needs to
+$\mathit{spawn}(x_l , y)$. i.,e., $y$ is a descendent process of $x$.
+The required property needs to
 assert about the transitive closure
 of the relation $\mathit{spawn}$. 
-\FLTL{} can be translated in a way similar to a standard translation of $LTL$ to monadic first order logic~\cite{Thomas}.
-Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
-E.g., $\Box \forall x \, p(x) \rightarrow q(y)$ will be translated into $\forall t \, \forall x \, ((p(x, t) \rightarrow \exists t' p(x, t') )$\footnote{The obtained model will have, in addition
-to the aforementioned relations, also
-the linear order $<$ over the naturals.}.
-It is possible to show that the transitive closure of $\mathit{spawn}$ cannot be expressed in first order setting. The proof is based on the
+\FLTL{} can be translated (in a way similar to the standard translation of $LTL$ into monadic first order logic formula~\cite{Thomas}) to a first order formula, with explicit occurrences of time variables
+over the naturals and the linear
+order relation $<$ (or $\leq$) between them.
+%Here, the relations will be written with their explicit last ``time'' parameter, and the temporal operators are replaced with first order quantification.
+For example, $\Box \forall x \, (p(x) \rightarrow \Diamond q(x))$ will be translated into $\forall x \, \forall t \, (p(x, t) \rightarrow  \, \exists t' \, (t \leq t' \wedge q(x, t') ) )$.
+%The obtained model has, in addition
+%to the aforementioned relations, also
+%the linear order relations $\leq$ over the naturals.}.
+However, the transitive closure of $\mathit{spawn}$ cannot be expressed in first order setting. This can be
+shown based on the
 compactness theory of first order logic~\cite{Flum}.
 
-\subsubsection{Extending \FLTL{} with dynamic quantification.}
-
-Extending \FLTL{} (\PFLTL{}) with quantifiers over temporal relations, we obtain \QFLTL{}
+\vspace{1.5ex}
+\noindent {\bf Extending \FLTL{} with dynamic
+quantification.} Relations play in \FLTL{} a
+similar role to propositions in LTL.
+Hence, in correspondence with the relation
+between LTL and \QLTL{}, we
+extend \FLTL{} (\PFLTL{}) with dynamic quantification over relations, obtaining \QFLTL{}
 (and the past-restricted version \QPFLTL{}). 
 The syntax includes $\Exists p \, \varphi$, 
-where $p$ denotes an auxiliary relation.
-We also allow $\Forall p\ \varphi = \neg \Exists \neg p \varphi$.
+where $p$ denotes a relation.
+We also allow $\Forall p\ \varphi = \neg \Exists p \neg  \varphi$.
 The semantics is as follows.
@@
 Consequently, quantification over relations effectively extends the model ${\cal R}$ into a model ${\cal R}'$ within the scope of the quantifier.
-Note that quantification here is dynamic (as in \QLTL{} and \QPLTL{}), since the relations are temporal and can have different sets of tuples in different states.
-
-\subsubsection{Extending \FLTL{}  with rules.}
-
+Note that quantification here is dynamic (as in \QLTL{} and \QPLTL{}) since the relations are temporal and can have different sets of tuples in different states.
+
+\vspace{1.5ex}
+\noindent {\rm Extending \FLTL{}  with rules.}
 We now extend \FLTL{} into \EFLTL{} in a way that is motivated by the  propositional extension from LTL (\PLTL{}) to \ELTL{} (\EPLTL{}). We allow the following formula:
@@
 free variable $x_j$,
-\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
+\item $r_j$ is an auxiliary temporal relation with two parameters: the first parameter is of the same 
 type as $x_j$ and the second one is, as usual,
@@
 \noindent
-We define\footnote{ Formal semantics 
+We define the semantics for the \EFLTL{} (\EPFLTL{})
+specification (\ref{EFLTL}) 
+by using the following equivalent
+\QFLTL{} (\QPFLTL{}, respectively) formula\footnote{ Formal semantics 
 can also be given by constructing a set of temporal relations extended
 with the auxiliary ones
-inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the \EFLTL{} (\EPFLTL{})
-specification (\ref{EFLTL}) 
-by using the following equivalent
-\QFLTL{} (\QPFLTL{}, respectively) formula:
+inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.}:
 
@@
 
-\begin{lemma} \label{samesame}
-The auxiliary temporal relations $r_1, \ldots r_n$ are
-uniquely defined by the rules for each model ${\cal R}$.
+\iffalse
+[Well foundedness of auxiliary propositions] \label{fourone}
+The values of the the
+auxiliary variables 
+of an \ELTL{} formula $\varphi$
+are uniquely defined in a state of an execution
+by the prefix of the execution up to and including that state.
+\fi
+
+\begin{lemma}[Well foundedness of auxiliary relations] \label{samesame}
+The auxiliary temporal relations of an \EFLTL{} formula 
+at state $i$ are
+uniquely defined by the prefix of the execution up to and including that state.
 \end{lemma}
@@
 The proof of this theorem includes encoding of a property that
-observes sets of data elements $a$, appearing separately, one per state as $v (a)$,
+observes sets of data elements, where 
+elements $a$, appears separately, i.e., one per state, as
+$v (a)$,
 in between states where $r$ appears. The domain of data elements is unbounded.
@@
 The property asserts that there are no two consecutive data sets that are equivalent.
-This property can be expressed in \QPFLTL{}. The details appear in the appendix.
-
-We use a combinatorial argument to show by contradiction that one cannot express this property
-in \EPFLTL{}. The reason is that every prefix of a model for an \EPFLTL{} property is extended uniquely with
+This property can be expressed in \QPFLTL{}. The details appear in an appendix.
+
+We use a combinatorial argument to show by contradiction that one cannot express this property using any
+\EPFLTL{} formula $\varphi$. The reason is that every prefix of a model for an \EPFLTL{} property is extended uniquely with
 auxiliary relations, according to Lemma~\ref{samesame}. Each prefix can be summarized by a finite number of relations: the ones in
@@
 $2^m$.
-This means that with large enough number of different values, each \EPFLTL{} formula over the models of this property can have two prefixes with the same summary where at least one of them has a data set that the other one does not.
+This means that with large enough number of different values, each \EPFLTL{} formula $\varphi$ over the models of this property can have two prefixes with the same summary, where one of them has a data set that the other one does not.
 The semantics of \EPFLTL{} implies that extending
-two models with the same summary in the same way would have the same truth value. So, we can extend
-the two models with a data set that appears in one of the
-prefixes but not  in the other, and the \EPFLTL{} property will not be able to distinguish between them. \qed
+two prefixes with the same summary in the same way would have the same truth value. Consequently, we can extend
+the two prefixes where some data set  appears in one of them
+but not  in the other into a complete model, and $\varphi$ will not be able to distinguish between these models. \qed
 
@@
 \begin{corollary}
-Restricting the quantification of \QPFLTL{} to existential quantification, strictly weakens its expressive power.
+Restricting the quantification of \QPFLTL{} to existential quantification, strictly weakens its expressive power\footnote{It is interesting to note that for~\QPLTL{}, restriction to existential quantification does not change the expressive power.}.
 \end{corollary}
@@
 The RV algorithm will also make
-use of sets of assignments over a set of variables for satisfying a subformula at
-some state, corresponding to $\old$ and $\current$, which had Boolean values in the 
-algorithm for propositional LTL in Section~\ref{LTLruntime}.
+use of sets of assignments over a set of variables, for satisfying a subformula at
+some state, corresponding to $\old$ and $\current$, also represented as relations
+(instead of propositions, as used for LTL in Section~\ref{LTLruntime}).
 
@@
 
-\klaus{Note that $\psi$ above is not defined as a syntactic category. It should really be $\varphi$. Likewise in other parts of the paper. Perhaps too big a change now? Also, there should be
+\klaus{NOte t that $\psi$ above is not defined as a syntactic category. It should really be $\varphi$. Likewise in other parts of the paper. Perhaps too big a change now? Also, there should be
 a greek letter for a specification. E.g.:
@@
 
-\subsubsection{Set Semantics.}
-
+\vspace{1.4ex} \noindent {\bf Set Semantics.}
 The RV algorithm for
-(E)\PLTL{}, presented in~Section \ref{LTLruntime}
+(R)\PLTL{}, presented in~Section \ref{LTLruntime}
 calculates
-$\current ( \eta )$, for $\eta$ a
+$\current ( \varphi )$, for $\varphi$ a
 subformula of the monitored property, to be the
 truth value of $\varphi$ over the
-prefix inspected by the RV algorithm so far. For (E)\PFLTL{},
-$\current (\eta )$ consists of the
+prefix inspected by the RV algorithm so far. For (R)\PFLTL{},
+$\current (\varphi)$ consists of the
 sets of assignments (in the form of relations
@@
 
-We redefine the semantics for \EPFLTL{} in
-equivalent way, which will be more directly related to the calculation of values
+We provide an alternative semantics for 
+the logic \EPFLTL{}, without
+changing its interpretation, in a
+way will be more directly related to the calculation of values
 in $\current$ by the
@@
 iff $( \gamma , \sigma , i ) \models \varphi$.
-
 We present here only two simple cases of the set semantics. The full set semantics appears in the appendix. 
@@
 
-\subsubsection{Runtime verification algorithm for \PFLTL{}.}
-
+\noindent {\bf Runtime verification algorithm for \PFLTL{}.}
+%
 We start by describing an algorithm for monitoring 
-\PFLTL{} properties, first presented in~\cite{HPU} and implemented
+\PFLTL{} properties, presented in~\cite{HPU} and implemented
 in the tool~\dejavu. 
@@
 represent relations over the Boolean
-encodings of these enumeration, rather than over the data values themselves. A hash function is used to connect the data values to their enumerations to maintain consistency
+encoding of these enumeration, rather than over the data values themselves. A hash function is used to connect the data values to their enumerations to maintain consistency
 between these two representations.
@@
 represented as BDDs~\cite{Bryant}. 
-For example, if the runtime verifier sees the input 
+For example, if the runtime-verifier sees the input 
 events 
@@
 A Boolean representation of the {\em set} of values 
-$\{\text{``a''},\text{``b''}\} $ would be equivalent to a Boolean function $(\neg b_1 \wedge \neg b_2)$ that returns 1 for $000$ and $001$ (the value
-of $b_0$ can be arbitrary).
+$\{\text{``a''},\text{``b''}\} $ would be equivalent to a Boolean function $(\neg b_1 \wedge \neg b_2)$ that returns 1 for $000$ and $001$ ($b_0$ can be arbitrary).
 
@@
 
-The use of BDDs can be replaced by other representations that
-can compactly and efficiently represent sets of values (e.g., ZDDs), and to which one can apply set operations like complementation, intersection, union (which are simply negation, conjunction and disjunction
-in BDDs) and projection (for the quantification).
-The marriage of BDDs and Boolean enumeration is in particular
-efficient, since collections of adjacent Boolean enumerations tend to compact well.
-
 \iffalse %%%%%
@@
 %complicated enumeration mechanism.
-
+%
 The function $\bddset (x, A)$ returns
@@
 Intersection and union of sets of assignments are translated simply
-to conjunction and disjunction of their BDD representation,
-respectively, and complementation
+into conjunction and disjunction of their BDD representation,
+respectively; complementation
 becomes BDD negation. We will denote
@@
 Finally, $\bfalse$ and $\btrue$  are the BDDs that return always
-$0$ or $1$, respectively.
+$0$ or $1$, respectively. The algorithm is as follows.
 
@@
 
-\subsubsection{Runtime verification algorithm for \EPFLTL{}}
-
+%The use of BDDs can be replaced by other representations that
+%can compactly and efficiently represent sets of values (e.g., ZDDs), and to which one can apply set operations like complementation, intersection, union (which are simply negation, conjunction and disjunction
+%in BDDs) and projection (for the quantification).
+
+\noindent {\bf RV algorithm for \EPFLTL{}}
 We extend now the algorithm to capture \EPFLTL{}.
@@
 relations $r_j$ may be defined in a rule with respect to
-a variable $x_j$ (as $r_j (x_j ):= \varphi_j (x_j)$ (this can be generalized 
+a variable $x_j$ as in $r_j (x_j ):= \varphi_j (x_j)$ (this can be generalized 
 to any number of variables), but $r_j$ can be used
@@
 $\rename ( r_j (x_j ) , y)$.
-
+%
 We then have to add
@@
 both on the left and the right of a definition
-such as $r ( x ) := p ( x ) \vee \ominus r ( x )$. For that,
-we need to use the {\em mixed} evaluation order, described in Section~\ref{LTLruntime}.
+such as $r ( x ) := p ( x ) \vee \ominus r ( x )$. 
+We need to use the {\em mixed evaluation order}, described in Section~\ref{LTLruntime}.
+
+\vspace{1.5ex}
+\noindent {\bf Complexity.} BDDs were first introduced to model checking~\cite{McMillan} since they can often (but not always) allow
+a very compact representation of states.
+In our context, each BDD in $\old$ or $\current$
+represents a relation with $k$ parameters, which summarizes
+the value of a subformula of the checked \PFLTL{} or
+\EPFLTL{} property
+with $k$ free variables
+over the prefix observed so
+far. Hence, it can grow up to a size that is polynomial in the number of values appearing
+in the prefix, and exponential in $k$ (with $k$ being
+typically very small). However, the marriage of BDDs and Boolean enumeration is in particular
+efficient, since collections of adjacent Boolean enumerations tend to compact well.
